### 2016-11-21 manuscript.Rmd
add first draft of manuscript

---
#author: Fran\c{c}ois Michonneau
title: A data-driven approach to identifying knowledge gaps in marine invertebrates biodiversity
output: pdf_document
---

# Summary

Species identification from DNA sequences is poised to revolutionize our
understanding of biodiversity, to facilitate environmental monitoring, and to
disentangle trophic networks. However, this approach relies on comprehensive
databases of accurately identified sequences.  Since its introduction in the
early 2000's, DNA barcoding, the sequencing of a common locus across taxa
bearing species-specific information, has revitalized taxonomy by facilitating
species delineation. More recently, advances in high-throughput sequencing
technologies have enabled the parallel sequencing of thousands of species that
occur in environmental samples. The approach initially developed to study
microbial communities such as the ones from guts or soil, has been modified and
extended to be applied to metazoan communities. This approach opens new research
venues as it becomes possible to estimate the diversity from species-rich
samples that were virtually inaccessible before. This is particularly true in
marine communities that include multiple phyla and organisms varying in size by
order of magnitude, making identification challenging. However, the utility of
the data generated by these projects is enhanced by the possibility of matching
the DNA sequences obtained to previously sequenced and correctly identified
species. Here, we combined information available in public biodiversity
databases to assess the proportion of identified species that have been
sequenced in the continental United States. We show that the effort is
consistent across the better-studied taxa, with `r average_prop_species_seq`% of
the fauna sequenced. Using a DNA barcode dataset of `r n_larvae` planktonic
larvae, representing `r n_esus` species, collected in coastal waters of Florida,
we find a similar proportion of species that can be assigned to a previously
sequenced species ( `r format_output(100 * prop_plankton_with_match)`%).  Our
results show that most marine species remain to be sequenced. Achieving this
goal will required a concerted effort, and resources.

<!---
the content of insect traps, marine sediments, plankton
samples, or communities settling on artificial substratum. When applied to
metazoan communities, these sequencing approaches typically target the barcoding
gene and theefore has been dubbed "meta-barcoding". These applications are also
making the public databases for these DNA barcodes all the more valuable. To
deliver on the meta-barcoding promises, it is crucial to have comprehensive
taxonomic coverage represented in the DNA sequences public databases. Here, we
investigate the proportion of species for each marine phylum that have been
sequenced at various spatial scale.
--->

# Main text

With increased anthropogenic pressure on ecosystems, it is crucial to become
efficient at documenting biodiversity before species go extinct. This is
especially important in marine ecosystems where an undefined, but large,
fraction of the biodiversity remains to be discovered especially among
invertebrates. However, taxonomic expertise is dwindling, and there are now only
a handful of natural history museums with active marine invertebrates
collections. Museum collections have been recording our knowledge about species
and their geographical ranges, but until recently, this information was
fragmented across institutions and difficult to synthesize. With collection
digitization and other large-scale projects have aimed at gathering biodiversity
information, public databases can now be used to enable synthesis and highlight
knowledge gaps. Here we use these data to synthetic the state of knowledge for
marine invertebrates, and demonstrate that the stage is set to combine these
data sources to inform where future efforts for documenting biodiversity.

We combined `r nrow(global_idigbio_species_list)` museum records for marine
invertebrates collected in the continental US representing
`r length(store_idigbio_records()$list())` phyla and
`r length(unique(global_idigbio_species_with_worms_info$worms_valid_name))`
species. Most samples were collected on the continental
shelf (Fig. 1). For the East coast, species richness is the highest in South Florida but
it is relatively uniform along the coastline. For the West coast, the Salish
Sea, the Bay Area, and Southern California have the highest richness (Fig. 2).

The number of sampled species is vastly underestimated.
`r prop_singleton_species`% of the species are known from a single
record. Futhermore, the strong correlation between the number of samples
collected and the species richness suggests that any new sampling effort
will lead to new records (Fig. 3).

DNA barcoding is one approach that can be used to alleviate the discrepancy
between the increased need for biodiversity inventories while taxonomic
expertise is dwindling. DNA barcoding aims at assembling a comprehensive
database of DNA sequences that are associated with accurately identified
vouchered specimens. Future identifications can then rely on this database. DNA
barcoding can also facilitate species delineation, in turn, increasing the
accuracy of the identifications.

Recent developments in high-throughput sequencing technologies are giving new
applications to these DNA barcode databases. Meta-barcoding enables the parallel
sequencing of thousands of species that occur in environmental samples.  This
approach opens new research venues as it becomes possible to estimate the
composition of species-rich samples that were virtually inaccessible before
(e.g., fish gut content, marine sediment samples, plankton tows). The ability to
use DNA to identify members of these communities relies on comprehensive DNA
barcode databases. To assess the proportion of recorded species in the US that
has been sequenced, we searched for each species occurring in museum collections
for possible matches in BOLD (which also includes GenBank records). We found
that the proportion of sequenced species is consistent across the better-studied
taxa (overall `r average_prop_species_seq`%, Fig. 4). This proportion is consistent
across phyla despite important differences in species richness, and taxonomic
tradition. We also found that phyla representative of the meiofauna are
under-represented, and very few specimens and sequences are available.

Identified specimens from museum collections could be used to complete these
datasets. Traditional approaches relying on Sanger sequencing might be limiting as
many species have not been collected recently enough to guarantee good quality DNA
(`r prop_species_not_collected_since_1990`% have not been collected since
1990). However, high-throughput sequencing could offer alternatives to recover
sequences from older specimens.

Besides data-mining museum records to identify knowledge gaps, new sampling
approaches can be used in the field to rapidly assess how much of the fauna is
known. Meta-barcoding and environmental DNA approaches appear promising to
sample _en masse_ species that are found in a given location. Because most
marine invertebrates have a planktonic larval stage, they may co-occur in the
water column, allowing to sample species across habitats. To illustrate this
approach, we individually DNA barcoded `r n_larvae` planktonic larvae for DNA
barcoding, collected over the course of year, representing `r n_esus` species,
collected in coastal waters of Florida. We found that
(`r format_output(100 * prop_plankton_with_match)`%, Fig. 4) can be assigned
to a previously sequenced species. Similar proportions have been obtained from
individually barcoded species that occur on ARMS. However, this proportion
dropped to less than 5% for smaller organisms sequenced with meta-barcoding.

The species identifications in BOLD are far from perfect. Of the
`r sum(prop_plankton_with_species_match_by_phylum$n_esu_match)` species that have a
match in public database,
`r format_output(prop_plankton_with_multi_species_match * 100)`% have been
identified as more than one taxon in BOLD. A few of these records might represent
species complexes, but most illustrate the difficulty to identify precisely
marine invertebrates.

<!-- Studying biodiversity and identifying knowledge gaps about species and their -->
<!-- distributions need to become more data-driven to target under-studided areas and -->
<!-- taxonomic groups. To survive the biodiversity crisis, we need -->
<!-- inter-disciplinarity among -->
<!-- scientists  only need to be taxonomists that can reliably -->
<!-- identify species, but also molecular biologists. -->

![Number of museum records for marine inverterbrates](figures/map_sampling_effort.pdf)

![Number of species of marine invertebrates](figures/map_diversity.pdf)

![Relation between the number of specimen and the number of species collected](figures/plot_sampling_effort.png)

![Proportion of species occuring in natural history collections that have been DNA barcoded](figures/proportion_barcoded.pdf)

<!--

The number of species that occurs in the sea remains uncertain.

While plankton is recognized for its biological importance (dispersal, food
source for many species), and can be used as an ecological indicator for
pollution or global change, identification of planktonic larvae at the species
level from morphological information is generally not possible because of the
paucity of informative characters to base the identifications on, and the lack
of knowledge of the appearance of the larval stages for most species. DNA
barcoding can bridge this knowledge gap by using genetic information to identify
larvae at the species level. These identifications are only possible if the
species have already been sequenced, and the sequences deposited in public
repositories. The accuracy of the identification relies on the expertise of the
taxonomists who identified the specimens that have been sequenced.

We collected over the course of a year planktonic larvae of marine invertebrates
in a coastal marsh of the Atlantic coast of Florida. One of our goals was to
identify these larvae at the species level using DNA barcoding. We photographed
and sequenced xxx larvae, representing an estimated xxx species. These data are
available from BOLD and can be browsed interactively at
http://plankton-atlas.net. Using records from public databases (BOLD, GenBank),
we could confidently assign only xxx% to species that had already been
sequenced. On average, these taxa were sequenced xxx km away from our study
site.

Florida is one of the most populated states, with a long tradition of studying
its marine life.  xxx universities have marine research programs, and xxx marine
laboratories. How does this proportion of xxx% identified species compare with
other places in Florida, the East coast of the US, the continental US, and the
world. In other words, how much of the species of marine invertebrates we know
about have publicly available DNA barcodes?

To estimate this number, we assembled species lists of marine invertebrates for
the continental US by querying iDigBio for each marine phylum. We only kept
records for specimens identified with a valid binomial name (specimens
identified at the genus level or higher were excluded from the analysis). We
used WoRMS to ensure that the species names were valid, and to disambiguate
possible synonyms. For each species, we retrieved avaiable DNA barcoding
sequence data from BOLD, which also includes DNA barcodes deposited in
GenBank. Of the xxx marine invertebrate species found in iDigBio, xxx% had DNA
barcodes available in public databases. These proportions varied by phylum, but
for the conspicuous and better studied groups (Porifera, Cnidaria, Mollusca,
Echinodermata), they were surprisingly consistent, and is a close estimate to
the proportion of species that have been sequenced for the United
States. Furthermore, we also looked for the proportion of species known to occur
in the Indian River Lagoon for which DNA barcodes are available, and again the
proportion of sequenced species (xxx%) is comparable.

Except for the few phyla that are known to include only a small number of
species (Cycliophora, Phoronida, Chaetognatha), microscopic groups, typically
component of the meiofauna, are disproportionally under-represented in museum
collections, and none or few species are available in public databases.

Our results highlight that most of our knowledge of marine biodiversity come
from large expeditions carried in the late 19th century and after World War
II. Since then, sampling effort has dramatically decreased, and taxonomic
expertise is waning. Yet, our ability to assign DNA sequences to species names
will greatly improve the value of the data generated by meta-barcoding studies.

and reflect that a large fraction of the diversity of
marine invertebrates remain to be sequenced.

Marine invertebrates are the biodiversity. All but one metazoan phyla occur in
the sea. Of the xx described species, yy are marine invertebrates. They
represent a multiple array of shapes, forms, and adaptations to all marine
environments. Yet, they are proportionally understudied. Here, we analyze
publicly available data from biodiversity databases to demonstrate that the
abandonment of traditional taxonomic practices, the lack of support for natural
history collections, is hampering our ability to use marine invertebrates as
indicators of ecosystem health, the effect of global change, the identification
of model organisms, and ... Technological advances are however making easier to
identify knowledge gaps, the potential to identify where new species occur, and
document diversity.

# The role of DNA barcoding

The use of molecular data for species identification and delineation is arguably
one of the most important advances in taxonomy. Especially with DNA barcoding
that proposes to use a common genetic marker across large taxonomic groups (the
mt marker COI for animals, ITS for fungi, and xx for plants). Because these
markers have typically low intra-specific variation, and differ noticeably
between even closely related species, they can be used to identify species and
unravel species complexes.

Species identification relies on reference databases, where sequences are
associated with accurately identified individuals, held in natural history
collections. When taxonomic sampling is comprehensive, and enough individuals of
each species are sampled to be representative of the standing genetic variation
in the population, DNA barcoding can be a very accurate tool with xx-yy% error
rate.

The utility of these reference databases is getting revived with the development
of meta-barcoding techniques, where thousands of species can be sequenced in
parallel on high-throughput sequencing platforms. If methods exist to estimate
the number of species these samples contain based on the level of genetic
divergence observed among the sequences, being able to match these sequences to
identified species based on their match to a sequence in a public database would
increase the potential applications for these approaches.

Marine invertebrates are however disproportionally under-represented in DNA
sequence databses. And they have a higher rate of misidentification. There are
however some pitfalls to this number. It might actually be even higher given
that they have been identified by a single taxonomist given the lack of
redundancy in expertise.

To illustrate the consequences of having so few barcodes for invertebrates, if
we take the faunistic list of IRL, only about 20% of the species known to be
occuring in the area actually have DNA sequences available for them. On average,
each species has been sequenced xx times, and from locations on average yy km
away from IRL. Given the high propencity for cryptic species among marine
invertebrates, it is possible that ...

We also sequenced xx larvae occurring in the plankton at the Whitney Marine
Lab, representing yy species (estimated from genetic distances). Of these yy
species, only zz% had good matches in BOLD/GenBank. In both cases, taxa1, taxa2,
and taxa3 are the most underrepresented in BOLD.

# The waining of taxonomy

Total number of collections in the US, how many have marine invertebrate
collections, how many have active curators.

We don't know how many species are in the sea. Estimates vary widely. Difficult
to sample because diversity of habitats, each requiring special sampling
techniques. Meanwhile, taxonomic expertise is decreasing despite efforts to keep
it alive (PEET).

To get around issue of sampling different habitat, sample coastal
meroplankton. Most species go through planktonic phase, therefore common habitat
to sample them all. The issue is that they look very different from adults, and
only DNA can be used to identify them.

Barcoding is not perfect, but can be very powerful to identify the larval forms
(e.g. one of the first paper was sampling caterpillars). Potentially differences
in larva that are not seen in adults and could be additional characters to tell
species appart.

Advantage of sampling plankton is that a single habitat is sampled. Not enough
known about heterogeneity in spatial distribution.

We sampled plankton bi-weekly for a year, and sequenced the larvae
found. Sequenced the barcoding gene and tried to match how many species we could
identify based on records in public databases. Overall, low percentage of
match. And matches also led to several species that have been collected but that
have not been identified to an appropriate level.

To compare the results we obtained here, and see how they could compare to other
habitats, we downloaded the species list from Indian River lagoon inventory. It
is the most comprehensive checklist in Florida, and has been developed over many
years. We asked how many of the species occurring in the Indian River Lagoon have
already been sequenced. We did the matching based on the species names. If we
find higher proportion of sequenced species, it means that plankton gives access
to species that are difficult to sample as adults. If lower proportion, it means
that we don't have many of the adults sequenced at Whitney.

Using plankton to sample biodiversity could lead to a speedup of
documentation. It would allow the identification of taxa/habitats that need
special attention. If representative of adult communities, could also be a
useful tool to monitor status of ecosystem through time.

Better database, can also be used to estimate the proportion of really dark
planktonic taxa.

Understand better population connectivity, currents

Importance of plankton in connecting populations. Important as food source.

Swimming in the dark in the age of big data

Biodiversity is in the sea. All phyla but one. New lineages of eukaryotes,
viruses...

We are at a time where it becomes cheaper and easier to delineate species. To
sort through specimens to identify species. So taxonomists can focus on
studying/describing new species.

Granularity of data not sufficient to be able to detect changes in the
environment, shifts in species ranges, changes in trophic structure. Understand
effect of small animals.

Rely on high school students, citizen science initiatives to help with labor
intensive part.

Oceans are fascinating.

Bioprospection

focus on a few taxa but lots of groups we know nothing about.

Strong focus on vertebrates. We are building theories and making conservation
approaches on a small fraction of the tree of life.

Trait data

Species Distribution

--->

### 2016-11-22 R/idigbio.R
create function to check spp in iDigBio but not from the US

@@ -46,6 +46,23 @@ spp_not_in_idigbio <- function(wrm, idig_qry) {
 }
 
 
+is_in_us <- function(country, lon, lat) {
+    identical(tolower(country), "united states") ||
+        ((lon > us_coords()$lon[1] && lon < us_coords()$lon[2]) &&
+         (lat > us_coords()$lat[1] && lat < us_coords()$lat[2]))
+}
+
+
+## takes the output of fetch_spp_from_idigbio
+spp_not_in_us <- function(idig) {
+    idig %>%
+        dplyr::group_by(scientificname) %>%
+        summarize(
+            in_us = any(mapply(is_in_us, country, geopoint.lon, geopoint.lat))
+        )
+}
+
+
 calc_prop_spp_not_in_idigbio <- function(not_idig) {
     sum(is.na(not_idig$uuid_lst))/nrow(not_idig)
 }

### 2016-11-22 R/utils.R

@@ -50,3 +50,12 @@ format_output <- function(x) {
 render <- function(file, ...) {
     rmarkdown::render(file, ...)
 }
+
+## US coordinates
+## pretty generous, but easier
+us_coords <- function() {
+    list(
+        lon = c(-128, -60),
+        lat = c(22, 51)
+    )
+}

### 2016-11-22 manuscript.Rmd
begin rewrite intro

@@
 fraction of the biodiversity remains to be discovered especially among
-invertebrates. However, taxonomic expertise is dwindling, and there are now only
+invertebrates. One approach would be to build on existing knowledge to identify
+gaps and target understudied areas or taxonomic groups. This is currently
+difficult because the knowledge is fragmented and not always
+accessible. However, recent several large-scale initiatives have been developed
+to aggregate and synthesize various facets of biodiversity knowledge. For
+instance, the Biodiversity Heritage Library aims at making the biodiversity
+litterature openly available, the World Registry of Marine Species provides a
+database of the validity and relationships for most marine species taxonomic
+names, and iDigBio digitizes and makes available on the web museum
+specimen-associated data. Our knowledge about species and their geographical
+range comes from:
+
+1. **Faunistic lists** synthesize and evaluate expert knowledge about a specific
+   location. While limited in geographical scope, they tend to be accurate and
+   comprehensive.
+1. **Museum specimens** are physical proofs of the presence of a species at a
+   given point in time and space. The identification can be re-assessed, but
+   their contribution to biodiversity knowledge is greatly enhanced if
+   associated data and metadata are digitized.
+1. **DNA barcodes** useful to delineate and identify species. It requires the
+   presence of a museum voucher. Only as good as the identification. Has been
+   very useful to identify species complexes, especially among marine
+   invertebrates.
+
+To assess the level of digitization for biodiversity, we compared two very
+comprehensive faunistic lists for the marine fauna of the US, one from the Puget
+Sound (West coast) by Kozloff and the Gulf of Mexico checklist (East
+coast). These areas are well studied. Overall, we found that xx% of the species
+included in these lists were not found in iDigBio, and xx% in GBIF. Microscopic,
+poorly studied metaozans were less represented than larger groups, but for the
+latter, proportions on the East and West coast were very similar. Most species
+are likely to be undersampled as xx% of the species are represented by less than
+5 lots.
+
+However, taxonomic expertise is dwindling, and there are now only
 a handful of natural history museums with active marine invertebrates

### 2017-03-08 R/time_plot.R
fix legend in plot

@@ -155,7 +155,9 @@ plot_cum_samples_through_time <- function(knowledge_through_time, facet = TRUE) 
         filter(phylum %in% phy_to_keep) %>%
         ggplot(aes(x = year, y = cum_n_samples, colour = phylum)) +
         geom_line() +
-        xlim(c(1850, 2017))
+        xlim(c(1850, 2017)) +
+        ylab("Cumulative Number of Samples") +
+        xlab("Year")
 
      if (facet)
         res <- res + facet_grid(fauna ~ database)
@@ -173,7 +175,9 @@ plot_cum_spp_through_time <- function(knowledge_through_time, facet = TRUE) {
                !is.na(cum_n_new_spp)) %>%
         ggplot(aes(x = year, y = cum_n_new_spp, colour = phylum)) +
         geom_line() +
-        xlim(c(1850, 2017))
+        xlim(c(1850, 2017)) +
+        ylab("Cumulative Number of Species") +
+        xlab("Year")
 
     if (facet)
         res <- res + facet_grid(fauna ~ database)

### 2017-03-08 manuscript.Rmd
edits to manuscript

--- 2016-11-21 manuscript.Rmd
@@
 #author: Fran\c{c}ois Michonneau
-title: A data-driven approach to identifying knowledge gaps in marine invertebrates biodiversity
-output: pdf_document
+title: "Data-driven approaches to identifying knowledge gaps in marine invertebrates biodiversity"
+output:
+  pdf_document #:
+    #pandoc_args: ["-F","pandoc-crossref"]
 ---
 
-# Summary
+<!-- use Pimm+2014 intro as model on using short informative sentences to build -->
+<!-- 1st paragraph. -->
 
-Species identification from DNA sequences is poised to revolutionize our
-understanding of biodiversity, to facilitate environmental monitoring, and to
-disentangle trophic networks. However, this approach relies on comprehensive
-databases of accurately identified sequences.  Since its introduction in the
-early 2000's, DNA barcoding, the sequencing of a common locus across taxa
-bearing species-specific information, has revitalized taxonomy by facilitating
-species delineation. More recently, advances in high-throughput sequencing
-technologies have enabled the parallel sequencing of thousands of species that
-occur in environmental samples. The approach initially developed to study
-microbial communities such as the ones from guts or soil, has been modified and
-extended to be applied to metazoan communities. This approach opens new research
-venues as it becomes possible to estimate the diversity from species-rich
-samples that were virtually inaccessible before. This is particularly true in
-marine communities that include multiple phyla and organisms varying in size by
-order of magnitude, making identification challenging. However, the utility of
-the data generated by these projects is enhanced by the possibility of matching
-the DNA sequences obtained to previously sequenced and correctly identified
-species. Here, we combined information available in public biodiversity
-databases to assess the proportion of identified species that have been
-sequenced in the continental United States. We show that the effort is
-consistent across the better-studied taxa, with `r average_prop_species_seq`% of
-the fauna sequenced. Using a DNA barcode dataset of `r n_larvae` planktonic
-larvae, representing `r n_esus` species, collected in coastal waters of Florida,
-we find a similar proportion of species that can be assigned to a previously
-sequenced species ( `r format_output(100 * prop_plankton_with_match)`%).  Our
-results show that most marine species remain to be sequenced. Achieving this
-goal will required a concerted effort, and resources.
+<!-- Species identification from DNA sequences is poised to revolutionize our understanding of biodiversity, to facilitate environmental monitoring, and to disentangle trophic networks.
+However, this approach relies on comprehensive databases of accurately identified sequences.
+Since its introduction in the early 2000's, DNA barcoding, the sequencing of a common locus across taxa bearing species-specific information, has revitalized taxonomy by facilitating species delineation.
+More recently, advances in high-throughput sequencing technologies have enabled the parallel sequencing of thousands of species that occur in environmental samples.
+The approach initially developed to study microbial communities such as the ones from guts or soil, has been modified and extended to be applied to metazoan communities.
+This approach opens new research venues as it becomes possible to estimate the diversity from species-rich samples that were virtually inaccessible before.
+This is particularly true in marine communities that include multiple phyla and organisms varying in size by order of magnitude, making identification challenging.
+However, the utility of the data generated by these projects is enhanced by the possibility of matching the DNA sequences obtained to previously sequenced and correctly identified species.
+Here, we combined information available in public biodiversity databases to assess the proportion of identified species that have been sequenced in the continental United States.
+We show that the effort is consistent across the better-studied taxa, with ` average_prop_species_seq`% of the fauna sequenced.
+Using a DNA barcode dataset of ` n_larvae` planktonic larvae, representing ` n_esus` species, collected in coastal waters of Florida, we find a similar proportion of species that can be assigned to a previously sequenced species ( ` format_output(100 * prop_plankton_with_match)`%).
+Our results show that most marine species remain to be sequenced.
+Achieving this goal will required a concerted effort, and resources.
+-->
 
@@
 metazoan communities, these sequencing approaches typically target the barcoding
-gene and theefore has been dubbed "meta-barcoding". These applications are also
+gene and therefore has been dubbed "meta-barcoding". These applications are also
 making the public databases for these DNA barcodes all the more valuable. To
@@
 
+<!-- studies that have summary statistics on what has been sequenced for various marine phyla:
+- for Antarctica: Grant+Linse2009-Polar Biology:
+
+We know that some perturbations have large effects because they are visible, but what about perturbations for which we don't see the effects directly because they affect small species?
+:: checkout nature paper on sea grasses.
+
+- Describing new species is crucial but not sufficient.
+Without knowing their ranges, and their habitats, we can't study or predict how perturbations of the environment are going to affect their distributions and their susceptibility to perturbations.
+The process of figuring out the ranges of species is more time consuming than possibly finding new species; and it requires shared knowledge in the community about these new species.
+Using barcoding and other moleuclar techniques to improve the documentation of species distributions is going to be critical.
+
+There can be public involvement in species description, but the exploration and collecting may remain an affair of specialists as it requires manipulating technical gear of have expertise.
+There are still many species that are easy to access to discover, but the bigger gap is in the species that are difficult to access, and small.
+
+We also need to better coordination in describing new species; so that efforts are not duplicated.
+
+Possibly also need to set guidelines on how to treat cases where the molecular data alone is too ambiguous.
+
+-->
+
 # Main text
 
-With increased anthropogenic pressure on ecosystems, it is crucial to become
-efficient at documenting biodiversity before species go extinct. This is
-especially important in marine ecosystems where an undefined, but large,
-fraction of the biodiversity remains to be discovered especially among
-invertebrates. However, taxonomic expertise is dwindling, and there are now only
-a handful of natural history museums with active marine invertebrates
-collections. Museum collections have been recording our knowledge about species
-and their geographical ranges, but until recently, this information was
-fragmented across institutions and difficult to synthesize. With collection
-digitization and other large-scale projects have aimed at gathering biodiversity
-information, public databases can now be used to enable synthesis and highlight
-knowledge gaps. Here we use these data to synthetic the state of knowledge for
-marine invertebrates, and demonstrate that the stage is set to combine these
-data sources to inform where future efforts for documenting biodiversity.
+<!--  study focused on USA but conclusions more general given that it's arguably one of the better studied area + data availability -->
+<!-- * barcoding has shown that typically marine invertebrates are complex of cryptic species with small ranges; therefore potentially more vulnerable to extinction risk. -->
+<!-- * need collaboration among researchers to share data standards, protocols -->
+<!-- * use existing knowledge to fill missing gaps -->
+<!-- * running against time -->
+<!-- * we need to document species diversity better for conservation (knowing what we protect and to measure the effectiveness of the protection) -->
+<!-- * how do we conserve animal species that we can't see, we don't find pretty, don't have eyes -->
+<!-- * recent advances in technologies make possibility of developing a global monitoring program possible: reduced costs of sequencing, detection of rare species, etc.... -->
+<!-- * Using this kind of approaches at a large scale might also be a good way to have a better sense of how much of the biodiversity we are missing. -->
+<!-- * Most species are rare, especially in very diverse places,  and the degradation of habitats makes it even more difficult to find these rare species. -->
 
-We combined `r nrow(global_idigbio_species_list)` museum records for marine
-invertebrates collected in the continental US representing
-`r length(store_idigbio_records()$list())` phyla and
-`r length(unique(global_idigbio_species_with_worms_info$worms_valid_name))`
-species. Most samples were collected on the continental
-shelf (Fig. 1). For the East coast, species richness is the highest in South Florida but
-it is relatively uniform along the coastline. For the West coast, the Salish
-Sea, the Bay Area, and Southern California have the highest richness (Fig. 2).
+With increasing impact of human activities on the biosphere, we need efficient and accurate approaches to document changes in species ranges, detect changes in the composition of communities, and more generally accelerate the documentation of species diversity.
+This is especially pressing in marine ecosystems where a significant portion of the biodiversity remains to be discovered and described.
 
-The number of sampled species is vastly underestimated.
-`r prop_singleton_species`% of the species are known from a single
-record. Futhermore, the strong correlation between the number of samples
-collected and the species richness suggests that any new sampling effort
-will lead to new records (Fig. 3).
+Until recently our knowledge of species and their distributions was difficult to synthesize because it was scattered and difficult to access.
+However, recent advances in digitization and public databases efforts have made widely available the content of natural history collections, publications about species, their geographical distributions, and their genetic data.
+Given the urgency with which we need to document biodiversity and the limited means available to do so, we need to use and build on this existing knowledge to identify gaps, and to be more efficient at collecting information and understanding our natural world.
 
-DNA barcoding is one approach that can be used to alleviate the discrepancy
-between the increased need for biodiversity inventories while taxonomic
-expertise is dwindling. DNA barcoding aims at assembling a comprehensive
-database of DNA sequences that are associated with accurately identified
-vouchered specimens. Future identifications can then rely on this database. DNA
-barcoding can also facilitate species delineation, in turn, increasing the
-accuracy of the identifications.
+The incorporation of molecular data in the study of biodiversity has highlighted the pervasiveness of cryptic species, and generally the fact that species have smaller ranges than their taxonomic identity implied.
+Notably, the wide adoption of DNA barcoding for the study of marine diversity has clarified our understanding of species limits in many groups, but has also enabled the identification of species during their larval stages (when the lack of morphological features make species identification difficult).
+DNA barcoding has also been seen as a possible solution to the widening gap between the increased need for biodiversity inventories while taxonomic expertise is dwindling.
+By matching DNA barcodes of unidentified species to a public database of accurately identified and vouchered specimens, one can greatly speed up the process of species identification by matching them to already identified species.
+Sequences with no match make it easier to flag specimens that warrant further taxonomic attention.
 
-Recent developments in high-throughput sequencing technologies are giving new
-applications to these DNA barcode databases. Meta-barcoding enables the parallel
-sequencing of thousands of species that occur in environmental samples.  This
-approach opens new research venues as it becomes possible to estimate the
-composition of species-rich samples that were virtually inaccessible before
-(e.g., fish gut content, marine sediment samples, plankton tows). The ability to
-use DNA to identify members of these communities relies on comprehensive DNA
-barcode databases. To assess the proportion of recorded species in the US that
-has been sequenced, we searched for each species occurring in museum collections
-for possible matches in BOLD (which also includes GenBank records). We found
-that the proportion of sequenced species is consistent across the better-studied
-taxa (overall `r average_prop_species_seq`%, Fig. 4). This proportion is consistent
-across phyla despite important differences in species richness, and taxonomic
-tradition. We also found that phyla representative of the meiofauna are
-under-represented, and very few specimens and sequences are available.
+Recent developments in high-throughput sequencing technologies are giving new applications to these DNA barcode databases.
+Meta-barcoding enables the parallel sequencing of thousands of DNA barcodes for species that occur in environmental samples.
+This approach opens new research venues as it becomes possible to estimate the composition of species-rich samples that were virtually inaccessible before (e.g., gut content, marine sediment samples, plankton tows).
+For instance, by sequencing the plankton of coastal areas, it becomes possible to sample the larval stages of many species that would otherwise be difficult to sample.
+However, to accurately identify the species found in these samples, comprehensive DNA barcode databases are needed.
 
-Identified specimens from museum collections could be used to complete these
-datasets. Traditional approaches relying on Sanger sequencing might be limiting as
-many species have not been collected recently enough to guarantee good quality DNA
-(`r prop_species_not_collected_since_1990`% have not been collected since
-1990). However, high-throughput sequencing could offer alternatives to recover
-sequences from older specimens.
+With the cost of sequencing going down, and the prospect of having an automated sampling process, these technologies will soon be used for ecosystem and biodiversity monitoring.
+However, to interpret the results and understand how species distributions shift with global change or other perturbations, we cannot only rely on the identity of the species, but we also need to take advantage of existing knowledge about these species that is coming from natural history collections.
+In any case, the insights we will gain from these approaches are only going to be as good as the level of completeness and accuracy of the information included public databases.
 
-Besides data-mining museum records to identify knowledge gaps, new sampling
-approaches can be used in the field to rapidly assess how much of the fauna is
-known. Meta-barcoding and environmental DNA approaches appear promising to
-sample _en masse_ species that are found in a given location. Because most
-marine invertebrates have a planktonic larval stage, they may co-occur in the
-water column, allowing to sample species across habitats. To illustrate this
-approach, we individually DNA barcoded `r n_larvae` planktonic larvae for DNA
-barcoding, collected over the course of year, representing `r n_esus` species,
-collected in coastal waters of Florida. We found that
-(`r format_output(100 * prop_plankton_with_match)`%, Fig. 4) can be assigned
-to a previously sequenced species. Similar proportions have been obtained from
-individually barcoded species that occur on ARMS. However, this proportion
-dropped to less than 5% for smaller organisms sequenced with meta-barcoding.
+<!-- here there are 3 things: data availability (having good APIs to access the -->
+<!-- data); data comprehensiveness (how many species are captured in the -->
+<!-- database); data accuracy (can we trust the data that is found in the -->
+<!-- databases) -->
 
-The species identifications in BOLD are far from perfect. Of the
-`r sum(prop_plankton_with_species_match_by_phylum$n_esu_match)` species that have a
-match in public database,
-`r format_output(prop_plankton_with_multi_species_match * 100)`% have been
-identified as more than one taxon in BOLD. A few of these records might represent
-species complexes, but most illustrate the difficulty to identify precisely
-marine invertebrates.
+Here, we assess the level of completeness of biodiversity public databases for marine invertebrates that occur within the boundaries of the EEZ of the US.
+Specifically, we use comprehensive regional faunistic lists to assess how much of this knowledge is captured by current digitization efforts of museum collections and occurrence records; and we estimate how many of the species have genetic data publicly available that would be useful in the context of identifying samples from meta-barcoding studies.
 
-<!-- Studying biodiversity and identifying knowledge gaps about species and their -->
-<!-- distributions need to become more data-driven to target under-studided areas and -->
-<!-- taxonomic groups. To survive the biodiversity crisis, we need -->
-<!-- inter-disciplinarity among -->
-<!-- scientists  only need to be taxonomists that can reliably -->
-<!-- identify species, but also molecular biologists. -->
+We extracted `r sum(gom_worms$is_marine, na.rm = TRUE)` marine invertebrates species from the faunistic list of the Gulf of Mexico, and `r sum(kozloff_worms$is_marine, na.rm=TRUE)` species from the Salish Sea.
+Species are unevenly represented in databases, with 25% to 50% of species listed to occur in these locations, that are not represented in digital records of museum collections.
+Typically a higher proportion of species from the Pacific Coast than the Gulf of Mexico are not represented.
+Comparatively, all but 2 of the `r nrow(mammals_species_asm)` mammal species known to occur in the US have digitized museum specimens.
 
-![Number of museum records for marine inverterbrates](figures/map_sampling_effort.pdf)
+The proportion of species that have DNA barcodes available is surprisingly constant across the better studied phyla.
+About 20% of species known to occur in the Gulf of Mexico and in the Salish Sea have DNA barcode records.
+Because geographic information is missing for most records (`r format_output(100*stat_bold$p_no_coords)`% of the `r format(stat_bold$n_total_records)` records analyzed here have no geographical coordinates information associated with them), and many marine invertebrate species are known to be cryptic complexes that are not as widely distributed as previously thought.
+Therefore specimens identified under the same species but sampled in distant locations, could represent cryptic species.
+It is thus crucial to ensure that accurate geographic information is provided when DNA barcodes are published, and that additional samples are collected, to ensure that the extent of their genetic diversity is captured.
+This high level of cryptic diversity is reflected in the fact that `r stat_bold$n_spp_muti_bin` species out of `r stat_bold$n_spp` include more than 1 BIN.
+This number is confounded by some taxonomic identification issues as `r stat_bold$n_shared_bins` BINs are shared across more than one species.
+
+<!-- However, among these records, xx% have been sampled in the US (xx%), Mexico (xx%) or Canada (xx%).  -->
+<!-- Paragraph about the same numbers but based on species list from iDigBio: -->
+
+<!-- - how many species are in iDigBio and don't show up in species lists from Kozloff and GOM -->
+<!-- - how many species in iDigBio have barcodes -->
+<!-- - how many species in iDigBio have OBIS and GBIF records -->
+
+We combined `r nrow(idigbio_records)` museum records for marine invertebrates collected in the continental US representing `r length(unique(tolower(idigbio_records[["data.dwc:phylum"]])))` phyla and `r length(unique(idigbio_records$worms_valid_name))` species.
+Most samples were collected on the continental shelf (Fig.
+1).
+For the East coast, species richness is the highest in South Florida but it is relatively uniform along the coastline.
+For the West coast, the Salish Sea, the Bay Area, and Southern California have the highest richness (Fig.
+2).
+
+These patterns highlight our lack of knowledge about the marine fauna: `r prop_singleton_species`% of the species are known from a single record.
+Furthermore, the strong correlation between the number of samples collected and the species richness suggests that any new sampling effort will lead to new records (Fig.
+3).
+This is also reflected in the cumulative numbers of species recorded for the first time and the number of samples through time (Fig.
+4).
+The slowing down of collections in the 1970 has been accompanied with a slowing down in the discovery of new records: `r n_spp_comparison$n_spp_2017_2000` species have been documented for the first time since 2000, while `r n_spp_comparison$n_spp_1970_1960` have been documented between 1960 and 1970.
+
+There are important discrepancies in our level of knowledge across phyla.
+While larger, and accessible groups are well represented in museum collections, and have a reasonable proportion of their known species represented in DNA barcode databases, the small organisms of the meiofauna are virtually absent: 12 of the 25 phyla represented in the dataset have less than 500 specimens in museums.
+
+However, taxonomic expertise is dwindling, and there are now only a handful of US natural history museums with active marine invertebrates collections.
+Museum collections have been recording our knowledge about species and their geographical ranges, but until recently, this information was fragmented across institutions and difficult to synthesize.
+With collection digitization and other large-scale projects have aimed at gathering biodiversity information, public databases can now be used to enable synthesis and highlight knowledge gaps.
+Here we use these data to synthetic the state of knowledge for marine invertebrates, and demonstrate that the stage is set to combine these data sources to inform where future efforts for documenting biodiversity.
+
+Based on these results, we recommend that additional field expeditions are organized to sample more thoroughly the marine environment.
+The data from museum collections make clear that any additional samples will contain new species and new records.
+However, taxonomists are needed to identify and describe these new species.
+The samples collected during these expeditions need to be photographed to document live appearance that would otherwise be lost during preservation.
+We also need to capture metadata associated with the specimens to be able to understand better the physical characteristics of the habitats for these species.
+The specimens need to be sub-sampled and have tissues suitable for genomic analysis preserved, so we can sequence their DNA barcode, or other genomic resources.
+
+Using metabarcoding is a promising approach to screen an habitat for its potential for species that are not found in public database, to maximize the usefulness in the future of the data generated by an expedition.
+Metabarcoding approaches need to complement and take advantage of the DNA barcodes already available.
+Additionally, in the future, data generated by metabarcoding efforts can be re-interpreted as more DNA barcodes become available.
+Building the library of DNA barcodes in parallel of metabarcoding efforts also ensures that matches in the database will correspond to actual specimens deposited in natural history collections; ruling out possible artifacts of the method.
+The knowledge will also build on existing natural history data rather than having to restart from scratch.
+Beyond their utility in documenting diversity, metabarcoding is also promising for understanding the functioning of ecosystems.
+
+<!-- Paragraph on how sampling effort has changed through time. -->
+
+<!-- - get numbers of species described by year, look if there is a correlation between sample numbers and number of new species (possible lag with time for description) -->
+<!-- - see if proportion of species identified at genus (or higher) only has increased recently -->
+
+![Number of museum records for marine invertebrates](figures/map_sampling_effort.pdf)
 
@@
 
-![Relation between the number of specimen and the number of species collected](figures/plot_sampling_effort.png)
+![Relation between the number of specimen and the number of species collected](figures/plot_sampling_effort.pdf)
 
-![Proportion of species occuring in natural history collections that have been DNA barcoded](figures/proportion_barcoded.pdf)
+![Cumulative number of samples through time](figures/cum_samples_through_time.pdf)
+
+![Cumulative number of species recorded for the first time through time](figures/cum_spp_through_time.pdf)
+
+![Proportion of species occurring in natural history collections that have been DNA barcoded](figures/proportion_barcoded.pdf)
+
+![Level of digitization for the Salish Sea](figures/plot_richness_per_db_koz.pdf)
+
+![Level of digitization for the Gulf of Mexico](figures/plot_richness_per_db_gom.pdf)
 
@@
 
-While plankton is recognized for its biological importance (dispersal, food
-source for many species), and can be used as an ecological indicator for
-pollution or global change, identification of planktonic larvae at the species
-level from morphological information is generally not possible because of the
-paucity of informative characters to base the identifications on, and the lack
-of knowledge of the appearance of the larval stages for most species. DNA
-barcoding can bridge this knowledge gap by using genetic information to identify
-larvae at the species level. These identifications are only possible if the
-species have already been sequenced, and the sequences deposited in public
-repositories. The accuracy of the identification relies on the expertise of the
-taxonomists who identified the specimens that have been sequenced.
+While plankton is recognized for its biological importance (dispersal, food source for many species), and can be used as an ecological indicator for pollution or global change, identification of planktonic larvae at the species level from morphological information is generally not possible because of the paucity of informative characters to base the identifications on, and the lack of knowledge of the appearance of the larval stages for most species.
+DNA barcoding can bridge this knowledge gap by using genetic information to identify larvae at the species level.
+These identifications are only possible if the species have already been sequenced, and the sequences deposited in public repositories.
+The accuracy of the identification relies on the expertise of the taxonomists who identified the specimens that have been sequenced.
 
-We collected over the course of a year planktonic larvae of marine invertebrates
-in a coastal marsh of the Atlantic coast of Florida. One of our goals was to
-identify these larvae at the species level using DNA barcoding. We photographed
-and sequenced xxx larvae, representing an estimated xxx species. These data are
-available from BOLD and can be browsed interactively at
-http://plankton-atlas.net. Using records from public databases (BOLD, GenBank),
-we could confidently assign only xxx% to species that had already been
-sequenced. On average, these taxa were sequenced xxx km away from our study
-site.
+We collected over the course of a year planktonic larvae of marine invertebrates in a coastal marsh of the Atlantic coast of Florida.
+One of our goals was to identify these larvae at the species level using DNA barcoding.
+We photographed and sequenced xxx larvae, representing an estimated xxx species.
+These data are available from BOLD and can be browsed interactively at http://plankton-atlas.net.
+Using records from public databases (BOLD, GenBank), we could confidently assign only xxx% to species that had already been sequenced.
+On average, these taxa were sequenced xxx km away from our study site.
 
-Florida is one of the most populated states, with a long tradition of studying
-its marine life.  xxx universities have marine research programs, and xxx marine
-laboratories. How does this proportion of xxx% identified species compare with
-other places in Florida, the East coast of the US, the continental US, and the
-world. In other words, how much of the species of marine invertebrates we know
-about have publicly available DNA barcodes?
+Florida is one of the most populated states, with a long tradition of studying its marine life.
+xxx universities have marine research programs, and xxx marine laboratories.
+How does this proportion of xxx% identified species compare with other places in Florida, the East coast of the US, the continental US, and the world.
+In other words, how much of the species of marine invertebrates we know about have publicly available DNA barcodes?
 
-To estimate this number, we assembled species lists of marine invertebrates for
-the continental US by querying iDigBio for each marine phylum. We only kept
-records for specimens identified with a valid binomial name (specimens
-identified at the genus level or higher were excluded from the analysis). We
-used WoRMS to ensure that the species names were valid, and to disambiguate
-possible synonyms. For each species, we retrieved avaiable DNA barcoding
-sequence data from BOLD, which also includes DNA barcodes deposited in
-GenBank. Of the xxx marine invertebrate species found in iDigBio, xxx% had DNA
-barcodes available in public databases. These proportions varied by phylum, but
-for the conspicuous and better studied groups (Porifera, Cnidaria, Mollusca,
-Echinodermata), they were surprisingly consistent, and is a close estimate to
-the proportion of species that have been sequenced for the United
-States. Furthermore, we also looked for the proportion of species known to occur
-in the Indian River Lagoon for which DNA barcodes are available, and again the
-proportion of sequenced species (xxx%) is comparable.
+To estimate this number, we assembled species lists of marine invertebrates for the continental US by querying iDigBio for each marine phylum.
+We only kept records for specimens identified with a valid binomial name (specimens identified at the genus level or higher were excluded from the analysis).
+We used WoRMS to ensure that the species names were valid, and to disambiguate possible synonyms.
+For each species, we retrieved avaiable DNA barcoding sequence data from BOLD, which also includes DNA barcodes deposited in GenBank.
+Of the xxx marine invertebrate species found in iDigBio, xxx% had DNA barcodes available in public databases.
+These proportions varied by phylum, but for the conspicuous and better studied groups (Porifera, Cnidaria, Mollusca, Echinodermata), they were surprisingly consistent, and is a close estimate to the proportion of species that have been sequenced for the United States.
+Furthermore, we also looked for the proportion of species known to occur in the Indian River Lagoon for which DNA barcodes are available, and again the proportion of sequenced species (xxx%) is comparable.
 
-Except for the few phyla that are known to include only a small number of
-species (Cycliophora, Phoronida, Chaetognatha), microscopic groups, typically
-component of the meiofauna, are disproportionally under-represented in museum
-collections, and none or few species are available in public databases.
+Except for the few phyla that are known to include only a small number of species (Cycliophora, Phoronida, Chaetognatha), microscopic groups, typically component of the meiofauna, are disproportionally under-represented in museum collections, and none or few species are available in public databases.
 
-Our results highlight that most of our knowledge of marine biodiversity come
-from large expeditions carried in the late 19th century and after World War
-II. Since then, sampling effort has dramatically decreased, and taxonomic
-expertise is waning. Yet, our ability to assign DNA sequences to species names
-will greatly improve the value of the data generated by meta-barcoding studies.
+Our results highlight that most of our knowledge of marine biodiversity come from large expeditions carried in the late 19th century and after World War II.
+Since then, sampling effort has dramatically decreased, and taxonomic expertise is waning.
+Yet, our ability to assign DNA sequences to species names will greatly improve the value of the data generated by meta-barcoding studies.
 
-and reflect that a large fraction of the diversity of
-marine invertebrates remain to be sequenced.
+and reflect that a large fraction of the diversity of marine invertebrates remain to be sequenced.
 
-Marine invertebrates are the biodiversity. All but one metazoan phyla occur in
-the sea. Of the xx described species, yy are marine invertebrates. They
-represent a multiple array of shapes, forms, and adaptations to all marine
-environments. Yet, they are proportionally understudied. Here, we analyze
-publicly available data from biodiversity databases to demonstrate that the
-abandonment of traditional taxonomic practices, the lack of support for natural
-history collections, is hampering our ability to use marine invertebrates as
-indicators of ecosystem health, the effect of global change, the identification
-of model organisms, and ... Technological advances are however making easier to
-identify knowledge gaps, the potential to identify where new species occur, and
-document diversity.
+Marine invertebrates are the biodiversity.
+All but one metazoan phyla occur in the sea.
+Of the xx described species, yy are marine invertebrates.
+They represent a multiple array of shapes, forms, and adaptations to all marine environments.
+Yet, they are proportionally understudied.
+Here, we analyze publicly available data from biodiversity databases to demonstrate that the abandonment of traditional taxonomic practices, the lack of support for natural history collections, is hampering our ability to use marine invertebrates as indicators of ecosystem health, the effect of global change, the identification of model organisms, and ...
+Technological advances are however making easier to identify knowledge gaps, the potential to identify where new species occur, and document diversity.
 
@@
 
-The use of molecular data for species identification and delineation is arguably
-one of the most important advances in taxonomy. Especially with DNA barcoding
-that proposes to use a common genetic marker across large taxonomic groups (the
-mt marker COI for animals, ITS for fungi, and xx for plants). Because these
-markers have typically low intra-specific variation, and differ noticeably
-between even closely related species, they can be used to identify species and
-unravel species complexes.
+The use of molecular data for species identification and delineation is arguably one of the most important advances in taxonomy.
+Especially with DNA barcoding that proposes to use a common genetic marker across large taxonomic groups (the mt marker COI for animals, ITS for fungi, and xx for plants).
+Because these markers have typically low intra-specific variation, and differ noticeably between even closely related species, they can be used to identify species and unravel species complexes.
 
-Species identification relies on reference databases, where sequences are
-associated with accurately identified individuals, held in natural history
-collections. When taxonomic sampling is comprehensive, and enough individuals of
-each species are sampled to be representative of the standing genetic variation
-in the population, DNA barcoding can be a very accurate tool with xx-yy% error
-rate.
+Species identification relies on reference databases, where sequences are associated with accurately identified individuals, held in natural history collections.
+When taxonomic sampling is comprehensive, and enough individuals of each species are sampled to be representative of the standing genetic variation in the population, DNA barcoding can be a very accurate tool with xx-yy% error rate.
 
-The utility of these reference databases is getting revived with the development
-of meta-barcoding techniques, where thousands of species can be sequenced in
-parallel on high-throughput sequencing platforms. If methods exist to estimate
-the number of species these samples contain based on the level of genetic
-divergence observed among the sequences, being able to match these sequences to
-identified species based on their match to a sequence in a public database would
-increase the potential applications for these approaches.
+The utility of these reference databases is getting revived with the development of meta-barcoding techniques, where thousands of species can be sequenced in parallel on high-throughput sequencing platforms.
+If methods exist to estimate the number of species these samples contain based on the level of genetic divergence observed among the sequences, being able to match these sequences to identified species based on their match to a sequence in a public database would increase the potential applications for these approaches.
 
-Marine invertebrates are however disproportionally under-represented in DNA
-sequence databses. And they have a higher rate of misidentification. There are
-however some pitfalls to this number. It might actually be even higher given
-that they have been identified by a single taxonomist given the lack of
-redundancy in expertise.
+Marine invertebrates are however disproportionally under-represented in DNA sequence databses.
+And they have a higher rate of misidentification.
+There are however some pitfalls to this number.
+It might actually be even higher given that they have been identified by a single taxonomist given the lack of redundancy in expertise.
 
-To illustrate the consequences of having so few barcodes for invertebrates, if
-we take the faunistic list of IRL, only about 20% of the species known to be
-occuring in the area actually have DNA sequences available for them. On average,
-each species has been sequenced xx times, and from locations on average yy km
-away from IRL. Given the high propencity for cryptic species among marine
-invertebrates, it is possible that ...
+To illustrate the consequences of having so few barcodes for invertebrates, if we take the faunistic list of IRL, only about 20% of the species known to be occuring in the area actually have DNA sequences available for them.
+On average, each species has been sequenced xx times, and from locations on average yy km away from IRL.
+Given the high propencity for cryptic species among marine invertebrates, it is possible that ...
 
-We also sequenced xx larvae occurring in the plankton at the Whitney Marine
-Lab, representing yy species (estimated from genetic distances). Of these yy
-species, only zz% had good matches in BOLD/GenBank. In both cases, taxa1, taxa2,
-and taxa3 are the most underrepresented in BOLD.
+We also sequenced xx larvae occurring in the plankton at the Whitney Marine Lab, representing yy species (estimated from genetic distances).
+Of these yy species, only zz% had good matches in BOLD/GenBank.
+In both cases, taxa1, taxa2, and taxa3 are the most underrepresented in BOLD.
 
@@
 
-Total number of collections in the US, how many have marine invertebrate
-collections, how many have active curators.
+Total number of collections in the US, how many have marine invertebrate collections, how many have active curators.
 
-We don't know how many species are in the sea. Estimates vary widely. Difficult
-to sample because diversity of habitats, each requiring special sampling
-techniques. Meanwhile, taxonomic expertise is decreasing despite efforts to keep
-it alive (PEET).
+We don't know how many species are in the sea.
+Estimates vary widely.
+Difficult to sample because diversity of habitats, each requiring special sampling techniques.
+Meanwhile, taxonomic expertise is decreasing despite efforts to keep it alive (PEET).
 
-To get around issue of sampling different habitat, sample coastal
-meroplankton. Most species go through planktonic phase, therefore common habitat
-to sample them all. The issue is that they look very different from adults, and
-only DNA can be used to identify them.
+To get around issue of sampling different habitat, sample coastal meroplankton.
+Most species go through planktonic phase, therefore common habitat to sample them all.
+The issue is that they look very different from adults, and only DNA can be used to identify them.
 
-Barcoding is not perfect, but can be very powerful to identify the larval forms
-(e.g. one of the first paper was sampling caterpillars). Potentially differences
-in larva that are not seen in adults and could be additional characters to tell
-species appart.
+Barcoding is not perfect, but can be very powerful to identify the larval forms (e.g.
+one of the first paper was sampling caterpillars).
+Potentially differences in larva that are not seen in adults and could be additional characters to tell species appart.
 
-Advantage of sampling plankton is that a single habitat is sampled. Not enough
-known about heterogeneity in spatial distribution.
+Advantage of sampling plankton is that a single habitat is sampled. Not enough known about heterogeneity in spatial distribution.
 
-We sampled plankton bi-weekly for a year, and sequenced the larvae
-found. Sequenced the barcoding gene and tried to match how many species we could
-identify based on records in public databases. Overall, low percentage of
-match. And matches also led to several species that have been collected but that
-have not been identified to an appropriate level.
+We sampled plankton bi-weekly for a year, and sequenced the larvae found.
+Sequenced the barcoding gene and tried to match how many species we could identify based on records in public databases.
+Overall, low percentage of match.
+And matches also led to several species that have been collected but that have not been identified to an appropriate level.
 
-To compare the results we obtained here, and see how they could compare to other
-habitats, we downloaded the species list from Indian River lagoon inventory. It
-is the most comprehensive checklist in Florida, and has been developed over many
-years. We asked how many of the species occurring in the Indian River Lagoon have
-already been sequenced. We did the matching based on the species names. If we
-find higher proportion of sequenced species, it means that plankton gives access
-to species that are difficult to sample as adults. If lower proportion, it means
-that we don't have many of the adults sequenced at Whitney.
+To compare the results we obtained here, and see how they could compare to other habitats, we downloaded the species list from Indian River lagoon inventory.
+It is the most comprehensive checklist in Florida, and has been developed over many years.
+We asked how many of the species occurring in the Indian River Lagoon have already been sequenced.
+We did the matching based on the species names.
+If we find higher proportion of sequenced species, it means that plankton gives access to species that are difficult to sample as adults.
+If lower proportion, it means that we don't have many of the adults sequenced at Whitney.
 
-Using plankton to sample biodiversity could lead to a speedup of
-documentation. It would allow the identification of taxa/habitats that need
-special attention. If representative of adult communities, could also be a
-useful tool to monitor status of ecosystem through time.
+Using plankton to sample biodiversity could lead to a speedup of documentation.
+It would allow the identification of taxa/habitats that need special attention.
+If representative of adult communities, could also be a useful tool to monitor status of ecosystem through time.
 
-Better database, can also be used to estimate the proportion of really dark
-planktonic taxa.
+Better database, can also be used to estimate the proportion of really dark planktonic taxa.
 
@@
 
-Biodiversity is in the sea. All phyla but one. New lineages of eukaryotes,
-viruses...
+Biodiversity is in the sea. All phyla but one. New lineages of eukaryotes, viruses...
 
-We are at a time where it becomes cheaper and easier to delineate species. To
-sort through specimens to identify species. So taxonomists can focus on
-studying/describing new species.
+We are at a time where it becomes cheaper and easier to delineate species. To sort through specimens to identify species. So taxonomists can focus on studying/describing new species.
 
-Granularity of data not sufficient to be able to detect changes in the
-environment, shifts in species ranges, changes in trophic structure. Understand
-effect of small animals.
+Granularity of data not sufficient to be able to detect changes in the environment, shifts in species ranges, changes in trophic structure. Understand effect of small animals.
 
-Rely on high school students, citizen science initiatives to help with labor
-intensive part.
+Rely on high school students, citizen science initiatives to help with labor intensive part.
 
@@
 
-Strong focus on vertebrates. We are building theories and making conservation
-approaches on a small fraction of the tree of life.
+Strong focus on vertebrates. We are building theories and making conservation approaches on a small fraction of the tree of life.
 
@@
 
+Identified specimens from museum collections could be used to complete these datasets.
+Traditional approaches relying on Sanger sequencing might be limiting as many species have not been collected recently enough to guarantee good quality DNA (` prop_species_not_collected_since_1990`% have not been collected since 1990).
+However, high-throughput sequencing could offer alternatives to recover sequences from older specimens.
+
+Besides data-mining museum records to identify knowledge gaps, new sampling approaches can be used in the field to rapidly assess how much of the fauna is known.
+Meta-barcoding and environmental DNA approaches appear promising to sample _en masse_ species that are found in a given location.
+Because most marine invertebrates have a planktonic larval stage, they may co-occur in the water column, allowing to sample species across habitats.
+To illustrate this approach, we individually DNA barcoded ` n_larvae` planktonic larvae for DNA barcoding, collected over the course of year, representing ` n_esus` species, collected in coastal waters of Florida.
+We found that (` format_output(100 * prop_plankton_with_match)`%, Fig.
+4) can be assigned to a previously sequenced species.
+Similar proportions have been obtained from individually barcoded species that occur on ARMS.
+However, this proportion dropped to less than 5% for smaller organisms sequenced with meta-barcoding.
+
+The species identifications in BOLD are far from perfect.
+Of the ` sum(prop_plankton_with_species_match_by_phylum$n_esu_match)` species that have a match in public database, ` format_output(prop_plankton_with_multi_species_match * 100)`% have been identified as more than one taxon in BOLD.
+A few of these records might represent species complexes, but most illustrate the difficulty to identify precisely marine invertebrates.
+
 --->

### 2017-03-13 manuscript.Rmd
edit manuscript

@@
 #author: Fran\c{c}ois Michonneau
-title: "Data-driven approaches to identifying knowledge gaps in marine invertebrates biodiversity"
+title: "Data-driven approaches to identifying knowledge gaps in marine invertebrate biodiversity"
 output:
@@
 
-With increasing impact of human activities on the biosphere, we need efficient and accurate approaches to document changes in species ranges, detect changes in the composition of communities, and more generally accelerate the documentation of species diversity.
-This is especially pressing in marine ecosystems where a significant portion of the biodiversity remains to be discovered and described.
+With increasing impact of human activities on the biosphere, we need efficient and accurate approaches to document changes in species ranges and community composition.
+While terrestrial ecosystems have received considerable attention, marine ecosystems are much less understood, and what constitutes the baseline of an healthy and resilient system is not fully characterized.
+Even basic information such as the number of species that live in the sea is not agreed on, and a significant portion of marine biodiversity remains to be discovered and described.
+There is therefore a pressing need to improve our knowledge of this environment before species go extinct and its functionning is irreversibly altered.
+We also have an heterogenous understanding of marine species distributions which limits our understanding of population sizes and connectivity, in turn making the modelization of marine resources and ecosystem resilience less accurate.
+Until recently, this knowledge was difficult to synthesize because it was scattered and difficult to access.
+However, recent advances in digitization and public databases efforts have made available the content of natural history collections, information about species occurrences, and genetic data.
+Given the urgency with which we need to document biodiversity and the limited means available to do so, we argue that we need to build on our existing knowledge to identify gaps, and we need to be more efficient at collecting information to improve our understanding of the natural world.
 
-Until recently our knowledge of species and their distributions was difficult to synthesize because it was scattered and difficult to access.
-However, recent advances in digitization and public databases efforts have made widely available the content of natural history collections, publications about species, their geographical distributions, and their genetic data.
-Given the urgency with which we need to document biodiversity and the limited means available to do so, we need to use and build on this existing knowledge to identify gaps, and to be more efficient at collecting information and understanding our natural world.
-
-The incorporation of molecular data in the study of biodiversity has highlighted the pervasiveness of cryptic species, and generally the fact that species have smaller ranges than their taxonomic identity implied.
-Notably, the wide adoption of DNA barcoding for the study of marine diversity has clarified our understanding of species limits in many groups, but has also enabled the identification of species during their larval stages (when the lack of morphological features make species identification difficult).
-DNA barcoding has also been seen as a possible solution to the widening gap between the increased need for biodiversity inventories while taxonomic expertise is dwindling.
-By matching DNA barcodes of unidentified species to a public database of accurately identified and vouchered specimens, one can greatly speed up the process of species identification by matching them to already identified species.
+The incorporation of molecular data into the study of biodiversity has highlighted the pervasiveness of cryptic species, and contributed to the recognition that species have smaller ranges than their taxonomic identity had implied.
+Notably, despite early criticisms, the wide adoption of DNA barcoding to study marine diversity has invigorated taxonomy.
+It was also suggested as a possible solution to the widening gap between the increased need for biodiversity inventories while taxonomic expertise is dwindling.
+By matching DNA barcodes of unidentified species to a reference public database of accurately identified and vouchered specimens, DNA barcoding can greatly speed up the process of species identification by matching them to already identified species.
 Sequences with no match make it easier to flag specimens that warrant further taxonomic attention.
 
+<!-- clarified our understanding of species limits in many groups, but has also enabled the identification of species during their larval stages (when the lack of morphological features make species identification difficult).
+DNA barcoding has also been seen as -->
+
 Recent developments in high-throughput sequencing technologies are giving new applications to these DNA barcode databases.
-Meta-barcoding enables the parallel sequencing of thousands of DNA barcodes for species that occur in environmental samples.
-This approach opens new research venues as it becomes possible to estimate the composition of species-rich samples that were virtually inaccessible before (e.g., gut content, marine sediment samples, plankton tows).
-For instance, by sequencing the plankton of coastal areas, it becomes possible to sample the larval stages of many species that would otherwise be difficult to sample.
+Metabarcoding has been developed to sequence in parallel the "barcoding gene" across the species included in environmental samples such as plankton tows, fish guts, or artificial structure scrappings.
 However, to accurately identify the species found in these samples, comprehensive DNA barcode databases are needed.
 
-With the cost of sequencing going down, and the prospect of having an automated sampling process, these technologies will soon be used for ecosystem and biodiversity monitoring.
-However, to interpret the results and understand how species distributions shift with global change or other perturbations, we cannot only rely on the identity of the species, but we also need to take advantage of existing knowledge about these species that is coming from natural history collections.
-In any case, the insights we will gain from these approaches are only going to be as good as the level of completeness and accuracy of the information included public databases.
+With decreasing sequencing costs, and the prospect of automating sample collection, these technologies will soon be used for ecosystem and biodiversity monitoring.
+DNA barcoding provides the possibility to link these newly generated data with historical natural history knowledge about species and their distributions.
+However, the insights we will gain from these approaches are only going to be as good as the level of completeness and accuracy of the information included in  reference databases.
 
-<!-- here there are 3 things: data availability (having good APIs to access the -->
-<!-- data); data comprehensiveness (how many species are captured in the -->
-<!-- database); data accuracy (can we trust the data that is found in the -->
-<!-- databases) -->
+<!-- here there are 3 things: data availability (having good APIs to access the data); data comprehensiveness (how many species are captured in the  database); data accuracy (can we trust the data that is found in the databases) -->
+
+Our knowledge about species and their distributions is gathered in regional faunistic lists.
+They are typically assembled by taxonomic experts.
+They are comprehensive but usually limited in their geographic scope.
+Natural history collections have more comprehensive geographic coverage, but the quality of the taxonomic identifications can vary.
 
 Here, we assess the level of completeness of biodiversity public databases for marine invertebrates that occur within the boundaries of the EEZ of the US.
-Specifically, we use comprehensive regional faunistic lists to assess how much of this knowledge is captured by current digitization efforts of museum collections and occurrence records; and we estimate how many of the species have genetic data publicly available that would be useful in the context of identifying samples from meta-barcoding studies.
+Specifically, we use comprehensive regional faunistic lists to assess how much of this knowledge is captured by current digitization efforts of museum collections and occurrence records; and we estimate how many of the species have genetic data publicly available that would be useful in the context of identifying samples from metabarcoding studies.
 
@@
 
-We combined `r nrow(idigbio_records)` museum records for marine invertebrates collected in the continental US representing `r length(unique(tolower(idigbio_records[["data.dwc:phylum"]])))` phyla and `r length(unique(idigbio_records$worms_valid_name))` species.
+We combined `r nrow(idigbio_records)` museum records for marine invertebrates collected in the EEZ waters of the continental US representing `r length(unique(tolower(idigbio_records[["data.dwc:phylum"]])))` phyla and `r length(unique(idigbio_records$worms_valid_name))` species.
 Most samples were collected on the continental shelf (Fig.

### 2017-03-16 manuscript.Rmd
edit manuscript

@@
 With increasing impact of human activities on the biosphere, we need efficient and accurate approaches to document changes in species ranges and community composition.
-While terrestrial ecosystems have received considerable attention, marine ecosystems are much less understood, and what constitutes the baseline of an healthy and resilient system is not fully characterized.
-Even basic information such as the number of species that live in the sea is not agreed on, and a significant portion of marine biodiversity remains to be discovered and described.
-There is therefore a pressing need to improve our knowledge of this environment before species go extinct and its functionning is irreversibly altered.
-We also have an heterogenous understanding of marine species distributions which limits our understanding of population sizes and connectivity, in turn making the modelization of marine resources and ecosystem resilience less accurate.
-Until recently, this knowledge was difficult to synthesize because it was scattered and difficult to access.
-However, recent advances in digitization and public databases efforts have made available the content of natural history collections, information about species occurrences, and genetic data.
-Given the urgency with which we need to document biodiversity and the limited means available to do so, we argue that we need to build on our existing knowledge to identify gaps, and we need to be more efficient at collecting information to improve our understanding of the natural world.
+While terrestrial ecosystems have received considerable attention, the baseline of healthy and resilient marine ecosystems are not well understood.
+A major hinderance is that we do not have enough information about which species live where, and there is potentially a large portion of marine biodiversity that remains to be discovered and described.
+Additionally, marine ecosystem monitoring are lacking to detect, and even less understand, how and why communities change through time.
+There is therefore a pressing need to improve our knowledge of marine systems before species go extinct, and their functionning are irreversibly altered.
+Our lack of knowledge about marine invertebrates make the modelisation of trophic networks less accurate, as diverse species assemblages are often reduced as simple functional categories that do no account for their ecological differences.
+Until very recently, our knowledge of biodiversity was scattered among heterogeneous sources (e.g.
+natural history collections, publications, expert knowledge), and difficult to access.
+This heterogeneity and the difficulty to accessing this data have made synthesis and the identification of knowledge gaps difficult.
+However, with the development of linked, publicly accessible, comprehensive databases of natural history records (e.g.
+iDigBio), species occurrences (e.g., GBIF, OBIS), genetic data (e.g., GenBank, BOLD), and registers of species names (e.g., WoRMS), it is now easier to quantitavely assess the state of our knowledge for biodiversity data.
+Here, we argue that given the urgency of the biodiversity crisis, we need to develop a data-driven approach to speed up the documentation of biodiversity and build on our existing knowledge to be more efficient at collecting information to improve our understanding of our natural world.
+Additionally, we argue that our understanding biodiversity in the marine environment and the impact of human activities on communities requires not only to get a more accurate list of the species that inhabits the oceans, but also to understand better their distributions.
+This can only be done with additional sampling and a commitment to training more taxonomists and experts that can identify the samples collected.
+To some extant, we are at a time when it has never been easier and cost effective to survey comprehensively the extant of marine biodiversity.
 
-The incorporation of molecular data into the study of biodiversity has highlighted the pervasiveness of cryptic species, and contributed to the recognition that species have smaller ranges than their taxonomic identity had implied.
-Notably, despite early criticisms, the wide adoption of DNA barcoding to study marine diversity has invigorated taxonomy.
-It was also suggested as a possible solution to the widening gap between the increased need for biodiversity inventories while taxonomic expertise is dwindling.
-By matching DNA barcodes of unidentified species to a reference public database of accurately identified and vouchered specimens, DNA barcoding can greatly speed up the process of species identification by matching them to already identified species.
+Biodiversity discovery and documentation has been transformed in the 21st century by using molecular data to facilitate species identification, delineate species limits, and more recently to quantify diversity and uncover large scale patterns of biodiversity organization.
+The wide adoption of DNA barcoding (where homologous DNA sequences are sequenced across the main domains of the Eukaryotes) has unraveled the pervasiveness of cryptic species, invigorating taxonomy and the description of new species since the early 2000s.
+Unraveling these cryptic species has also contributed to the recognition that species have smaller ranges than their taxonomic identity had implied making them potentially more susceptible to extinction or overexploitation.
+And despite early criticisms, DNA barcoding has also been suggested as a possible solution to the widening gap between the increased need to document biodiversity while taxonomic expertise is dwindling.
+Thus, large databases of DNA barcodes have been collaboratively assembled and made pubicly available.
+By matching DNA barcodes of unidentified species to a reference public database of accurately identified and vouchered specimens, DNA barcoding can greatly speed up species identification.
 Sequences with no match make it easier to flag specimens that warrant further taxonomic attention.
+
+Barcoding/metabarcoding certainly not perfect methods, but much better than current understanding.
 
@@
 Recent developments in high-throughput sequencing technologies are giving new applications to these DNA barcode databases.
-Metabarcoding has been developed to sequence in parallel the "barcoding gene" across the species included in environmental samples such as plankton tows, fish guts, or artificial structure scrappings.
-However, to accurately identify the species found in these samples, comprehensive DNA barcode databases are needed.
-
+Metabarcoding has been developed to sequence in parallel the barcoding gene across the species included in environmental samples such as plankton tows, fish guts, or artificial structure scrappings.
 With decreasing sequencing costs, and the prospect of automating sample collection, these technologies will soon be used for ecosystem and biodiversity monitoring.
@@
 However, the insights we will gain from these approaches are only going to be as good as the level of completeness and accuracy of the information included in reference databases.
+However, to accurately identify the species found in these samples, comprehensive DNA barcode databases are needed.
+Here, we focus mostly on museum specimens because we can double check their identification and we can potentially extract DNA from them to complete existing databases of DNA barcodes.
 
-<!-- here there are 3 things: data availability (having good APIs to access the data); data comprehensiveness (how many species are captured in the  database); data accuracy (can we trust the data that is found in the databases) -->
+<!-- here there are 3 things: data availability (having good APIs to access the
+data); data comprehensiveness (how many species are captured in the database);
+data accuracy (can we trust the data that is found in the databases) -->
 
@@
 Typically a higher proportion of species from the Pacific Coast than the Gulf of Mexico are not represented.
-Comparatively, all but 2 of the `r nrow(mammals_species_asm)` mammal species known to occur in the US have digitized museum specimens.
+Comparatively, all but 2 of the `r nrow(mammals_species_asm)` mammal species and `r sum(amphibian_summary$count == 0)` of the `r nrow(amphibian_summary)` amphibian species known to occur in the US have digitized museum specimens.
 
@@
 Therefore specimens identified under the same species but sampled in distant locations, could represent cryptic species.
-It is thus crucial to ensure that accurate geographic information is provided when DNA barcodes are published, and that additional samples are collected, to ensure that the extent of their genetic diversity is captured.
+It is thus crucial to ensure that accurate geographic information is provided when DNA barcodes are published, and that additional samples are collected, to capture more accurately their geographic range and the extent of their genetic diversity.
 This high level of cryptic diversity is reflected in the fact that `r stat_bold$n_spp_muti_bin` species out of `r stat_bold$n_spp` include more than 1 BIN.
@@
 
+<!-- Paragraph on how sampling effort has changed through time. -->
+
+<!-- - get numbers of species described by year, look if there is a correlation between sample numbers and number of new species (possible lag with time for description) -->
+<!-- - see if proportion of species identified at genus (or higher) only has increased recently -->
+
 Using metabarcoding is a promising approach to screen an habitat for its potential for species that are not found in public database, to maximize the usefulness in the future of the data generated by an expedition.
@@
 
-<!-- Paragraph on how sampling effort has changed through time. -->
-
-<!-- - get numbers of species described by year, look if there is a correlation between sample numbers and number of new species (possible lag with time for description) -->
-<!-- - see if proportion of species identified at genus (or higher) only has increased recently -->
+![Number of species of marine invertebrates](figures/map_diversity.pdf)
 
@@
 
+![Relation between the number of specimen and the number of species collected](figures/plot_sampling_effort.pdf)
+
 ![Level of digitization for the Salish Sea](figures/plot_richness_per_db_koz.pdf)
@@
 <!--
-
-The number of species that occurs in the sea remains uncertain.
 

### 2017-04-04 manuscript.Rmd
edit manuscript

@@
 With increasing impact of human activities on the biosphere, we need efficient and accurate approaches to document changes in species ranges and community composition.
-While terrestrial ecosystems have received considerable attention, the baseline of healthy and resilient marine ecosystems are not well understood.
-A major hinderance is that we do not have enough information about which species live where, and there is potentially a large portion of marine biodiversity that remains to be discovered and described.
-Additionally, marine ecosystem monitoring are lacking to detect, and even less understand, how and why communities change through time.
+While terrestrial ecosystems have received considerable attention, the baseline of healthy and resilient marine ecosystems are not well understood, and modeling the impact of human activities is limited to a few taxonomic groups.
+A major hinderance is that we do not have enough information about which species live where, and an unknown but potentially significant portion of marine biodiversity remains to be discovered and described.
+Additionally, marine ecosystem monitoring initiatives are lacking to detect and understand, how and why communities change through time [@Duffy2013,@Borja2016].
 There is therefore a pressing need to improve our knowledge of marine systems before species go extinct, and their functionning are irreversibly altered.
-Our lack of knowledge about marine invertebrates make the modelisation of trophic networks less accurate, as diverse species assemblages are often reduced as simple functional categories that do no account for their ecological differences.
+Our lack of knowledge about marine invertebrates make the modelisation of trophic networks less accurate, as diverse species assemblages are often reduced as simple functional categories that do no account for their ecological diversity.
 Until very recently, our knowledge of biodiversity was scattered among heterogeneous sources (e.g.
 natural history collections, publications, expert knowledge), and difficult to access.
-This heterogeneity and the difficulty to accessing this data have made synthesis and the identification of knowledge gaps difficult.
-However, with the development of linked, publicly accessible, comprehensive databases of natural history records (e.g.
-iDigBio), species occurrences (e.g., GBIF, OBIS), genetic data (e.g., GenBank, BOLD), and registers of species names (e.g., WoRMS), it is now easier to quantitavely assess the state of our knowledge for biodiversity data.
-Here, we argue that given the urgency of the biodiversity crisis, we need to develop a data-driven approach to speed up the documentation of biodiversity and build on our existing knowledge to be more efficient at collecting information to improve our understanding of our natural world.
-Additionally, we argue that our understanding biodiversity in the marine environment and the impact of human activities on communities requires not only to get a more accurate list of the species that inhabits the oceans, but also to understand better their distributions.
-This can only be done with additional sampling and a commitment to training more taxonomists and experts that can identify the samples collected.
-To some extant, we are at a time when it has never been easier and cost effective to survey comprehensively the extant of marine biodiversity.
+This heterogeneity and the difficulty to access this data have made synthesis and the identification of knowledge gaps difficult.
+However, with the development of linked, public, comprehensive databases of natural history records (e.g.
+iDigBio), species occurrences (e.g., GBIF, OBIS), genetic data (e.g., GenBank, BOLD), and registers of species names (e.g., WoRMS), it is now easier to quantitavely evaluate the state of our knowledge for biodiversity data.
+Here, we aggregate publicly available data for marine invertebrates that occur in the United States to highligth gaps in our knowledge.
+Given the urgency of the biodiversity crisis, we need to develop data-driven initiatives to efficiently document biodiversity, and build on our existing knowledge to collect information to improve our understanding of marine systems.
+Additionally, we emphasize that our understanding of marine biodiversity, and the impact of human activities on communities require not only to get a more accurate list of the species that inhabits the oceans, but we also need better data on species distributions and abundances.
+This can only be achieved with additional sampling and a commitment to training more taxonomists and experts that can identify the material collected.
+We live at a time when it has never been easier and cost effective to survey comprehensively the extant of marine biodiversity.
 
-Biodiversity discovery and documentation has been transformed in the 21st century by using molecular data to facilitate species identification, delineate species limits, and more recently to quantify diversity and uncover large scale patterns of biodiversity organization.
-The wide adoption of DNA barcoding (where homologous DNA sequences are sequenced across the main domains of the Eukaryotes) has unraveled the pervasiveness of cryptic species, invigorating taxonomy and the description of new species since the early 2000s.
+Biodiversity discovery and documentation has been transformed in the 21st century by using molecular data to facilitate species identification, delineate species.
+And more recently, the development of high throughput sequencing technologies is used to quantify diversity and uncover large scale patterns of biodiversity organization.
+The wide adoption of DNA barcoding (where homologous DNA sequences are sequenced across the main domains of the Eukaryotes) has unraveled the pervasiveness of cryptic species, invigorating taxonomy and the description of new species, and enabled collaborative research programs such as the Census of Marine Life.
 Unraveling these cryptic species has also contributed to the recognition that species have smaller ranges than their taxonomic identity had implied making them potentially more susceptible to extinction or overexploitation.
-And despite early criticisms, DNA barcoding has also been suggested as a possible solution to the widening gap between the increased need to document biodiversity while taxonomic expertise is dwindling.
+Despite early criticisms and now well-understood limitations, DNA barcoding has been suggested as a possible solution to the widening gap between the increased need to document biodiversity while taxonomic expertise is dwindling.
 Thus, large databases of DNA barcodes have been collaboratively assembled and made pubicly available.
@@
 
-Barcoding/metabarcoding certainly not perfect methods, but much better than current understanding.
-
 <!-- clarified our understanding of species limits in many groups, but has also enabled the identification of species during their larval stages (when the lack of morphological features make species identification difficult).
@@
 
-Recent developments in high-throughput sequencing technologies are giving new applications to these DNA barcode databases.
-Metabarcoding has been developed to sequence in parallel the barcoding gene across the species included in environmental samples such as plankton tows, fish guts, or artificial structure scrappings.
-With decreasing sequencing costs, and the prospect of automating sample collection, these technologies will soon be used for ecosystem and biodiversity monitoring.
+Recent developments in high-throughput sequencing technologies are giving new applications to these DNA barcode databases [@Leray2016].
+Metabarcoding has been developed to sequence in parallel the barcoding gene across the species included in environmental samples such as plankton tows, fish guts, or artificial structure scrappings [@Leray2015].
+With decreasing sequencing costs, and the prospect of automating sample collection, these technologies will soon be used routinely for ecosystem and biodiversity monitoring [@Bohan2017].
 DNA barcoding provides the possibility to link these newly generated data with historical natural history knowledge about species and their distributions.
@@
 However, to accurately identify the species found in these samples, comprehensive DNA barcode databases are needed.
-Here, we focus mostly on museum specimens because we can double check their identification and we can potentially extract DNA from them to complete existing databases of DNA barcodes.
+Here, we mostly focus on museum specimens because we can validate their identification, and a fraction of these samples (if they have not been preserved in formalin) can be used to obtain DNA from them to complete existing databases of DNA barcodes.
+
+<!-- Barcoding/metabarcoding certainly not perfect methods, but much better than current understanding. -->
 
@@
 
-Our knowledge about species and their distributions is gathered in regional faunistic lists.
-They are typically assembled by taxonomic experts.
-They are comprehensive but usually limited in their geographic scope.
-Natural history collections have more comprehensive geographic coverage, but the quality of the taxonomic identifications can vary.
-
-Here, we assess the level of completeness of biodiversity public databases for marine invertebrates that occur within the boundaries of the EEZ of the US.
-Specifically, we use comprehensive regional faunistic lists to assess how much of this knowledge is captured by current digitization efforts of museum collections and occurrence records; and we estimate how many of the species have genetic data publicly available that would be useful in the context of identifying samples from metabarcoding studies.
+Our knowledge about species and their distributions is gathered in regional faunistic lists, typically assembled by taxonomic experts.
+They are comprehensive but limited in their geographic scope.
+On the other hand, natural history collections have more comprehensive geographic coverage, but the quality of the taxonomic identifications can vary, and some common species might be underrepresented.
+We assessed the level of completeness of biodiversity public databases for marine invertebrates that occur within the boundaries of the EEZ of the US.
+Specifically, we compared comprehensive regional faunistic lists to assess how much of this knowledge is captured by current digitization efforts of museum collections and occurrence records; and we estimate how many of the species have genetic data publicly available that would be useful in the context of identifying samples from metabarcoding studies.
 

### 2017-05-19 R/idigbio_by_geo.R
generalise idigbio query to maps with only 1 geo

@@ -3,8 +3,8 @@ get_bounding_box <- function(map, cellsize = .5) {
     box <- sp::bbox(map)
     gt <- sp::GridTopology(c(box[1,1], box[2,1]),
                            cellsize = rep(cellsize, 2), rep(100, 2))
-    gr <- as(as(sp::SpatialGrid(gt), "SpatialPixels"), "SpatialPolygons")
-    proj4string(gr) <- CRS(proj4string(map))
+    gr <- as(as(sp::SpatialGrid(gt, CRS(proj4string(map))), "SpatialPixels"),
+             "SpatialPolygons")
     it <- rgeos::gIntersects(map, gr, byid = TRUE)
     gr[it]
 }
@@ -32,30 +32,21 @@ generate_bounding_boxes <- function(map_usa, cellsize = .5) {
     map_usa_sp_df <- geojson_sp(map_usa)
 
     map_usa_fort <- ggplot2::fortify(map_usa_sp_df)
-
-    map_usa_east_sp <-
+    res <- lapply(levels(map_usa_fort$piece), function(i) {
         SpatialPolygons(
             list(
                 Polygons(
-                    list(Polygon(subset(map_usa_fort, piece == 1)[, c("long", "lat")])),
+                    list(Polygon(subset(map_usa_fort, piece == i)[, c("long", "lat")])),
                     ID = 1)
             ), proj4string = CRS(proj4string(map_usa_sp_df)))
+    }) %>%
+        lapply(get_bounding_box, cellsize = cellsize) %>%
+        lapply(bb_to_df)
+    names(res) <- levels(map_usa_fort$piece)
 
-    map_usa_west_sp <-
-        SpatialPolygons(
-            list(
-                Polygons(
-                    list(Polygon(subset(map_usa_fort, piece == 2)[, c("long", "lat")])),
-                    ID = 1)
-            ), proj4string = CRS(proj4string(map_usa_sp_df)))
-
-    east_bb <- get_bounding_box(map_usa_east_sp, cellsize = cellsize)
-    west_bb <- get_bounding_box(map_usa_west_sp, cellsize = cellsize)
-    east_df <- bb_to_df(east_bb)
-    west_df <- bb_to_df(west_bb)
-
-    dplyr::bind_rows(list(east = east_df, west = west_df), .id = "coast")
+    dplyr::bind_rows(res, .id = "piece")
 }
+
 
 ## build the list of iDigBio queries based on the data frame that
[…]

### 2017-05-19 R/comparison_by_geo.R
first attempt at plot by phylum

@@ -63,5 +63,103 @@ generate_upsetr_csv <- function(..., file) {
 
 plot_upsetr <- function(csv_file, ...) {
     readr::read_csv(csv_file) %>%
+        as.data.frame() %>%
         UpSetR::upset(order.by="freq", ...)
+}
+
+###
+
+if (FALSE) {
+
+    idig_gom <- compare_with_geo(gom_worms, idigbio_gom_records)
+    obis_gom <- compare_with_geo(gom_worms, obis_gom_records)
+
+    comp_gom <- dplyr::bind_rows(list(idigbio = idig_gom, obis = obis_gom),
+                                 .id = "database")
+
+    idig_koz <- compare_with_geo(kozloff_worms, idigbio_kozloff_records)
+    obis_koz <- compare_with_geo(kozloff_worms, obis_kozloff_records)
+
+    comp_koz <- dplyr::bind_rows(list(idigbio = idig_koz, obis = obis_koz),
+                                 .id = "database")
+
+    comp_db <- dplyr::bind_rows(list(gom = comp_gom,
+                                     kozloff = comp_koz), .id = "region") %>%
+        dplyr::filter(data_source %in% c("not_in_geo", "not_in_list"))
+
+    phyla_to_keep <- distinct(comp_db, region, phylum) %>%
+        count(phylum) %>%
+        filter(n == 2) %>%
+        .[["phylum"]]
+
+    comp_db <- filter(comp_db,
+                      phylum %in% phyla_to_keep)
+
+    rr <- comp_db %>%
+        count(region, database, data_source, phylum) %>%
+        mutate(number = ifelse(data_source == "not_in_geo", -n, n)) %>%
+        select(-n) %>%
+        #tidyr::spread(data_source, number) %>%
+        mutate(phylum = capitalize(phylum))
+
+    XX <- rr %>%
+        ungroup() %>%
+        arrange(number) %>%
+        distinct(phylum) %>%
+        .[[1]]
+
+    rr$phylum <- factor(rr$phylum, levels = XX)
+
+    pdf("figures/test_compare_dbs.pdf", height = 8, width = 15)
+    offset <- 160
+    dodge_width <- .6
+    ggplot(rr, aes(x = phylum,
+                   color = database)) +
+        geom_linerange(data = filter(rr, data_source == "not_in_geo"),
+                       aes(ymin = -offset, ymax = -offset + number),
+                       position = position_dodge(width = dodge_width)) +
[…]

### 2017-05-19 manuscript.Rmd
more edits to manuscript

@@
 A major hinderance is that we do not have enough information about which species live where, and an unknown but potentially significant portion of marine biodiversity remains to be discovered and described.
-Additionally, marine ecosystem monitoring initiatives are lacking to detect and understand, how and why communities change through time [@Duffy2013,@Borja2016].
+Additionally, coordinated marine ecosystem monitoring initiatives are lacking [@Duffy2013,@Borja2016].
 There is therefore a pressing need to improve our knowledge of marine systems before species go extinct, and their functionning are irreversibly altered.
-Our lack of knowledge about marine invertebrates make the modelisation of trophic networks less accurate, as diverse species assemblages are often reduced as simple functional categories that do no account for their ecological diversity.
+
 Until very recently, our knowledge of biodiversity was scattered among heterogeneous sources (e.g.
 natural history collections, publications, expert knowledge), and difficult to access.
-This heterogeneity and the difficulty to access this data have made synthesis and the identification of knowledge gaps difficult.
-However, with the development of linked, public, comprehensive databases of natural history records (e.g.
-iDigBio), species occurrences (e.g., GBIF, OBIS), genetic data (e.g., GenBank, BOLD), and registers of species names (e.g., WoRMS), it is now easier to quantitavely evaluate the state of our knowledge for biodiversity data.
-Here, we aggregate publicly available data for marine invertebrates that occur in the United States to highligth gaps in our knowledge.
-Given the urgency of the biodiversity crisis, we need to develop data-driven initiatives to efficiently document biodiversity, and build on our existing knowledge to collect information to improve our understanding of marine systems.
-Additionally, we emphasize that our understanding of marine biodiversity, and the impact of human activities on communities require not only to get a more accurate list of the species that inhabits the oceans, but we also need better data on species distributions and abundances.
+These factors have contributed to the lack of synthesis and have made the identification of knowledge gaps difficult.
+However, with the development of public databases for natural history records (e.g.
+iDigBio [@Page2015]), species occurrences (e.g., GBIF [@Edwards2004], OBIS [@Zhang2002]), genetic data (e.g., GenBank [@Benson2013], BOLD [@Ratnasingham2007a]), and registers of species names (e.g., WoRMS [@Costello2013b]), we can quantitavely evaluate what we know.
+Here, we aggregate publicly available data for marine invertebrates that occur in the United States and in the Gulf of Mexico to highlight gaps in our knowledge.
+We show that public databases are relatively incomplete, and that most species are known from a few speciemens and a few locations.
+We emphasize that our understanding of marine biodiversity, and the impact of human activities on communities require not only to get a more accurate list of the species that inhabits the oceans, but we also need more comprehensive data on species distributions and abundances.
 This can only be achieved with additional sampling and a commitment to training more taxonomists and experts that can identify the material collected.
-We live at a time when it has never been easier and cost effective to survey comprehensively the extant of marine biodiversity.
 
-Biodiversity discovery and documentation has been transformed in the 21st century by using molecular data to facilitate species identification, delineate species.
-And more recently, the development of high throughput sequencing technologies is used to quantify diversity and uncover large scale patterns of biodiversity organization.
-The wide adoption of DNA barcoding (where homologous DNA sequences are sequenced across the main domains of the Eukaryotes) has unraveled the pervasiveness of cryptic species, invigorating taxonomy and the description of new species, and enabled collaborative research programs such as the Census of Marine Life.
-Unraveling these cryptic species has also contributed to the recognition that species have smaller ranges than their taxonomic identity had implied making them potentially more susceptible to extinction or overexploitation.
-Despite early criticisms and now well-understood limitations, DNA barcoding has been suggested as a possible solution to the widening gap between the increased need to document biodiversity while taxonomic expertise is dwindling.
-Thus, large databases of DNA barcodes have been collaboratively assembled and made pubicly available.
-By matching DNA barcodes of unidentified species to a reference public database of accurately identified and vouchered specimens, DNA barcoding can greatly speed up species identification.
-Sequences with no match make it easier to flag specimens that warrant further taxonomic attention.
+Regional faunistic lists assembled and curated by taxonomic experts typically provide a comprehensive and detailed account of species that are known to occur in a given geographical area.
+We assessed how faunistic lists assembled over decades by specialists compare with records in public databases.
+We parsed faunistic lists for marine invertebrates from the Gulf of Mexico [@Tunnell2009] and from the Pacific Northwest [@Kozloff1987].
+Our goals were (1) to compare  the species lists for these geographical areas assembled by experts with the records found in databases; (2) to assess the proportion of species included in the expert lists that have records in public databases.
 
-<!-- clarified our understanding of species limits in many groups, but has also enabled the identification of species during their larval stages (when the lack of morphological features make species identification difficult).
-DNA barcoding has also been seen as -->
+To compare the faunistic lists from the Gulf of Mexico and the Pacific Northwest with records in public databases, we queried the iDigBio and OBIS databases for all records they included in the same geographic boundaries as the lists.
+After filtering these lists to only include the same higher taxonomic level taxa as the species lists, we had for the Gulf of Mexico `r nrow(idigbio_gom_records)` and `r nrow(obis_gom_records)` records for iDigBio and OBIS respectively representing `r dplyr::n_distinct(idigbio_gom_records, worms_valid_name)` and `r dplyr::n_distinct(obis_gom_records, worms_valid_name)` species.
+For the Gulf of Mexico, most of the species included in the expert lists are recorded in both iDigBio and OBIS (xx %), or OBIS (xx %).
+However, for the Pacific Northwest, only xx% of the species from the expert list have records in public databases.
 
-Recent developments in high-throughput sequencing technologies are giving new applications to these DNA barcode databases [@Leray2016].
-Metabarcoding has been developed to sequence in parallel the barcoding gene across the species included in environmental samples such as plankton tows, fish guts, or artificial structure scrappings [@Leray2015].
-With decreasing sequencing costs, and the prospect of automating sample collection, these technologies will soon be used routinely for ecosystem and biodiversity monitoring [@Bohan2017].
-DNA barcoding provides the possibility to link these newly generated data with historical natural history knowledge about species and their distributions.
-However, the insights we will gain from these approaches are only going to be as good as the level of completeness and accuracy of the information included in reference databases.
-However, to accurately identify the species found in these samples, comprehensive DNA barcode databases are needed.
-Here, we mostly focus on museum specimens because we can validate their identification, and a fraction of these samples (if they have not been preserved in formalin) can be used to obtain DNA from them to complete existing databases of DNA barcodes.
-
-<!-- Barcoding/metabarcoding certainly not perfect methods, but much better than current understanding. -->
-
-<!-- here there are 3 things: data availability (having good APIs to access the
-data); data comprehensiveness (how many species are captured in the database);
-data accuracy (can we trust the data that is found in the databases) -->
-
-Our knowledge about species and their distributions is gathered in regional faunistic lists, typically assembled by taxonomic experts.
-They are comprehensive but limited in their geographic scope.
-On the other hand, natural history collections have more comprehensive geographic coverage, but the quality of the taxonomic identifications can vary, and some common species might be underrepresented.
-We assessed the level of completeness of biodiversity public databases for marine invertebrates that occur within the boundaries of the EEZ of the US.
-Specifically, we compared comprehensive regional faunistic lists to assess how much of this knowledge is captured by current digitization efforts of museum collections and occurrence records; and we estimate how many of the species have genetic data publicly available that would be useful in the context of identifying samples from metabarcoding studies.
+assess the proportion of species included in these lists that (1) are recorded in databases from the same geographical area; (2) have any records in databases
+, with the records that occur in the same geographical areas in iDigBio and OBIS.
+Our goal was to assess the proportion of species known to occur in a given area by specialists that have been captured by digitization efforts.
+First, to estimate the proportion of species included in these faunistic lists whose occurences have been digitized, we queried biodiversity databases using the same geographical limits as the ones used by the authors of these faunistic lists.
+We found ...
+Second, to estimate whether listed species are known to at least occur in these databases, we queried them for the presence of the species within the databases.
 
@@
 
+; and we estimate how many of the species have genetic data publicly available that would be useful in the context of identifying samples from metabarcoding studies
+
 Using metabarcoding is a promising approach to screen an habitat for its potential for species that are not found in public database, to maximize the usefulness in the future of the data generated by an expedition.
@@
 Beyond their utility in documenting diversity, metabarcoding is also promising for understanding the functioning of ecosystems.
+
+Given the urgency of the biodiversity crisis, we need to develop data-driven initiatives to efficiently document biodiversity, and build on our existing knowledge to collect information to improve our understanding of marine systems.
+Additionally,
+
+Biodiversity discovery and documentation has been transformed in the 21st century by molecular data to facilitate species identification, delineate species.
+And more recently, the development of high throughput sequencing technologies is used to quantify diversity and uncover large scale patterns of biodiversity organization.
+The wide adoption of DNA barcoding (where homologous DNA sequences are sequenced across the main domains of the Eukaryotes) has unraveled the pervasiveness of cryptic species, invigorating taxonomy and the description of new species, and enabled collaborative research programs such as the Census of Marine Life.
+Unraveling these cryptic species has also contributed to the recognition that species have smaller ranges than their taxonomic identity had implied making them potentially more susceptible to extinction or overexploitation.
+Despite early criticisms and now well-understood limitations, DNA barcoding has been suggested as a possible solution to the widening gap between the increased need to document biodiversity while taxonomic expertise is dwindling.
+Thus, large databases of DNA barcodes have been collaboratively assembled and made pubicly available.
+By matching DNA barcodes of unidentified species to a reference public database of accurately identified and vouchered specimens, DNA barcoding can greatly speed up species identification: sequences with no match make it easier to flag specimens that warrant further taxonomic attention.
+
+<!-- clarified our understanding of species limits in many groups, but has also enabled the identification of species during their larval stages (when the lack of morphological features make species identification difficult).
+DNA barcoding has also been seen as -->
+
+Recent developments in high-throughput sequencing technologies are giving new applications to these DNA barcode databases [@Leray2016].
+Metabarcoding has been developed to sequence in parallel the barcoding gene across the species included in environmental samples such as plankton tows, fish guts, or artificial structure scrappings [@Leray2015].
+With decreasing sequencing costs, and the prospect of automating sample collection, these technologies will soon be used routinely for ecosystem and biodiversity monitoring [@Bohan2017].
+DNA barcoding provides the possibility to link these newly generated data with historical natural history knowledge about species and their distributions.
+However, the insights we will gain from these approaches are only going to be as good as the level of completeness and accuracy of the information included in reference databases.
+Here, we mostly focus on museum specimens because we can validate their identification, and a fraction of these samples (if they have not been preserved in formalin) can be used to obtain DNA from them to complete existing databases of DNA barcodes.
+
+<!-- Barcoding/metabarcoding certainly not perfect methods, but much better than current understanding. -->
+
+<!-- here there are 3 things: data availability (having good APIs to access the
+data); data comprehensiveness (how many species are captured in the database);
+data accuracy (can we trust the data that is found in the databases) -->
 
@@
 <!--
+
+It has never been easier and cost effective to survey comprehensively the extant of marine biodiversity.
 

### 2017-05-26 R/usa_maps.R
extend south ward PWN map

@@ -16,7 +16,7 @@ get_felder_gom_map <- function(file) {
 }
 
 get_kozloff_map <- function() {
-    wkt_string <- "POLYGON((-129.43 51.29, -126.9 51.29, -122.2 49.18, -122.2 47, -124.3 46.9, -129.43 51.29))"
+    wkt_string <- "POLYGON((-129.43 51.29, -126.9 51.29, -122.2 49.18, -122.2 47, -123.5 46.8, -123.5 43.3, -124 43.3, -124.3 46.9, -129.43 51.29))"
     rgeos::readWKT(wkt_string, p4s = "+proj=longlat +datum=WGS84") %>%
         geojsonio::geojson_json()
 }

### 2017-05-26 manuscript.Rmd
edit manuscript

@@
 
-# Main text
-
 <!--  study focused on USA but conclusions more general given that it's arguably one of the better studied area + data availability -->
@@
 
+## Abstract
+
 With increasing impact of human activities on the biosphere, we need efficient and accurate approaches to document changes in species ranges and community composition.
@@
 However, with the development of public databases for natural history records (e.g.
-iDigBio [@Page2015]), species occurrences (e.g., GBIF [@Edwards2004], OBIS [@Zhang2002]), genetic data (e.g., GenBank [@Benson2013], BOLD [@Ratnasingham2007a]), and registers of species names (e.g., WoRMS [@Costello2013b]), we can quantitavely evaluate what we know.
-Here, we aggregate publicly available data for marine invertebrates that occur in the United States and in the Gulf of Mexico to highlight gaps in our knowledge.
+iDigBio [@Page2015]), species occurrences (e.g., GBIF [@Edwards2004], OBIS [@Zhang2002]), genetic data (e.g., GenBank [@Benson2013], BOLD [@Ratnasingham2007a]), and registers of species names (e.g., WoRMS [@Costello2013b]), we can now evaluate what we know.
+Here, we aggregate publicly available data for marine invertebrates that occur in the United States, in the Gulf of Mexico, and in the Pacific North-West, to highlight gaps in our knowledge.
 We show that public databases are relatively incomplete, and that most species are known from a few speciemens and a few locations.
@@
 
-Regional faunistic lists assembled and curated by taxonomic experts typically provide a comprehensive and detailed account of species that are known to occur in a given geographical area.
-We assessed how faunistic lists assembled over decades by specialists compare with records in public databases.
-We parsed faunistic lists for marine invertebrates from the Gulf of Mexico [@Tunnell2009] and from the Pacific Northwest [@Kozloff1987].
-Our goals were (1) to compare  the species lists for these geographical areas assembled by experts with the records found in databases; (2) to assess the proportion of species included in the expert lists that have records in public databases.
+## Results
+
+### Comparison with Regional Lists by Geography
+
+Regional faunistic lists assembled and curated by taxonomic experts typically provide a comprehensive and detailed account of species that are known to occur in a given geographic area.
+We assessed how faunistic lists assembled over decades by specialists compare with records in public databases for the same geographic area.
+We parsed faunistic lists to extract marine invertebrates from the Gulf of Mexico [@Tunnell2009] and from the Pacific Northwest [@Kozloff1987].
+Our goals were (1) to compare  the species lists for these geographic areas assembled by experts with the records found in databases; (2) to assess the proportion of species included in the expert lists that have records in public databases.
 
 To compare the faunistic lists from the Gulf of Mexico and the Pacific Northwest with records in public databases, we queried the iDigBio and OBIS databases for all records they included in the same geographic boundaries as the lists.
-After filtering these lists to only include the same higher taxonomic level taxa as the species lists, we had for the Gulf of Mexico `r nrow(idigbio_gom_records)` and `r nrow(obis_gom_records)` records for iDigBio and OBIS respectively representing `r dplyr::n_distinct(idigbio_gom_records, worms_valid_name)` and `r dplyr::n_distinct(obis_gom_records, worms_valid_name)` species.
+After filtering these lists to only include the same higher taxonomic level taxa as the species lists, we had for the Gulf of Mexico `r nrow(idigbio_gom_records)` and `r nrow(obis_gom_records)` records for iDigBio and OBIS respectively representing `r dplyr::n_distinct(idigbio_gom_records$worms_valid_name)` and `r dplyr::n_distinct(obis_gom_records$worms_valid_name)` species.
 For the Gulf of Mexico, most of the species included in the expert lists are recorded in both iDigBio and OBIS (xx %), or OBIS (xx %).
-However, for the Pacific Northwest, only xx% of the species from the expert list have records in public databases.
+However, for the Pacific Northwest, only xx% of the species from the expert list have records in public databases (Fig.
+1).
 
-assess the proportion of species included in these lists that (1) are recorded in databases from the same geographical area; (2) have any records in databases
-, with the records that occur in the same geographical areas in iDigBio and OBIS.
-Our goal was to assess the proportion of species known to occur in a given area by specialists that have been captured by digitization efforts.
-First, to estimate the proportion of species included in these faunistic lists whose occurences have been digitized, we queried biodiversity databases using the same geographical limits as the ones used by the authors of these faunistic lists.
-We found ...
-Second, to estimate whether listed species are known to at least occur in these databases, we queried them for the presence of the species within the databases.
+Species that are included in the list but do not have records in databases could be (1) rare: they have never been collected or have been collected only a few times, and the corresponding records have not been digitized yet; (2) difficult to preserve: large, very small, or fragile species are preserved less frequently; (3) are in the database under a different name: the species has been synonimized or re-identified since the publication of the list under a different name that may be included in the database; (4) at the limit of its range: it was included in the list because it is known to occur at the limit of its geographic zone, but has never been collected within the boundaries.
+
+Species that are in the database but not in the list could be: (1) records not considered by the authors of the checklist: either because they were not accessible or known to the authors; (2) records erronously excluded the records (2) records that have been discovered since the list has been assembled; (3) records with incorrect identification; (4) records with incorrect geographic coordinates: their associated geographic information in the database places them in the Gulf of Mexico or the Pacific Northwest, but they were actually collected elsewhere.
+
+### Comparison with Regional Lists by Species Names
+
+To assess whether species included in regional lists but absent from the same geographic area nevertheless occur elsewhere, we queried the iDigBio and OBIS database for all the marine invertebrate species included in these lists.
 
 We extracted `r sum(gom_worms$is_marine, na.rm = TRUE)` marine invertebrates species from the faunistic list of the Gulf of Mexico, and `r sum(kozloff_worms$is_marine, na.rm=TRUE)` species from the Salish Sea.
+We then looked for the number of species listed that are known to occur in the same geographic area, the United States, or anywhere else in the world.
+Overall, OBIS had records for most of the species that are included in the species lists.
+However, even for the better studied phyla about 75% of the species have records from the same geographic area, while it is less than 50% for the Pacific Northwest.
+
+These results suggest that even when considering species lists that have been curated and that represent possibly the most accurate knowledge for this geographic area, it would not be possible to reconstruct these lists based on currently digitized information.
+It highlights that additional digitization but even more importantly additional collecting is needed to validate the occurences of the species included in these lists.
+
 Species are unevenly represented in databases, with 25% to 50% of species listed to occur in these locations, that are not represented in digital records of museum collections.
@@
 
-The proportion of species that have DNA barcodes available is surprisingly constant across the better studied phyla.
-About 20% of species known to occur in the Gulf of Mexico and in the Salish Sea have DNA barcode records.
-Because geographic information is missing for most records (`r format_output(100*stat_bold$p_no_coords)`% of the `r format(stat_bold$n_total_records)` records analyzed here have no geographical coordinates information associated with them), and many marine invertebrate species are known to be cryptic complexes that are not as widely distributed as previously thought.
-Therefore specimens identified under the same species but sampled in distant locations, could represent cryptic species.
-It is thus crucial to ensure that accurate geographic information is provided when DNA barcodes are published, and that additional samples are collected, to capture more accurately their geographic range and the extent of their genetic diversity.
-This high level of cryptic diversity is reflected in the fact that `r stat_bold$n_spp_muti_bin` species out of `r stat_bold$n_spp` include more than 1 BIN.
-This number is confounded by some taxonomic identification issues as `r stat_bold$n_shared_bins` BINs are shared across more than one species.
-
-<!-- However, among these records, xx% have been sampled in the US (xx%), Mexico (xx%) or Canada (xx%).  -->
-<!-- Paragraph about the same numbers but based on species list from iDigBio: -->
-
-<!-- - how many species are in iDigBio and don't show up in species lists from Kozloff and GOM -->
-<!-- - how many species in iDigBio have barcodes -->
-<!-- - how many species in iDigBio have OBIS and GBIF records -->
+### What do we know about marine invertebrates in the US?
 
@@
 2).
+
+Based on these data, there is no evidence of a diversity gradient.
 
@@
 However, taxonomic expertise is dwindling, and there are now only a handful of US natural history museums with active marine invertebrates collections.
-Museum collections have been recording our knowledge about species and their geographical ranges, but until recently, this information was fragmented across institutions and difficult to synthesize.
+Museum collections have been recording our knowledge about species and their geographic ranges, but until recently, this information was fragmented across institutions and difficult to synthesize.
 With collection digitization and other large-scale projects have aimed at gathering biodiversity information, public databases can now be used to enable synthesis and highlight knowledge gaps.
 Here we use these data to synthetic the state of knowledge for marine invertebrates, and demonstrate that the stage is set to combine these data sources to inform where future efforts for documenting biodiversity.
+
+### Current genetic data and the future of biodiveristy discovery and monitoring
+
+The proportion of species that have DNA barcodes available is surprisingly constant across the better studied phyla.
+About 20% of species known to occur in the Gulf of Mexico and in the Salish Sea have DNA barcode records.
+Because geographic information is missing for most records (`r format_output(100*stat_bold$p_no_coords)`% of the `r format(stat_bold$n_total_records)` records analyzed here have no geographic coordinates information associated with them), and many marine invertebrate species are known to be cryptic complexes that are not as widely distributed as previously thought.
+Therefore specimens identified under the same species but sampled in distant locations, could represent cryptic species.
+It is thus crucial to ensure that accurate geographic information is provided when DNA barcodes are published, and that additional samples are collected, to capture more accurately their geographic range and the extent of their genetic diversity.
+This high level of cryptic diversity is reflected in the fact that `r stat_bold$n_spp_muti_bin` species out of `r stat_bold$n_spp` include more than 1 BIN.
+This number is confounded by some taxonomic identification issues as `r stat_bold$n_shared_bins` BINs are shared across more than one species.
+
+## What to do next?
+
+* Improve data quality:
+  - easy: fix mispelling in identifications
+  - improve identification of unidentified lots (requires taxonomists/specialists to be able to identify the specimens)
+  - add and validate geographic information associated with specimens
+  - develop infrastructure and tools to continuously check and validate data
+    quality in databases.
 
@@
 Beyond their utility in documenting diversity, metabarcoding is also promising for understanding the functioning of ecosystems.
+
+## Conclusions
 

### 2017-05-26 R/gradient.R
add code for looking into diversity gradients

@@ -1,0 +1,17 @@
+plot_gradient <- function(idig) {
+
+    idig %>%
+        dplyr::mutate(coast = if_else(decimallongitude > -100, "east", "west")) %>%
+        dplyr::mutate(
+                   lat_grad = ntile(decimallatitude, 100)
+               ) %>%
+        dplyr::group_by(coast, clean_phylum, lat_grad) %>%
+        dplyr::summarize(n_spp = n_distinct(cleaned_scientificname),
+                         n_obs = n()) %>%
+        dplyr::filter(clean_phylum %in% c("arthropoda", "mollusca", "annelida", "echinodermata", "cnidaria")) %>%
+        dplyr::mutate(p_samp = n_spp/n_obs) %>%
+        ggplot(aes(x = lat_grad, y = n_spp, color = clean_phylum, shape = coast, size = n_obs)) +
+        geom_point() + facet_grid(clean_phylum ~ coast) +
+        geom_smooth(aes(weight = n_obs))
+
+}

### 2017-05-28 manuscript.Rmd
update manuscript

@@
 title: "Data-driven approaches to identifying knowledge gaps in marine invertebrate biodiversity"
+date: '`r format(Sys.time())`'
 output:
-  pdf_document #:
-    #pandoc_args: ["-F","pandoc-crossref"]
+  bookdown::pdf_document2:
+    #includes:
+    #  before_body: ../template/doc_prefix.tex
+    #  in_header: ../template/preamble.tex
+    keep_tex: yes
+    latex_engine: pdflatex
+    number_sections: no
+    toc: no
+  bookdown::word_document2: null
+fontsize: 12pt
+link-citations: yes
+csl: nature.csl
+bibliography: /home/francois/Library/dark-plankton.bib
 ---
@@
 
-<!---
+<!--
 the content of insect traps, marine sediments, plankton
@@
 sequenced at various spatial scale.
---->
+-->
 
@@
 To compare the faunistic lists from the Gulf of Mexico and the Pacific Northwest with records in public databases, we queried the iDigBio and OBIS databases for all records they included in the same geographic boundaries as the lists.
-After filtering these lists to only include the same higher taxonomic level taxa as the species lists, we had for the Gulf of Mexico `r nrow(idigbio_gom_records)` and `r nrow(obis_gom_records)` records for iDigBio and OBIS respectively representing `r dplyr::n_distinct(idigbio_gom_records$worms_valid_name)` and `r dplyr::n_distinct(obis_gom_records$worms_valid_name)` species.
-For the Gulf of Mexico, most of the species included in the expert lists are recorded in both iDigBio and OBIS (xx %), or OBIS (xx %).
-However, for the Pacific Northwest, only xx% of the species from the expert list have records in public databases (Fig.
+After filtering these lists to only include the same higher taxonomic level taxa as the species lists, and keeping only records for which the species could be match to a species name in WoRMS, we had for the Gulf of Mexico `r nrow(idigbio_gom_records)` and `r nrow(obis_gom_records)` records for iDigBio and OBIS respectively representing `r dplyr::n_distinct(idigbio_gom_records$worms_valid_name)` and `r dplyr::n_distinct(obis_gom_records$worms_valid_name)` species.
+For the Pacific Northwest, we retrieved `r nrow(idigbio_kozloff_records)` and `r nrow(obis_kozloff_records)` records for iDigBio and OBIS respectively representing `r dplyr::n_distinct(idigbio_kozloff_records$worms_valid_name)` and `r dplyr::n_distinct(obis_kozloff_records$worms_valid_name)` species.
+For the Gulf of Mexico, the majority of the species included in the expert lists are recorded in both iDigBio and OBIS (`r format_output(species_in_common$n_spp_gom_idigbio_obis/species_in_common$n_spp_gom*100)` %), or OBIS (`r format_output(species_in_common$n_spp_gom_obis/species_in_common$n_spp_gom*100)` %) (Fig.
+\@ref(fig:database-overlap)).
+However, for the Pacific Northwest, only `r format_output(species_in_common$n_spp_kozloff_idigbio_obis/species_in_common$n_spp_kozloff*100)` and `r format_output(species_in_common$n_spp_kozloff_obis/species_in_common$n_spp_kozloff*100)` of the species from the regional list had records in OBIS and iDigBio or OBIS respectively (Fig.
 1).
+
+```{r compare_database_overlap, fig.cap='Overlap between the regional species list (list), records in iDigBio (idigbio), records in OBIS (obis), and records in BOLD (bold) for the Gulf of Mexico (left) and the Pacific Northwest (right).
+(\\#fig:database-overlap)', echo=FALSE}
+knitr::include_graphics("figures/database_coverage.pdf")
+```
 
@@
 We then looked for the number of species listed that are known to occur in the same geographic area, the United States, or anywhere else in the world.
-Overall, OBIS had records for most of the species that are included in the species lists.
+Overall, OBIS had georeferenced records for most of the species that are included in the species lists.
 However, even for the better studied phyla about 75% of the species have records from the same geographic area, while it is less than 50% for the Pacific Northwest.
 
-These results suggest that even when considering species lists that have been curated and that represent possibly the most accurate knowledge for this geographic area, it would not be possible to reconstruct these lists based on currently digitized information.
-It highlights that additional digitization but even more importantly additional collecting is needed to validate the occurences of the species included in these lists.
+These results suggest that even when considering species lists that have been curated and that represent possibly the most accurate knowledge for this geographic area, today we could only reconstruct these list at 50 to 70% completedness based on digitized information.
+It highlights that additional digitization is necessary but even more importantly additional collecting is needed to validate the occurences of the species included in these lists.
 

### 2017-05-28 manuscript.Rmd
add % sign in text

@@
 \@ref(fig:database-overlap)).
-However, for the Pacific Northwest, only `r format_output(species_in_common$n_spp_kozloff_idigbio_obis/species_in_common$n_spp_kozloff*100)` and `r format_output(species_in_common$n_spp_kozloff_obis/species_in_common$n_spp_kozloff*100)` of the species from the regional list had records in OBIS and iDigBio or OBIS respectively (Fig.
+However, for the Pacific Northwest, only `r format_output(species_in_common$n_spp_kozloff_idigbio_obis/species_in_common$n_spp_kozloff*100)`% and `r format_output(species_in_common$n_spp_kozloff_obis/species_in_common$n_spp_kozloff*100)`% of the species from the regional list had records in OBIS and iDigBio or OBIS respectively (Fig.
 1).

### 2017-05-30 manuscript.Rmd
use cowplot to combine richness per database plots + include in ms

@@
 ---
+
+```{r setup, echo=FALSE, include=FALSE}
+knitr::opts_chunk$set(echo = FALSE)
+```
 
@@
 
-To assess whether species included in regional lists but absent from the same geographic area nevertheless occur elsewhere, we queried the iDigBio and OBIS database for all the marine invertebrate species included in these lists.
+To assess whether species included in regional lists but absent from the same geographic area nevertheless occur elsewhere, we queried the iDigBio and OBIS databases for all the marine invertebrate species included in these lists.
 
@@
 However, even for the better studied phyla about 75% of the species have records from the same geographic area, while it is less than 50% for the Pacific Northwest.
+
+```{r database_richness, fig.cap='Comparison of the fraction of species listed in regional lists that are present in biodiversity databases for the Gulf of Mexico (A) and the Pacific Northwest (B).'}
+knitr::include_graphics("figures/richness_per_db.pdf")
+```
 
@@
 ![Relation between the number of specimen and the number of species collected](figures/plot_sampling_effort.pdf)
-
-![Level of digitization for the Salish Sea](figures/plot_richness_per_db_koz.pdf)
-
-![Level of digitization for the Gulf of Mexico](figures/plot_richness_per_db_gom.pdf)
 

### 2017-05-30 R/gom_list.R

@@ -185,7 +185,11 @@ plot_richness_per_db <- function(smry_db, data_source, region = c("gom", "pnw"))
 
     full_region <- c(gom = "GoM", pnw = "PNW")
 
+    phyla_to_plot <-  c("Arthropoda", "Mollusca", "Annelida", "Cnidaria", "Echinodermata",
+                        "Platyhelminthes", "Porifera", "Bryozoa", "Chordata")
+
     smry_db %>%
+        dplyr::filter(taxon_name %in% phyla_to_plot) %>%
         dplyr::mutate(
                    n_idigbio_us_diff = n_idigbio - n_idigbio_in_us,
                    n_obis_us_diff = n_obis - n_obis_in_us,
@@ -206,7 +210,7 @@ plot_richness_per_db <- function(smry_db, data_source, region = c("gom", "pnw"))
                       -n_idigbio_in_us, -n_obis_in_us, -n_gbif_in_us
                       ) %>%
         tidyr::gather(source_n_spp, n_spp, -taxon_name) %>%
-        dplyr::mutate(source_db = gsub("(n_[a-z]+)(_.+)?", "\\1", source_n_spp)) %>% #write_csv("/tmp/nb.csv"); return(NULL)
+        dplyr::mutate(source_db = gsub("(n_[a-z]+)(_.+)?", "\\1", source_n_spp)) %>%
         ggplot(aes(x = factor(source_db,
                               levels = rev(c("n_total", "n_obis", "n_gbif", "n_idigbio", "n_barcoded")),
                               labels = c("BOLD", "iDigBio", "GBIF", "OBIS", "Total")
@@ -222,23 +226,37 @@ plot_richness_per_db <- function(smry_db, data_source, region = c("gom", "pnw"))
                                             "n_barcoded"
                                             ))),
                                  labels = rev(
-                                     gsub("XXX", full_region[region],
-                                          c("Total",
-                                            "OBIS in XXX", "OBIS in US EEZ", "OBIS global",
-                                            "GBIF in XXX", "GBIF in US EEZ", "GBIF global",
-                                            "iDigBio in XXX", "iDigBio in US EEZ", "iDigBio global",
-                                            "BOLD"
-                                            )))
+                                     #full_region[region],
+                                     c("Species in list",
+                                       "OBIS in region", "OBIS in US EEZ", "OBIS global",
+                                       "GBIF in region", "GBIF in US EEZ", "GBIF global",
[…]

### 2017-05-30 manuscript.Rmd
add proportions of species with obis, idigbio, gbif records

@@
 
-These results suggest that even when considering species lists that have been curated and that represent possibly the most accurate knowledge for this geographic area, today we could only reconstruct these list at 50 to 70% completedness based on digitized information.
+These results suggest that even when considering species lists that have been curated and that represent possibly the most accurate knowledge for this geographic area, today we could only reconstruct these lists partially based on digitized information.
+For instance, based on OBIS data the list would be at `r format_output(summary_richness_per_db_gom$prop_obis_in_gom[summary_richness_per_db_gom$taxon_name == "Porifera"]*100)`% complete for sponges to `r format_output(summary_richness_per_db_gom$prop_obis_in_gom[summary_richness_per_db_gom$taxon_name == "Mollusca"]*100)`% for molluscs.
+However, these levels drop significantly if we consider species for which museum specimens are available from  `r format_output(summary_richness_per_db_gom$prop_idigbio_in_gom[summary_richness_per_db_gom$taxon_name == "Mollusca"]*100)`% for molluscs to `r format_output(summary_richness_per_db_gom$prop_idigbio_in_gom[summary_richness_per_db_gom$taxon_name == "Annelida"]*100)`% for annelids.
 It highlights that additional digitization is necessary but even more importantly additional collecting is needed to validate the occurences of the species included in these lists.

### 2017-05-30 R/gom_list.R

@@ -176,7 +176,31 @@ summarize_richness_per_db <- function(bold_db, idig_db, obis_db, gbif_db, region
         dplyr::filter(taxon_name %in% c("Arthropoda", "Mollusca", "Annelida",
                                         "Cnidaria", "Echinodermata", "Platyhelminthes",
                                         "Porifera", "Bryozoa", "Nematoda", "Chordata"
-                                        ))
+                                        )) %>%
+        dplyr::mutate_(.dots = setNames(list(
+                           lazyeval::interp("n_idigbio_in_region/n_total",
+                                            n_idigbio_in_region = as.name(paste0("n_idigbio_in_", region))),
+                           "n_idigbio_in_us/n_total",
+                           "n_idigbio/n_total",
+                           lazyeval::interp("n_obis_in_region/n_total",
+                                            n_obis_in_region = as.name(paste0("n_obis_in_", region))),
+                          "n_obis_in_us/n_total",
+                           "n_obis/n_total",
+                           lazyeval::interp("n_gbif_in_region/n_total",
+                                            n_gbif_in_region = as.name(paste0("n_gbif_in_", region))),
+                           "n_gbif_in_us/n_total",
+                           "n_gbif/n_total"
+                       ),
+                       c(paste0("prop_idigbio_in_", region),
+                         "prop_idigbio_in_us",
+                         "prop_idigbio",
+                         paste0("prop_obis_in_", region),
+                         "prop_obis_in_us",
+                         "prop_obis",
+                         paste0("prop_gbif_in_", region),
+                         "prop_gbif_in_us",
+                         "prop_gbif"))
+                       )
 
 }
 
@@ -207,13 +231,14 @@ plot_richness_per_db <- function(smry_db, data_source, region = c("gom", "pnw"))
                                 ))
                        )  %>%
         dplyr::select(-n_idigbio, -n_obis, -n_gbif,
-                      -n_idigbio_in_us, -n_obis_in_us, -n_gbif_in_us
+                      -n_idigbio_in_us, -n_obis_in_us, -n_gbif_in_us,
+                      -starts_with("prop")
                       ) %>%
         tidyr::gather(source_n_spp, n_spp, -taxon_name) %>%
         dplyr::mutate(source_db = gsub("(n_[a-z]+)(_.+)?", "\\1", source_n_spp)) %>%
         ggplot(aes(x = factor(source_db,
[…]

### 2017-05-31 manuscript.Rmd
remove section about vertebrate coverage

@@
 It highlights that additional digitization is necessary but even more importantly additional collecting is needed to validate the occurences of the species included in these lists.
-
-Species are unevenly represented in databases, with 25% to 50% of species listed to occur in these locations, that are not represented in digital records of museum collections.
-Typically a higher proportion of species from the Pacific Coast than the Gulf of Mexico are not represented.
-Comparatively, all but 2 of the `r nrow(mammals_species_asm)` mammal species and `r sum(amphibian_summary$count == 0)` of the `r nrow(amphibian_summary)` amphibian species known to occur in the US have digitized museum specimens.
 

### 2017-05-31 R/gom_list.R
add more options to combine_cowplots

@@ -271,16 +271,22 @@ plot_richness_per_db <- function(smry_db, data_source, region = c("gom", "pnw"))
         coord_flip()
 }
 
-combine_cowplots <- function(g1, g2, ...) {
-    legend <- get_legend(g1)
+combine_cowplots <- function(g1, g2, nrow = 1, common_legend = TRUE, ...) {
+    if (common_legend) {
+        g1 <- g1 + theme(legend.position = "none")
+        g2 <- g2 + theme(legend.position = "none")
+        legend <- get_legend(g1)
+    }
     prow <- cowplot::plot_grid(
-                         g1 + theme(legend.position = "none"),
-                         g2 + theme(legend.position = "none"),
+                         g1,
+                         g2,
                  align = 'vh',
                  labels = c("A", "B"),
                  hjust = -1,
-                 nrow = 1
+                 nrow = nrow
                  )
-    cowplot::plot_grid(prow, legend, rel_widths = c(3, .4))
+    if (common_legend) {
+        cowplot::plot_grid(prow, legend, rel_widths = c(3, .4), ...)
+    } else prow
 
 }

### 2017-05-31 manuscript.Rmd
simplify code/dependencies for sampling + diversity maps

@@
 These results suggest that even when considering species lists that have been curated and that represent possibly the most accurate knowledge for this geographic area, today we could only reconstruct these lists partially based on digitized information.
-For instance, based on OBIS data the list would be at `r format_output(summary_richness_per_db_gom$prop_obis_in_gom[summary_richness_per_db_gom$taxon_name == "Porifera"]*100)`% complete for sponges to `r format_output(summary_richness_per_db_gom$prop_obis_in_gom[summary_richness_per_db_gom$taxon_name == "Mollusca"]*100)`% for molluscs.
+For instance, based on OBIS data the list would be `r format_output(summary_richness_per_db_gom$prop_obis_in_gom[summary_richness_per_db_gom$taxon_name == "Porifera"]*100)`% complete for sponges to `r format_output(summary_richness_per_db_gom$prop_obis_in_gom[summary_richness_per_db_gom$taxon_name == "Mollusca"]*100)`% for molluscs.
 However, these levels drop significantly if we consider species for which museum specimens are available from  `r format_output(summary_richness_per_db_gom$prop_idigbio_in_gom[summary_richness_per_db_gom$taxon_name == "Mollusca"]*100)`% for molluscs to `r format_output(summary_richness_per_db_gom$prop_idigbio_in_gom[summary_richness_per_db_gom$taxon_name == "Annelida"]*100)`% for annelids.
-It highlights that additional digitization is necessary but even more importantly additional collecting is needed to validate the occurences of the species included in these lists.
+Similar patterns can be observed for the Pacific Northwest, but overall the levels of data availability are lower: from `r format_output(summary_richness_per_db_koz$prop_obis_in_pnw[summary_richness_per_db_koz$taxon_name == "Porifera"]*100)`% for sponges to `r format_output(summary_richness_per_db_koz$prop_obis_in_pnw[summary_richness_per_db_koz$taxon_name == "Annelida"]*100)`% for annelids for the proportion of species that have records in OBIS, and from  `r format_output(summary_richness_per_db_koz$prop_idigbio_in_pnw[summary_richness_per_db_koz$taxon_name == "Porifera"]*100)`% for sponges to `r format_output(summary_richness_per_db_koz$prop_idigbio_in_pnw[summary_richness_per_db_koz$taxon_name == "Mollusca"]*100)`% for molluscs for the proportion of species that have records in iDigBio.
+Museum specimens are critical to validate the occurence of species [@Rocha2014], especially for marine invertebrates for which species limits based on morphology have been known to be challenged by other lines of evidence including molecular data [@Knowlton2000].
+Some of the discrepencies observed between the number of species for which occurence data and museum specimens are available suggest that additional digitization of collections (including capturing their geographic information) are needed.
 
@@
 2).
+
+```{r map-diversity-sampling, fig.cap='Spatial distribution of the number of samples (A) and species richness (B) for marine invertebrates based on specimen data available from iDigBio within the limits of the Exclusive Economic Zone for the contiental USA.'}
+knitr::include_graphics("figures/map_diversity_sampling.pdf")
+```
 
@@
 data accuracy (can we trust the data that is found in the databases) -->
-
-![Number of species of marine invertebrates](figures/map_diversity.pdf)
-
-![Number of museum records for marine invertebrates](figures/map_sampling_effort.pdf)
-
-![Number of species of marine invertebrates](figures/map_diversity.pdf)
 

### 2017-05-31 R/correlation_specimens_species.R
more small tweaks to figure

@@ -9,8 +9,8 @@ plot_sampling_effort <- function(id) {
     id %>%
         prepare_sampling_effort_data %>%
         ggplot(aes(x = n_specimen, y = n_species, shape = east_west, colour = latitude)) +
-        scale_colour_viridis(option = "magma", direction = -1, end = .95) +
-        geom_point(alpha = .6, position = position_jitter(width = .01, height = .01)) +
+        scale_colour_viridis(option = "magma", direction = -1, end = .95, name = "Latitude") +
+        geom_point(alpha = .6, position = position_jitter(width = .02, height = .02)) +
         geom_abline(slope = 1, intercept = 0) +
         xlab("Number of records") + ylab("Number of species") +
         scale_x_log10() + scale_y_log10() +

### 2017-05-31 manuscript.Rmd
write some words

@@
 Most samples were collected on the continental shelf (Fig.
-1).
+\@ref(fig:map-diversity-sampling)A).
 For the East coast, species richness is the highest in South Florida but it is relatively uniform along the coastline.
 For the West coast, the Salish Sea, the Bay Area, and Southern California have the highest richness (Fig.
-2).
+\@ref(fig:map-diversity-sampling)B).
 
-```{r map-diversity-sampling, fig.cap='Spatial distribution of the number of samples (A) and species richness (B) for marine invertebrates based on specimen data available from iDigBio within the limits of the Exclusive Economic Zone for the contiental USA.'}
+```{r map-diversity-sampling, fig.cap='Spatial distribution of the number of samples (A) and species richness (B) for marine invertebrates based on specimen data available from iDigBio within the limits of the Exclusive Economic Zone for the contiental USA.
+(\\#fig:map-diversity-sampling)'}
 knitr::include_graphics("figures/map_diversity_sampling.pdf")
@@
 
-Based on these data, there is no evidence of a diversity gradient.
+These patterns of spatial distribution of the species richness is strongly confounded by the sampling.
+When investigating the relation between the number of samples and the number of species for each cell of the map (\@ref(fig:correlation-sampling)), we can see that until reaching about 100 records per 0.5° there is little redundancy in the sampled species.
+
+```{r correlation-sampling, fig.cap='Relation between number of records and number of species for each cell of the map used in \\@ref(fig:map-diversity-sampling)'}
+knitr::include_graphics("figures/plot_sampling_effort.pdf")
+```
 
@@
 data accuracy (can we trust the data that is found in the databases) -->
-
-![Relation between the number of specimen and the number of species collected](figures/plot_sampling_effort.pdf)
 

### 2017-06-01 R/time_plot.R
fix bug: the function wasn't using cumsum

@@ -128,10 +128,8 @@ make_knowledge_through_time_idigbio <- function(idigbio) {
 calc_n_spp_comparison <- function(idig_time) {
     res <- idig_time %>%
         group_by(year) %>%
-        summarize(
-            n_new_spp = sum(n_new_spp),
-            cum_n_new_spp = sum(cum_n_new_spp)
-        )
+        summarize(n_new_spp = sum(n_new_spp)) %>%
+        mutate(cum_n_new_spp = cumsum(n_new_spp))
 
     list(
         n_spp_1970_1960 = res[res$year == 1970, "cum_n_new_spp"] - res[res$year == 1960, "cum_n_new_spp"],

### 2017-06-01 R/time_plot.R
use the same style for knowledge through time plots

@@ -146,17 +146,18 @@ filter_phyla <- function(ktt, n_min = 100) {
         unique(x = .$phylum)
 }
 
-
 plot_cum_samples_through_time <- function(knowledge_through_time, facet = TRUE)  {
     phy_to_keep <- filter_phyla(knowledge_through_time)
 
     res <- knowledge_through_time %>%
         filter(phylum %in% phy_to_keep) %>%
         ggplot(aes(x = year, y = cum_n_samples, colour = phylum)) +
-        geom_line() +
+        geom_line(size = .7) +
         xlim(c(1850, 2017)) +
         ylab("Cumulative Number of Samples") +
-        xlab("Year")
+        xlab("Year") +
+        theme_ipsum(base_family = "Ubuntu Condensed") +
+        scale_colour_hc(name = "", labels = capitalize(phy_to_keep))
 
      if (facet)
         res <- res + facet_grid(fauna ~ database)
@@ -173,10 +174,12 @@ plot_cum_spp_through_time <- function(knowledge_through_time, facet = TRUE) {
         filter(phylum %in% phy_to_keep,
                !is.na(cum_n_new_spp)) %>%
         ggplot(aes(x = year, y = cum_n_new_spp, colour = phylum)) +
-        geom_line() +
+        geom_line(size = .7) +
         xlim(c(1850, 2017)) +
         ylab("Cumulative Number of Species") +
-        xlab("Year")
+        xlab("Year") +
+        theme_ipsum(base_family = "Ubuntu Condensed")  +
+        scale_colour_hc()
 
     if (facet)
         res <- res + facet_grid(fauna ~ database)

### 2017-06-01 manuscript.Rmd
add more text

@@
 
-These patterns of spatial distribution of the species richness is strongly confounded by the sampling.
-When investigating the relation between the number of samples and the number of species for each cell of the map (\@ref(fig:correlation-sampling)), we can see that until reaching about 100 records per 0.5° there is little redundancy in the sampled species.
+These patterns of spatial distribution of the species richness are strongly confounded by the sampling effort.
+When investigating the relation between the number of samples and the number of species for each cell of the map (Fig.
+\@ref(fig:correlation-sampling)), we can see that until reaching about 100 records per cell there is little redundancy in the sampled species.
 
-```{r correlation-sampling, fig.cap='Relation between number of records and number of species for each cell of the map used in \\@ref(fig:map-diversity-sampling)'}
+```{r correlation-sampling, fig.cap='Relation between number of records and number of species for each cell of the map used in Fig. \\@ref(fig:map-diversity-sampling). (\\#fig:correlation-sampling)'}
 knitr::include_graphics("figures/plot_sampling_effort.pdf")
@@
 
-These patterns highlight our lack of knowledge about the marine fauna: `r prop_singleton_species`% of the species are known from a single record.
-Furthermore, the strong correlation between the number of samples collected and the species richness suggests that any new sampling effort will lead to new records (Fig.
-3).
-This is also reflected in the cumulative numbers of species recorded for the first time and the number of samples through time (Fig.
-4).
-The slowing down of collections in the 1970 has been accompanied with a slowing down in the discovery of new records: `r n_spp_comparison$n_spp_2017_2000` species have been documented for the first time since 2000, while `r n_spp_comparison$n_spp_1970_1960` have been documented between 1960 and 1970.
+The inadequacy of the sampling effort compared to the diversity of the fauna is also reflected by the fact that `r prop_singleton_species`% of the species are known from a single record.
+The strong correlation between the number of samples collected and the species richness suggests that any new sampling effort will lead to new records (Fig.
+\@ref(fig:map-diversity-sampling)) and new species.
+Another indication of this inadequacy is the relation between the number of collected specimen and the number of species recorded (Fig.
+\@ref(fig:idigbio-through-time)).
+Initial explorations in the late ninteeth century lead to an initial exponential growth in the knowledge of the fauna, that tapered off, until the 1970's when intensive collecting efforts led to an increase in the recording of new taxa (Fig.
+\@ref(fig:idigbio-through-time)).
+Since then, the slowing down of collections in the 1970 has been accompanied with a slowing down in the discovery of new records: `r n_spp_comparison$n_spp_2017_2000` species have been documented for the first time since 2000 (`r format_output(n_spp_comparison$n_spp_2017_2000/18)` species.year^-1^), while `r n_spp_comparison$n_spp_1970_1960` have been documented between 1960 and 1970 (`r format_output(n_spp_comparison$n_spp_1970_1960/11)` species.year^-1^).
+Coordinated large-scale biodiversity expeditions can produce a wealth of data and new records.
+For instance, a recent study in Western Australia reported that 135 samples led to 805 species of crustaceans and polychates, 94.6% and 72% of which were undescribed [@Poore2014].
+
+```{r idigbio-through-time, fig.cap='Cumulative number of samples (A) and species (B) of marine invertebrates recorded through time based on iDigBio records. (\\#fig:idigbio-through-time)'}
+knitr::include_graphics("figures/idigbio_through_time.pdf")
+```
+
+Modelling the geographic distributions of species, and how they can be impacted by global change has undergone important advances recently.
+However, it is generally recognized that to obtain reliable results such efforts require to have accurate geographic information for a minimum of 10 specimens but having at least 30 is preferred.
+Based on iDigBio data, `r prop_10_specimens_species`% of species have less than 10 specimens and `r prop_30_specimens_species`% have less than 30 specimens.
+This suggests that our understanding of the  geographic range of species is limited as most species are known from only a few locations, while efforts to estimate their current and future ranges based on occurrence data are rendered moot by the low number of data points available.
 
 There are important discrepancies in our level of knowledge across phyla.
-While larger, and accessible groups are well represented in museum collections, and have a reasonable proportion of their known species represented in DNA barcode databases, the small organisms of the meiofauna are virtually absent: 12 of the 25 phyla represented in the dataset have less than 500 specimens in museums.
+While larger, and accessible groups are well represented in museum collections, the small organisms of the meiofauna are virtually absent: 12 of the 25 phyla represented in the dataset have less than 500 specimens in museums.
 
@@
 
-### Current genetic data and the future of biodiveristy discovery and monitoring
+### Genetic data for the future of biodiversity discovery and monitoring
 
@@
 
-![Cumulative number of samples through time](figures/cum_samples_through_time.pdf)
-
-![Cumulative number of species recorded for the first time through time](figures/cum_spp_through_time.pdf)
-
 ![Proportion of species occurring in natural history collections that have been DNA barcoded](figures/proportion_barcoded.pdf)
 
-![Relation between the number of specimen and the number of species collected](figures/plot_sampling_effort.pdf)
+## References
 

### 2017-06-01 manuscript.Rmd
add info about low number of records for rare phyla

@@
 There are important discrepancies in our level of knowledge across phyla.
-While larger, and accessible groups are well represented in museum collections, the small organisms of the meiofauna are virtually absent: 12 of the 25 phyla represented in the dataset have less than 500 specimens in museums.
-
-However, taxonomic expertise is dwindling, and there are now only a handful of US natural history museums with active marine invertebrates collections.
-Museum collections have been recording our knowledge about species and their geographic ranges, but until recently, this information was fragmented across institutions and difficult to synthesize.
-With collection digitization and other large-scale projects have aimed at gathering biodiversity information, public databases can now be used to enable synthesis and highlight knowledge gaps.
-Here we use these data to synthetic the state of knowledge for marine invertebrates, and demonstrate that the stage is set to combine these data sources to inform where future efforts for documenting biodiversity.
+While larger, and accessible groups represent the bulk of the records for marine invertebrates in museum collections, the small organisms of the meiofauna are virtually absent.
+Of the `r n_idigbio_phyla` phyla represented in iDigBio, `r records_rare_phyla$n_rare_phyla` are represented by less than 500 records, and make up `r records_rare_phyla$prop_rare_phyla`% of the total number of records.
 
@@
     quality in databases.
+
+However, taxonomic expertise is dwindling, and there are now only a handful of US natural history museums with active marine invertebrates collections.
+Museum collections have been recording our knowledge about species and their geographic ranges, but until recently, this information was fragmented across institutions and difficult to synthesize.
+With collection digitization and other large-scale projects have aimed at gathering biodiversity information, public databases can now be used to enable synthesis and highlight knowledge gaps.
+Here we use these data to synthetic the state of knowledge for marine invertebrates, and demonstrate that the stage is set to combine these data sources to inform where future efforts for documenting biodiversity.
 

### 2017-06-01 R/correlation_specimens_species.R

@@ -51,3 +51,18 @@ calc_prop_species_not_collected_since <- function(idig, max_year) {
         )
     format_output(nrow(dplyr::filter(res, last_collected < max_year))/nrow(res)*100)
 }
+
+calc_records_rare_phyla <- function(idig) {
+    res <- idig %>%
+        dplyr::count(clean_phylum, sort = TRUE) %>%
+        dplyr::mutate(is_rare = dplyr::if_else(n < 500, n, 0L))
+
+    list(
+        n_rare_phyla = dplyr::filter(res, is_rare > 0) %>% nrow(),
+        prop_rare_phyla = format_output((sum(res$is_rare)/sum(res$n))*100)
+    )
+}
+
+calc_n_idigbio_phyla <- function(idig) {
+    n_distinct(idig$clean_phylum)
+}

### 2017-06-01 R/bold.R
make proportion barcoded graph look like the others

@@ -166,12 +166,14 @@ summary_bold_status <- function(gom_bold, koz_bold, idig_bold) {
                                      )
                ) %>%
             dplyr::mutate(taxon_name = capitalize(taxon_name)) %>%
-            ggplot(aes(x = reorder(taxon_name, p_has_bold), y = p_has_bold, fill = data_source)) +
+            ggplot(aes(x = reorder(taxon_name, p_has_bold), y = p_has_bold,
+                       fill = data_source)) +
             geom_col(position = "dodge") +
             xlab("") + ylab("Proportion of species with available DNA barcodes") +
             scale_fill_viridis(discrete = TRUE,
-                               name = "Data source",
-                               labels = c("all iDigBio", "Gulf of Mexico", "Salish Sea")) +
+                               name = "",
+                               labels = c("US EEZ", "Gulf of Mexico", "Pacific Northwest")) +
+            theme_ipsum(base_family = "Ubuntu Condensed") +
             coord_flip()
 }
 

### 2017-06-01 manuscript.Rmd
add text to manuscript

--- 2017-06-01 manuscript.Rmd
@@
 
-<!-- Species identification from DNA sequences is poised to revolutionize our understanding of biodiversity, to facilitate environmental monitoring, and to disentangle trophic networks.
-However, this approach relies on comprehensive databases of accurately identified sequences.
-Since its introduction in the early 2000's, DNA barcoding, the sequencing of a common locus across taxa bearing species-specific information, has revitalized taxonomy by facilitating species delineation.
-More recently, advances in high-throughput sequencing technologies have enabled the parallel sequencing of thousands of species that occur in environmental samples.
-The approach initially developed to study microbial communities such as the ones from guts or soil, has been modified and extended to be applied to metazoan communities.
-This approach opens new research venues as it becomes possible to estimate the diversity from species-rich samples that were virtually inaccessible before.
-This is particularly true in marine communities that include multiple phyla and organisms varying in size by order of magnitude, making identification challenging.
-However, the utility of the data generated by these projects is enhanced by the possibility of matching the DNA sequences obtained to previously sequenced and correctly identified species.
-Here, we combined information available in public biodiversity databases to assess the proportion of identified species that have been sequenced in the continental United States.
-We show that the effort is consistent across the better-studied taxa, with ` average_prop_species_seq`% of the fauna sequenced.
+<!-- We show that the effort is consistent across the better-studied taxa, with ` average_prop_species_seq`% of the fauna sequenced.
 Using a DNA barcode dataset of ` n_larvae` planktonic larvae, representing ` n_esus` species, collected in coastal waters of Florida, we find a similar proportion of species that can be assigned to a previously sequenced species ( ` format_output(100 * prop_plankton_with_match)`%).
@@
 This can only be achieved with additional sampling and a commitment to training more taxonomists and experts that can identify the material collected.
+
+<!-- Species identification from DNA sequences is poised to revolutionize our understanding of biodiversity, to facilitate environmental monitoring, and to disentangle trophic networks.
+However, this approach relies on comprehensive databases of accurately identified sequences.
+-->
 
@@
 There are important discrepancies in our level of knowledge across phyla.
-While larger, and accessible groups are well represented in museum collections, the small organisms of the meiofauna are virtually absent: 12 of the 25 phyla represented in the dataset have less than 500 specimens in museums.
+While larger, and accessible groups represent the bulk of the records for marine invertebrates in museum collections, the small organisms, including representant of the meiofauna are virtually absent.
+Of the `r n_idigbio_phyla` phyla occurring in iDigBio, `r records_rare_phyla$n_rare_phyla` are represented by less than 500 records, and make up `r records_rare_phyla$prop_rare_phyla`% of the total number of records.
 
-However, taxonomic expertise is dwindling, and there are now only a handful of US natural history museums with active marine invertebrates collections.
-Museum collections have been recording our knowledge about species and their geographic ranges, but until recently, this information was fragmented across institutions and difficult to synthesize.
-With collection digitization and other large-scale projects have aimed at gathering biodiversity information, public databases can now be used to enable synthesis and highlight knowledge gaps.
-Here we use these data to synthetic the state of knowledge for marine invertebrates, and demonstrate that the stage is set to combine these data sources to inform where future efforts for documenting biodiversity.
+<!-- - get numbers of species described by year, look if there is a correlation between sample numbers and number of new species (possible lag with time for description) -->
+<!-- - see if proportion of species identified at genus (or higher) only has increased recently -->
 
@@
 
-The proportion of species that have DNA barcodes available is surprisingly constant across the better studied phyla.
-About 20% of species known to occur in the Gulf of Mexico and in the Salish Sea have DNA barcode records.
-Because geographic information is missing for most records (`r format_output(100*stat_bold$p_no_coords)`% of the `r format(stat_bold$n_total_records)` records analyzed here have no geographic coordinates information associated with them), and many marine invertebrate species are known to be cryptic complexes that are not as widely distributed as previously thought.
+Since its introduction in the early 2000's, DNA barcoding, the sequencing of a common locus across taxa bearing species-specific information, has revitalized taxonomy by facilitating species delineation [@Meyer2005-plos].
+More recently, advances in high-throughput sequencing technologies have enabled the parallel sequencing of thousands of species that occur in environmental samples [@Leray2015].
+This approach opens up new research venues it becomes possible to estimate the diversity from species-rich samples that were virtually inaccessible before.
+This is particularly true in marine communities that include multiple phyla and organisms varying in size by order of magnitude, making identification challenging.
+However, the utility of the data generated by these projects is enhanced by the possibility of matching the DNA sequences obtained to previously sequenced and correctly identified species [@Leray2015].
+With decreasing sequencing costs, and the prospect of automating sample collection, these technologies may soon be used routinely to monitory biodiversity [@Bohan2017].
+DNA barcoding provides the possibility to link these newly generated data with historical natural history knowledge about species and their distributions.
+However, the insights we will gain from these approaches are only going to be as good as the level of completeness and accuracy of the information included in reference databases.
+
+To estimate the proportion of the marine invertebrate diversity for which DNA barcodes are available, we queried BOLD (Barcode of Life Datasystem) for all the species that are listed in the regional lists for the Gulf of Mexico, the Pacific Northwest, as well as all the records for marine invertebrates that occur in the EEZ of the US found in iDigBio.
+Here, we focused on records based on museum specimens as some of them could potentially be used to complete current DNA barcoding efforts.
+
+The proportion of species that have DNA barcodes available is relatively constant across the better studied phyla at the scale of the EEZ.
+It varies between 12% for bryozoans to 28% for cnidarians.
+The proportions for species known to occur in the Gulf of Mexico are usually similar than for the US, while typically a higher fraction of species from the Pacific Northwest are available.
+Geographic information is missing for most records (`r format_output(100*stat_bold$p_no_coords)`% of the `r format(stat_bold$n_total_records)` BOLD records analyzed here have no geographic coordinates information associated with them), and many marine invertebrate species are known to be cryptic complexes that are not as widely distributed as previously thought.
 Therefore specimens identified under the same species but sampled in distant locations, could represent cryptic species.
@@
 This number is confounded by some taxonomic identification issues as `r stat_bold$n_shared_bins` BINs are shared across more than one species.
+
+```{r proportion-barcoded, fig.cap='Proportion of species known to occur in the US EEZ based on iDigBio records, in the Gulf of Mexico [@Tunnell2009], in the Pacific Northwest[@Kozloff1987], that have publicly available DNA barcodes.'}
+knitr::include_graphics("figures/proportion_barcoded.pdf")
+```
 
@@
 
+However, taxonomic expertise is dwindling, and there are now only a handful of US natural history museums with active marine invertebrates collections.
+Museum collections have been recording our knowledge about species and their geographic ranges, but until recently, this information was fragmented across institutions and difficult to synthesize.
+With collection digitization and other large-scale projects have aimed at gathering biodiversity information, public databases can now be used to enable synthesis and highlight knowledge gaps.
+Here we use these data to synthetic the state of knowledge for marine invertebrates, and demonstrate that the stage is set to combine these data sources to inform where future efforts for documenting biodiversity.
+
 Based on these results, we recommend that additional field expeditions are organized to sample more thoroughly the marine environment.
@@
 The specimens need to be sub-sampled and have tissues suitable for genomic analysis preserved, so we can sequence their DNA barcode, or other genomic resources.
-
-<!-- Paragraph on how sampling effort has changed through time. -->
-
-<!-- - get numbers of species described by year, look if there is a correlation between sample numbers and number of new species (possible lag with time for description) -->
-<!-- - see if proportion of species identified at genus (or higher) only has increased recently -->
 
@@
 DNA barcoding has also been seen as -->
-
-Recent developments in high-throughput sequencing technologies are giving new applications to these DNA barcode databases [@Leray2016].
-Metabarcoding has been developed to sequence in parallel the barcoding gene across the species included in environmental samples such as plankton tows, fish guts, or artificial structure scrappings [@Leray2015].
-With decreasing sequencing costs, and the prospect of automating sample collection, these technologies will soon be used routinely for ecosystem and biodiversity monitoring [@Bohan2017].
-DNA barcoding provides the possibility to link these newly generated data with historical natural history knowledge about species and their distributions.
-However, the insights we will gain from these approaches are only going to be as good as the level of completeness and accuracy of the information included in reference databases.
-Here, we mostly focus on museum specimens because we can validate their identification, and a fraction of these samples (if they have not been preserved in formalin) can be used to obtain DNA from them to complete existing databases of DNA barcodes.
 

### 2017-06-02 R/bold.R
use namespace for function

@@ -94,19 +94,19 @@ make_stat_bold <- function(gom_bld, koz_bld, gom_wrm, koz_wrm) {
     bin_data <- get_bold_bins(gom_wrm, koz_wrm)
 
     n_shared_bins <- bin_data %>%
-        group_by(bins) %>%
-        summarize(
+        dplyr::group_by(bins) %>%
+        dplyr::summarize(
             n_shared_bins = n_distinct(worms_valid_name)
         ) %>%
-        filter(n_shared_bins > 1) %>%
+        dplyr::filter(n_shared_bins > 1) %>%
         nrow()
 
     n_spp_multi_bin <- bin_data %>%
-        group_by(worms_valid_name) %>%
-        summarize(
+        dplyr::group_by(worms_valid_name) %>%
+        dplyr::summarize(
             n_spp_multi_bin = n_distinct(bins)
         ) %>%
-        filter(n_spp_multi_bin > 1) %>%
+        dplyr::filter(n_spp_multi_bin > 1) %>%
         nrow()
 
     list(

### 2017-06-02 manuscript.Rmd
add text, fix typos, polish things up

@@
     toc: no
-  bookdown::word_document2: null
 fontsize: 12pt
@@
 
+Museum collections have been recording our knowledge about species and their geographic ranges, but until recently, this information was fragmented across institutions and difficult to synthesize.
+The digitization of natural history collections and other biodiversity information systems can now be used to enable synthesis and highlight knowledge gaps.
+Here we use these data to synthesize the state of our knowledge for marine invertebrates in the continental US.
+We show that current knowledge and sampling efforts are inadequate given the diversity of the fauna and the geographic extent of US.
+While recent high-throughput sequencing technologies promise to greatly speed up the documentation of marine species and their ranges, its value would be greatly improved by anchoring the results into existing DNA sequence data and associated voucher.
+However, the best documented phyla have at most 25% of their species sequenced for COI.
+More sampling, coordinated at the scale of the US, is needed to improve our knowledge of species and their geographic ranges.
+This data is needed to detect the impact of global change and other perturbations.
+Advances in computing, information systems, and sequencing technologies, have set the stage to inform where future efforts for documenting biodiversity should be applied.
+
+## Introduction
+
 With increasing impact of human activities on the biosphere, we need efficient and accurate approaches to document changes in species ranges and community composition.
 While terrestrial ecosystems have received considerable attention, the baseline of healthy and resilient marine ecosystems are not well understood, and modeling the impact of human activities is limited to a few taxonomic groups.
-A major hinderance is that we do not have enough information about which species live where, and an unknown but potentially significant portion of marine biodiversity remains to be discovered and described.
+A major hindrance is that we do not have enough information about which species live where, and an unknown but potentially significant portion of marine biodiversity remains to be discovered and described.
 Additionally, coordinated marine ecosystem monitoring initiatives are lacking [@Duffy2013,@Borja2016].
-There is therefore a pressing need to improve our knowledge of marine systems before species go extinct, and their functionning are irreversibly altered.
+There is therefore a pressing need to improve our knowledge of marine systems before species go extinct, and their functioning is irreversibly altered.
 
@@
 However, with the development of public databases for natural history records (e.g.
-iDigBio [@Page2015]), species occurrences (e.g., GBIF [@Edwards2004], OBIS [@Zhang2002]), genetic data (e.g., GenBank [@Benson2013], BOLD [@Ratnasingham2007a]), and registers of species names (e.g., WoRMS [@Costello2013b]), we can now evaluate what we know.
-Here, we aggregate publicly available data for marine invertebrates that occur in the United States, in the Gulf of Mexico, and in the Pacific North-West, to highlight gaps in our knowledge.
-We show that public databases are relatively incomplete, and that most species are known from a few speciemens and a few locations.
+iDigBio [@Page2015]), species occurrences (e.g., GBIF [@Edwards2004], OBIS [@Zhang2002]), genetic data (e.g., GenBank [@Benson2013], BOLD [@Ratnasingham2007a]), and registers of species names (e.g., WoRMS [@Costello2013b]), we can better evaluate what and how much we know.
+Here, we aggregate publicly available data for marine invertebrates that occur in the continental United States, the Gulf of Mexico, and the Pacific Northwest, to highlight gaps in our knowledge.
+We show that public databases are relatively incomplete, and that most species are known from a few specimens and a few locations.
 We emphasize that our understanding of marine biodiversity, and the impact of human activities on communities require not only to get a more accurate list of the species that inhabits the oceans, but we also need more comprehensive data on species distributions and abundances.
 This can only be achieved with additional sampling and a commitment to training more taxonomists and experts that can identify the material collected.
-
-<!-- Species identification from DNA sequences is poised to revolutionize our understanding of biodiversity, to facilitate environmental monitoring, and to disentangle trophic networks.
-However, this approach relies on comprehensive databases of accurately identified sequences.
--->
 
@@
 To compare the faunistic lists from the Gulf of Mexico and the Pacific Northwest with records in public databases, we queried the iDigBio and OBIS databases for all records they included in the same geographic boundaries as the lists.
-After filtering these lists to only include the same higher taxonomic level taxa as the species lists, and keeping only records for which the species could be match to a species name in WoRMS, we had for the Gulf of Mexico `r nrow(idigbio_gom_records)` and `r nrow(obis_gom_records)` records for iDigBio and OBIS respectively representing `r dplyr::n_distinct(idigbio_gom_records$worms_valid_name)` and `r dplyr::n_distinct(obis_gom_records$worms_valid_name)` species.
+After filtering these lists to only include the same higher taxonomic level taxa as the species lists, and keeping only records for which the species could be matched to a species name in WoRMS, we had for the Gulf of Mexico `r nrow(idigbio_gom_records)` and `r nrow(obis_gom_records)` records for iDigBio and OBIS respectively, representing `r dplyr::n_distinct(idigbio_gom_records$worms_valid_name)` and `r dplyr::n_distinct(obis_gom_records$worms_valid_name)` species.
 For the Pacific Northwest, we retrieved `r nrow(idigbio_kozloff_records)` and `r nrow(obis_kozloff_records)` records for iDigBio and OBIS respectively representing `r dplyr::n_distinct(idigbio_kozloff_records$worms_valid_name)` and `r dplyr::n_distinct(obis_kozloff_records$worms_valid_name)` species.
@@
 
-Species that are in the database but not in the list could be: (1) records not considered by the authors of the checklist: either because they were not accessible or known to the authors; (2) records erronously excluded the records (2) records that have been discovered since the list has been assembled; (3) records with incorrect identification; (4) records with incorrect geographic coordinates: their associated geographic information in the database places them in the Gulf of Mexico or the Pacific Northwest, but they were actually collected elsewhere.
+Species that are in the database but not in the list could be: (1) records not considered by the authors of the checklist: either because they were not accessible or known to the authors; (2) records erroneously excluded the records (2) records that have been discovered since the list has been assembled; (3) records with incorrect identification; (4) records with incorrect geographic coordinates: their associated geographic information in the database places them in the Gulf of Mexico or the Pacific Northwest, but they were actually collected elsewhere.
 
@@
 Similar patterns can be observed for the Pacific Northwest, but overall the levels of data availability are lower: from `r format_output(summary_richness_per_db_koz$prop_obis_in_pnw[summary_richness_per_db_koz$taxon_name == "Porifera"]*100)`% for sponges to `r format_output(summary_richness_per_db_koz$prop_obis_in_pnw[summary_richness_per_db_koz$taxon_name == "Annelida"]*100)`% for annelids for the proportion of species that have records in OBIS, and from  `r format_output(summary_richness_per_db_koz$prop_idigbio_in_pnw[summary_richness_per_db_koz$taxon_name == "Porifera"]*100)`% for sponges to `r format_output(summary_richness_per_db_koz$prop_idigbio_in_pnw[summary_richness_per_db_koz$taxon_name == "Mollusca"]*100)`% for molluscs for the proportion of species that have records in iDigBio.
-Museum specimens are critical to validate the occurence of species [@Rocha2014], especially for marine invertebrates for which species limits based on morphology have been known to be challenged by other lines of evidence including molecular data [@Knowlton2000].
-Some of the discrepencies observed between the number of species for which occurence data and museum specimens are available suggest that additional digitization of collections (including capturing their geographic information) are needed.
+
+Museum specimens are critical to validate the occurrence of species [@Rocha2014], especially for marine invertebrates for which species limits based on morphology have been known to be challenged by other lines of evidence including molecular data [@Knowlton2000].
+Some of the discrepancies observed between the number of species for which occurrence data and museum specimens are available suggest that additional digitization of collections (including capturing their geographic information) are needed.
 
@@
 
-```{r map-diversity-sampling, fig.cap='Spatial distribution of the number of samples (A) and species richness (B) for marine invertebrates based on specimen data available from iDigBio within the limits of the Exclusive Economic Zone for the contiental USA.
+```{r map-diversity-sampling, fig.cap='Spatial distribution of the number of samples (A) and species richness (B) for marine invertebrates based on specimen data available from iDigBio within the limits of the Exclusive Economic Zone for the continental USA.
 (\\#fig:map-diversity-sampling)'}
@@
 When investigating the relation between the number of samples and the number of species for each cell of the map (Fig.
-\@ref(fig:correlation-sampling)), we can see that until reaching about 100 records per cell there is little redundancy in the sampled species.
+\@ref(fig:correlation-sampling)), we can see that until reaching about 100 records per cell there is little redundancy in the sampled species (each record represent a different species).
 
@@
 The inadequacy of the sampling effort compared to the diversity of the fauna is also reflected by the fact that `r prop_singleton_species`% of the species are known from a single record.
-The strong correlation between the number of samples collected and the species richness suggests that any new sampling effort will lead to new records (Fig.
+The strong correlation between the number of samples collected and the species richness suggests that any new sampling effort will lead to new observations (Fig.
 \@ref(fig:map-diversity-sampling)) and new species.
-Another indication of this inadequacy is the relation between the number of collected specimen and the number of species recorded (Fig.
+Another indication of this inadequacy is the relation between the number of collected specimen and the number of species recorded through time (Fig.
 \@ref(fig:idigbio-through-time)).
-Initial explorations in the late ninteeth century lead to an initial exponential growth in the knowledge of the fauna, that tapered off, until the 1970's when intensive collecting efforts led to an increase in the recording of new taxa (Fig.
+Initial explorations in the late ninetieth century lead to an initial exponential growth in the knowledge of the fauna, that tapered off, until the 1970's when intensive collecting efforts led to an increase in the recording of new taxa (Fig.
 \@ref(fig:idigbio-through-time)).
@@
 Coordinated large-scale biodiversity expeditions can produce a wealth of data and new records.
-For instance, a recent study in Western Australia reported that 135 samples led to 805 species of crustaceans and polychates, 94.6% and 72% of which were undescribed [@Poore2014].
+For instance, a recent study in Western Australia reported that 135 samples led to 805 species of crustaceans and polychaetes, 94.6% and 72% of which were undescribed [@Poore2014].
 
@@
 
-Modelling the geographic distributions of species, and how they can be impacted by global change has undergone important advances recently.
+Modeling the geographic distributions of species, and how they can be impacted by global change has undergone important advances recently.
 However, it is generally recognized that to obtain reliable results such efforts require to have accurate geographic information for a minimum of 10 specimens but having at least 30 is preferred.
@@
 There are important discrepancies in our level of knowledge across phyla.
-While larger, and accessible groups represent the bulk of the records for marine invertebrates in museum collections, the small organisms, including representant of the meiofauna are virtually absent.
+While larger, and accessible groups represent the bulk of the records for marine invertebrates in museum collections, the small organisms, including representatives of the meiofauna are virtually absent.
 Of the `r n_idigbio_phyla` phyla occurring in iDigBio, `r records_rare_phyla$n_rare_phyla` are represented by less than 500 records, and make up `r records_rare_phyla$prop_rare_phyla`% of the total number of records.
@@
 More recently, advances in high-throughput sequencing technologies have enabled the parallel sequencing of thousands of species that occur in environmental samples [@Leray2015].
-This approach opens up new research venues it becomes possible to estimate the diversity from species-rich samples that were virtually inaccessible before.
+This approach opens up new research venues as it becomes possible to estimate the diversity from species-rich samples that were virtually inaccessible before.
 This is particularly true in marine communities that include multiple phyla and organisms varying in size by order of magnitude, making identification challenging.
-However, the utility of the data generated by these projects is enhanced by the possibility of matching the DNA sequences obtained to previously sequenced and correctly identified species [@Leray2015].
-With decreasing sequencing costs, and the prospect of automating sample collection, these technologies may soon be used routinely to monitory biodiversity [@Bohan2017].
+However, the utility of the data generated by these projects is enhanced by the possibility of matching DNA sequences obtained to previously sequenced and correctly identified species [@Leray2015].
+With decreasing sequencing costs, and the prospect of automating sample collection, these technologies may soon be used routinely to monitor biodiversity [@Bohan2017].
 DNA barcoding provides the possibility to link these newly generated data with historical natural history knowledge about species and their distributions.
@@
 It is thus crucial to ensure that accurate geographic information is provided when DNA barcodes are published, and that additional samples are collected, to capture more accurately their geographic range and the extent of their genetic diversity.
-This high level of cryptic diversity is reflected in the fact that `r stat_bold$n_spp_muti_bin` species out of `r stat_bold$n_spp` include more than 1 BIN.
-This number is confounded by some taxonomic identification issues as `r stat_bold$n_shared_bins` BINs are shared across more than one species.
+This cryptic diversity is reflected in the fact that `r stat_bold$n_spp_multi_bin` species out of `r stat_bold$n_spp` include more than 1 BIN (BOLD Index Number, operational taxonomic units that closely resemble species [@Ratnasingham2013]).
 
@@
 
-## What to do next?
+## Where do we go from there?
 
-* Improve data quality:
-  - easy: fix mispelling in identifications
-  - improve identification of unidentified lots (requires taxonomists/specialists to be able to identify the specimens)
-  - add and validate geographic information associated with specimens
-  - develop infrastructure and tools to continuously check and validate data
-    quality in databases.
+Based on the results of this study, we recommend a multi-level approach to not only improve the quality of the existing data, but also to efficiently collect new data.
 
-However, taxonomic expertise is dwindling, and there are now only a handful of US natural history museums with active marine invertebrates collections.
-Museum collections have been recording our knowledge about species and their geographic ranges, but until recently, this information was fragmented across institutions and difficult to synthesize.
-With collection digitization and other large-scale projects have aimed at gathering biodiversity information, public databases can now be used to enable synthesis and highlight knowledge gaps.
-Here we use these data to synthetic the state of knowledge for marine invertebrates, and demonstrate that the stage is set to combine these data sources to inform where future efforts for documenting biodiversity.
+### Improve existing data
 
-Based on these results, we recommend that additional field expeditions are organized to sample more thoroughly the marine environment.
-The data from museum collections make clear that any additional samples will contain new species and new records.
-However, taxonomists are needed to identify and describe these new species.
-The samples collected during these expeditions need to be photographed to document live appearance that would otherwise be lost during preservation.
+With data aggregation of natural history collections through initiatives like iDigBio, and the compilation of a nearly complete register of marine species with WoRMS, in combination with the development of algorithms for text manipulation, it is increasingly easier to validate and spell check the scientific names of the identifications of museum specimens.
+Additionally, because WoRMS links the history of the names and their synonymies, it is possible to dissociate the concepts of species identification with the concept of species identity, allowing to compare specimens identified at different time periods.
+
+Similarly, a significant fraction of museum records do not have or have inaccurate geographic information associated with them.
+While tools have been developed to convert names of locales into GPS coordinates on land, similar approaches may not work for locales at sea.
+However, geographic information is available in other forms, and can be digitized.
+
+For both the spelling of scientific names, and the geographic information, computing tools can be developed to detect issues and propose solutions.
+There are opportunities for services using publicly available data to be developed to make curators and collection managers aware of these issues so that they can fixed at the source.
+It however requires that resources are available to museum collections to address these issues.
+
+Increasing the computing literacy of the staff of museum in order for them to manipulate and use the data from their collection database is equally critical.
+As data across institutions get pooled into the large biodiversity information systems, it is crucial that data standards such as Darwin Core are embraced by the community.
+
+<!-- - develop infrastructure and tools to continuously check and validate data quality in databases. -->
+
+### Collect better data faster
+
+This study highlights that we have too little data about too few species of marine invertebrates, we recommend that additional coordinated field expeditions are organized to sample more thoroughly the marine environment.
+Data from museum collections make clear that any additional samples will contain new species and new records.
+Samples collected during these expeditions need to be photographed to document live appearance that would otherwise be lost during preservation.
 We also need to capture metadata associated with the specimens to be able to understand better the physical characteristics of the habitats for these species.
-The specimens need to be sub-sampled and have tissues suitable for genomic analysis preserved, so we can sequence their DNA barcode, or other genomic resources.
+Specimens need to be sub-sampled and have tissues suitable for genomic analysis preserved for their DNA barcode, or any other genomic resources.
+Some of these parts of data and specimen collection can be automated and captured digitally directly in the field speeding up data collection.
 
-; and we estimate how many of the species have genetic data publicly available that would be useful in the context of identifying samples from metabarcoding studies
+Coupling these more traditional approaches of biodiversity documentation with metabarcoding is a promising approach to screen an habitat for its potential for species that do not yet occur in public databases.
+Metabarcoding can complement and take advantage of the DNA barcodes already available.
+Additionally, in the future, data generated by metabarcoding efforts can be re-interpreted as more DNA barcodes associated with correctly identified specimens become available.
+Building the library of DNA barcodes in parallel of metabarcoding efforts ensures that matches in the database will correspond to actual specimens deposited in natural history collections; ruling out possible artifacts of the method.
+Information about species identity and distribution will thus build on existing natural history data rather than being limited to matching DNA sequences across studies.
 
-Using metabarcoding is a promising approach to screen an habitat for its potential for species that are not found in public database, to maximize the usefulness in the future of the data generated by an expedition.
-Metabarcoding approaches need to complement and take advantage of the DNA barcodes already available.
-Additionally, in the future, data generated by metabarcoding efforts can be re-interpreted as more DNA barcodes become available.
-Building the library of DNA barcodes in parallel of metabarcoding efforts also ensures that matches in the database will correspond to actual specimens deposited in natural history collections; ruling out possible artifacts of the method.
-The knowledge will also build on existing natural history data rather than having to restart from scratch.
-Beyond their utility in documenting diversity, metabarcoding is also promising for understanding the functioning of ecosystems.
+<!-- Beyond their utility in documenting diversity, metabarcoding is also promising for understanding the functioning of ecosystems. -->
 
-## Conclusions
+However, taxonomists and experts are needed to collect, sort, process, and identify these samples and the data that come along with them.
+Even if technological advances can greatly speed up the documentation of species and their distributions, trained personnel is needed.
+Taxonomic expertise is dwindling, and there are now only a handful of US natural history museums with active marine invertebrates collections.
 
 Given the urgency of the biodiversity crisis, we need to develop data-driven initiatives to efficiently document biodiversity, and build on our existing knowledge to collect information to improve our understanding of marine systems.
-Additionally,
 
-Biodiversity discovery and documentation has been transformed in the 21st century by molecular data to facilitate species identification, delineate species.
+<!-- Biodiversity discovery and documentation has been transformed in the 21st century by molecular data to facilitate species identification, delineate species.
 And more recently, the development of high throughput sequencing technologies is used to quantify diversity and uncover large scale patterns of biodiversity organization.
@@
 By matching DNA barcodes of unidentified species to a reference public database of accurately identified and vouchered specimens, DNA barcoding can greatly speed up species identification: sequences with no match make it easier to flag specimens that warrant further taxonomic attention.
+-->
 
@@
 
+<!-- widening gap with other groups (vertebrates, plants) for which more comprehensive data is available -->
+
 <!-- here there are 3 things: data availability (having good APIs to access the
@@
 data accuracy (can we trust the data that is found in the databases) -->
-
-![Proportion of species occurring in natural history collections that have been DNA barcoded](figures/proportion_barcoded.pdf)
 
@@
 
-Except for the few phyla that are known to include only a small number of species (Cycliophora, Phoronida, Chaetognatha), microscopic groups, typically component of the meiofauna, are disproportionally under-represented in museum collections, and none or few species are available in public databases.
-
-Our results highlight that most of our knowledge of marine biodiversity come from large expeditions carried in the late 19th century and after World War II.
-Since then, sampling effort has dramatically decreased, and taxonomic expertise is waning.
-Yet, our ability to assign DNA sequences to species names will greatly improve the value of the data generated by meta-barcoding studies.
-
-and reflect that a large fraction of the diversity of marine invertebrates remain to be sequenced.
-
 Marine invertebrates are the biodiversity.
@@
 
-Marine invertebrates are however disproportionally under-represented in DNA sequence databses.
-And they have a higher rate of misidentification.
-There are however some pitfalls to this number.
-It might actually be even higher given that they have been identified by a single taxonomist given the lack of redundancy in expertise.
-
 To illustrate the consequences of having so few barcodes for invertebrates, if we take the faunistic list of IRL, only about 20% of the species known to be occuring in the area actually have DNA sequences available for them.
@@
 
-We also sequenced xx larvae occurring in the plankton at the Whitney Marine Lab, representing yy species (estimated from genetic distances).
-Of these yy species, only zz% had good matches in BOLD/GenBank.
-In both cases, taxa1, taxa2, and taxa3 are the most underrepresented in BOLD.
-
 # The waining of taxonomy
-
-Total number of collections in the US, how many have marine invertebrate collections, how many have active curators.
 

### 2017-06-02 manuscript.Rmd
better title (maybe)

@@
 #author: Fran\c{c}ois Michonneau
-title: "Data-driven approaches to identifying knowledge gaps in marine invertebrate biodiversity"
-date: '`r format(Sys.time())`'
+title: "The state of knowledge for marine invertebrate biodiversity in the continental US"
+date: ' ' #'`r format(Sys.time())`'
 output:

### 2017-06-05 manuscript.Rmd
re-organize discussion about barcoding results

@@
 The proportions for species known to occur in the Gulf of Mexico are usually similar than for the US, while typically a higher fraction of species from the Pacific Northwest are available.
-Geographic information is missing for most records (`r format_output(100*stat_bold$p_no_coords)`% of the `r format(stat_bold$n_total_records)` BOLD records analyzed here have no geographic coordinates information associated with them), and many marine invertebrate species are known to be cryptic complexes that are not as widely distributed as previously thought.
+In 2010, it was estimated that a global scale, less than 10% of the species for each marine phylum had been sequenced [@Radulovici2010], suggesting that the US fauna is generally better characterized than the rest of the world, combined with many other species have been sequenced since this report.
+
+Using the fraction of known species that have a barcode as a yard stick of the progress made is not sufficient.
+Many marine invertebrate species are known to be cryptic complexes that are not as widely distributed as currently reported [@Knowlton2000,@Meyer2005-evolution].
 Therefore specimens identified under the same species but sampled in distant locations, could represent cryptic species.
-It is thus crucial to ensure that accurate geographic information is provided when DNA barcodes are published, and that additional samples are collected, to capture more accurately their geographic range and the extent of their genetic diversity.
-This cryptic diversity is reflected in the fact that `r stat_bold$n_spp_multi_bin` species out of `r stat_bold$n_spp` include more than 1 BIN (BOLD Index Number, operational taxonomic units that closely resemble species [@Ratnasingham2013]).
+Geographic information is missing for most records (`r format_output(100*stat_bold$p_no_coords)`% of the `r format(stat_bold$n_total_records)` BOLD records analyzed here have no geographic coordinates information associated with them).
+It is thus crucial to ensure that accurate geographic information is provided when DNA barcodes are published, and that enough samples for each species are sequenced, to capture more accurately their geographic range and the extent of their genetic diversity.
+This cryptic diversity is also at least partially reflected in the fact that `r stat_bold$n_spp_multi_bin` species out of `r stat_bold$n_spp` include more than 1 BIN (BOLD Index Number, operational taxonomic units that closely resemble species [@Ratnasingham2013]).
 

### 2017-06-09 R/comparison_by_geo.R
only display a few phyla for data source comparison

@@ -114,13 +114,13 @@ compare_taxonomy_database_overlap <- function(database_overlap) {
 
 
 plot_database_overlap <- function(comp_db) {
-    phyla_to_keep <- comp_db %>%
-        dplyr::distinct(region, phylum) %>%
-        dplyr::count(phylum) %>%
-        dplyr::filter(n == 2) %>%
-        .[["phylum"]]
+
+    phyla_to_keep <- c("mollusca", "arthropoda", "annelida",
+                       "cnidaria", "bryozoa", "porifera",
+                       "echinodermata", "chordata")
 
     comp_db <- comp_db %>%
+        dplyr::mutate(phylum = tolower(phylum)) %>%
         dplyr::filter(phylum %in% phyla_to_keep)
 
     ## to have 2 sided plot, we take the negative value of the count
@@ -151,7 +151,7 @@ plot_database_overlap <- function(comp_db) {
                        position = position_dodge(width = dodge_width)) +
         geom_linerange(aes(x = phylum, ymin = offset, ymax = offset + not_in_list),
                        position = position_dodge(width = dodge_width)) +
-         geom_point(aes(y = offset + not_in_list), size = .7,
+        geom_point(aes(y = offset + not_in_list), size = .7,
                         position = position_dodge(width = dodge_width)) +
         geom_text(aes(x = phylum, label = phylum, y = 0),
                   family = "Ubuntu Condensed",
@@ -177,7 +177,7 @@ plot_database_overlap <- function(comp_db) {
                axis.text.x = element_text(size = 10),
                axis.text.y = element_blank(),
                strip.text = element_text(size = 18, face = "bold", hjust = 0.030),
-               plot.background = element_rect(),
+               plot.background = element_blank(),
                legend.text  = element_text(family = "Ubuntu Condensed", size = 10),
                panel.spacing.x = unit(5, "lines")
                ) +

### 2017-06-09 R/time_plot.R
[fix #10] make plot of instiutions through time

@@ -254,3 +254,32 @@ plot_change_trend_through_time <- function(knowledge_through_time) {
         geom_point() +
         facet_grid(database ~ fauna)
 }
+
+
+plot_institutions_through_time <- function(idig_records) {
+    idig <- idig_records %>%
+        idigbio_add_year() %>%
+        dplyr::filter(is_marine == TRUE, is_binomial == TRUE, !is.na(worms_id))
+
+    inst_to_keep <- idig %>%
+        dplyr::count(institutioncode, sort = TRUE) %>%
+        dplyr::filter(n  >= 1000) %>%
+        dplyr::distinct() %>%
+        .[["institutioncode"]]
+
+    tmp_data <- data_frame(institutioncode = inst_to_keep)
+
+    idig %>%
+        dplyr::filter(institutioncode %in% inst_to_keep) %>%
+        dplyr::count(institutioncode, year) %>%
+        tidyr::complete(year = 1850:2017, institutioncode,  fill = list(n = 0)) %>%
+        dplyr::filter(year >= 1999) %>%
+        dplyr::group_by(institutioncode) %>%
+        dplyr::arrange(year) %>%
+        dplyr::mutate(cum_n = cumsum(n)) %>% filter(year == 2017) %>% arrange(desc(cum_n))
+        dplyr::ungroup() %>%
+        ggplot(aes(x = year, y = cum_n, #color = institutioncode,
+                   fill = institutioncode)) +
+        geom_col() + facet_wrap(~ institutioncode, scales = "free_y")
+
+}

### 2017-06-09 manuscript.Rmd
edits to manuscript

@@
 With increasing impact of human activities on the biosphere, we need efficient and accurate approaches to document changes in species ranges and community composition.
-While terrestrial ecosystems have received considerable attention, the baseline of healthy and resilient marine ecosystems are not well understood, and modeling the impact of human activities is limited to a few taxonomic groups.
-A major hindrance is that we do not have enough information about which species live where, and an unknown but potentially significant portion of marine biodiversity remains to be discovered and described.
+While terrestrial ecosystems have received considerable attention, the baseline of healthy and resilient marine ecosystems are not well understood [@Worm2006], and modeling the impact of human activities is limited to a few taxonomic groups [@Tittensor2010].
+A major hindrance for marine species, is that we do not have enough information about which species live where, and an unknown but potentially significant portion of marine biodiversity remains to be discovered and described [@Mora2011].
 Additionally, coordinated marine ecosystem monitoring initiatives are lacking [@Duffy2013,@Borja2016].
@@
 Until very recently, our knowledge of biodiversity was scattered among heterogeneous sources (e.g.
-natural history collections, publications, expert knowledge), and difficult to access.
-These factors have contributed to the lack of synthesis and have made the identification of knowledge gaps difficult.
-However, with the development of public databases for natural history records (e.g.
-iDigBio [@Page2015]), species occurrences (e.g., GBIF [@Edwards2004], OBIS [@Zhang2002]), genetic data (e.g., GenBank [@Benson2013], BOLD [@Ratnasingham2007a]), and registers of species names (e.g., WoRMS [@Costello2013b]), we can better evaluate what and how much we know.
+natural history collections, publications, expert knowledge), difficult to access, and therefore difficult to synthesize.
+The development of public databases for natural history records (e.g.
+iDigBio [@Page2015]), species occurrences (e.g., OBIS [@Zhang2002]), genetic data (e.g., GenBank [@Benson2013], BOLD [@Ratnasingham2007a]), and registers of species names (e.g., WoRMS [@Costello2013b]), now allow us to evaluate what and how much we know.
 Here, we aggregate publicly available data for marine invertebrates that occur in the continental United States, the Gulf of Mexico, and the Pacific Northwest, to highlight gaps in our knowledge.
-We show that public databases are relatively incomplete, and that most species are known from a few specimens and a few locations.
+We show that public databases are relatively incomplete, and that most species are known from only a few specimens and a few locations.
 We emphasize that our understanding of marine biodiversity, and the impact of human activities on communities require not only to get a more accurate list of the species that inhabits the oceans, but we also need more comprehensive data on species distributions and abundances.
@@
 For the Pacific Northwest, we retrieved `r nrow(idigbio_kozloff_records)` and `r nrow(obis_kozloff_records)` records for iDigBio and OBIS respectively representing `r dplyr::n_distinct(idigbio_kozloff_records$worms_valid_name)` and `r dplyr::n_distinct(obis_kozloff_records$worms_valid_name)` species.
-For the Gulf of Mexico, the majority of the species included in the expert lists are recorded in both iDigBio and OBIS (`r format_output(species_in_common$n_spp_gom_idigbio_obis/species_in_common$n_spp_gom*100)` %), or OBIS (`r format_output(species_in_common$n_spp_gom_obis/species_in_common$n_spp_gom*100)` %) (Fig.
+For the Gulf of Mexico, the majority of the species included in the expert lists are recorded in both iDigBio and OBIS (`r format_output(species_in_common$n_spp_gom_idigbio_obis/species_in_common$n_spp_gom*100)` %), or only OBIS (`r format_output(species_in_common$n_spp_gom_obis/species_in_common$n_spp_gom*100)` %) (Fig.
 \@ref(fig:database-overlap)).
-However, for the Pacific Northwest, only `r format_output(species_in_common$n_spp_kozloff_idigbio_obis/species_in_common$n_spp_kozloff*100)`% and `r format_output(species_in_common$n_spp_kozloff_obis/species_in_common$n_spp_kozloff*100)`% of the species from the regional list had records in OBIS and iDigBio or OBIS respectively (Fig.
+For the Pacific Northwest, these proportions were much lower with only `r format_output(species_in_common$n_spp_kozloff_idigbio_obis/species_in_common$n_spp_kozloff*100)`% of the species from the regional list had records in OBIS and iDigBio, and `r format_output(species_in_common$n_spp_kozloff_obis/species_in_common$n_spp_kozloff*100)`%  in OBIS only (Fig.
 1).
+Overall, there is more overlap among the data sources for the Gulf of Mexico than the Pacific Northwest.
 
-```{r compare_database_overlap, fig.cap='Overlap between the regional species list (list), records in iDigBio (idigbio), records in OBIS (obis), and records in BOLD (bold) for the Gulf of Mexico (left) and the Pacific Northwest (right).
+```{r compare_database_coverage, fig.cap='Overlap between the regional species list (list), records in iDigBio (idigbio), records in OBIS (obis), and records in BOLD (bold) for the Gulf of Mexico (left) and the Pacific Northwest (right).
 (\\#fig:database-overlap)', echo=FALSE}
 knitr::include_graphics("figures/database_coverage.pdf")
+```
+
+```{r compare_database_overlap, fig.cap='Comparison of species by phylum across the public databases and the faunistic list.', ech=FALSE}
+knitr::include_graphics("figures/database_overlap.pdf")
 ```
@@
 
-```{r proportion-barcoded, fig.cap='Proportion of species known to occur in the US EEZ based on iDigBio records, in the Gulf of Mexico [@Tunnell2009], in the Pacific Northwest[@Kozloff1987], that have publicly available DNA barcodes.'}
+```{r proportion-barcoded, fig.cap='Proportion of species known to occur in the US EEZ based on iDigBio records, in the Gulf of Mexico [@Tunnell2009], in the Pacific Northwest[@Kozloff1987], that have publicly available DNA barcodes.
+Numbers next to the bars indicate the number of species.'}
 knitr::include_graphics("figures/proportion_barcoded.pdf")
@@
 With data aggregation of natural history collections through initiatives like iDigBio, and the compilation of a nearly complete register of marine species with WoRMS, in combination with the development of algorithms for text manipulation, it is increasingly easier to validate and spell check the scientific names of the identifications of museum specimens.
-Additionally, because WoRMS links the history of the names and their synonymies, it is possible to dissociate the concepts of species identification with the concept of species identity, allowing to compare specimens identified at different time periods.
+Linking species names to identifiers in taxonomic databases would facilitate data linking, specimen discoverability, and data usability.
+Additionally, because species names registers like WoRMS link the history of the names and their synonymies, it is possible to dissociate the concepts of species identification with the concept of species identity, allowing researchers to compare specimens identified at different time periods, a tedious process that would have required expert knowledge in the past.
 
 Similarly, a significant fraction of museum records do not have or have inaccurate geographic information associated with them.
-While tools have been developed to convert names of locales into GPS coordinates on land, similar approaches may not work for locales at sea.
+While tools have been developed to convert names of locales into GPS coordinates on land [@rios2010geolocate], similar approaches may not work for locales at sea.
 However, geographic information is available in other forms, and can be digitized.

### 2017-06-09 manuscript.Rmd
[fix #24] include prop of species collected since list publication

@@
 \@ref(fig:database-overlap)).
-For the Pacific Northwest, these proportions were much lower with only `r format_output(species_in_common$n_spp_kozloff_idigbio_obis/species_in_common$n_spp_kozloff*100)`% of the species from the regional list had records in OBIS and iDigBio, and `r format_output(species_in_common$n_spp_kozloff_obis/species_in_common$n_spp_kozloff*100)`%  in OBIS only (Fig.
-1).
-Overall, there is more overlap among the data sources for the Gulf of Mexico than the Pacific Northwest.
+For the Pacific Northwest, these proportions were much lower with only `r format_output(species_in_common$n_spp_kozloff_idigbio_obis/species_in_common$n_spp_kozloff*100)`% of the species from the regional list with records in OBIS and iDigBio, and `r format_output(species_in_common$n_spp_kozloff_obis/species_in_common$n_spp_kozloff*100)`%  in OBIS only (Fig.
+\@ref(fig:database-coverage)).
 
@@
 
-```{r compare_database_overlap, fig.cap='Comparison of species by phylum across the public databases and the faunistic list.', ech=FALSE}
+We investigated the number of species for each phylum that contributed to the discrepancies between the regional species lists and the database records.
+We found that the better studied phyla contributed the most to the differences across the data sources.
+OBIS tends to be closer to the full record set, but iDigBio has more species for the Gulf of Mexico that are not included in the list (Fig.
+\@ref(fig:database-overlap))
+
+Overall, there is more overlap among the data sources for the Gulf of Mexico than the Pacific Northwest.
+
+```{r compare_database_overlap, fig.cap='Comparison of species by phylum across the public databases and the faunistic list. \\#fig:database-overlap', ech=FALSE}
 knitr::include_graphics("figures/database_overlap.pdf")
@@
 
-Species that are included in the list but do not have records in databases could be (1) rare: they have never been collected or have been collected only a few times, and the corresponding records have not been digitized yet; (2) difficult to preserve: large, very small, or fragile species are preserved less frequently; (3) are in the database under a different name: the species has been synonimized or re-identified since the publication of the list under a different name that may be included in the database; (4) at the limit of its range: it was included in the list because it is known to occur at the limit of its geographic zone, but has never been collected within the boundaries.
+Species that are included in the list but do not have records in databases could be (1) rare: they have never been collected or have been collected only a few times, and the corresponding records have not been digitized yet; (2) difficult to preserve: large, very small, or fragile species are preserved less frequently; (3) are in the database under a different name: the species has been synonimized or re-identified since the publication of the list under a different name that may be included in the database; (4) at the limit of its range: it was included in the list because it is known to occur at the limit of its geographic zone, but has never been collected within the boundaries; (5) does not have geographic coordinates associated with it: it can be in the database but without georeferenced datapoints that made it possible to discover them.
 
-Species that are in the database but not in the list could be: (1) records not considered by the authors of the checklist: either because they were not accessible or known to the authors; (2) records erroneously excluded the records (2) records that have been discovered since the list has been assembled; (3) records with incorrect identification; (4) records with incorrect geographic coordinates: their associated geographic information in the database places them in the Gulf of Mexico or the Pacific Northwest, but they were actually collected elsewhere.
+Species that are in the database but not in the list could be: (1) records not considered by the authors of the checklist: either because they were not accessible or known to the authors; (2) records erroneously excluded the records (2) records that have been discovered since the list has been assembled (for iDigbio records from the Gulf of Mexico, this represent `r format_output(mean(recently_collected$gom_in_idig$min_year > 2001))`%, n=`r nrow(recently_collected$gom_in_idig)` and `r format_output(mean(recently_collected$pnw_in_idig$min_year > 1987))`%, n=`r nrow(recently_collected$pnw_in_idig)`); (3) records with incorrect identification; (4) records with incorrect geographic coordinates: their associated geographic information in the database places them in the Gulf of Mexico or the Pacific Northwest, but they were actually collected elsewhere.
 

### 2017-06-09 R/comparison_by_geo.R

@@ -189,7 +189,7 @@ plot_database_overlap <- function(comp_db) {
                 xmin = -.6,
                 xmax = -.6) +
             annotation_custom(
-                grob = grid::textGrob(label = "Not in database", hjust = 1,
+                grob = grid::textGrob(label = "Not in databases", hjust = 1,
                                       gp = gpar(fontfamily = "Ubuntu Condensed")
                                       ),
                 ymin = -100 - offset,
@@ -234,4 +234,43 @@ get_species_in_common <- function(gom_worms, idigbio_gom_records, obis_gom_recor
             dplyr::n_distinct()
 
         )
+}
+
+not_in_list_collected_recently <- function(database_overlap) {
+
+    .f <- . %>% fetch_spp_from_idigbio() %>%
+        idigbio_parse_year() %>%
+        dplyr::mutate(
+                   cleaned_scientificname = cleanup_species_names(scientificname),
+                   is_binomial = is_binomial(scientificname)
+               ) %>%
+        add_worms() %>%
+        group_by(phylum, worms_valid_name) %>%
+        summarize(min_year = min(year))
+
+
+    ## for idigbio gom
+    do_idig_gom <- database_overlap %>%
+        dplyr::filter(
+                   region == "gom",
+                   data_source == "not_in_list",
+                   database == "idigbio"
+               )
+
+    spp_gom_in_idig <- do_idig_gom %>% .f
+
+    ## for idigbio pnw
+    do_idig_pnw <- database_overlap %>%
+        dplyr::filter(
+                   region == "kozloff",
+                   data_source == "not_in_list",
+                   database == "idigbio"
+               )
+    spp_pnw_in_idig <- do_idig_pnw %>% .f
+
+    list(
+        gom_in_idig = spp_gom_in_idig,
+        pnw_in_idig = spp_pnw_in_idig
+    )
+
 }

### 2017-06-10 manuscript.Rmd
fix target names

@@
 After filtering these lists to only include the same higher taxonomic level taxa as the species lists, and keeping only records for which the species could be matched to a species name in WoRMS, we had for the Gulf of Mexico `r nrow(idigbio_gom_records)` and `r nrow(obis_gom_records)` records for iDigBio and OBIS respectively, representing `r dplyr::n_distinct(idigbio_gom_records$worms_valid_name)` and `r dplyr::n_distinct(obis_gom_records$worms_valid_name)` species.
-For the Pacific Northwest, we retrieved `r nrow(idigbio_kozloff_records)` and `r nrow(obis_kozloff_records)` records for iDigBio and OBIS respectively representing `r dplyr::n_distinct(idigbio_kozloff_records$worms_valid_name)` and `r dplyr::n_distinct(obis_kozloff_records$worms_valid_name)` species.
+For the Pacific Northwest, we retrieved `r nrow(idigbio_pnw_records)` and `r nrow(obis_kozloff_records)` records for iDigBio and OBIS respectively representing `r dplyr::n_distinct(idigbio_pnw_records$worms_valid_name)` and `r dplyr::n_distinct(obis_kozloff_records$worms_valid_name)` species.
 For the Gulf of Mexico, the majority of the species included in the expert lists are recorded in both iDigBio and OBIS (`r format_output(species_in_common$n_spp_gom_idigbio_obis/species_in_common$n_spp_gom*100)` %), or only OBIS (`r format_output(species_in_common$n_spp_gom_obis/species_in_common$n_spp_gom*100)` %) (Fig.

### 2017-06-14 manuscript.Rmd
[fix #22] incorporate Mark's comments

@@
 We show that current knowledge and sampling efforts are inadequate given the diversity of the fauna and the geographic extent of US.
-While recent high-throughput sequencing technologies promise to greatly speed up the documentation of marine species and their ranges, its value would be greatly improved by anchoring the results into existing DNA sequence data and associated voucher.
-However, the best documented phyla have at most 25% of their species sequenced for COI.
+While recent high-throughput sequencing technologies promise to greatly speed up the documentation of marine species and their ranges, its value would be greatly improved by anchoring the results into existing DNA sequence data and associated vouchers.
+However, the best documented phyla have at most 25% of their species sequenced for the most widely used DNA barcoding gene, cytochrome oxidase I (COI).
 More sampling, coordinated at the scale of the US, is needed to improve our knowledge of species and their geographic ranges.
-This data is needed to detect the impact of global change and other perturbations.
+This data is needed to detect the impact of climate changes and other anthropogenic strains to global environments.
 Advances in computing, information systems, and sequencing technologies, have set the stage to inform where future efforts for documenting biodiversity should be applied.
@@
 With increasing impact of human activities on the biosphere, we need efficient and accurate approaches to document changes in species ranges and community composition.
-While terrestrial ecosystems have received considerable attention, the baseline of healthy and resilient marine ecosystems are not well understood [@Worm2006], and modeling the impact of human activities is limited to a few taxonomic groups [@Tittensor2010].
-A major hindrance for marine species, is that we do not have enough information about which species live where, and an unknown but potentially significant portion of marine biodiversity remains to be discovered and described [@Mora2011].
+While terrestrial ecosystems have received considerable attention, the baseline of healthy and resiliency of marine ecosystems are not well understood [@Worm2006], and modeling the impact of human activities is limited to a few taxonomic groups [@Tittensor2010].
+A major hindrance for marine species, due to their complex life history strategies, and their size range, is that we do not have enough information about which species live where, or what portion of marine biodiversity remains to be discovered and described [@Mora2011].
 Additionally, coordinated marine ecosystem monitoring initiatives are lacking [@Duffy2013,@Borja2016].
@@
 iDigBio [@Page2015]), species occurrences (e.g., OBIS [@Zhang2002]), genetic data (e.g., GenBank [@Benson2013], BOLD [@Ratnasingham2007a]), and registers of species names (e.g., WoRMS [@Costello2013b]), now allow us to evaluate what and how much we know.
-Here, we aggregate publicly available data for marine invertebrates that occur in the continental United States, the Gulf of Mexico, and the Pacific Northwest, to highlight gaps in our knowledge.
+Here, we aggregate publicly available data for marine invertebrates that occur within the limits of the exclusive economic zone (EEZ) of the continental United States, as well as the Gulf of Mexico and the Pacific Northwest, to highlight gaps in our knowledge that provide a roadmap for important new directions of global research.
 We show that public databases are relatively incomplete, and that most species are known from only a few specimens and a few locations.
 We emphasize that our understanding of marine biodiversity, and the impact of human activities on communities require not only to get a more accurate list of the species that inhabits the oceans, but we also need more comprehensive data on species distributions and abundances.
-This can only be achieved with additional sampling and a commitment to training more taxonomists and experts that can identify the material collected.
+This can only be achieved with additional cost-effective sampling and a commitment to training more taxonomists and experts that can identify and interpret the material collected.
 
@@
 
-To compare the faunistic lists from the Gulf of Mexico and the Pacific Northwest with records in public databases, we queried the iDigBio and OBIS databases for all records they included in the same geographic boundaries as the lists.
-After filtering these lists to only include the same higher taxonomic level taxa as the species lists, and keeping only records for which the species could be matched to a species name in WoRMS, we had for the Gulf of Mexico `r nrow(idigbio_gom_records)` and `r nrow(obis_gom_records)` records for iDigBio and OBIS respectively, representing `r dplyr::n_distinct(idigbio_gom_records$worms_valid_name)` and `r dplyr::n_distinct(obis_gom_records$worms_valid_name)` species.
-For the Pacific Northwest, we retrieved `r nrow(idigbio_pnw_records)` and `r nrow(obis_kozloff_records)` records for iDigBio and OBIS respectively representing `r dplyr::n_distinct(idigbio_pnw_records$worms_valid_name)` and `r dplyr::n_distinct(obis_kozloff_records$worms_valid_name)` species.
+To compare the faunistic lists from the Gulf of Mexico and the Pacific Northwest with records in public databases, we queried the iDigBio (https://portal.idigbio.org) and OBIS (http://www.iobis.org/) databases for all records they included in the same geographic boundaries as the lists.
+After filtering these lists to only include the same higher taxonomic level taxa as the species lists, and keeping only records for which the species could be matched to a species name in WoRMS, we had for the Gulf of Mexico `r format(nrow(idigbio_gom_records), big.mark=",")` and `r format(nrow(obis_gom_records), big.mark=",")` records for iDigBio and OBIS respectively, representing `r format(dplyr::n_distinct(idigbio_gom_records$worms_valid_name), big.mark=",")` and `r format(dplyr::n_distinct(obis_gom_records$worms_valid_name), big.mark=",")` species.
+For the Pacific Northwest, we retrieved `r format(nrow(idigbio_pnw_records), big.mark=",")` and `r format(nrow(obis_kozloff_records), big.mark=",")` records for iDigBio and OBIS respectively representing `r format(dplyr::n_distinct(idigbio_pnw_records$worms_valid_name), big.mark=",")` and `r format(dplyr::n_distinct(obis_kozloff_records$worms_valid_name), big.mark=",")` species.
 For the Gulf of Mexico, the majority of the species included in the expert lists are recorded in both iDigBio and OBIS (`r format_output(species_in_common$n_spp_gom_idigbio_obis/species_in_common$n_spp_gom*100)` %), or only OBIS (`r format_output(species_in_common$n_spp_gom_obis/species_in_common$n_spp_gom*100)` %) (Fig.
@@
 ```{r compare_database_coverage, fig.cap='Overlap between the regional species list (list), records in iDigBio (idigbio), records in OBIS (obis), and records in BOLD (bold) for the Gulf of Mexico (left) and the Pacific Northwest (right).
-(\\#fig:database-overlap)', echo=FALSE}
+(\\#fig:database-coverage)', echo=FALSE}
 knitr::include_graphics("figures/database_coverage.pdf")
@@
 
-We extracted `r sum(gom_worms$is_marine, na.rm = TRUE)` marine invertebrates species from the faunistic list of the Gulf of Mexico, and `r sum(kozloff_worms$is_marine, na.rm=TRUE)` species from the Salish Sea.
+We extracted `r format(sum(gom_worms$is_marine, na.rm = TRUE), big.mark=",")` marine invertebrates species from the faunistic list of the Gulf of Mexico, and `r format(sum(kozloff_worms$is_marine, na.rm=TRUE), big.mark=",")` species from the Pacific Northwest.
 We then looked for the number of species listed that are known to occur in the same geographic area, the United States, or anywhere else in the world.
@@
 
-These results suggest that even when considering species lists that have been curated and that represent possibly the most accurate knowledge for this geographic area, today we could only reconstruct these lists partially based on digitized information.
+These results suggest that even when considering species lists that have been curated and that represent possibly the most accurate knowledge for any particular geographic area, today we can only partially reconstruct these lists based on digitized information.
 For instance, based on OBIS data the list would be `r format_output(summary_richness_per_db_gom$prop_obis_in_gom[summary_richness_per_db_gom$taxon_name == "Porifera"]*100)`% complete for sponges to `r format_output(summary_richness_per_db_gom$prop_obis_in_gom[summary_richness_per_db_gom$taxon_name == "Mollusca"]*100)`% for molluscs.
@@
 
-We combined `r nrow(idigbio_records)` museum records for marine invertebrates collected in the EEZ waters of the continental US representing `r length(unique(tolower(idigbio_records[["data.dwc:phylum"]])))` phyla and `r length(unique(idigbio_records$worms_valid_name))` species.
+We combined `r format(nrow(idigbio_records), big.mark=",")` museum records for marine invertebrates collected in the EEZ waters of the continental US representing `r length(unique(tolower(idigbio_records[["data.dwc:phylum"]])))` phyla and `r format(length(unique(idigbio_records$worms_valid_name)), big.mark=",")` species.
 Most samples were collected on the continental shelf (Fig.
@@
 For the East coast, species richness is the highest in South Florida but it is relatively uniform along the coastline.
-For the West coast, the Salish Sea, the Bay Area, and Southern California have the highest richness (Fig.
+For the West coast, the Pacific Northwest, the San Francisco Bay Area, and Southern California have the highest richness (Fig.
 \@ref(fig:map-diversity-sampling)B).
@@
 
-The inadequacy of the sampling effort compared to the diversity of the fauna is also reflected by the fact that `r prop_singleton_species`% of the species are known from a single record.
+The inadequacy of the sampling effort compared to the diversity of the fauna is also reflected by the fact that only `r prop_singleton_species`% of the species are known from a single record.
 The strong correlation between the number of samples collected and the species richness suggests that any new sampling effort will lead to new observations (Fig.
@@
 \@ref(fig:idigbio-through-time)).
-Since then, the slowing down of collections in the 1970 has been accompanied with a slowing down in the discovery of new records: `r n_spp_comparison$n_spp_2017_2000` species have been documented for the first time since 2000 (`r format_output(n_spp_comparison$n_spp_2017_2000/18)` species.year^-1^), while `r n_spp_comparison$n_spp_1970_1960` have been documented between 1960 and 1970 (`r format_output(n_spp_comparison$n_spp_1970_1960/11)` species.year^-1^).
-Coordinated large-scale biodiversity expeditions can produce a wealth of data and new records.
+Since then, the slowing down of collections after the 1970s has been accompanied with a slowing down in the discovery of new records: `r n_spp_comparison$n_spp_2017_2000` species have been documented for the first time since 2000 (`r format_output(n_spp_comparison$n_spp_2017_2000/18)` species.year^-1^), while `r n_spp_comparison$n_spp_1970_1960` have been documented between 1960 and 1970 (`r format_output(n_spp_comparison$n_spp_1970_1960/11)` species.year^-1^).
+Therefore, coordinated large-scale biodiversity expeditions will continue to generate a wealth of data and new records.
 For instance, a recent study in Western Australia reported that 135 samples led to 805 species of crustaceans and polychaetes, 94.6% and 72% of which were undescribed [@Poore2014].
@@
 
-Modeling the geographic distributions of species, and how they can be impacted by global change has undergone important advances recently.
+Predicting the geographic range of species, and how they can be impacted by global change has undergone important advances recently.
 However, it is generally recognized that to obtain reliable results such efforts require to have accurate geographic information for a minimum of 10 specimens but having at least 30 is preferred.
@@
 
-There are important discrepancies in our level of knowledge across phyla.
+Furtheremore, there are important discrepancies in our level of knowledge across phyla.
 While larger, and accessible groups represent the bulk of the records for marine invertebrates in museum collections, the small organisms, including representatives of the meiofauna are virtually absent.
 Of the `r n_idigbio_phyla` phyla occurring in iDigBio, `r records_rare_phyla$n_rare_phyla` are represented by less than 500 records, and make up `r records_rare_phyla$prop_rare_phyla`% of the total number of records.
+
+<!-- From Mark: These previously undersampled taxa are likely to be much more important for the detection of subtle changes in environmental quality due to their size and physiological vulnerability.
+(?).
+-->
+<!-- also, the impact of changes in the composition in their communties is largely unexplored, look into Bik's paper? -->
 
@@
 
-### Genetic data for the future of biodiversity discovery and monitoring
+### Molecular data for the future of biodiversity discovery and monitoring
 
 Since its introduction in the early 2000's, DNA barcoding, the sequencing of a common locus across taxa bearing species-specific information, has revitalized taxonomy by facilitating species delineation [@Meyer2005-plos].
-More recently, advances in high-throughput sequencing technologies have enabled the parallel sequencing of thousands of species that occur in environmental samples [@Leray2015].
+More recently, advances in high-throughput sequencing technologies have enabled the parallel sequencing of thousands of species that occur in environmental samples cost effectively [@Leray2015].
 This approach opens up new research venues as it becomes possible to estimate the diversity from species-rich samples that were virtually inaccessible before.
@@
 However, the utility of the data generated by these projects is enhanced by the possibility of matching DNA sequences obtained to previously sequenced and correctly identified species [@Leray2015].
-With decreasing sequencing costs, and the prospect of automating sample collection, these technologies may soon be used routinely to monitor biodiversity [@Bohan2017].
+With decreasing sequencing costs, and the prospect of automating sample collection, these technologies may soon be used routinely to monitor biodiversity at defined temporal frequency [@Bohan2017].
 DNA barcoding provides the possibility to link these newly generated data with historical natural history knowledge about species and their distributions.
@@
 Therefore specimens identified under the same species but sampled in distant locations, could represent cryptic species.
-Geographic information is missing for most records (`r format_output(100*stat_bold$p_no_coords)`% of the `r format(stat_bold$n_total_records)` BOLD records analyzed here have no geographic coordinates information associated with them).
+Geographic information is missing for most records (`r format_output(100*stat_bold$p_no_coords)`% of the `r format(stat_bold$n_total_records, big.mark=",")` BOLD records analyzed here have no geographic coordinate information associated with them).
 It is thus crucial to ensure that accurate geographic information is provided when DNA barcodes are published, and that enough samples for each species are sequenced, to capture more accurately their geographic range and the extent of their genetic diversity.
-This cryptic diversity is also at least partially reflected in the fact that `r stat_bold$n_spp_multi_bin` species out of `r stat_bold$n_spp` include more than 1 BIN (BOLD Index Number, operational taxonomic units that closely resemble species [@Ratnasingham2013]).
+This cryptic diversity is also at least partially reflected in the fact that `r stat_bold$n_spp_multi_bin` species out of `r format(stat_bold$n_spp, big.mark=",")` include more than 1 BIN (BOLD Index Number, operational taxonomic units that closely resemble species [@Ratnasingham2013]).
 
@@
 
-Based on the results of this study, we recommend a multi-level approach to not only improve the quality of the existing data, but also to efficiently collect new data.
+Based on the results of this study, we recommend a multi-level approach for both improving the quality of existing data, as well as for the efficient collection of new data.
 
@@
 
-With data aggregation of natural history collections through initiatives like iDigBio, and the compilation of a nearly complete register of marine species with WoRMS, in combination with the development of algorithms for text manipulation, it is increasingly easier to validate and spell check the scientific names of the identifications of museum specimens.
+With data aggregation of natural history collections through initiatives like iDigBio, and the compilation of a nearly complete register of marine species with WoRMS, in combination with the development of algorithms for text analysis, it is increasingly easier to validate and spell check the scientific names of the identifications of digitized museum specimens.
 Linking species names to identifiers in taxonomic databases would facilitate data linking, specimen discoverability, and data usability.
@@
 
-### Collect better data faster
+### Collect better data faster and more cost effectively
 
@@
 Coupling these more traditional approaches of biodiversity documentation with metabarcoding is a promising approach to screen an habitat for its potential for species that do not yet occur in public databases.
-Metabarcoding can complement and take advantage of the DNA barcodes already available.
+<!-- explain why better --> Metabarcoding can complement and take advantage of the DNA barcodes already available.
 Additionally, in the future, data generated by metabarcoding efforts can be re-interpreted as more DNA barcodes associated with correctly identified specimens become available.
@@
 data accuracy (can we trust the data that is found in the databases) -->
+
+## Material and methods
+
+### Data sources
+
+iDigBio:
+OBIS:
+WoRMS:
 

### 2017-06-15 manuscript.Rmd
[fix #23] get info for spp listed as "not in database"

@@
 Species that are included in the list but do not have records in databases could be (1) rare: they have never been collected or have been collected only a few times, and the corresponding records have not been digitized yet; (2) difficult to preserve: large, very small, or fragile species are preserved less frequently; (3) are in the database under a different name: the species has been synonimized or re-identified since the publication of the list under a different name that may be included in the database; (4) at the limit of its range: it was included in the list because it is known to occur at the limit of its geographic zone, but has never been collected within the boundaries; (5) does not have geographic coordinates associated with it: it can be in the database but without georeferenced datapoints that made it possible to discover them.
+By querying iDigBio we found that for the `r format(not_really_in_database$idig_gom_n_spp_not_in_db, big.mark=",")` species that do not appear within the geographic range of the Gulf of Mexico but are on the list, `r format_output(not_really_in_database$idig_gom_p_spp_not_in_db*100)`% are truly absent from iDigBio, while for the `r format(not_really_in_database$idig_gom_n_spp_no_coords, big.mark=",")` species that appear in iDigBio `r format_output(not_really_in_database$idig_gom_p_spp_no_coords*100)`%  do not have geographic coordinates associated with them, and for the ones that do `r format_output(not_really_in_database$idig_gom_p_spp_within_usa*100)`% fall within the US waters (n = `r format(not_really_in_database$idig_gom_n_spp_within_usa, big.mark=",")`).
 

### 2017-06-15 R/comparison_by_geo.R

@@ -232,7 +232,6 @@ get_species_in_common <- function(gom_worms, idigbio_gom_records, obis_gom_recor
             dplyr::semi_join(idigbio_kozloff_records, by = "worms_valid_name") %>%
             dplyr::select(worms_valid_name) %>%
             dplyr::n_distinct()
-
         )
 }
 
@@ -273,4 +272,64 @@ not_in_list_collected_recently <- function(database_overlap) {
         pnw_in_idig = spp_pnw_in_idig
     )
 
+}
+
+## Species that appear to not be in iDigbio for the geographical list,
+## either because they do not have GPS coordinates associated with
+## their records, or they have records that fall outside the
+## geographic area.
+## Only GOM for now.
+get_not_really_in_database <- function(database_overlap, map_usa) {
+
+    idig_gom_not_in_db <- database_overlap %>%
+        dplyr::filter(region == "gom",
+                      data_source == "not_in_geo",
+                      database == "idigbio")
+
+    idig_gom_spp_records <- idig_gom_not_in_db %>%
+        fetch_spp_from_idigbio() %>%
+        is_within_eez_records(map_usa)
+
+    ## species that are trully absent
+    idig_gom_n_spp_not_in_db <- length(setdiff(tolower(idig_gom_not_in_db$worms_valid_name),
+                                               unique(idig_gom_spp_records$scientificname)))
+
+    ## species that have no geographic coordinates
+    has_no_coords <- function(lat, lon) {
+        is.na(lat) & is.na(lon)
+    }
+    idig_gom_spp_no_coords <- idig_gom_spp_records %>%
+        dplyr::group_by(scientificname) %>%
+        dplyr::summarize(
+                   n_no_coords = sum(has_no_coords(decimallatitude, decimallongitude)),
+                   p_no_coords = mean(has_no_coords(decimallatitude, decimallongitude))
+               )
+
+    ## total number of species that have records in idigbio
+    idig_gom_n_spp_no_coords <- nrow(idig_gom_spp_no_coords)
+    idig_gom_p_spp_no_coords <- mean(idig_gom_spp_no_coords$p_no_coords == 1)
+
+    ## proportion of species that have at least one record within US EEZ
+    idig_gom_within_usa <- idig_gom_spp_records %>%
[…]

### 2017-06-15 manuscript.Rmd
[fix #26] add FWRI mention in discussion

--- 2017-06-14 manuscript.Rmd
@@
 Species that are included in the list but do not have records in databases could be (1) rare: they have never been collected or have been collected only a few times, and the corresponding records have not been digitized yet; (2) difficult to preserve: large, very small, or fragile species are preserved less frequently; (3) are in the database under a different name: the species has been synonimized or re-identified since the publication of the list under a different name that may be included in the database; (4) at the limit of its range: it was included in the list because it is known to occur at the limit of its geographic zone, but has never been collected within the boundaries; (5) does not have geographic coordinates associated with it: it can be in the database but without georeferenced datapoints that made it possible to discover them.
+By querying iDigBio we found that for the `r format(not_really_in_database$idig_gom_n_spp_not_in_db, big.mark=",")` species that do not appear within the geographic range of the Gulf of Mexico but are on the list, `r format_output(not_really_in_database$idig_gom_p_spp_not_in_db*100)`% are truly absent from iDigBio, while for the `r format(not_really_in_database$idig_gom_n_spp_no_coords, big.mark=",")` species that appear in iDigBio `r format_output(not_really_in_database$idig_gom_p_spp_no_coords*100)`%  do not have geographic coordinates associated with them, and for the ones that dok `r format_output(not_really_in_database$idig_gom_p_spp_within_usa*100)`% fall within the US waters (n = `r format(not_really_in_database$idig_gom_n_spp_within_usa, big.mark=",")`).
 
@@
 Some of the discrepancies observed between the number of species for which occurrence data and museum specimens are available suggest that additional digitization of collections (including capturing their geographic information) are needed.
+For instance, one can expect that the digitization coverage for the Gulf of Mexico will greatly increase when records from the Florida Wildlife Research Institute collection (St Petersburg, FL) become integrated into the biodiversity information systems.
 

### 2017-06-15 manuscript.Rmd
edit manuscript...

@@
 
-<!-- We show that the effort is consistent across the better-studied taxa, with ` average_prop_species_seq`% of the fauna sequenced.
-Using a DNA barcode dataset of ` n_larvae` planktonic larvae, representing ` n_esus` species, collected in coastal waters of Florida, we find a similar proportion of species that can be assigned to a previously sequenced species ( ` format_output(100 * prop_plankton_with_match)`%).
-Our results show that most marine species remain to be sequenced.
-Achieving this goal will required a concerted effort, and resources.
--->
-
 <!--
-the content of insect traps, marine sediments, plankton
-samples, or communities settling on artificial substratum. When applied to
-metazoan communities, these sequencing approaches typically target the barcoding
-gene and therefore has been dubbed "meta-barcoding". These applications are also
-making the public databases for these DNA barcodes all the more valuable. To
-deliver on the meta-barcoding promises, it is crucial to have comprehensive
-taxonomic coverage represented in the DNA sequences public databases. Here, we
-investigate the proportion of species for each marine phylum that have been
-sequenced at various spatial scale.
--->
-
-<!-- studies that have summary statistics on what has been sequenced for various marine phyla:
-- for Antarctica: Grant+Linse2009-Polar Biology:
-
-We know that some perturbations have large effects because they are visible, but what about perturbations for which we don't see the effects directly because they affect small species?
-:: checkout nature paper on sea grasses.
-
-- Describing new species is crucial but not sufficient.
-Without knowing their ranges, and their habitats, we can't study or predict how perturbations of the environment are going to affect their distributions and their susceptibility to perturbations.
-The process of figuring out the ranges of species is more time consuming than possibly finding new species; and it requires shared knowledge in the community about these new species.
-Using barcoding and other moleuclar techniques to improve the documentation of species distributions is going to be critical.
 
@@
 
-Furtheremore, there are important discrepancies in our level of knowledge across phyla.
+Furthermore, there are important discrepancies in our level of knowledge across phyla.
 While larger, and accessible groups represent the bulk of the records for marine invertebrates in museum collections, the small organisms, including representatives of the meiofauna are virtually absent.
@@
 The proportions for species known to occur in the Gulf of Mexico are usually similar than for the US, while typically a higher fraction of species from the Pacific Northwest are available.
-In 2010, it was estimated that a global scale, less than 10% of the species for each marine phylum had been sequenced [@Radulovici2010], suggesting that the US fauna is generally better characterized than the rest of the world, combined with many other species have been sequenced since this report.
+In 2010, it was estimated that a global scale, less than 10% of the species for each marine phylum had been sequenced [@Radulovici2010], and that this number was even lower in some areas (e.g.
+2.6% for the Antarctic [@Grant2009]).
+This suggests that even as many additional species have had their DNA barcodes made available since these studies, the US fauna is generally better characterized than other parts of the world.
 
@@
 This study highlights that we have too little data about too few species of marine invertebrates, we recommend that additional coordinated field expeditions are organized to sample more thoroughly the marine environment.
-Data from museum collections make clear that any additional samples will contain new species and new records.
+The relation between sampling effort and diversity from museum collections makes clear that any additional samples will improve greatly our knowledge of marine biodiversity.
+
 Samples collected during these expeditions need to be photographed to document live appearance that would otherwise be lost during preservation.
 We also need to capture metadata associated with the specimens to be able to understand better the physical characteristics of the habitats for these species.
-Specimens need to be sub-sampled and have tissues suitable for genomic analysis preserved for their DNA barcode, or any other genomic resources.
-Some of these parts of data and specimen collection can be automated and captured digitally directly in the field speeding up data collection.
+Specimens need to be sub-sampled to create material suitable for genomic analysis.
+Parts of data and specimen collection can be automated and captured digitally directly in the field speeding up data collection.
 
@@
 
-<!-- Beyond their utility in documenting diversity, metabarcoding is also promising for understanding the functioning of ecosystems. -->
+In order to predict the impact of climate change and other anthropogenic perturbations on species distributions, we need to collect more occurences for species.
+One potential challenge for this is to be able to convey taxonomic concepts for mostly unamed species.
+Developing robust identifiers that can be used across biodiversity information systems, and potentially anchored in results from DNA barcoding or other molecular techniques, is needed to collaboratively build our knowledge of species distributions.
+
+Beyond their utility in documenting diversity, metabarcoding is also a promising technology to monitor changes in communities and for understanding the functioning of ecosystems.
+For instance, despite our limited knowledge of the diversity of communities in the meiofauna, metabarcoding could be used to detect and documents changes in species assemblages following the Deepwater Horizon Oil Spill [@Bik2012].
+Metabarcoding has also been used to reveal important differences in the diet of 3 related species of fishes that co-occur [@Leray2015b], providing a much more detailed understanding of trophic relations.
 

### 2017-06-19 manuscript.Rmd
edit manuscript

@@
 With increasing impact of human activities on the biosphere, we need efficient and accurate approaches to document changes in species ranges and community composition.
-While terrestrial ecosystems have received considerable attention, the baseline of healthy and resiliency of marine ecosystems are not well understood [@Worm2006], and modeling the impact of human activities is limited to a few taxonomic groups [@Tittensor2010].
-A major hindrance for marine species, due to their complex life history strategies, and their size range, is that we do not have enough information about which species live where, or what portion of marine biodiversity remains to be discovered and described [@Mora2011].
-Additionally, coordinated marine ecosystem monitoring initiatives are lacking [@Duffy2013,@Borja2016].
-There is therefore a pressing need to improve our knowledge of marine systems before species go extinct, and their functioning is irreversibly altered.
+While terrestrial ecosystems have received considerable attention, the baseline of healthy and resilient marine ecosystems are not well understood [@Worm2006], and modeling the impact of human activities is limited to a few taxonomic groups [@Tittensor2010].
+For marine organisms, because of the diversity of their complex life history strategies and their size range, we have limited information about which species live where, or what portion of their diversity remains to be discovered and described [@Mora2011].
+Additionally, coordinated marine ecosystem monitoring initiatives are lacking [@Duffy2013,@Borja2016], making it difficult to monitor changes in species distributions and abundances.
+There is therefore a pressing need to improve our knowledge of marine systems before their functioning is irreversibly altered, and species go extinct.
+
+The continental US includes a wide diversity of marine ecosystems from the Florida Keys, to the Gulf of Maine on the East Coast; and from the arid Southern California to the oceanic coast of the Pacific Northwest.
+With estuaries, bays, sandy beaches, rocky shores, and coral reefs, the US coastlines harbors many biotas.
+While the continental shelf extends for several hundred kilometers on the East coast, it is narrow on the West coast.
+This diversity of habitats and depths, combined with the exploitation of natural resources as early as the 1880, has led to extensive exploration and documentation of marine biodiversity [@Fautin2011].
+However, apart from the synthesis that emerged from the Census of Marine Life efforts [@Fauntin2011], no study has attempted to synthesize the current state of knowledge of marine biodiversity in the US.
 
@@
 iDigBio [@Page2015]), species occurrences (e.g., OBIS [@Zhang2002]), genetic data (e.g., GenBank [@Benson2013], BOLD [@Ratnasingham2007a]), and registers of species names (e.g., WoRMS [@Costello2013b]), now allow us to evaluate what and how much we know.
-Here, we aggregate publicly available data for marine invertebrates that occur within the limits of the exclusive economic zone (EEZ) of the continental United States, as well as the Gulf of Mexico and the Pacific Northwest, to highlight gaps in our knowledge that provide a roadmap for important new directions of global research.
+Here, we aggregate publicly available data for marine invertebrates that occur within the limits of the exclusive economic zone (EEZ) of the continental United States (US), as well as the Gulf of Mexico and the Pacific Northwest, to highlight gaps in our knowledge that provide a roadmap for important new directions of global research.
 We show that public databases are relatively incomplete, and that most species are known from only a few specimens and a few locations.
@@
 \@ref(fig:database-coverage)).
+Previous esimates for species richness in the Gulf of Mexico for marine invertebrates were of `r format(11150-1541-434, big.mark=",")` species, and `r format(8374-909-134)` for the Pacific coast (including California) [@Fautin2010].
 
@@
 
+Another potential venue to improve the quality of the data in biodiversity information systems, is better incitives for scientisits to publish their occurrence records alongside their papers as it is done for genetic data with GenBank [@Huang2011].
+
 <!-- - develop infrastructure and tools to continuously check and validate data quality in databases. -->
@@
 
+Assembling species lists is a necessary but not sufficient step to understand biodiversity.
+Beyond counting the number of species, we need to understand better the geographical distribution of species, their abundance, and their biology.
 In order to predict the impact of climate change and other anthropogenic perturbations on species distributions, we need to collect more occurences for species.
@@
 Given the urgency of the biodiversity crisis, we need to develop data-driven initiatives to efficiently document biodiversity, and build on our existing knowledge to collect information to improve our understanding of marine systems.
+Digitization efforts are catching up on an already fragmented and vastly incomplete records.
 

### 2017-06-19 manuscript.Rmd
edit manuscript

@@
 While the continental shelf extends for several hundred kilometers on the East coast, it is narrow on the West coast.
-This diversity of habitats and depths, combined with the exploitation of natural resources as early as the 1880, has led to extensive exploration and documentation of marine biodiversity [@Fautin2011].
-However, apart from the synthesis that emerged from the Census of Marine Life efforts [@Fauntin2011], no study has attempted to synthesize the current state of knowledge of marine biodiversity in the US.
+This diversity of habitats and depths, combined with the exploitation of natural resources as early as the 1880, has led to extensive exploration and documentation of marine biodiversity [@Fautin2010].
+However, apart from the synthesis that emerged from the Census of Marine Life efforts [@Fautin2010], no study has attempted to synthesize the current state of knowledge of marine biodiversity across the continental US.
 
@@
 \@ref(fig:database-coverage)).
-Previous esimates for species richness in the Gulf of Mexico for marine invertebrates were of `r format(11150-1541-434, big.mark=",")` species, and `r format(8374-909-134)` for the Pacific coast (including California) [@Fautin2010].
+Previous esimates for species richness in the Gulf of Mexico for marine invertebrates were of `r format(11150-1541-434, big.mark=",")` species, and `r format(8374-909-134, big.mark=",")` for the Pacific coast (including California) [@Fautin2010].
 
@@
 
-```{r compare_database_overlap, fig.cap='Comparison of species by phylum across the public databases and the faunistic list. \\#fig:database-overlap', ech=FALSE}
+```{r compare_database_overlap, fig.cap='Comparison of species by phylum across the public databases and the faunistic list. (\\#fig:database-overlap)', ech=FALSE}
 knitr::include_graphics("figures/database_overlap.pdf")
@@
 
-Increasing the computing literacy of the staff of museum in order for them to manipulate and use the data from their collection database is equally critical.
+Increasing the computing literacy of the staff of museums in order for them to manipulate and use the data from their collection database is equally critical.
 As data across institutions get pooled into the large biodiversity information systems, it is crucial that data standards such as Darwin Core are embraced by the community.
 
-Another potential venue to improve the quality of the data in biodiversity information systems, is better incitives for scientisits to publish their occurrence records alongside their papers as it is done for genetic data with GenBank [@Huang2011].
+Another potential venue to improve the quality of the data in biodiversity information systems, is better incitives from journal editors and funding agencies for scientists to publish their occurrence records in databases alongside their papers as it is done for genetic data with GenBank [@Huang2011e].
 
@@
 The relation between sampling effort and diversity from museum collections makes clear that any additional samples will improve greatly our knowledge of marine biodiversity.
+Concerted and collaborative efforts are needed to share expertise and limited resources.
 
@@
 Specimens need to be sub-sampled to create material suitable for genomic analysis.
-Parts of data and specimen collection can be automated and captured digitally directly in the field speeding up data collection.
+Parts of data acquisition associated with specimens and their biota can be automated and captured digitally directly in the field, speeding up their ingestion in database collections.
 
-Coupling these more traditional approaches of biodiversity documentation with metabarcoding is a promising approach to screen an habitat for its potential for species that do not yet occur in public databases.
-<!-- explain why better --> Metabarcoding can complement and take advantage of the DNA barcodes already available.
-Additionally, in the future, data generated by metabarcoding efforts can be re-interpreted as more DNA barcodes associated with correctly identified specimens become available.
-Building the library of DNA barcodes in parallel of metabarcoding efforts ensures that matches in the database will correspond to actual specimens deposited in natural history collections; ruling out possible artifacts of the method.
-Information about species identity and distribution will thus build on existing natural history data rather than being limited to matching DNA sequences across studies.
+Integrating metabarcoding as part of the data collection conducted in biodiversity surveys is a promising approach to screen an habitat for its potential to lead to new species or species that have not yet public DNA barcodes.
+For instance, by analyzing metabarcoding samples that contain planktonic larvae from a given habitat, we could get an indication of which taxonomic group would benefit most from additional sampling effort or give an indication of the proportion of species that has not been sampled yet.
+Additionally, data generated by metabarcoding efforts can be re-interpreted as more DNA barcodes associated with correctly identified specimens become available.
+Building the library of DNA barcodes in parallel of metabarcoding efforts ensures that database matches correspond to actual specimens deposited in natural history collections, ruling out possible artifacts of the method.
+With such approaches, the geographic distribution of species can build on and linked to existing records rather than being limited to matching DNA sequences across studies.
 
-Assembling species lists is a necessary but not sufficient step to understand biodiversity.
-Beyond counting the number of species, we need to understand better the geographical distribution of species, their abundance, and their biology.
-In order to predict the impact of climate change and other anthropogenic perturbations on species distributions, we need to collect more occurences for species.
-One potential challenge for this is to be able to convey taxonomic concepts for mostly unamed species.
-Developing robust identifiers that can be used across biodiversity information systems, and potentially anchored in results from DNA barcoding or other molecular techniques, is needed to collaboratively build our knowledge of species distributions.
+Assembling species lists is a necessary but not sufficient step in our quest to understand biodiversity.
+We also need to understand better the geographic distribution of biodiversity, how the abundance of species vary in space and time, and their role in the functionning of ecosystems.
+In order to predict the impact of climate change and other anthropogenic perturbations on species distributions, we need to collect additional occurences for species as well as the characteristics of their habitats.
+Among the challenges for this enterprise, we need to be able to convey taxonomic concepts for species which, for the most part, have not formally been named yet.
+Developing robust identifiers that can be shared across biodiversity information systems, and anchored in results from DNA barcoding or other molecular techniques, is needed to collaboratively build our knowledge of species distributions.
 
-Beyond their utility in documenting diversity, metabarcoding is also a promising technology to monitor changes in communities and for understanding the functioning of ecosystems.
-For instance, despite our limited knowledge of the diversity of communities in the meiofauna, metabarcoding could be used to detect and documents changes in species assemblages following the Deepwater Horizon Oil Spill [@Bik2012].
-Metabarcoding has also been used to reveal important differences in the diet of 3 related species of fishes that co-occur [@Leray2015b], providing a much more detailed understanding of trophic relations.
+Metabarcoding also offers to be a promising technology to monitor changes in communities and for understanding the functioning of ecosystems.
+For instance, despite our limited knowledge of the diversity of communities in the meiofauna, metabarcoding has been successfully used to detect and documents changes in species assemblages following the Deepwater Horizon Oil Spill [@Bik2012].
+Metabarcoding has also revealed important differences in the diet of 3 related species of fishes that co-occur at small spatial scales, providing a much more detailed understanding of trophic relations [@Leray2015b].
 
-However, taxonomists and experts are needed to collect, sort, process, and identify these samples and the data that come along with them.
-Even if technological advances can greatly speed up the documentation of species and their distributions, trained personnel is needed.
-Taxonomic expertise is dwindling, and there are now only a handful of US natural history museums with active marine invertebrates collections.
+Even if technological advances can greatly speed up the documentation of species and their distributions, taxonomists and experts are needed  to collect, sort, process, and identify samples.
+Taxonomic expertise within the US is dwindling, and there are now only a handful of US natural history museums with active marine invertebrates collections, and curatorial staff.
 

### 2017-06-19 manuscript.Rmd
[fix #27] add mention of how to prioritize specimens for digitization

@@
 For both the spelling of scientific names, and the geographic information, computing tools can be developed to detect issues and propose solutions.
-There are opportunities for services using publicly available data to be developed to make curators and collection managers aware of these issues so that they can fixed at the source.
-It however requires that resources are available to museum collections to address these issues.
+There are opportunities for services using publicly available data to be developed to make curators and collection managers aware of these issues so that they can be fixed at the source.
+It however requires that resources are made available to museum collections to address these issues.
 
@@
 
-Another potential venue to improve the quality of the data in biodiversity information systems, is better incitives from journal editors and funding agencies for scientists to publish their occurrence records in databases alongside their papers as it is done for genetic data with GenBank [@Huang2011e].
+Another venue to improve the quality of the data in biodiversity information systems, is better incitives from journal editors and funding agencies for scientists to publish their occurrence records in databases alongside their papers as it is done for genetic data with GenBank [@Huang2011e].
+
+Specimens with accurate identifications and specimens that have been preserved in conditions such that it is possible to get DNA out of them should be prioritized, as they yield the most data that are relevant to biodiversity knowledge.
 

### 2017-06-20 manuscript.Rmd
[fix #20] mention extinction of marine inverts
[fix #21] restructure intro + talk more about DNA

@@
 Additionally, coordinated marine ecosystem monitoring initiatives are lacking [@Duffy2013,@Borja2016], making it difficult to monitor changes in species distributions and abundances.
+While most documented global extinctions of marine species are vertebrates, many others may have been lost unnoticed because they were not known [@Sala2006].
 There is therefore a pressing need to improve our knowledge of marine systems before their functioning is irreversibly altered, and species go extinct.
 
-The continental US includes a wide diversity of marine ecosystems from the Florida Keys, to the Gulf of Maine on the East Coast; and from the arid Southern California to the oceanic coast of the Pacific Northwest.
-With estuaries, bays, sandy beaches, rocky shores, and coral reefs, the US coastlines harbors many biotas.
-While the continental shelf extends for several hundred kilometers on the East coast, it is narrow on the West coast.
-This diversity of habitats and depths, combined with the exploitation of natural resources as early as the 1880, has led to extensive exploration and documentation of marine biodiversity [@Fautin2010].
-However, apart from the synthesis that emerged from the Census of Marine Life efforts [@Fautin2010], no study has attempted to synthesize the current state of knowledge of marine biodiversity across the continental US.
+Research attempting to understand how much we know about marine biodiversity has focused on estimating the total number of species that inhabits the oceans, and which fraction of this diversity has been named [@Appeltans2012,@Costello2013].
+Undoubtely, the development of molecular methods to investigate species limits has reinvigorated taxonomy, and contributed to elucidating cryptic species complexes [@Knowlton2000].
+Thus, it has been argued that molecular data can speed up the enterprise of discovering and naming species [@Blaxter2016].
+As species remain the most convenient unit to measure biological diversity, understanding how much has been done, how much remains to be done, and how we can speed up the description of species diversity is crucial.
+However, to understand the role of these species in their environment, we need to document the ecology of these species [@Cernansky2017].
+For instance, to be able to predict how their geographic ranges will be affected by global change, we need detailed information about their environmental requirements.
+Occurrence data for marine species could be used for environmental niche modelling of their geographic and bathymetric distributions but the proportion of species for which we have enough data to do so is unknown.
 
@@
 This can only be achieved with additional cost-effective sampling and a commitment to training more taxonomists and experts that can identify and interpret the material collected.
+
+The continental US includes a wide diversity of marine ecosystems from the Florida Keys, to the Gulf of Maine on the East Coast; and from the arid Southern California to the oceanic coast of the Pacific Northwest.
+With estuaries, bays, sandy beaches, rocky shores, and coral reefs, the US coastlines harbors many biotas.
+While the continental shelf extends for several hundred kilometers on the East coast, it is narrow on the West coast.
+This diversity of habitats and depths, combined with the exploitation of natural resources as early as the 1880, has led to extensive exploration and documentation of marine biodiversity [@Fautin2010].
+However, apart from the synthesis that emerged from the Census of Marine Life efforts [@Fautin2010], no study has attempted to synthesize the current state of knowledge of marine biodiversity across the continental US.
 
@@
 The proportions for species known to occur in the Gulf of Mexico are usually similar than for the US, while typically a higher fraction of species from the Pacific Northwest are available.
-In 2010, it was estimated that a global scale, less than 10% of the species for each marine phylum had been sequenced [@Radulovici2010], and that this number was even lower in some areas (e.g.
+In 2012, it was estimated that a global scale, less than 3% of the marine species had been sequenced [@Appeltans2012], and that this number was even lower in some areas (e.g.
 2.6% for the Antarctic [@Grant2009]).
@@
 
+Considering additional linked data with species such as DNA has been a first step in a digital biodiversity world where other data on species can be gathered in databases and analyzed to better understand species distributions.
+This has paved the way to more linked data/databases.
+Still issues though with connecting identifiers [@Guralnick2014].
+
 Even if technological advances can greatly speed up the documentation of species and their distributions, taxonomists and experts are needed  to collect, sort, process, and identify samples.
@@
 Digitization efforts are catching up on an already fragmented and vastly incomplete records.
+Here, we only reported results from the US where there is at least two centuries of biodiversity exploration data and high-level of digitization.
+Many other parts of the world also need additional attention, especially in some developing countries where environmental pressure on ecosystems is even higher than in the US.
 

### 2017-06-20 manuscript.Rmd
[fix #19] more mentions about other studies

@@
 
-<!--  study focused on USA but conclusions more general given that it's arguably one of the better studied area + data availability -->
-<!-- * barcoding has shown that typically marine invertebrates are complex of cryptic species with small ranges; therefore potentially more vulnerable to extinction risk. -->
-<!-- * need collaboration among researchers to share data standards, protocols -->
-<!-- * use existing knowledge to fill missing gaps -->
-<!-- * running against time -->
-<!-- * we need to document species diversity better for conservation (knowing what we protect and to measure the effectiveness of the protection) -->
 <!-- * how do we conserve animal species that we can't see, we don't find pretty, don't have eyes -->
-<!-- * recent advances in technologies make possibility of developing a global monitoring program possible: reduced costs of sequencing, detection of rare species, etc.... -->
-<!-- * Using this kind of approaches at a large scale might also be a good way to have a better sense of how much of the biodiversity we are missing. -->
 <!-- * Most species are rare, especially in very diverse places,  and the degradation of habitats makes it even more difficult to find these rare species. -->
@@
 
-The continental US includes a wide diversity of marine ecosystems from the Florida Keys, to the Gulf of Maine on the East Coast; and from the arid Southern California to the oceanic coast of the Pacific Northwest.
+Here, we focus on the continental US because the marine biodiversity is extensively documented as the result of a long history of studies by governmental, academic, and private institutions [@Fautin2010].
+The continental US includes a wide diversity of marine ecosystems from the Florida Keys, to the Gulf of Maine on the East Coast; and from the arid Southern California to the oceanic coast of the Pacific Northwest on the West coast.
 With estuaries, bays, sandy beaches, rocky shores, and coral reefs, the US coastlines harbors many biotas.
 While the continental shelf extends for several hundred kilometers on the East coast, it is narrow on the West coast.
-This diversity of habitats and depths, combined with the exploitation of natural resources as early as the 1880, has led to extensive exploration and documentation of marine biodiversity [@Fautin2010].
-However, apart from the synthesis that emerged from the Census of Marine Life efforts [@Fautin2010], no study has attempted to synthesize the current state of knowledge of marine biodiversity across the continental US.
+This diversity of habitats and bathymetric profiles, combined with efforts at the national level to make data publicly available, make this region especially suitable to evaluate how much we know about marine biodiversity using biodiversity databases.
+However, apart from the syntheses that emerged from the Census of Marine Life efforts [@Costello2010,@Fautin2010], no study has yet attempted to synthesize the current state of knowledge of marine biodiversity across the continental US, and this study is the first using data from public databases.
 

### 2017-06-20 R/idigbio_by_geo.R
separate database creation from filtering

@@ -127,8 +127,13 @@ get_coords_idigbio_query <- function(map_usa, cellsize = .5) {
 ## database that will store the results use_cache: if TRUE, this uses
 ## the results from the iDigBio storr; if false, the entire storr is
 ## destroyed before
-fill_store_idigbio_by_geo <- function(coords, db_table, use_cache) {
+connect_idigbio_db <- function() {
+    dplyr::src_postgres("idigbio", host = "localhost",
+                        user = "marinediversity",
+                        password = "password")
+}
 
+create_idigbio_db <- function(coords, db_table, use_cache) {
     idig_types <- structure(c("TEXT", "TEXT", "TEXT", "TEXT", "TEXT",
                               "TEXT", "TEXT", "TEXT", "TEXT", "TEXT",
                               "TEXT", "TEXT", "REAL", "REAL", "TEXT",
@@ -142,12 +147,12 @@ fill_store_idigbio_by_geo <- function(coords, db_table, use_cache) {
                                   "geopoint.lon", "geopoint.lat",
                                   "clean_phylum", "clean_class", "clean_family"))
 
-    idig_data_db <- dplyr::src_postgres("idigbio", host = "localhost",
-                                        user = "marinediversity",
-                                        password = "password")
+    idig_data_db <- connect_idigbio_db()
     con <- idig_data_db$con
+
     db_begin(con)
     on.exit(db_rollback(con))
+
     if (db_has_table(con, db_table))
         db_drop_table(con, db_table)
 
@@ -169,7 +174,7 @@ fill_store_idigbio_by_geo <- function(coords, db_table, use_cache) {
             ## that will take care of plants, fungi, and records with no
             ## specified phylum
             filter(clean_phylum %in% gom_phyla())
-        message(" DONE")
+        message(" DONE.")
         db_insert_into(con, db_table, r)
     })
     db_create_indexes(con, db_table, indexes = list(c("clean_phylum", "clean_class", "clean_family"),
@@ -177,12 +182,17 @@ fill_store_idigbio_by_geo <- function(coords, db_table, use_cache) {
     db_analyze(con, db_table)
     db_commit(con)
     on.exit(NULL)
+}
+
+
+fill_store_idigbio_by_geo <- function(db_table) {
 
     ## We only want:
     ## - unique records
     ## - invertebrates
     ## - marine
-    ## - within the EEZ boundaries (optional)
+
[…]

### 2017-06-20 R/utils.R
more comprehensive species name cleanup

@@ -1,4 +1,9 @@
 cleanup_species_names <- function(nm, rm_subgenus = FALSE) {
+    if (length(nm) > 1) browser()
+    ## remove author/year
+    if (grepl("[0-9]{4}", nm)) {
+        nm <- gsub("([a-z]+)\\s([a-z]+)?\\s?.+", "\\1 \\2", nm)
+    }
     ## remove words that contain numbers
     nm <- gsub("\\S*\\d\\S*", "", nm)
     ## remove extra spaces
@@ -7,10 +12,12 @@ cleanup_species_names <- function(nm, rm_subgenus = FALSE) {
     nm <- gsub("\\sspp?\\.", "", nm)
     ## remove words fully capitalized
     nm <- gsub("\\s[A-Z]+", "", nm)
+    ## remove cf., aff. and ?
+    nm <- gsub("\\s?(cf\\.|f\\.|aff\\.|\\?|(ex\\.? gr\\.))\\s?", " ", nm)
+    ## remove varieties
+    nm <- gsub("var. [a-z]+$", "", nm)
     ## remove trailing spaces
     nm <- gsub("\\s+$", "", nm)
-    ## remove cf., aff. and ?
-    nm <- gsub("\\s(cf\\.|f\\.|aff\\.|\\?|(ex\\.? gr\\.))\\s", " ", nm)
     ## remove subgenus
     if (rm_subgenus)
         nm <- gsub("\\s?\\([^)]*\\)\\s?", " ", nm)

### 2017-06-20 manuscript.Rmd
edit manuscript

@@
 There are still many species that are easy to access to discover, but the bigger gap is in the species that are difficult to access, and small.
-
-We also need to better coordination in describing new species; so that efforts are not duplicated.
-
-Possibly also need to set guidelines on how to treat cases where the molecular data alone is too ambiguous.
 
@@
 
-Regional faunistic lists assembled and curated by taxonomic experts typically provide a comprehensive and detailed account of species that are known to occur in a given geographic area.
+Regional faunistic lists assembled and curated by taxonomic experts provide a comprehensive and detailed account of species that are known to occur in a given geographic area.
 We assessed how faunistic lists assembled over decades by specialists compare with records in public databases for the same geographic area.

### 2017-06-28 manuscript.Rmd
[fix #29] plot proportions of lot identified at species level

@@
 
+## Supplementary material
+
+```{r id_though_time, fig.cap='Proportion of specimens collected through time identified at the species level.
+Trend lines are  B-splines of the 3rd degree polynomial.
+For all groups, we observe a decline in the proportion of specimens collected identified at the species level suggesting a lack of expertise to identify precisely what is collected.'}
+knitr::include_graphics("figures/idigbio_identification_level_through_time.pdf")
+```
+
 <!--

### 2017-06-28 R/time_plot.R

@@ -325,9 +325,36 @@ test_id_level <- function() {
 
 }
 
-plot_identification_level_through_time <- function(idig_records) {
-    idig_records %>%
+data_identification_level_through_time <- function(idig_records) {
+    res <- idig_records %>%
         mutate(id_level = vapply(cleaned_scientificname, get_id_level, character(1))) %>%
+        filter(id_level != "unknown", !is.na(id_level)) %>%
+        idigbio_add_year()
+
+    res %>%
+        distinct(cleaned_scientificname, .keep_all = TRUE) %>%
+        sample_n(500) %>%
         write_csv("/tmp/res.csv")
+
+    res
+}
+
+
+plot_identification_level_through_time <- function(id_level) {
+    id_level %>%
+        count(year, clean_phylum, id_level) %>%
+        group_by(year, clean_phylum) %>%
+        mutate(p = n/sum(n),
+               n_lots = sum(n)) %>%
+        filter(id_level == "species",
+               clean_phylum %in% c("annelida", "arthropoda", "chordata",
+                                   "cnidaria", "echinodermata", "mollusca")) %>%
+        ggplot(aes(x = year, y = p, colour = id_level)) +
+        geom_point(aes(size = n_lots)) +
+        geom_hline(yintercept = 1) +
+        geom_smooth(aes(fill = id_level), method = "lm", formula = y ~ splines::bs(x, degree = 3), show_guide = FALSE) +
+        facet_wrap(~ clean_phylum) +
+        guides(color = FALSE) +
+        scale_size_continuous(name = "Number of specimens")
 
 }

### 2017-07-06 manuscript.Rmd
edit manuscript

--- 2017-06-20 manuscript.Rmd
@@
 <!-- 1st paragraph. -->
-
-<!--
-
-There can be public involvement in species description, but the exploration and collecting may remain an affair of specialists as it requires manipulating technical gear of have expertise.
-There are still many species that are easy to access to discover, but the bigger gap is in the species that are difficult to access, and small.
-
--->
 
@@
 Of the `r n_idigbio_phyla` phyla occurring in iDigBio, `r records_rare_phyla$n_rare_phyla` are represented by less than 500 records, and make up `r records_rare_phyla$prop_rare_phyla`% of the total number of records.
+Yet, these undersampled taxa are likely to play an important role in the detection of changes in environmental quality due to their size, and physiological vulnerability [@Bik2012].
+Additionally, the impact of these changes on higher trophic levels and ecosystems is largely unexplored.
 
-<!-- From Mark: These previously undersampled taxa are likely to be much more important for the detection of subtle changes in environmental quality due to their size and physiological vulnerability.
-(?).
--->
-<!-- also, the impact of changes in the composition in their communties is largely unexplored, look into Bik's paper? -->
+To assess the status of taxonomic expertise in the US, we calculated the proportion of specimens identified at the species level through time.
+For all taxa, excepts for molluscs, there is a consistent decline, indicating that specimens remain for the most part unidentified in collections \@ref{fig:identification-through-time}.
+For diverse groups such as annelids or arthropods, the proportion of specimens identified at the species-level is now below 30%.
+The number of scientists describing new species might be increasing globally [@Costello2013], but local experts are needed to keep identifying the species that get collected.
 
-<!-- - get numbers of species described by year, look if there is a correlation between sample numbers and number of new species (possible lag with time for description) -->
-<!-- - see if proportion of species identified at genus (or higher) only has increased recently -->
+```{r id_though_time, fig.cap='Proportion of marine invertebrate specimens collected through time identified at the species level.
+Trend lines are  B-splines of the 3rd degree polynomial.
+For all groups, we observe a decline in the proportion of specimens collected identified at the species level suggesting a lack of expertise to identify precisely what is collected.
+(\\#fig:identification-through-time)'}
+knitr::include_graphics("figures/idigbio_identification_level_through_time.pdf")
+```
 
@@
 
+Crowd-sourced data and citizen science initiatives have been proposed to augment data about species abundance and distributions [@Pimm2014].
+For marine taxa, beyond the application to near-shore habitats on easily recognizable species [@Delanay2008], large-scale efforts are currently not possible given the specialized equipment needed to sample most marine habitats and the expertise required to identify species.
+There however opportunities to utilize citizen science to improve biodiversity data and engage the general public into marine conservation issues.
+For instance it has been successfully used to classify plankton images [@Robinson2017], transcribe hand-written museum labels [@Hill2012], or be involved in bioblitzes [@Ballard2017].
+These initiatives however involve a lot less people compared to projects related to birds.
+
 Given the urgency of the biodiversity crisis, we need to develop data-driven initiatives to efficiently document biodiversity, and build on our existing knowledge to collect information to improve our understanding of marine systems.
@@
 
-iDigBio:
-OBIS:
-WoRMS:
+## References
 
-## References
+<div id="refs"></div>
+
+\newpage
+
+## Supplementary material
 

### 2017-07-06 manuscript.Rmd
[fix #28] write methods section

@@
 
-### Data sources
+The list of species for the Gulf of Mexico was obtained by downloading CSV files for each animal phylum from http://e-gulf.org/ [@egulf].
+This database includes the species listed in a comprehensive biotic inventory of the Gulf of Mexico [@Felder2009].
+
+The list of species for the Pacific Northwest was downloaded from:  https://www.eopugetsound.org/species/custom-lists/307 (accessed on July 5th, 2017) as a Microsoft Excel spreadsheet.
+This is a digital version of the species included in [@Kozloff1974].
+
+To search by geographic area, we obtained the boundaries of the Exclusive Economic Zone for the United States from marineregions.org (mrgid:8456) [@mregions].
+For the Pacific Northwest, Kozloff did not specify the exact geographic extent covered by his book.
+We infered a polygon that encompass part of the Oregon coast to the northern parts of Vancouver island.
+For the Gulf of Mexico, we used the same geographic area as the one defined in [@Tunnell2009].
+These geographic areas were split into squares of 0.2°x0.2° which were used to query the web API of iDigBio and OBIS using the R packages ridigbio and robis respectively.
+The resulting occurences were then clipped to only include records that fell within the EEZ boundaries.
+
+The species names used in queries or obtained as results were checked against the World Registry of Marine Species (WoRMS).
+Whenever possible, species names were matched and the current valid names for the species were used.
+This ensured that the same species listed under different synonyms was only counted once.
+Species that could not be matched in WoRMS, and species listed as non-marine in WoRMS were not included in subsequent analyses.
+The R package taxizesoap was used to query the WoRMS database.
+
+To assess the proportion of species for which DNA barcodes are available, we used the current valid names for all records to query the Barcode of Life Datasystem (BOLD) using the R package bold.
+When no records where found for a species, we looked up for alternative matches using its synonyms as listed in WoRMS.
+
+This study is fully reproducible and the code needed to perform the analyses included here is available from: https://github.com/fmichonneau/sok-marine-biodiversity, and archived on Zenodo (doi:xxx.xxx.xxxx).
+The data downloaded from the databases at the time of analysis are archived on Zenodo (doi:xxx.xxx.xxx)/
 

### 2017-07-06 manuscript.Rmd
[fix #16] explain how to read upset plots + add refs

@@
 ```{r compare_database_coverage, fig.cap='Overlap between the regional species list (list), records in iDigBio (idigbio), records in OBIS (obis), and records in BOLD (bold) for the Gulf of Mexico (left) and the Pacific Northwest (right).
+This plot is generated using the UpSet technique, which uses a matrix-based layout to represent intersections of set and their sizes.
+The rows represent the sets, the columns represent the intersections, and each set that is part of the intersection is marked with a black filled circle in the matrix cell.
+Black lines connect the sets that are part of the intersections.
 (\\#fig:database-coverage)', echo=FALSE}
@@
 We infered a polygon that encompass part of the Oregon coast to the northern parts of Vancouver island.
-For the Gulf of Mexico, we used the same geographic area as the one defined in [@Tunnell2009].
+For the Gulf of Mexico, we used the same geographic area as the one defined in [@Tunnell2009] by modifying manually the eastern boundaries of the shape file for the Gulf of Mexico provided by marineregions.org (mrgid:4288).
 These geographic areas were split into squares of 0.2°x0.2° which were used to query the web API of iDigBio and OBIS using the R packages ridigbio and robis respectively.
@@
 When no records where found for a species, we looked up for alternative matches using its synonyms as listed in WoRMS.
+
+Figure \@ref(fig:database-coverage) was generated using the R package UpSetR [@Conway2017] based on the UpSet technique [@Lex2014].
 

### 2017-07-06 R/gom_list.R
remove GBIF from graph + better theme for paper

@@ -239,9 +239,10 @@ plot_richness_per_db <- function(smry_db, region = c("gom", "pnw")) {
                       ) %>%
         tidyr::gather(source_n_spp, n_spp, -taxon_name) %>%
         dplyr::mutate(source_db = gsub("(n_[a-z]+)(_.+)?", "\\1", source_n_spp)) %>%
+        dplyr::filter(source_db != "n_gbif") %>%
         ggplot(aes(x = factor(source_db,
-                              levels = rev(c("n_total", "n_obis", "n_gbif", "n_idigbio", "n_barcoded")),
-                              labels = c("BOLD", "iDigBio", "GBIF", "OBIS", "List")
+                              levels = rev(c("n_total", "n_obis", "n_idigbio", "n_barcoded")),
+                              labels = c("BOLD", "iDigBio", "OBIS", "List")
                               ),
                    y = n_spp,
                    fill = factor(source_n_spp,
@@ -249,14 +250,14 @@ plot_richness_per_db <- function(smry_db, region = c("gom", "pnw")) {
                                      gsub("XXX", region,
                                           c("n_total",
                                             "n_obis_in_XXX", "n_obis_XXX_diff",
-                                            "n_gbif_in_XXX", "n_gbif_XXX_diff",
+                                            #"n_gbif_in_XXX", "n_gbif_XXX_diff",
                                             "n_idigbio_in_XXX", "n_idigbio_XXX_diff",
                                             "n_barcoded"
                                             ))),
                                  labels = rev(
                                      c("Species in list",
                                        "OBIS in region","OBIS global",
-                                       "GBIF in region", "GBIF global",
+                                       #"GBIF in region", "GBIF global",
                                        "iDigBio in region", "iDigBio global",
                                        "BOLD"
                                        ))
@@ -270,7 +271,7 @@ plot_richness_per_db <- function(smry_db, region = c("gom", "pnw")) {
         labs(title = "") +
         scale_fill_viridis(name = "Data source", option = "viridis", discrete = TRUE,
                            guide = guide_legend(reverse=TRUE)) +
-        theme_gray(base_family = "Ubuntu Condensed") +
+        theme_bw(base_family = "Ubuntu Condensed") +
         coord_flip()
 }
 

### 2017-07-06 manuscript.Rmd
many edits to manuscript

@@
 The digitization of natural history collections and other biodiversity information systems can now be used to enable synthesis and highlight knowledge gaps.
-Here we use these data to synthesize the state of our knowledge for marine invertebrates in the continental US.
+Here we use these data to synthesize the state of our knowledge for marine invertebrates in the continental United States.
 We show that current knowledge and sampling efforts are inadequate given the diversity of the fauna and the geographic extent of US.
-While recent high-throughput sequencing technologies promise to greatly speed up the documentation of marine species and their ranges, its value would be greatly improved by anchoring the results into existing DNA sequence data and associated vouchers.
+While recent high-throughput sequencing technologies promise to greatly speed up the documentation of marine species and their ranges, its value will be greatly improved by anchoring the results into existing DNA sequence data and associated vouchers.
 However, the best documented phyla have at most 25% of their species sequenced for the most widely used DNA barcoding gene, cytochrome oxidase I (COI).
 More sampling, coordinated at the scale of the US, is needed to improve our knowledge of species and their geographic ranges.
-This data is needed to detect the impact of climate changes and other anthropogenic strains to global environments.
+This data is needed to detect the impact of climate changes and other anthropogenic strains to the environment.
 Advances in computing, information systems, and sequencing technologies, have set the stage to inform where future efforts for documenting biodiversity should be applied.
@@
 With increasing impact of human activities on the biosphere, we need efficient and accurate approaches to document changes in species ranges and community composition.
-While terrestrial ecosystems have received considerable attention, the baseline of healthy and resilient marine ecosystems are not well understood [@Worm2006], and modeling the impact of human activities is limited to a few taxonomic groups [@Tittensor2010].
-For marine organisms, because of the diversity of their complex life history strategies and their size range, we have limited information about which species live where, or what portion of their diversity remains to be discovered and described [@Mora2011].
+While terrestrial ecosystems have received considerable attention, the baseline of healthy and resilient marine ecosystems are not well understood [@Worm2006], and modeling the impact of human activities has been limited to a few taxonomic groups [@Tittensor2010].
+For marine organisms, because of the diversity of their complex life history strategies, their size range, and the difficulties associated with sampling their habitats, we have limited information about which species live where, or what portion of their diversity remains to be discovered and described [@Mora2011].
 Additionally, coordinated marine ecosystem monitoring initiatives are lacking [@Duffy2013,@Borja2016], making it difficult to monitor changes in species distributions and abundances.
-While most documented global extinctions of marine species are vertebrates, many others may have been lost unnoticed because they were not known [@Sala2006].
-There is therefore a pressing need to improve our knowledge of marine systems before their functioning is irreversibly altered, and species go extinct.
+While most documented global extinctions of marine species are vertebrates, many others species may have been lost unnoticed because they were not known [@Sala2006].
+There is therefore a pressing need to improve our knowledge of marine systems before their functioning are irreversibly altered, and species go extinct.
 
@@
 iDigBio [@Page2015]), species occurrences (e.g., OBIS [@Zhang2002]), genetic data (e.g., GenBank [@Benson2013], BOLD [@Ratnasingham2007a]), and registers of species names (e.g., WoRMS [@Costello2013b]), now allow us to evaluate what and how much we know.
-Here, we aggregate publicly available data for marine invertebrates that occur within the limits of the exclusive economic zone (EEZ) of the continental United States (US), as well as the Gulf of Mexico and the Pacific Northwest, to highlight gaps in our knowledge that provide a roadmap for important new directions of global research.
-We show that public databases are relatively incomplete, and that most species are known from only a few specimens and a few locations.
-We emphasize that our understanding of marine biodiversity, and the impact of human activities on communities require not only to get a more accurate list of the species that inhabits the oceans, but we also need more comprehensive data on species distributions and abundances.
-This can only be achieved with additional cost-effective sampling and a commitment to training more taxonomists and experts that can identify and interpret the material collected.
+Here, we aggregate publicly available data for marine invertebrates that occur within the limits of the exclusive economic zone (EEZ) of the continental United States (US), as well as the Gulf of Mexico and the Pacific Northwest, to highlight gaps in our knowledge that provide a roadmap for important new directions in global research.
+We show that while public databases are capturing a significant portion of our knowledge about species richenss, most species are known from only a few specimens and a few locations.
+We emphasize that our understanding of marine biodiversity, and the impact of human activities on communities require not only to get a more accurate list of the species that inhabits the oceans, but more comprehensive data on species distributions and abundances are needed.
+These goals can only be achieved with additional cost-effective sampling and a commitment to training more taxonomists and experts that can identify and interpret the material collected.
 
 Here, we focus on the continental US because the marine biodiversity is extensively documented as the result of a long history of studies by governmental, academic, and private institutions [@Fautin2010].
-The continental US includes a wide diversity of marine ecosystems from the Florida Keys, to the Gulf of Maine on the East Coast; and from the arid Southern California to the oceanic coast of the Pacific Northwest on the West coast.
+The continental US includes a wide diversity of marine ecosystems from the Florida Keys, to the Gulf of Maine on the East Coast; and from the arid Southern California to the oceanic shores of the Pacific Northwest on the West coast.
 With estuaries, bays, sandy beaches, rocky shores, and coral reefs, the US coastlines harbors many biotas.
@@
 This diversity of habitats and bathymetric profiles, combined with efforts at the national level to make data publicly available, make this region especially suitable to evaluate how much we know about marine biodiversity using biodiversity databases.
-However, apart from the syntheses that emerged from the Census of Marine Life efforts [@Costello2010,@Fautin2010], no study has yet attempted to synthesize the current state of knowledge of marine biodiversity across the continental US, and this study is the first using data from public databases.
+Additionally, apart from the syntheses that emerged from the Census of Marine Life efforts [@Costello2010,@Fautin2010], no study has yet attempted to synthesize the current state of knowledge of marine biodiversity across the continental US, and this study is the first using data from public databases.
 
@@
 We assessed how faunistic lists assembled over decades by specialists compare with records in public databases for the same geographic area.
-We parsed faunistic lists to extract marine invertebrates from the Gulf of Mexico [@Tunnell2009] and from the Pacific Northwest [@Kozloff1987].
+We parsed faunistic lists to extract marine invertebrates from the Gulf of Mexico [@Tunnell2009] and from the Pacific Northwest [@Kozloff1974], leading to `r format(sum(gom_worms$is_marine, na.rm=TRUE), big.mark=",")` and `r format(sum(kozloff_worms$is_marine, na.rm=TRUE), big.mark=",")` species respectively.
 Our goals were (1) to compare  the species lists for these geographic areas assembled by experts with the records found in databases; (2) to assess the proportion of species included in the expert lists that have records in public databases.
@@
 To compare the faunistic lists from the Gulf of Mexico and the Pacific Northwest with records in public databases, we queried the iDigBio (https://portal.idigbio.org) and OBIS (http://www.iobis.org/) databases for all records they included in the same geographic boundaries as the lists.
-After filtering these lists to only include the same higher taxonomic level taxa as the species lists, and keeping only records for which the species could be matched to a species name in WoRMS, we had for the Gulf of Mexico `r format(nrow(idigbio_gom_records), big.mark=",")` and `r format(nrow(obis_gom_records), big.mark=",")` records for iDigBio and OBIS respectively, representing `r format(dplyr::n_distinct(idigbio_gom_records$worms_valid_name), big.mark=",")` and `r format(dplyr::n_distinct(obis_gom_records$worms_valid_name), big.mark=",")` species.
-For the Pacific Northwest, we retrieved `r format(nrow(idigbio_pnw_records), big.mark=",")` and `r format(nrow(obis_kozloff_records), big.mark=",")` records for iDigBio and OBIS respectively representing `r format(dplyr::n_distinct(idigbio_pnw_records$worms_valid_name), big.mark=",")` and `r format(dplyr::n_distinct(obis_kozloff_records$worms_valid_name), big.mark=",")` species.
-For the Gulf of Mexico, the majority of the species included in the expert lists are recorded in both iDigBio and OBIS (`r format_output(species_in_common$n_spp_gom_idigbio_obis/species_in_common$n_spp_gom*100)` %), or only OBIS (`r format_output(species_in_common$n_spp_gom_obis/species_in_common$n_spp_gom*100)` %) (Fig.
-\@ref(fig:database-overlap)).
+We filtered the results to only include the same higher taxonomic level taxa as in the faunistic lists, and kept only records for which the species could be matched to a species name in WoRMS.
+For the Gulf of Mexico, we retrieved  `r format(dplyr::n_distinct(idigbio_gom_records$worms_valid_name), big.mark=",")` species in iDigBio (from $n =$  `r format(nrow(idigbio_gom_records), big.mark=",")` records), and `r format(dplyr::n_distinct(obis_gom_records$worms_valid_name), big.mark=",")` species in OBIS ($n=$ `r format(nrow(obis_gom_records), big.mark=",")` records).
+For the Pacific Northwest, we retrieved `r format(dplyr::n_distinct(idigbio_pnw_records$worms_valid_name), big.mark=",")` species in iDigBio ($n=$ and `r format(nrow(idigbio_pnw_records), big.mark=",")`) and `r format(dplyr::n_distinct(obis_kozloff_records$worms_valid_name), big.mark=",")` species in OBIS ($n=$ `r format(nrow(obis_kozloff_records), big.mark=",")`).
+Overall, the number of species known from the Gulf of Mexico match the number estimated by experts (`r format(11150-1541-434, big.mark=",")` species [@Fautin2010]).
+For the Pacific coast, this number is lower than the estimate (`r format(8374-909-134, big.mark=",")` species [@Fautin2010]), but this estimate included species from California and from deeper areas that were not included in Kozloff's faunistic list.
+
+For the Gulf of Mexico, the majority of the species included in the expert faunistic list are recorded in both iDigBio and OBIS (`r format_output(species_in_common$n_spp_gom_idigbio_obis/species_in_common$n_spp_gom*100)` %), or only OBIS (`r format_output(species_in_common$n_spp_gom_obis/species_in_common$n_spp_gom*100)` %) (Fig.
+\@ref(fig:database-coverage)).
 For the Pacific Northwest, these proportions were much lower with only `r format_output(species_in_common$n_spp_kozloff_idigbio_obis/species_in_common$n_spp_kozloff*100)`% of the species from the regional list with records in OBIS and iDigBio, and `r format_output(species_in_common$n_spp_kozloff_obis/species_in_common$n_spp_kozloff*100)`%  in OBIS only (Fig.
 \@ref(fig:database-coverage)).
-Previous esimates for species richness in the Gulf of Mexico for marine invertebrates were of `r format(11150-1541-434, big.mark=",")` species, and `r format(8374-909-134, big.mark=",")` for the Pacific coast (including California) [@Fautin2010].
+Actually, the majority of the species in the faunistic list are not recorded in any of the public databases.
+Overall, there is more overlap among the data sources for the Gulf of Mexico than the Pacific Northwest.
 
 ```{r compare_database_coverage, fig.cap='Overlap between the regional species list (list), records in iDigBio (idigbio), records in OBIS (obis), and records in BOLD (bold) for the Gulf of Mexico (left) and the Pacific Northwest (right).
-This plot is generated using the UpSet technique, which uses a matrix-based layout to represent intersections of set and their sizes.
-The rows represent the sets, the columns represent the intersections, and each set that is part of the intersection is marked with a black filled circle in the matrix cell.
+This plot is generated using the UpSet technique[@Lex2014], which uses a matrix-based layout to represent intersections of sets and their sizes.
+The rows represent the sets, the columns represent the intersections, and each set that is part of the intersection is marked with a black-filled circle in the cells underneath.
 Black lines connect the sets that are part of the intersections.
@@
 OBIS tends to be closer to the full record set, but iDigBio has more species for the Gulf of Mexico that are not included in the list (Fig.
-\@ref(fig:database-overlap))
-
-Overall, there is more overlap among the data sources for the Gulf of Mexico than the Pacific Northwest.
+\@ref(fig:database-overlap)).
+Many species of mollusks, arthropods, and annelids, known to occur in the Gulf of Mexico or in the Pacific Northwest, have not digital records for them in these geographic areas.
 
@@
 
-Species that are included in the list but do not have records in databases could be (1) rare: they have never been collected or have been collected only a few times, and the corresponding records have not been digitized yet; (2) difficult to preserve: large, very small, or fragile species are preserved less frequently; (3) are in the database under a different name: the species has been synonimized or re-identified since the publication of the list under a different name that may be included in the database; (4) at the limit of its range: it was included in the list because it is known to occur at the limit of its geographic zone, but has never been collected within the boundaries; (5) does not have geographic coordinates associated with it: it can be in the database but without georeferenced datapoints that made it possible to discover them.
-By querying iDigBio we found that for the `r format(not_really_in_database$idig_gom_n_spp_not_in_db, big.mark=",")` species that do not appear within the geographic range of the Gulf of Mexico but are on the list, `r format_output(not_really_in_database$idig_gom_p_spp_not_in_db*100)`% are truly absent from iDigBio, while for the `r format(not_really_in_database$idig_gom_n_spp_no_coords, big.mark=",")` species that appear in iDigBio `r format_output(not_really_in_database$idig_gom_p_spp_no_coords*100)`%  do not have geographic coordinates associated with them, and for the ones that dok `r format_output(not_really_in_database$idig_gom_p_spp_within_usa*100)`% fall within the US waters (n = `r format(not_really_in_database$idig_gom_n_spp_within_usa, big.mark=",")`).
+Species that are included in the list but do not have records in databases could be (1) rare: they have never been collected or have been collected only a few times, and the corresponding records have not been digitized yet; (2) difficult to preserve: large, very small, or fragile species are deposited in museums less frequently; (3) are in the database under a different name: the species has been synonimized or re-identified since the publication of the list under a different name that may be included in the database; (4) at the limit of their range: the species were included in the list because they are known to occur within the boundaries of the geographic area, but they have never been collected within these geographic boundaries; (5) do not have geographic coordinates associated with them: some of these species can be in the database but without georeferenced datapoints and therefore were not returned in our queries.
 
-Species that are in the database but not in the list could be: (1) records not considered by the authors of the checklist: either because they were not accessible or known to the authors; (2) records erroneously excluded the records (2) records that have been discovered since the list has been assembled (for iDigbio records from the Gulf of Mexico, this represent `r format_output(mean(recently_collected$gom_in_idig$min_year > 2001))`%, n=`r nrow(recently_collected$gom_in_idig)` and `r format_output(mean(recently_collected$pnw_in_idig$min_year > 1987))`%, n=`r nrow(recently_collected$pnw_in_idig)`); (3) records with incorrect identification; (4) records with incorrect geographic coordinates: their associated geographic information in the database places them in the Gulf of Mexico or the Pacific Northwest, but they were actually collected elsewhere.
+Untangling the relative contribution of these factors to explain the discrepencies is difficult as it would require to know the criteria used by the experts who have assembled the lists.
+However, by querying iDigBio we found that for the `r format(not_really_in_database$idig_gom_n_spp_not_in_db, big.mark=",")` species that do not appear within the geographic range of the Gulf of Mexico but are on the list, `r format_output(not_really_in_database$idig_gom_p_spp_not_in_db*100)`% are truly absent from iDigBio, while for the `r format(not_really_in_database$idig_gom_n_spp_no_coords, big.mark=",")` species that appear in iDigBio, `r format_output(not_really_in_database$idig_gom_p_spp_no_coords*100)`%  do not have geographic coordinates associated with them, and for the ones that do have geographic coordinates, `r format_output(not_really_in_database$idig_gom_p_spp_within_usa*100)`% ($n=$  `r format(not_really_in_database$idig_gom_n_spp_within_usa, big.mark=",")`) fall within the US waters.
+These numbers indicate that the lack of geographic coordinates associated with the specimens in the database contribute at least significantly to the discrepency, but that a large fraction of these species have no digital records of their occurrence in the Gulf of Mexico.
+
+Species that are in the database but not in the faunistic list could be: (1) records not considered by the authors of the checklist: either because they were not accessible or known to the authors; (2) records erroneously excluded by the author of the checklist; (3) records that have been discovered since the list has been assembled; (4) records with incorrect identification: records identified as a species that does not occur in the area; (5) records with incorrect geographic coordinates: their associated geographic information in the database places them in the Gulf of Mexico or the Pacific Northwest, but they were actually collected elsewhere.
+
+We investigated the proportion of species discovered since the publication of the faunistic lists using iDigBio records.
+They represent `r sum(recently_collected$gom_in_idig$min_year > 2001)` species for the Gulf of Mexico, and `r sum(recently_collected$pnw_in_idig$min_year > 1987)` species for the Pacific Northwest.
+While errors in geographic coordinates cannot be completely ruled out, it is more likely that the main contributing factor was that the authors of these checklists were not aware of the presence of these species based only on museum records, as these lists were mostly assembled from publications, and only a small fraction of museum records are used in publications.
 
@@
 
-To assess whether species included in regional lists but absent from the same geographic area nevertheless occur elsewhere, we queried the iDigBio and OBIS databases for all the marine invertebrate species included in these lists.
+To assess whether species included in regional faunistic lists but absent from the geographic area in public databases nevertheless occur elsewhere, we queried the iDigBio and OBIS using the species names included in these faunistic lists.
 
-We extracted `r format(sum(gom_worms$is_marine, na.rm = TRUE), big.mark=",")` marine invertebrates species from the faunistic list of the Gulf of Mexico, and `r format(sum(kozloff_worms$is_marine, na.rm=TRUE), big.mark=",")` species from the Pacific Northwest.
-We then looked for the number of species listed that are known to occur in the same geographic area, the United States, or anywhere else in the world.
-Overall, OBIS had georeferenced records for most of the species that are included in the species lists.
+We extracted the `r format(sum(gom_worms$is_marine, na.rm = TRUE), big.mark=",")` marine invertebrates species from the faunistic list of the Gulf of Mexico, and `r format(sum(kozloff_worms$is_marine, na.rm=TRUE), big.mark=",")` species from the Pacific Northwest.
+We then looked for the number of species from these faunistic lists that are known to occur within the geographic area, or somewhere else in the world.
+Overall,
+OBIS had georeferenced records for most of the species that were included in the faunistic lists for the Gulf of Mexico.
 However, even for the better studied phyla about 75% of the species have records from the same geographic area, while it is less than 50% for the Pacific Northwest.

### 2017-07-06 manuscript.Rmd
remove section about comparison with species lists

@@
 While errors in geographic coordinates cannot be completely ruled out, it is more likely that the main contributing factor was that the authors of these checklists were not aware of the presence of these species based only on museum records, as these lists were mostly assembled from publications, and only a small fraction of museum records are used in publications.
-
-### Comparison with Regional Lists by Species Names
-
-To assess whether species included in regional faunistic lists but absent from the geographic area in public databases nevertheless occur elsewhere, we queried the iDigBio and OBIS using the species names included in these faunistic lists.
-
-We extracted the `r format(sum(gom_worms$is_marine, na.rm = TRUE), big.mark=",")` marine invertebrates species from the faunistic list of the Gulf of Mexico, and `r format(sum(kozloff_worms$is_marine, na.rm=TRUE), big.mark=",")` species from the Pacific Northwest.
-We then looked for the number of species from these faunistic lists that are known to occur within the geographic area, or somewhere else in the world.
-Overall,
-OBIS had georeferenced records for most of the species that were included in the faunistic lists for the Gulf of Mexico.
-However, even for the better studied phyla about 75% of the species have records from the same geographic area, while it is less than 50% for the Pacific Northwest.
-
-```{r database_richness, fig.cap='Comparison of the fraction of species listed in regional lists that are present in biodiversity databases for the Gulf of Mexico (A) and the Pacific Northwest (B).'}
-knitr::include_graphics("figures/richness_per_db.pdf")
-```
-
-These results suggest that even when considering species lists that have been curated and that represent possibly the most accurate knowledge for any particular geographic area, today we can only partially reconstruct these lists based on digitized information.
-For instance, based on OBIS data the list would be `r format_output(summary_richness_per_db_gom$prop_obis_in_gom[summary_richness_per_db_gom$taxon_name == "Porifera"]*100)`% complete for sponges to `r format_output(summary_richness_per_db_gom$prop_obis_in_gom[summary_richness_per_db_gom$taxon_name == "Mollusca"]*100)`% for molluscs.
-However, these levels drop significantly if we consider species for which museum specimens are available from  `r format_output(summary_richness_per_db_gom$prop_idigbio_in_gom[summary_richness_per_db_gom$taxon_name == "Mollusca"]*100)`% for molluscs to `r format_output(summary_richness_per_db_gom$prop_idigbio_in_gom[summary_richness_per_db_gom$taxon_name == "Annelida"]*100)`% for annelids.
-Similar patterns can be observed for the Pacific Northwest, but overall the levels of data availability are lower: from `r format_output(summary_richness_per_db_koz$prop_obis_in_pnw[summary_richness_per_db_koz$taxon_name == "Porifera"]*100)`% for sponges to `r format_output(summary_richness_per_db_koz$prop_obis_in_pnw[summary_richness_per_db_koz$taxon_name == "Annelida"]*100)`% for annelids for the proportion of species that have records in OBIS, and from  `r format_output(summary_richness_per_db_koz$prop_idigbio_in_pnw[summary_richness_per_db_koz$taxon_name == "Porifera"]*100)`% for sponges to `r format_output(summary_richness_per_db_koz$prop_idigbio_in_pnw[summary_richness_per_db_koz$taxon_name == "Mollusca"]*100)`% for molluscs for the proportion of species that have records in iDigBio.
 

### 2017-07-06 R/time_plot.R
use same color palette for prop of records ID'ed at species-level

@@ -378,6 +378,7 @@ plot_identification_level_through_time <- function(id_level) {
         guides(color = FALSE, fill = FALSE) +
         scale_size_continuous(name = "Number of specimens") +
         labs(x = "Year", y = "Proportion of specimens identified at the species level") +
-        theme(legend.position = "top")
+        theme(legend.position = "top") +
+        scale_colour_hc() + scale_fill_hc()
 
 }

### 2017-07-06 manuscript.Rmd
more text edits

@@
 These numbers indicate that the lack of geographic coordinates associated with the specimens in the database contribute at least significantly to the discrepency, but that a large fraction of these species have no digital records of their occurrence in the Gulf of Mexico.
+In other words, additional digitization of collections (including capturing their geographic coordinates information) are needed.
+For instance, one can expect that the digitization coverage for the Gulf of Mexico will greatly increase once records from the Florida Wildlife Research Institute collections (St Petersburg, FL) become integrated into the biodiversity information systems.
 
@@
 While errors in geographic coordinates cannot be completely ruled out, it is more likely that the main contributing factor was that the authors of these checklists were not aware of the presence of these species based only on museum records, as these lists were mostly assembled from publications, and only a small fraction of museum records are used in publications.
-
-Museum specimens are critical to validate the occurrence of species [@Rocha2014], especially for marine invertebrates for which species limits based on morphology have been known to be challenged by other lines of evidence including molecular data [@Knowlton2000].
-Some of the discrepancies observed between the number of species for which occurrence data and museum specimens are available suggest that additional digitization of collections (including capturing their geographic information) are needed.
-For instance, one can expect that the digitization coverage for the Gulf of Mexico will greatly increase when records from the Florida Wildlife Research Institute collection (St Petersburg, FL) become integrated into the biodiversity information systems.
+However, museum specimens are critical to validate the occurrence of species [@Rocha2014], especially for marine invertebrates for which species limits based on morphology have been known to be challenged by other lines of evidence including molecular data [@Knowlton2000].
 
@@
 
-We combined `r format(nrow(idigbio_records), big.mark=",")` museum records for marine invertebrates collected in the EEZ waters of the continental US representing `r length(unique(tolower(idigbio_records[["data.dwc:phylum"]])))` phyla and `r format(length(unique(idigbio_records$worms_valid_name)), big.mark=",")` species.
+We combined `r format(nrow(idigbio_records), big.mark=",")` museum records for marine invertebrates collected within the EEZ of the continental US representing `r length(unique(tolower(idigbio_records[["data.dwc:phylum"]])))` phyla and `r format(length(unique(idigbio_records$worms_valid_name)), big.mark=",")` species.
 Most samples were collected on the continental shelf (Fig.
@@
 When investigating the relation between the number of samples and the number of species for each cell of the map (Fig.
-\@ref(fig:correlation-sampling)), we can see that until reaching about 100 records per cell there is little redundancy in the sampled species (each record represent a different species).
+\@ref(fig:correlation-sampling)), it takes about 100 records per cell before there is some redundancy in the sampled species (each record represent a different species).
 
@@
 
-The inadequacy of the sampling effort compared to the diversity of the fauna is also reflected by the fact that only `r prop_singleton_species`% of the species are known from a single record.
-The strong correlation between the number of samples collected and the species richness suggests that any new sampling effort will lead to new observations (Fig.
-\@ref(fig:map-diversity-sampling)) and new species.
+The inadequacy of the sampling effort compared to the diversity of the fauna is also reflected by the fact that `r prop_singleton_species`% of the species are known from a single record.
+The strong correlation between the number of samples collected and the species richness suggests that any new sampling effort will lead to new observations and new species (Fig.
+\@ref(fig:correlation-sampling)).
 Another indication of this inadequacy is the relation between the number of collected specimen and the number of species recorded through time (Fig.
@@
 
-```{r idigbio-through-time, fig.cap='Cumulative number of samples (A) and species (B) of marine invertebrates recorded through time based on iDigBio records. (\\#fig:idigbio-through-time)'}
+```{r idigbio-through-time, fig.cap='Cumulative number of samples (A) and species (B) of marine invertebrates recorded through time based on iDigBio records.
+Relation between number of samples collected, and number of species recorded for the first time (C).
+(\\#fig:idigbio-through-time)'}
 knitr::include_graphics("figures/idigbio_through_time.pdf")
@@
 To assess the status of taxonomic expertise in the US, we calculated the proportion of specimens identified at the species level through time.
-For all taxa, excepts for molluscs, there is a consistent decline, indicating that specimens remain for the most part unidentified in collections \@ref{fig:identification-through-time}.
+For all taxa, excepts for molluscs, there is a consistent decline, indicating that specimens remain for the most part unidentified in collections (Fig.
+\@ref(fig:identification-through-time)).
 For diverse groups such as annelids or arthropods, the proportion of specimens identified at the species-level is now below 30%.

### 2017-07-07 manuscript.Rmd
edit manuscript

@@
 In other words, additional digitization of collections (including capturing their geographic coordinates information) are needed.
-For instance, one can expect that the digitization coverage for the Gulf of Mexico will greatly increase once records from the Florida Wildlife Research Institute collections (St Petersburg, FL) become integrated into the biodiversity information systems.
+For instance, one can expect that the digitization coverage for the Gulf of Mexico will greatly increase once records from the Florida Wildlife Research Institute collections (St Petersburg, FL) become integrated into biodiversity information systems.
 
@@
 
-We combined `r format(nrow(idigbio_records), big.mark=",")` museum records for marine invertebrates collected within the EEZ of the continental US representing `r length(unique(tolower(idigbio_records[["data.dwc:phylum"]])))` phyla and `r format(length(unique(idigbio_records$worms_valid_name)), big.mark=",")` species.
+We combined `r format(nrow(idigbio_records), big.mark=",")` museum records for marine invertebrates collected within the EEZ of the continental US representing `r length(unique(tolower(idigbio_records[["data.dwc:phylum"]])))` phyla and `r format(length(unique(idigbio_records$worms_valid_name)), big.mark=",")` species from iDigBio.
 Most samples were collected on the continental shelf (Fig.
@@
 
-These patterns of spatial distribution of the species richness are strongly confounded by the sampling effort.
-When investigating the relation between the number of samples and the number of species for each cell of the map (Fig.
-\@ref(fig:correlation-sampling)), it takes about 100 records per cell before there is some redundancy in the sampled species (each record represent a different species).
+These patterns of spatial distribution for the species richness are strongly confounded by the sampling effort.
+Investigating the relation between the number of samples and the number of species for each cell of the map (Fig.
+\@ref(fig:correlation-sampling)) reveals that it takes about 100 records per cell before there is clear redundancy in the sampled species (each record represent a different species).
 
-```{r correlation-sampling, fig.cap='Relation between number of records and number of species for each cell of the map used in Fig. \\@ref(fig:map-diversity-sampling). (\\#fig:correlation-sampling)'}
+```{r correlation-sampling, fig.cap='Relation between number of records and number of species for each cell of the map used in Fig.
+\\@ref(fig:map-diversity-sampling).
+The black line represents the 1:1 relation.
+(\\#fig:correlation-sampling)'}
 knitr::include_graphics("figures/plot_sampling_effort.pdf")
@@
 The inadequacy of the sampling effort compared to the diversity of the fauna is also reflected by the fact that `r prop_singleton_species`% of the species are known from a single record.
-The strong correlation between the number of samples collected and the species richness suggests that any new sampling effort will lead to new observations and new species (Fig.
-\@ref(fig:correlation-sampling)).
 Another indication of this inadequacy is the relation between the number of collected specimen and the number of species recorded through time (Fig.
@@
 Since then, the slowing down of collections after the 1970s has been accompanied with a slowing down in the discovery of new records: `r n_spp_comparison$n_spp_2017_2000` species have been documented for the first time since 2000 (`r format_output(n_spp_comparison$n_spp_2017_2000/18)` species.year^-1^), while `r n_spp_comparison$n_spp_1970_1960` have been documented between 1960 and 1970 (`r format_output(n_spp_comparison$n_spp_1970_1960/11)` species.year^-1^).
-Therefore, coordinated large-scale biodiversity expeditions will continue to generate a wealth of data and new records.
-For instance, a recent study in Western Australia reported that 135 samples led to 805 species of crustaceans and polychaetes, 94.6% and 72% of which were undescribed [@Poore2014].
+The strong correlations between (1) the number of samples collected and the species richness (Fig.
+\@ref(fig:correlation-sampling)).; (2) the number of specimens collected per year and the number of species recorded for the first time (Fig.
+\@ref(fig:idigbio-through-time)C), suggest that any new sampling effort will lead to new observations and new species.
+Coordinated large-scale biodiversity expeditions continue to be needed to improve our understanding of the diversity of species and their spatial distributions.
+While less explored than the US, a recent study in Western Australia reported that 135 samples led to 805 species of crustaceans and polychaetes, 94.6% and 72% of which were undescribed [@Poore2014], highlighting the potential for discovery in the oceans.
 
 ```{r idigbio-through-time, fig.cap='Cumulative number of samples (A) and species (B) of marine invertebrates recorded through time based on iDigBio records.
-Relation between number of samples collected, and number of species recorded for the first time (C).
+Relation between number of samples collected, and number of species recorded for the first time (C), each point represents numbers for a given year.
 (\\#fig:idigbio-through-time)'}
@@
 
-Predicting the geographic range of species, and how they can be impacted by global change has undergone important advances recently.
-However, it is generally recognized that to obtain reliable results such efforts require to have accurate geographic information for a minimum of 10 specimens but having at least 30 is preferred.
+Predicting the geographic range of species, and how they will be impacted by global change has undergone important advances recently [@xxxx].
+However, it is generally recognized that to obtain reliable results such efforts require to have accurate geographic information for a minimum of 10 specimens but having at least 30 is preferred [@xxxx].
 Based on iDigBio data, `r prop_10_specimens_species`% of species have less than 10 specimens and `r prop_30_specimens_species`% have less than 30 specimens.
@@
 To assess the status of taxonomic expertise in the US, we calculated the proportion of specimens identified at the species level through time.
-For all taxa, excepts for molluscs, there is a consistent decline, indicating that specimens remain for the most part unidentified in collections (Fig.
+For all taxa, excepts for molluscs, there is a consistent decline, indicating that specimens collected recently remain for the most part unidentified in collections (Fig.
 \@ref(fig:identification-through-time)).
 For diverse groups such as annelids or arthropods, the proportion of specimens identified at the species-level is now below 30%.
-The number of scientists describing new species might be increasing globally [@Costello2013], but local experts are needed to keep identifying the species that get collected.
+The number of scientists describing new species might be increasing globally [@Costello2013], but taxonomic experts continue to be needed to identify specimens recently collected.
 
@@
 This approach opens up new research venues as it becomes possible to estimate the diversity from species-rich samples that were virtually inaccessible before.
-This is particularly true in marine communities that include multiple phyla and organisms varying in size by order of magnitude, making identification challenging.
+This is particularly true in marine communities that include multiple phyla and organisms varying in size by orders of magnitude, making identification challenging.
 However, the utility of the data generated by these projects is enhanced by the possibility of matching DNA sequences obtained to previously sequenced and correctly identified species [@Leray2015].
-With decreasing sequencing costs, and the prospect of automating sample collection, these technologies may soon be used routinely to monitor biodiversity at defined temporal frequency [@Bohan2017].
+With decreasing sequencing costs, and the prospect of automating sample collection, these technologies may soon be used routinely to monitor biodiversity continuously [@Bohan2017].
 DNA barcoding provides the possibility to link these newly generated data with historical natural history knowledge about species and their distributions.
@@
 The proportion of species that have DNA barcodes available is relatively constant across the better studied phyla at the scale of the EEZ.
-It varies between 12% for bryozoans to 28% for cnidarians.
+It varies between 12% for bryozoans to 28% for cnidarians (\@ref(fig:proportion-barcoded)).
 The proportions for species known to occur in the Gulf of Mexico are usually similar than for the US, while typically a higher fraction of species from the Pacific Northwest are available.
-In 2012, it was estimated that a global scale, less than 3% of the marine species had been sequenced [@Appeltans2012], and that this number was even lower in some areas (e.g.
+In 2012, it was estimated that, at a global scale, less than 3% of the marine species had been sequenced [@Appeltans2012], and that this number was even lower in some areas (e.g.
 2.6% for the Antarctic [@Grant2009]).
@@
 ```{r proportion-barcoded, fig.cap='Proportion of species known to occur in the US EEZ based on iDigBio records, in the Gulf of Mexico [@Tunnell2009], in the Pacific Northwest[@Kozloff1987], that have publicly available DNA barcodes.
-Numbers next to the bars indicate the number of species.'}
+Numbers next to the bars indicate the number of species for which DNA barcodes are available.
+(\\#fig:proportion-barcoded)'}
 knitr::include_graphics("figures/proportion_barcoded.pdf")
@@
 
-Similarly, a significant fraction of museum records do not have or have inaccurate geographic information associated with them.
+Similarly, a significant fraction of museum records have missing or inaccurate geographic information associated with them.
 While tools have been developed to convert names of locales into GPS coordinates on land [@rios2010geolocate], similar approaches may not work for locales at sea.
@@
 For both the spelling of scientific names, and the geographic information, computing tools can be developed to detect issues and propose solutions.
-There are opportunities for services using publicly available data to be developed to make curators and collection managers aware of these issues so that they can be fixed at the source.
+There are opportunities to develop computing services that use publicly available data to flag and fix these issues directly in collection databases.
 It however requires that resources are made available to museum collections to address these issues.
@@
 
-<!-- - develop infrastructure and tools to continuously check and validate data quality in databases. -->
-
 ### Collect better data faster and more cost effectively
 
-This study highlights that we have too little data about too few species of marine invertebrates, we recommend that additional coordinated field expeditions are organized to sample more thoroughly the marine environment.
+This study highlights that we have too little data about too few species of marine invertebrates.
+We recommend that additional coordinated field expeditions are organized to sample more thoroughly the marine environment.
 The relation between sampling effort and diversity from museum collections makes clear that any additional samples will improve greatly our knowledge of marine biodiversity.
@@
 
-Integrating metabarcoding as part of the data collection conducted in biodiversity surveys is a promising approach to screen an habitat for its potential to lead to new species or species that have not yet public DNA barcodes.
+Integrating metabarcoding as part of the data collection conducted in biodiversity surveys is a promising approach to screen an habitat for its potential to lead to new discoveries.
 For instance, by analyzing metabarcoding samples that contain planktonic larvae from a given habitat, we could get an indication of which taxonomic group would benefit most from additional sampling effort or give an indication of the proportion of species that has not been sampled yet.
@@
 Building the library of DNA barcodes in parallel of metabarcoding efforts ensures that database matches correspond to actual specimens deposited in natural history collections, ruling out possible artifacts of the method.
-With such approaches, the geographic distribution of species can build on and linked to existing records rather than being limited to matching DNA sequences across studies.
+With such approaches, the geographic distribution of species can build on and be linked to existing records rather than being limited to matching DNA sequences across studies.
 
@@
 Even if technological advances can greatly speed up the documentation of species and their distributions, taxonomists and experts are needed  to collect, sort, process, and identify samples.
-Taxonomic expertise within the US is dwindling, and there are now only a handful of US natural history museums with active marine invertebrates collections, and curatorial staff.
+Taxonomic expertise within the US is dwindling, and there are now only a handful of US natural history museums with active marine invertebrates collections.
 
 Crowd-sourced data and citizen science initiatives have been proposed to augment data about species abundance and distributions [@Pimm2014].
-For marine taxa, beyond the application to near-shore habitats on easily recognizable species [@Delanay2008], large-scale efforts are currently not possible given the specialized equipment needed to sample most marine habitats and the expertise required to identify species.
-There however opportunities to utilize citizen science to improve biodiversity data and engage the general public into marine conservation issues.
+For marine taxa, beyond the application to near-shore habitats on easily recognizable species [@Delaney2008], large-scale efforts are currently not possible given the specialized equipment needed to sample most marine habitats and the expertise required to identify species.
+There are however opportunities for citizen science to improve biodiversity data, educate and engage the general public with marine conservation issues.
 For instance it has been successfully used to classify plankton images [@Robinson2017], transcribe hand-written museum labels [@Hill2012], or be involved in bioblitzes [@Ballard2017].

### 2017-07-07 manuscript.Rmd
more manuscript edits

@@
 Similarly, a significant fraction of museum records have missing or inaccurate geographic information associated with them.
-While tools have been developed to convert names of locales into GPS coordinates on land [@rios2010geolocate], similar approaches may not work for locales at sea.
-However, geographic information is available in other forms, and can be digitized.
+Geographic information might be available in other forms, but need to be digitized.
+While tools have been developed to convert names of locales into GPS coordinates on land [@rios2010geolocate], similar approaches need to be develop for the sea.
 
 For both the spelling of scientific names, and the geographic information, computing tools can be developed to detect issues and propose solutions.
-There are opportunities to develop computing services that use publicly available data to flag and fix these issues directly in collection databases.
-It however requires that resources are made available to museum collections to address these issues.
+There are opportunities to develop computing services that use publicly available data to flag and fix these issues directly from collection databases records.
+It however requires that resources are made available to museum collections in order to address these issues directly in collection databases.
 
-Increasing the computing literacy of the staff of museums in order for them to manipulate and use the data from their collection database is equally critical.
+Increasing the computing literacy and teaching data wrangling skills to staff of museums in order for them to manipulate and use the data from their collection database is equally critical.
 As data across institutions get pooled into the large biodiversity information systems, it is crucial that data standards such as Darwin Core are embraced by the community.
 
-Another venue to improve the quality of the data in biodiversity information systems, is better incitives from journal editors and funding agencies for scientists to publish their occurrence records in databases alongside their papers as it is done for genetic data with GenBank [@Huang2011e].
+Another venue to improve the quality of the data in biodiversity information systems, is to develop better incitives from journal editors and funding agencies for scientists to publish their occurrence records in databases alongside their papers as it is done for genetic data with GenBank [@Huang2011e].
 
-Specimens with accurate identifications and specimens that have been preserved in conditions such that it is possible to get DNA out of them should be prioritized, as they yield the most data that are relevant to biodiversity knowledge.
+<!-- Specimens with accurate identifications and specimens that have been preserved in conditions such that it is possible to get DNA out of them should be prioritized, as they yield the most data that are relevant to biodiversity knowledge.
+-->
 
@@
 This study highlights that we have too little data about too few species of marine invertebrates.
-We recommend that additional coordinated field expeditions are organized to sample more thoroughly the marine environment.
+We recommend that additional coordinated field expeditions are organized to sample more comprehensively the marine environment.
 The relation between sampling effort and diversity from museum collections makes clear that any additional samples will improve greatly our knowledge of marine biodiversity.
@@
 
-Samples collected during these expeditions need to be photographed to document live appearance that would otherwise be lost during preservation.
+Samples collected during these expeditions need to be photographed to document live appearance that would otherwise be lost during the preservation process.
 We also need to capture metadata associated with the specimens to be able to understand better the physical characteristics of the habitats for these species.

### 2017-07-07 manuscript.Rmd
moar edits

@@
 Research attempting to understand how much we know about marine biodiversity has focused on estimating the total number of species that inhabits the oceans, and which fraction of this diversity has been named [@Appeltans2012,@Costello2013].
-Undoubtely, the development of molecular methods to investigate species limits has reinvigorated taxonomy, and contributed to elucidating cryptic species complexes [@Knowlton2000].
+Undoubtedly, the development of molecular methods to investigate species limits has reinvigorated taxonomy, and contributed to elucidating cryptic species complexes [@Knowlton2000].
 Thus, it has been argued that molecular data can speed up the enterprise of discovering and naming species [@Blaxter2016].
@@
 For instance, to be able to predict how their geographic ranges will be affected by global change, we need detailed information about their environmental requirements.
-Occurrence data for marine species could be used for environmental niche modelling of their geographic and bathymetric distributions but the proportion of species for which we have enough data to do so is unknown.
+Occurrence data for marine species could be used for environmental niche modeling of their geographic and bathymetric distributions but the proportion of species for which we have enough data to do so is unknown.
 
@@
 iDigBio [@Page2015]), species occurrences (e.g., OBIS [@Zhang2002]), genetic data (e.g., GenBank [@Benson2013], BOLD [@Ratnasingham2007a]), and registers of species names (e.g., WoRMS [@Costello2013b]), now allow us to evaluate what and how much we know.
-Here, we aggregate publicly available data for marine invertebrates that occur within the limits of the exclusive economic zone (EEZ) of the continental United States (US), as well as the Gulf of Mexico and the Pacific Northwest, to highlight gaps in our knowledge that provide a roadmap for important new directions in global research.
-We show that while public databases are capturing a significant portion of our knowledge about species richenss, most species are known from only a few specimens and a few locations.
+Here, we aggregate publicly available data for marine invertebrates that occur within the limits of the exclusive economic zone (EEZ) of the continental United States (US), as well as the Gulf of Mexico and the Pacific Northwest, to highlight gaps in our knowledge that provide a road map for important new directions in global research.
+We show that while public databases are capturing a significant portion of our knowledge about species richness, most species are known from only a few specimens and a few locations.
 We emphasize that our understanding of marine biodiversity, and the impact of human activities on communities require not only to get a more accurate list of the species that inhabits the oceans, but more comprehensive data on species distributions and abundances are needed.
@@
 The continental US includes a wide diversity of marine ecosystems from the Florida Keys, to the Gulf of Maine on the East Coast; and from the arid Southern California to the oceanic shores of the Pacific Northwest on the West coast.
-With estuaries, bays, sandy beaches, rocky shores, and coral reefs, the US coastlines harbors many biotas.
+With estuaries, bays, sandy beaches, rocky shores, and coral reefs, the US coastlines harbors many biota.
 While the continental shelf extends for several hundred kilometers on the East coast, it is narrow on the West coast.
-This diversity of habitats and bathymetric profiles, combined with efforts at the national level to make data publicly available, make this region especially suitable to evaluate how much we know about marine biodiversity using biodiversity databases.
+This diversity of habitats and bathymetric profiles, combined with efforts at the national level to make data publicly available, make this region especially suitable to evaluate how much we know about marine biodiversity using biodiversity repositories.
 Additionally, apart from the syntheses that emerged from the Census of Marine Life efforts [@Costello2010,@Fautin2010], no study has yet attempted to synthesize the current state of knowledge of marine biodiversity across the continental US, and this study is the first using data from public databases.
@@
 For the Gulf of Mexico, the majority of the species included in the expert faunistic list are recorded in both iDigBio and OBIS (`r format_output(species_in_common$n_spp_gom_idigbio_obis/species_in_common$n_spp_gom*100)` %), or only OBIS (`r format_output(species_in_common$n_spp_gom_obis/species_in_common$n_spp_gom*100)` %) (Fig.
-\@ref(fig:database-coverage)).
+\@ref(fig:database-coverage)A).
 For the Pacific Northwest, these proportions were much lower with only `r format_output(species_in_common$n_spp_kozloff_idigbio_obis/species_in_common$n_spp_kozloff*100)`% of the species from the regional list with records in OBIS and iDigBio, and `r format_output(species_in_common$n_spp_kozloff_obis/species_in_common$n_spp_kozloff*100)`%  in OBIS only (Fig.
-\@ref(fig:database-coverage)).
+\@ref(fig:database-coverage)B).
 Actually, the majority of the species in the faunistic list are not recorded in any of the public databases.
@@
 \@ref(fig:database-overlap)).
-Many species of mollusks, arthropods, and annelids, known to occur in the Gulf of Mexico or in the Pacific Northwest, have not digital records for them in these geographic areas.
+Many species of mollusks, arthropods, and annelids, known to occur in the Gulf of Mexico or in the Pacific Northwest, have no digital records in these geographic areas.
 
-```{r compare_database_overlap, fig.cap='Comparison of species by phylum across the public databases and the faunistic list. (\\#fig:database-overlap)', ech=FALSE}
+```{r compare_database_overlap, fig.cap='Comparison of species by phylum across the public databases and the faunistic list. (\\#fig:database-overlap)'}
 knitr::include_graphics("figures/database_overlap.pdf")
@@
 
-Species that are included in the list but do not have records in databases could be (1) rare: they have never been collected or have been collected only a few times, and the corresponding records have not been digitized yet; (2) difficult to preserve: large, very small, or fragile species are deposited in museums less frequently; (3) are in the database under a different name: the species has been synonimized or re-identified since the publication of the list under a different name that may be included in the database; (4) at the limit of their range: the species were included in the list because they are known to occur within the boundaries of the geographic area, but they have never been collected within these geographic boundaries; (5) do not have geographic coordinates associated with them: some of these species can be in the database but without georeferenced datapoints and therefore were not returned in our queries.
+Species that are included in the list but do not have records in databases could be (1) rare: they have never been collected or have been collected only a few times, and the corresponding records have not been digitized yet; (2) difficult to preserve: large, very small, or fragile species are deposited in museums less frequently; (3) are in the database under a different name: the species has been synonimized or re-identified since the publication of the list under a different name that may be included in the database; (4) at the limit of their range: the species were included in the list because they are known to occur within the boundaries of the geographic area, but they have never been collected within these geographic boundaries; (5) do not have geographic coordinates associated with them: some of these species can be in the database but without georeferenced data points and therefore were not returned in our queries.
 
-Untangling the relative contribution of these factors to explain the discrepencies is difficult as it would require to know the criteria used by the experts who have assembled the lists.
+Untangling the relative contribution of these factors to explain the discrepancies is difficult as it would require to know the criteria used by the experts who have assembled the lists.
 However, by querying iDigBio we found that for the `r format(not_really_in_database$idig_gom_n_spp_not_in_db, big.mark=",")` species that do not appear within the geographic range of the Gulf of Mexico but are on the list, `r format_output(not_really_in_database$idig_gom_p_spp_not_in_db*100)`% are truly absent from iDigBio, while for the `r format(not_really_in_database$idig_gom_n_spp_no_coords, big.mark=",")` species that appear in iDigBio, `r format_output(not_really_in_database$idig_gom_p_spp_no_coords*100)`%  do not have geographic coordinates associated with them, and for the ones that do have geographic coordinates, `r format_output(not_really_in_database$idig_gom_p_spp_within_usa*100)`% ($n=$  `r format(not_really_in_database$idig_gom_n_spp_within_usa, big.mark=",")`) fall within the US waters.
-These numbers indicate that the lack of geographic coordinates associated with the specimens in the database contribute at least significantly to the discrepency, but that a large fraction of these species have no digital records of their occurrence in the Gulf of Mexico.
+These numbers indicate that the lack of geographic coordinates associated with the specimens in the database contribute at least significantly to the discrepancy, but that a large fraction of these species have no digital records of their occurrence in the Gulf of Mexico.
 In other words, additional digitization of collections (including capturing their geographic coordinates information) are needed.
@@
 \@ref(fig:idigbio-through-time)).
-Initial explorations in the late ninetieth century lead to an initial exponential growth in the knowledge of the fauna, that tapered off, until the 1970's when intensive collecting efforts led to an increase in the recording of new taxa (Fig.
+Initial explorations in the late ninetieth century led to an initial exponential growth in the knowledge of the fauna, that tapered off, until the 1970s when intensive collecting efforts led to an increase in the recording of new taxa (Fig.
 \@ref(fig:idigbio-through-time)).
-Since then, the slowing down of collections after the 1970s has been accompanied with a slowing down in the discovery of new records: `r n_spp_comparison$n_spp_2017_2000` species have been documented for the first time since 2000 (`r format_output(n_spp_comparison$n_spp_2017_2000/18)` species.year^-1^), while `r n_spp_comparison$n_spp_1970_1960` have been documented between 1960 and 1970 (`r format_output(n_spp_comparison$n_spp_1970_1960/11)` species.year^-1^).
+Since then, the slowing down of collections has been accompanied with a slowing down in the discovery of new records: `r n_spp_comparison$n_spp_2017_2000` species have been documented for the first time since 2000 (`r format_output(n_spp_comparison$n_spp_2017_2000/18)` species.year^-1^), while `r n_spp_comparison$n_spp_1970_1960` have been documented between 1960 and 1970 (`r format_output(n_spp_comparison$n_spp_1970_1960/11)` species.year^-1^).
 The strong correlations between (1) the number of samples collected and the species richness (Fig.
-\@ref(fig:correlation-sampling)).; (2) the number of specimens collected per year and the number of species recorded for the first time (Fig.
+\@ref(fig:correlation-sampling)), and (2) the number of specimens collected per year and the number of species recorded for the first time (Fig.
 \@ref(fig:idigbio-through-time)C), suggest that any new sampling effort will lead to new observations and new species.
 Coordinated large-scale biodiversity expeditions continue to be needed to improve our understanding of the diversity of species and their spatial distributions.
-While less explored than the US, a recent study in Western Australia reported that 135 samples led to 805 species of crustaceans and polychaetes, 94.6% and 72% of which were undescribed [@Poore2014], highlighting the potential for discovery in the oceans.
+While less explored than the US, a recent study in Western Australia reported that the collection of 135 locations led to the collection of 805 species of crustaceans and polychaetes, 94.6% and 72% of which were undescribed [@Poore2014], highlighting the potential for discovery in the oceans.
 
@@
 However, it is generally recognized that to obtain reliable results such efforts require to have accurate geographic information for a minimum of 10 specimens but having at least 30 is preferred [@xxxx].
-Based on iDigBio data, `r prop_10_specimens_species`% of species have less than 10 specimens and `r prop_30_specimens_species`% have less than 30 specimens.
+From the iDigBio data, we found that `r prop_10_specimens_species`% of species are known from less than 10 records and `r prop_30_specimens_species`% from less than 30 records.
 This suggests that our understanding of the geographic range of species is limited as most species are known from only a few locations, while efforts to estimate their current and future ranges based on occurrence data are rendered moot by the low number of data points available.
@@
 For diverse groups such as annelids or arthropods, the proportion of specimens identified at the species-level is now below 30%.
-The number of scientists describing new species might be increasing globally [@Costello2013], but taxonomic experts continue to be needed to identify specimens recently collected.
+The number of scientists describing new species might be increasing globally [@Costello2013], but taxonomic experts continue to be needed to identify specimens recently collected in the US.
 
@@
 
-Since its introduction in the early 2000's, DNA barcoding, the sequencing of a common locus across taxa bearing species-specific information, has revitalized taxonomy by facilitating species delineation [@Meyer2005-plos].
+Since its introduction in the early 2000s, DNA barcoding, the sequencing of a common locus across taxa bearing species-specific information, has revitalized taxonomy by facilitating species delineation [@Meyer2005-plos].
 More recently, advances in high-throughput sequencing technologies have enabled the parallel sequencing of thousands of species that occur in environmental samples cost effectively [@Leray2015].
@@
 The proportion of species that have DNA barcodes available is relatively constant across the better studied phyla at the scale of the EEZ.
-It varies between 12% for bryozoans to 28% for cnidarians (\@ref(fig:proportion-barcoded)).
+It varies between 12% for bryozoans to 28% for cnidarians (Fig.
+@ref(fig:proportion-barcoded)).
 The proportions for species known to occur in the Gulf of Mexico are usually similar than for the US, while typically a higher fraction of species from the Pacific Northwest are available.
@@
 It is thus crucial to ensure that accurate geographic information is provided when DNA barcodes are published, and that enough samples for each species are sequenced, to capture more accurately their geographic range and the extent of their genetic diversity.
-This cryptic diversity is also at least partially reflected in the fact that `r stat_bold$n_spp_multi_bin` species out of `r format(stat_bold$n_spp, big.mark=",")` include more than 1 BIN (BOLD Index Number, operational taxonomic units that closely resemble species [@Ratnasingham2013]).
+This cryptic diversity is also at least partially reflected in the fact that `r stat_bold$n_spp_multi_bin` species out of `r format(stat_bold$n_spp, big.mark=",")` in our dataset include more than 1 BIN (BOLD Index Number, operational taxonomic units that closely resemble species [@Ratnasingham2013]).
 
@@
 Similarly, a significant fraction of museum records have missing or inaccurate geographic information associated with them.
-Geographic information might be available in other forms, but need to be digitized.
-While tools have been developed to convert names of locales into GPS coordinates on land [@rios2010geolocate], similar approaches need to be develop for the sea.
+Geographic information might be available in other forms, but needs to be digitized.
+While tools have been developed to convert names of locales into GPS coordinates on land [@rios2010geolocate], similar approaches are lacking for the sea.
 
 For both the spelling of scientific names, and the geographic information, computing tools can be developed to detect issues and propose solutions.
-There are opportunities to develop computing services that use publicly available data to flag and fix these issues directly from collection databases records.
+There are opportunities to develop computing services that use publicly available data to flag and fix these issues directly from collection databases records (e.g.
+see [@Otegui2016] for vertebrate records).
 It however requires that resources are made available to museum collections in order to address these issues directly in collection databases.
@@
 Increasing the computing literacy and teaching data wrangling skills to staff of museums in order for them to manipulate and use the data from their collection database is equally critical.
-As data across institutions get pooled into the large biodiversity information systems, it is crucial that data standards such as Darwin Core are embraced by the community.
+As data across institutions get incorporated and pooled into large biodiversity information systems, it is crucial that data standards such as Darwin Core are embraced by the community.
 
-Another venue to improve the quality of the data in biodiversity information systems, is to develop better incitives from journal editors and funding agencies for scientists to publish their occurrence records in databases alongside their papers as it is done for genetic data with GenBank [@Huang2011e].
+Another venue to improve the quality of the data in biodiversity information systems, is to develop better incentives from journal editors and funding agencies for scientists to publish their occurrence records in databases alongside their papers as it is done for genetic data with GenBank [@Huang2011e].
 
@@
 Samples collected during these expeditions need to be photographed to document live appearance that would otherwise be lost during the preservation process.
-We also need to capture metadata associated with the specimens to be able to understand better the physical characteristics of the habitats for these species.
+We also need to capture metadata associated with specimens to be able to understand better the physical characteristics of the habitats and the life history for these species.
 Specimens need to be sub-sampled to create material suitable for genomic analysis.
@@
 Assembling species lists is a necessary but not sufficient step in our quest to understand biodiversity.
-We also need to understand better the geographic distribution of biodiversity, how the abundance of species vary in space and time, and their role in the functionning of ecosystems.
-In order to predict the impact of climate change and other anthropogenic perturbations on species distributions, we need to collect additional occurences for species as well as the characteristics of their habitats.
-Among the challenges for this enterprise, we need to be able to convey taxonomic concepts for species which, for the most part, have not formally been named yet.
+We also need to understand better the geographic distribution of biodiversity, how the abundance of species vary in space and time, and their role in the functioning of ecosystems.
+In order to predict the impact of climate change and other anthropogenic perturbations on species distributions, we need to collect additional occurrences for species as well as the characteristics of their habitats.
+Among the challenges of this enterprise, we need to be able to convey taxonomic concepts for species which, for the most part, have not formally been named yet.
 Developing robust identifiers that can be shared across biodiversity information systems, and anchored in results from DNA barcoding or other molecular techniques, is needed to collaboratively build our knowledge of species distributions.
@@
 Metabarcoding also offers to be a promising technology to monitor changes in communities and for understanding the functioning of ecosystems.
-For instance, despite our limited knowledge of the diversity of communities in the meiofauna, metabarcoding has been successfully used to detect and documents changes in species assemblages following the Deepwater Horizon Oil Spill [@Bik2012].
-Metabarcoding has also revealed important differences in the diet of 3 related species of fishes that co-occur at small spatial scales, providing a much more detailed understanding of trophic relations [@Leray2015b].
+For instance, despite our limited knowledge of the diversity of communities in the meiofauna, metabarcoding has been successfully used to detect and documents changes in species assemblages following the Deepwater Horizon oil spill [@Bik2012].
+Metabarcoding has also revealed important differences in the diet of three related species of fishes that co-occur at small spatial scales, providing a much more detailed understanding of trophic relations [@Leray2015b].
 
-Considering additional linked data with species such as DNA has been a first step in a digital biodiversity world where other data on species can be gathered in databases and analyzed to better understand species distributions.
-This has paved the way to more linked data/databases.
-Still issues though with connecting identifiers [@Guralnick2014].
+## Conclusion
 
+Given the urgency of the biodiversity crisis, we need to develop data-driven initiatives to efficiently document biodiversity, and build on our existing knowledge to collect information to improve our understanding of marine systems.
+
+DNA sequences have been one of the first test case on how to link biodiversity data across repositories.
+These initial efforts have revealed the lack of infrastructure to use data formats designed for linked data, the lack of ontologies, and the fragility of the identifiers used [@Stucky2014; @Walls2014; @Guralnick2014].
+As new types of data will be added and linked to existing repositories, the community needs to develop modern and robust ways to integrate heterogeneous biodiversity data sources [@Guralnick2015; @Deck2015].
+
+Digitization efforts are catching up on already fragmented and incomplete records.
 Even if technological advances can greatly speed up the documentation of species and their distributions, taxonomists and experts are needed  to collect, sort, process, and identify samples.
@@
 For marine taxa, beyond the application to near-shore habitats on easily recognizable species [@Delaney2008], large-scale efforts are currently not possible given the specialized equipment needed to sample most marine habitats and the expertise required to identify species.
-There are however opportunities for citizen science to improve biodiversity data, educate and engage the general public with marine conservation issues.
-For instance it has been successfully used to classify plankton images [@Robinson2017], transcribe hand-written museum labels [@Hill2012], or be involved in bioblitzes [@Ballard2017].
-These initiatives however involve a lot less people compared to projects related to birds.
+There are however opportunities for citizen science to help curate biodiversity data, while educating and engaging the public with marine conservation issues.
+For instance it has been successfully used to classify plankton images [@Robinson2017], or transcribe hand-written museum labels [@Hill2012].
 
-Given the urgency of the biodiversity crisis, we need to develop data-driven initiatives to efficiently document biodiversity, and build on our existing knowledge to collect information to improve our understanding of marine systems.
-Digitization efforts are catching up on an already fragmented and vastly incomplete records.
-Here, we only reported results from the US where there is at least two centuries of biodiversity exploration data and high-level of digitization.
-Many other parts of the world also need additional attention, especially in some developing countries where environmental pressure on ecosystems is even higher than in the US.
-
-<!-- Biodiversity discovery and documentation has been transformed in the 21st century by molecular data to facilitate species identification, delineate species.
-And more recently, the development of high throughput sequencing technologies is used to quantify diversity and uncover large scale patterns of biodiversity organization.
-The wide adoption of DNA barcoding (where homologous DNA sequences are sequenced across the main domains of the Eukaryotes) has unraveled the pervasiveness of cryptic species, invigorating taxonomy and the description of new species, and enabled collaborative research programs such as the Census of Marine Life.
-Unraveling these cryptic species has also contributed to the recognition that species have smaller ranges than their taxonomic identity had implied making them potentially more susceptible to extinction or overexploitation.
-Despite early criticisms and now well-understood limitations, DNA barcoding has been suggested as a possible solution to the widening gap between the increased need to document biodiversity while taxonomic expertise is dwindling.
-Thus, large databases of DNA barcodes have been collaboratively assembled and made pubicly available.
-By matching DNA barcodes of unidentified species to a reference public database of accurately identified and vouchered specimens, DNA barcoding can greatly speed up species identification: sequences with no match make it easier to flag specimens that warrant further taxonomic attention.
--->
-
-<!-- clarified our understanding of species limits in many groups, but has also enabled the identification of species during their larval stages (when the lack of morphological features make species identification difficult).
-DNA barcoding has also been seen as -->
-
-<!-- Barcoding/metabarcoding certainly not perfect methods, but much better than current understanding. -->
+Here, we reported results from the US where there is at least two centuries of biodiversity exploration data and high-levels of digitization of museum collections.
+Many other parts of the world need additional attention, especially in some developing countries where environmental pressure on ecosystems is even higher than in the US.
 
@@
 
-<!-- here there are 3 things: data availability (having good APIs to access the
+<!-- here there are 4 things: data availability (having good APIs to access the
 data); data comprehensiveness (how many species are captured in the database);
-data accuracy (can we trust the data that is found in the databases) -->
+data accuracy (can we trust the data that is found in the databases); data linkage -->
 
@@
 For the Pacific Northwest, Kozloff did not specify the exact geographic extent covered by his book.
-We infered a polygon that encompass part of the Oregon coast to the northern parts of Vancouver island.
+We inferred a polygon that encompass part of the Oregon coast to the northern parts of Vancouver island.
 For the Gulf of Mexico, we used the same geographic area as the one defined in [@Tunnell2009] by modifying manually the eastern boundaries of the shape file for the Gulf of Mexico provided by marineregions.org (mrgid:4288).
 These geographic areas were split into squares of 0.2°x0.2° which were used to query the web API of iDigBio and OBIS using the R packages ridigbio and robis respectively.
-The resulting occurences were then clipped to only include records that fell within the EEZ boundaries.
+The resulting occurrences were then clipped to only include records that fell within the EEZ boundaries.
 
@@
 
-## Supplementary material
-
-<!--
-
-It has never been easier and cost effective to survey comprehensively the extant of marine biodiversity.
-
-While plankton is recognized for its biological importance (dispersal, food source for many species), and can be used as an ecological indicator for pollution or global change, identification of planktonic larvae at the species level from morphological information is generally not possible because of the paucity of informative characters to base the identifications on, and the lack of knowledge of the appearance of the larval stages for most species.
-DNA barcoding can bridge this knowledge gap by using genetic information to identify larvae at the species level.
-These identifications are only possible if the species have already been sequenced, and the sequences deposited in public repositories.
-The accuracy of the identification relies on the expertise of the taxonomists who identified the specimens that have been sequenced.
-
-We collected over the course of a year planktonic larvae of marine invertebrates in a coastal marsh of the Atlantic coast of Florida.
-One of our goals was to identify these larvae at the species level using DNA barcoding.
-We photographed and sequenced xxx larvae, representing an estimated xxx species.
-These data are available from BOLD and can be browsed interactively at http://plankton-atlas.net.
-Using records from public databases (BOLD, GenBank), we could confidently assign only xxx% to species that had already been sequenced.
-On average, these taxa were sequenced xxx km away from our study site.
-
-Florida is one of the most populated states, with a long tradition of studying its marine life.
-xxx universities have marine research programs, and xxx marine laboratories.
-How does this proportion of xxx% identified species compare with other places in Florida, the East coast of the US, the continental US, and the world.
-In other words, how much of the species of marine invertebrates we know about have publicly available DNA barcodes?
-
-To estimate this number, we assembled species lists of marine invertebrates for the continental US by querying iDigBio for each marine phylum.
-We only kept records for specimens identified with a valid binomial name (specimens identified at the genus level or higher were excluded from the analysis).
-We used WoRMS to ensure that the species names were valid, and to disambiguate possible synonyms.
-For each species, we retrieved avaiable DNA barcoding sequence data from BOLD, which also includes DNA barcodes deposited in GenBank.
-Of the xxx marine invertebrate species found in iDigBio, xxx% had DNA barcodes available in public databases.
-These proportions varied by phylum, but for the conspicuous and better studied groups (Porifera, Cnidaria, Mollusca, Echinodermata), they were surprisingly consistent, and is a close estimate to the proportion of species that have been sequenced for the United States.
-Furthermore, we also looked for the proportion of species known to occur in the Indian River Lagoon for which DNA barcodes are available, and again the proportion of sequenced species (xxx%) is comparable.
-
-Marine invertebrates are the biodiversity.
-All but one metazoan phyla occur in the sea.
-Of the xx described species, yy are marine invertebrates.
-They represent a multiple array of shapes, forms, and adaptations to all marine environments.
-Yet, they are proportionally understudied.
-Here, we analyze publicly available data from biodiversity databases to demonstrate that the abandonment of traditional taxonomic practices, the lack of support for natural history collections, is hampering our ability to use marine invertebrates as indicators of ecosystem health, the effect of global change, the identification of model organisms, and ...
-Technological advances are however making easier to identify knowledge gaps, the potential to identify where new species occur, and document diversity.
-
-# The role of DNA barcoding
-
-The use of molecular data for species identification and delineation is arguably one of the most important advances in taxonomy.
-Especially with DNA barcoding that proposes to use a common genetic marker across large taxonomic groups (the mt marker COI for animals, ITS for fungi, and xx for plants).
-Because these markers have typically low intra-specific variation, and differ noticeably between even closely related species, they can be used to identify species and unravel species complexes.
-
-Species identification relies on reference databases, where sequences are associated with accurately identified individuals, held in natural history collections.
-When taxonomic sampling is comprehensive, and enough individuals of each species are sampled to be representative of the standing genetic variation in the population, DNA barcoding can be a very accurate tool with xx-yy% error rate.
-
-The utility of these reference databases is getting revived with the development of meta-barcoding techniques, where thousands of species can be sequenced in parallel on high-throughput sequencing platforms.
-If methods exist to estimate the number of species these samples contain based on the level of genetic divergence observed among the sequences, being able to match these sequences to identified species based on their match to a sequence in a public database would increase the potential applications for these approaches.
-
-To illustrate the consequences of having so few barcodes for invertebrates, if we take the faunistic list of IRL, only about 20% of the species known to be occuring in the area actually have DNA sequences available for them.
-On average, each species has been sequenced xx times, and from locations on average yy km away from IRL.
-Given the high propencity for cryptic species among marine invertebrates, it is possible that ...
-
-# The waining of taxonomy
-
-We don't know how many species are in the sea.
-Estimates vary widely.
-Difficult to sample because diversity of habitats, each requiring special sampling techniques.
-Meanwhile, taxonomic expertise is decreasing despite efforts to keep it alive (PEET).
-
-To get around issue of sampling different habitat, sample coastal meroplankton.
-Most species go through planktonic phase, therefore common habitat to sample them all.
-The issue is that they look very different from adults, and only DNA can be used to identify them.
-
-Barcoding is not perfect, but can be very powerful to identify the larval forms (e.g.
-one of the first paper was sampling caterpillars).
-Potentially differences in larva that are not seen in adults and could be additional characters to tell species appart.
-
-Advantage of sampling plankton is that a single habitat is sampled. Not enough known about heterogeneity in spatial distribution.
-
-We sampled plankton bi-weekly for a year, and sequenced the larvae found.
-Sequenced the barcoding gene and tried to match how many species we could identify based on records in public databases.
-Overall, low percentage of match.
-And matches also led to several species that have been collected but that have not been identified to an appropriate level.
-
-To compare the results we obtained here, and see how they could compare to other habitats, we downloaded the species list from Indian River lagoon inventory.
-It is the most comprehensive checklist in Florida, and has been developed over many years.
-We asked how many of the species occurring in the Indian River Lagoon have already been sequenced.
-We did the matching based on the species names.
-If we find higher proportion of sequenced species, it means that plankton gives access to species that are difficult to sample as adults.
-If lower proportion, it means that we don't have many of the adults sequenced at Whitney.
-
-Using plankton to sample biodiversity could lead to a speedup of documentation.
-It would allow the identification of taxa/habitats that need special attention.
-If representative of adult communities, could also be a useful tool to monitor status of ecosystem through time.
-
-Better database, can also be used to estimate the proportion of really dark planktonic taxa.
-
-Understand better population connectivity, currents
-
-Importance of plankton in connecting populations. Important as food source.
-
-Swimming in the dark in the age of big data
-
-Biodiversity is in the sea. All phyla but one. New lineages of eukaryotes, viruses...
-
-We are at a time where it becomes cheaper and easier to delineate species. To sort through specimens to identify species. So taxonomists can focus on studying/describing new species.
-
-Granularity of data not sufficient to be able to detect changes in the environment, shifts in species ranges, changes in trophic structure. Understand effect of small animals.
-
-Rely on high school students, citizen science initiatives to help with labor intensive part.
-
-Oceans are fascinating.
-
-Bioprospection
-
-focus on a few taxa but lots of groups we know nothing about.
-
-Strong focus on vertebrates. We are building theories and making conservation approaches on a small fraction of the tree of life.
-
-Trait data
-
-Species Distribution
-
-Identified specimens from museum collections could be used to complete these datasets.
-Traditional approaches relying on Sanger sequencing might be limiting as many species have not been collected recently enough to guarantee good quality DNA (` prop_species_not_collected_since_1990`% have not been collected since 1990).
-However, high-throughput sequencing could offer alternatives to recover sequences from older specimens.
-
-Besides data-mining museum records to identify knowledge gaps, new sampling approaches can be used in the field to rapidly assess how much of the fauna is known.
-Meta-barcoding and environmental DNA approaches appear promising to sample _en masse_ species that are found in a given location.
-Because most marine invertebrates have a planktonic larval stage, they may co-occur in the water column, allowing to sample species across habitats.
-To illustrate this approach, we individually DNA barcoded ` n_larvae` planktonic larvae for DNA barcoding, collected over the course of year, representing ` n_esus` species, collected in coastal waters of Florida.
-We found that (` format_output(100 * prop_plankton_with_match)`%, Fig.
-4) can be assigned to a previously sequenced species.
-Similar proportions have been obtained from individually barcoded species that occur on ARMS.
-However, this proportion dropped to less than 5% for smaller organisms sequenced with meta-barcoding.
-
-The species identifications in BOLD are far from perfect.
-Of the ` sum(prop_plankton_with_species_match_by_phylum$n_esu_match)` species that have a match in public database, ` format_output(prop_plankton_with_multi_species_match * 100)`% have been identified as more than one taxon in BOLD.
-A few of these records might represent species complexes, but most illustrate the difficulty to identify precisely marine invertebrates.
-
---->
+<!-- ## Supplementary material -->

### 2017-07-10 manuscript.Rmd
edit mansucript, fix refs

@@
 
-Predicting the geographic range of species, and how they will be impacted by global change has undergone important advances recently [@xxxx].
-However, it is generally recognized that to obtain reliable results such efforts require to have accurate geographic information for a minimum of 10 specimens but having at least 30 is preferred [@xxxx].
+Predicting the geographic range of species, and how they will be impacted by global change has undergone important advances recently [@Guillera-Arroita2017], but remain scarecely used in the marine realm [@Robinson2011].
+One limitation is the low number of observations available for each species: it is generally recognized that to obtain reliable results such efforts require to have accurate geographic information for a minimum of 10 specimens but having at least 30 is preferred [@Wisz2008].
 From the iDigBio data, we found that `r prop_10_specimens_species`% of species are known from less than 10 records and `r prop_30_specimens_species`% from less than 30 records.
@@
 
-Given the urgency of the biodiversity crisis, we need to develop data-driven initiatives to efficiently document biodiversity, and build on our existing knowledge to collect information to improve our understanding of marine systems.
+Given the urgency of the biodiversity crisis, we need to develop data-driven initiatives to document biodiversity more efficiently, and build on our existing knowledge to collect data that will improve our understanding of marine systems.
 
 DNA sequences have been one of the first test case on how to link biodiversity data across repositories.
-These initial efforts have revealed the lack of infrastructure to use data formats designed for linked data, the lack of ontologies, and the fragility of the identifiers used [@Stucky2014; @Walls2014; @Guralnick2014].
-As new types of data will be added and linked to existing repositories, the community needs to develop modern and robust ways to integrate heterogeneous biodiversity data sources [@Guralnick2015; @Deck2015].
+These initial efforts have revealed the lack of infrastructure to use data formats specifically designed for this task, the lack of ontologies available for biodiversity data, and the fragility of the identifiers [@Stucky2014; @Walls2014; @Guralnick2014].
+As new types of data will be collected, added, and linked across repositories, the community needs to develop modern and robust ways to integrate these heterogeneous biodiversity data sources [@Guralnick2015; @Deck2015].
 
@@
 Taxonomic expertise within the US is dwindling, and there are now only a handful of US natural history museums with active marine invertebrates collections.
+This lack of resources is not sustainable in the face of the challenges and discoveries that marine ecosystems have to offer.
 
 Crowd-sourced data and citizen science initiatives have been proposed to augment data about species abundance and distributions [@Pimm2014].
-For marine taxa, beyond the application to near-shore habitats on easily recognizable species [@Delaney2008], large-scale efforts are currently not possible given the specialized equipment needed to sample most marine habitats and the expertise required to identify species.
+Beyond the application to near-shore habitats on easily recognizable species [@Delaney2008], large-scale efforts are currently not possible for marine ecosystems given the specialized equipment needed to sample most habitats and the expertise required to identify species.
 There are however opportunities for citizen science to help curate biodiversity data, while educating and engaging the public with marine conservation issues.
-For instance it has been successfully used to classify plankton images [@Robinson2017], or transcribe hand-written museum labels [@Hill2012].
+For instance citizen science has been used successfully to classify plankton images [@Robinson2017], or transcribe hand-written museum labels [@Hill2012].
 

### 2017-07-11 R/classification.R
make classification code standalone

@@ -1,0 +1,33 @@
+get_classification_from_wid <- function(wid, verbose = FALSE) {
+    lapply(wid, function(i) {
+        if (verbose) message("getting classification for aphiaid ", i)
+        res <- store_worms_classification()$get(as.character(i))
+        res <- res[[as.character(i)]]
+        res %>% mutate_if(is.factor, as.character)
+    })
+}
+
+unfold_classification <- function(classif, rank) {
+    empty_classif <- data_frame(Phylum = character(0),
+                                Class = character(0),
+                                Order = character(0),
+                                Family = character(0))
+    cols_keep <- c("Phylum", "Class", "Order", "Family")
+    to_keep <- intersect(classif$rank, cols_keep)
+    dplyr::distinct(classif, rank, .keep_all = TRUE) %>%
+        tidyr::spread(rank, name) %>%
+            dplyr::select_(.dots = to_keep) %>%
+            dplyr::bind_rows(empty_classif)
+}
+
+
+add_classification <- function(data) {
+    stopifnot(exists("worms_id", data))
+    data %>%
+        dplyr::filter(!is.na(worms_id)) %>%
+        dplyr::mutate(classification = get_classification_from_wid(worms_id, verbose = FALSE),
+                      classification_df = purrr::map(classification, unfold_classification)) %>%
+        tidyr::unnest(classification_df) %>%
+        dplyr::select(-classification) %>%
+        dplyr::mutate(taxon_name = Phylum, rank = "Phylum")
+}

### 2017-07-11 R/comparison_by_geo.R

@@ -1,26 +1,4 @@
 compare_with_geo <- function(spp_list, geo_list, verbose = FALSE) {
-    get_classification <- function(x) {
-        lapply(x, function(i) {
-            if (verbose) message("getting classification for aphiaid ", i)
-            res <- store_worms_classification()$get(as.character(i))
-            res <- res[[as.character(i)]]
-            res %>% mutate_if(is.factor, as.character)
-        })
-    }
-
-    unfold_classification <- function(classif, rank) {
-        empty_classif <- data_frame(Phylum = character(0),
-                                    Class = character(0),
-                                    Order = character(0),
-                                    Family = character(0))
-        cols_keep <- c("Phylum", "Class", "Order", "Family")
-        to_keep <- intersect(classif$rank, cols_keep)
-        dplyr::distinct(classif, rank, .keep_all = TRUE) %>%
-            tidyr::spread(rank, name) %>%
-            dplyr::select_(.dots = to_keep) %>%
-            dplyr::bind_rows(empty_classif)
-    }
-
     list(
         ## in spp_list but not in geo_list
         not_in_list =
@@ -30,7 +8,7 @@ compare_with_geo <- function(spp_list, geo_list, verbose = FALSE) {
             dplyr::distinct(worms_valid_name, .keep_all = TRUE) %>%
             dplyr::select(worms_id, worms_valid_name,
                           phylum = taxon_name) %>%
-            dplyr::mutate(classification = get_classification(worms_id),
+            dplyr::mutate(classification = get_classification_from_wid(worms_id, verbose),
                           classification_df = map(classification, unfold_classification)) %>%
             tidyr::unnest(classification_df)
        ,
@@ -42,8 +20,8 @@ compare_with_geo <- function(spp_list, geo_list, verbose = FALSE) {
             dplyr::distinct(worms_valid_name, .keep_all = TRUE) %>%
             dplyr::select(worms_id, worms_valid_name,
                           phylum = taxon_name) %>%
-            dplyr::mutate(classification = get_classification(worms_id),
-                          classification_df = map(classification, unfold_classification)) %>%
+            dplyr::mutate(classification = get_classification_from_wid(worms_id, verbose),
+                          classification_df = purrr::map(classification, unfold_classification)) %>%
             tidyr::unnest(classification_df)
 
     ) %>% dplyr::bind_rows(.id = "data_source")

### 2017-07-11 manuscript.Rmd
[fix #34] use online version of kozloff (but keep GP's edits)

@@
 We assessed how faunistic lists assembled over decades by specialists compare with records in public databases for the same geographic area.
-We parsed faunistic lists to extract marine invertebrates from the Gulf of Mexico [@Tunnell2009] and from the Pacific Northwest [@Kozloff1974], leading to `r format(sum(gom_worms$is_marine, na.rm=TRUE), big.mark=",")` and `r format(sum(kozloff_worms$is_marine, na.rm=TRUE), big.mark=",")` species respectively.
+We parsed faunistic lists to extract marine invertebrates from the Gulf of Mexico [@Tunnell2009] and from the Pacific Northwest [@Kozloff1974], leading to `r format(sum(gom_worms$is_marine, na.rm=TRUE), big.mark=",")` and `r format(sum(kozloff_species$is_marine, na.rm=TRUE), big.mark=",")` species respectively.
 Our goals were (1) to compare  the species lists for these geographic areas assembled by experts with the records found in databases; (2) to assess the proportion of species included in the expert lists that have records in public databases.

### 2017-07-11 R/kozloff.R

@@ -1,19 +1,10 @@
 get_kozloff_species <- function(koz_raw) {
-    readxl::read_excel(koz_raw) %>%
-        dplyr::select(
-            counter = Counter,
-            phylum = Phylum,
-            class = Class,
-            order = Order,
-            scientific_name_verbatim = ScientificName) %>%
-        dplyr::mutate(rank = rep("phylum", nrow(.)),
-                      taxon_name = phylum,
-                      cleaned_scientificname = cleanup_species_names(scientific_name_verbatim,
-                                                                     rm_subgenus = TRUE),
+    readr::read_csv(koz_raw) %>%
+        dplyr::mutate(cleaned_scientificname = cleanup_species_names(scientificname_verbatim, rm_subgenus = TRUE),
                       is_binomial = is_binomial(cleaned_scientificname)) %>%
-        dplyr::filter(is_binomial == TRUE)
+            dplyr::distinct(cleaned_scientificname, .keep_all = TRUE) %>%
+            dplyr::filter(is_binomial == TRUE)
 }
-
 
 
 

### 2017-07-20 manuscript.Rmd
complete using prop. to display overlap between databases and lists

@@
 
-```{r compare_database_overlap, fig.cap='Comparison of species by phylum across the public databases and the faunistic list. (\\#fig:database-overlap)'}
-knitr::include_graphics("figures/database_overlap.pdf")
+```{r compare_database_overlap, fig.cap='Proportions of the total number of species by phylum that are absent from faunistic lists but are in iDigBio or OBIS ("Not in list"), and absent from iDigBio or OBIS but are in the faunistic lists ("Not in databases").
+The numbers represent the number of species.
+(\\#fig:database-overlap)'}
+knitr::include_graphics("figures/database_overlap_percent.pdf")
 ```

### 2017-07-20 R/comparison_by_geo.R

@@ -211,13 +211,27 @@ plot_database_overlap_percent <- function(comp_db, full_list) {
     ## to have 2 sided plot, we take the negative value of the count
     ## for the records that are NOT in the geographical range in the
     ## database.
-    data_for_plot <- comp_db %>%
+    data_for_plot_raw <- comp_db %>%
         dplyr::count(region, database, data_source, phylum) %>%
         dplyr::left_join(spp_per_phylum, by = c("region", "phylum")) %>%
         dplyr::mutate(p = n/n_spp) %>%
-        dplyr::mutate(number = ifelse(data_source == "not_in_geo", -p, p)) %>%
+        dplyr::mutate(number = ifelse(data_source == "not_in_geo", -p, p))
+
+    prop_for_plot <- data_for_plot_raw %>%
         dplyr::select(-p, -n, -n_spp) %>%
-        tidyr::spread(data_source, number) %>%
+        tidyr::spread(data_source, number)
+
+    nspp_for_plot <- data_for_plot_raw %>%
+        dplyr::select(-p, -number, -n_spp) %>%
+        tidyr::spread(data_source, n) %>%
+        dplyr::rename(
+                   "nspp_not_in_geo"  = "not_in_geo",
+                   "nspp_not_in_list" = "not_in_list"
+               )
+
+    data_for_plot <- dplyr::left_join(prop_for_plot,
+                                      nspp_for_plot,
+                                      by = c("region", "database", "phylum")) %>%
         dplyr::mutate(phylum = capitalize(phylum))
 
 
@@ -228,10 +242,14 @@ plot_database_overlap_percent <- function(comp_db, full_list) {
         distinct(phylum) %>%
         .[[1]]
 
-    data_for_plot$phylum <- factor(data_for_plot$phylum, levels = rev(phylum_label))
+    data_for_plot$phylum <- factor(data_for_plot$phylum,
+                                   ## levels = rev(phylum_label))
+                                   levels = rev(c("Arthropoda", "Mollusca", "Annelida",
+                                              "Cnidaria", "Echinodermata", "Chordata",
+                                              "Porifera", "Bryozoa")))
 
     offset <- .5
-    dodge_width <- .6
+    dodge_width <- .7
     gg <- ggplot(data_for_plot, aes(x = phylum, color = database)) +
         geom_linerange(aes(x = phylum, ymin = -offset, ymax = -offset + not_in_geo),
                        position = position_dodge(width = dodge_width)) +
[…]

### 2017-07-20 R/idigbio.R

@@ -18,6 +18,7 @@ idigbio_fields <- function() {
 internal_fetch_idigbio <- function(spp, split_size) {
     assertthat::assert_that(assertthat::is.number(split_size))
     assertthat::assert_that(is.character(spp))
+    assertthat::assert_that(length(spp) > 0)
 
     fields <- idigbio_fields()
     split_species <- split_by_n(na.omit(spp), split_size)

### 2017-07-20 R/time_plot.R
add sampling effort plot

@@ -102,6 +102,11 @@ make_knowledge_through_time_idigbio <- function(idigbio) {
         ) %>%
         dplyr::mutate(cum_n_samples = cumsum(n_samples))
 
+    n_spp <- idigbio %>%
+        dplyr::distinct(phylum, worms_valid_name, year) %>%
+        dplyr::count(phylum, year) %>%
+        dplyr::rename(n_spp = n)
+
     n_species <- idigbio %>%
         dplyr::group_by(phylum, worms_valid_name, year) %>%
         dplyr::arrange(year) %>%
@@ -119,8 +124,8 @@ make_knowledge_through_time_idigbio <- function(idigbio) {
         dplyr::mutate(n_new_spp = replace(n_new_spp, is.na(n_new_spp), 0)) %>%
         dplyr::mutate(cum_n_new_spp = cumsum(n_new_spp)) %>%
         dplyr::left_join(spp_total, by = "phylum") %>%
-        dplyr::mutate(cum_p_new_spp = cum_n_new_spp/n_spp_total) %>%
-        dplyr::rename(phylum = phylum)
+        dplyr::left_join(n_spp, by = c("phylum", "year")) %>%
+        dplyr::mutate(cum_p_new_spp = cum_n_new_spp/n_spp_total)
 
     n_species
 

### 2017-07-20 manuscript.Rmd
more changes

@@
 For marine organisms, because of the diversity of their complex life history strategies, their size range, and the difficulties associated with sampling their habitats, we have limited information about which species live where, or what portion of their diversity remains to be discovered and described [@Mora2011].
-Additionally, coordinated marine ecosystem monitoring initiatives are lacking [@Duffy2013,@Borja2016], making it difficult to monitor changes in species distributions and abundances.
+Additionally, coordinated marine ecosystem monitoring initiatives are lacking [@Duffy2013,@Borja2016] (but see MarineGEO and PISCO), making it difficult to monitor changes in species distributions and abundances.
 While most documented global extinctions of marine species are vertebrates, many others species may have been lost unnoticed because they were not known [@Sala2006].
@@
 natural history collections, publications, expert knowledge), difficult to access, and therefore difficult to synthesize.
-The development of public databases for natural history records (e.g.
+The development of public databases for specimen records (e.g.
 iDigBio [@Page2015]), species occurrences (e.g., OBIS [@Zhang2002]), genetic data (e.g., GenBank [@Benson2013], BOLD [@Ratnasingham2007a]), and registers of species names (e.g., WoRMS [@Costello2013b]), now allow us to evaluate what and how much we know.
-Here, we aggregate publicly available data for marine invertebrates that occur within the limits of the exclusive economic zone (EEZ) of the continental United States (US), as well as the Gulf of Mexico and the Pacific Northwest, to highlight gaps in our knowledge that provide a road map for important new directions in global research.
+Here, we aggregate publicly available data for marine invertebrates that occur within the limits of the exclusive economic zone (EEZ) of the continental United States (US), as well as the Gulf of Mexico (GOM) and the Pacific Northwest (PNW), to highlight gaps in our knowledge that provide a road map for important new directions in global research.
 We show that while public databases are capturing a significant portion of our knowledge about species richness, most species are known from only a few specimens and a few locations.
@@
 For the Gulf of Mexico, we retrieved  `r format(dplyr::n_distinct(idigbio_gom_records$worms_valid_name), big.mark=",")` species in iDigBio (from $n =$  `r format(nrow(idigbio_gom_records), big.mark=",")` records), and `r format(dplyr::n_distinct(obis_gom_records$worms_valid_name), big.mark=",")` species in OBIS ($n=$ `r format(nrow(obis_gom_records), big.mark=",")` records).
-For the Pacific Northwest, we retrieved `r format(dplyr::n_distinct(idigbio_pnw_records$worms_valid_name), big.mark=",")` species in iDigBio ($n=$ and `r format(nrow(idigbio_pnw_records), big.mark=",")`) and `r format(dplyr::n_distinct(obis_kozloff_records$worms_valid_name), big.mark=",")` species in OBIS ($n=$ `r format(nrow(obis_kozloff_records), big.mark=",")`).
+For the Pacific Northwest, we retrieved `r format(dplyr::n_distinct(idigbio_pnw_records$worms_valid_name), big.mark=",")` species in iDigBio ($n=$ and `r format(nrow(idigbio_pnw_records), big.mark=",")`) and `r format(dplyr::n_distinct(obis_pnw_records$worms_valid_name), big.mark=",")` species in OBIS ($n=$ `r format(nrow(obis_pnw_records), big.mark=",")`).
 Overall, the number of species known from the Gulf of Mexico match the number estimated by experts (`r format(11150-1541-434, big.mark=",")` species [@Fautin2010]).
@@
 ```{r idigbio-through-time, fig.cap='Cumulative number of samples (A) and species (B) of marine invertebrates recorded through time based on iDigBio records.
-Relation between number of samples collected, and number of species recorded for the first time (C), each point represents numbers for a given year.
+Relation between number of samples collected, and number of species recorded for the first time (C), each point represents numbers for a given year, dashed lines show the 1:1 and 1:100 ratios for number of species recorded for the first time and number of samples.
 (\\#fig:idigbio-through-time)'}

### 2017-07-20 R/time_plot.R

@@ -145,7 +145,7 @@ calc_n_spp_comparison <- function(idig_time) {
 
 }
 
-filter_phyla <- function(ktt, n_min = 100) {
+filter_phyla <- function(ktt, n_min = 200) {
     ktt %>%
         group_by(phylum) %>%
         filter(cum_n_new_spp ==  max(cum_n_new_spp, na.rm = TRUE)) %>%
@@ -218,6 +218,25 @@ plot_samples_vs_spp_through_time <- function(knowledge_through_time) {
     res
 }
 
+
+plot_samples_vs_spp_std <- function(knowledge_through_time) {
+
+    phy_to_keep <- filter_phyla(knowledge_through_time)
+
+    knowledge_through_time %>%
+        dplyr::ungroup() %>%
+        dplyr::filter(phylum %in% phy_to_keep) %>%
+        dplyr::mutate(prop_new_spp = n_new_spp/n_samples,
+                      mean_new_prop = roll_meanr(prop_new_spp, n = 5),
+                      prop_spp = roll_meanr(n_spp/n_samples, n = 5)
+                      ) %>%
+        ggplot(aes(x = year, y = prop_spp, colour = phylum)) +
+        geom_point() +
+        facet_wrap(~phylum) +
+        scale_colour_hc(name = "")  +
+        geom_smooth(se = FALSE)
+
+}
 
 plot_cum_frac_through_time <- function(knowledge_through_time) {
 

### 2017-07-25 manuscript.Rmd
incorporate GP's edits

@@
 For the Gulf of Mexico, we retrieved  `r format(dplyr::n_distinct(idigbio_gom_records$worms_valid_name), big.mark=",")` species in iDigBio (from $n =$  `r format(nrow(idigbio_gom_records), big.mark=",")` records), and `r format(dplyr::n_distinct(obis_gom_records$worms_valid_name), big.mark=",")` species in OBIS ($n=$ `r format(nrow(obis_gom_records), big.mark=",")` records).
-For the Pacific Northwest, we retrieved `r format(dplyr::n_distinct(idigbio_pnw_records$worms_valid_name), big.mark=",")` species in iDigBio ($n=$ and `r format(nrow(idigbio_pnw_records), big.mark=",")`) and `r format(dplyr::n_distinct(obis_pnw_records$worms_valid_name), big.mark=",")` species in OBIS ($n=$ `r format(nrow(obis_pnw_records), big.mark=",")`).
+For the Pacific Northwest, we retrieved `r format(dplyr::n_distinct(idigbio_pnw_records$worms_valid_name), big.mark=",")` species in iDigBio ($n=$ and `r format(nrow(idigbio_pnw_records), big.mark=",")` records) and `r format(dplyr::n_distinct(obis_pnw_records$worms_valid_name), big.mark=",")` species in OBIS ($n=$ `r format(nrow(obis_pnw_records), big.mark=",")` records).
 Overall, the number of species known from the Gulf of Mexico match the number estimated by experts (`r format(11150-1541-434, big.mark=",")` species [@Fautin2010]).
@@
 
+<!-- this paragraph needs overhaul to make it clearer -->
 For the Gulf of Mexico, the majority of the species included in the expert faunistic list are recorded in both iDigBio and OBIS (`r format_output(species_in_common$n_spp_gom_idigbio_obis/species_in_common$n_spp_gom*100)` %), or only OBIS (`r format_output(species_in_common$n_spp_gom_obis/species_in_common$n_spp_gom*100)` %) (Fig.
@@
 
+<!-- need to separate across all phyla statements and by phylum statements -->
 We investigated the number of species for each phylum that contributed to the discrepancies between the regional species lists and the database records.
-We found that the better studied phyla contributed the most to the differences across the data sources.
+We found that the more diverse phyla contributed the most to the differences across the data sources.
 OBIS tends to be closer to the full record set, but iDigBio has more species for the Gulf of Mexico that are not included in the list (Fig.
@@
 
-Species that are included in the list but do not have records in databases could be (1) rare: they have never been collected or have been collected only a few times, and the corresponding records have not been digitized yet; (2) difficult to preserve: large, very small, or fragile species are deposited in museums less frequently; (3) are in the database under a different name: the species has been synonimized or re-identified since the publication of the list under a different name that may be included in the database; (4) at the limit of their range: the species were included in the list because they are known to occur within the boundaries of the geographic area, but they have never been collected within these geographic boundaries; (5) do not have geographic coordinates associated with them: some of these species can be in the database but without georeferenced data points and therefore were not returned in our queries.
+<!-- need to separate better baises for which I have data and those for which I don't -->
+Species that are included in the list but do not have records in databases could be (1) rare: they have never been collected or have been collected only a few times, and the corresponding records have not been digitized yet; (2) difficult to preserve: large, very small, or fragile species are deposited in museums less frequently; (3) are in the database under a different name: the species has been synonymized or re-identified since the publication of the list under a different name that may be included in the database; (4) at the limit of their range: the species were included in the list because they are known to occur within the boundaries of the geographic area, but they have never been collected within these geographic boundaries; (5) do not have geographic coordinates associated with them: some of these species can be in the database but without georeferenced data points and therefore were not returned in our queries.
 
 Untangling the relative contribution of these factors to explain the discrepancies is difficult as it would require to know the criteria used by the experts who have assembled the lists.
-However, by querying iDigBio we found that for the `r format(not_really_in_database$idig_gom_n_spp_not_in_db, big.mark=",")` species that do not appear within the geographic range of the Gulf of Mexico but are on the list, `r format_output(not_really_in_database$idig_gom_p_spp_not_in_db*100)`% are truly absent from iDigBio, while for the `r format(not_really_in_database$idig_gom_n_spp_no_coords, big.mark=",")` species that appear in iDigBio, `r format_output(not_really_in_database$idig_gom_p_spp_no_coords*100)`%  do not have geographic coordinates associated with them, and for the ones that do have geographic coordinates, `r format_output(not_really_in_database$idig_gom_p_spp_within_usa*100)`% ($n=$  `r format(not_really_in_database$idig_gom_n_spp_within_usa, big.mark=",")`) fall within the US waters.
+However, by querying iDigBio we found that for the `r format(not_really_in_database$idig_gom_n_spp_not_in_db, big.mark=",")` species that do not appear within the geographic range of the Gulf of Mexico but are on the list, `r format_output(not_really_in_database$idig_gom_p_spp_not_in_db*100)`% are completely absent from iDigBio across all areas, while for the `r format(not_really_in_database$idig_gom_n_spp_no_coords, big.mark=",")` species that appear in iDigBio, `r format_output(not_really_in_database$idig_gom_p_spp_no_coords*100)`%  do not have geographic coordinates associated with them, and for the ones that do have geographic coordinates, `r format_output(not_really_in_database$idig_gom_p_spp_within_usa*100)`% ($n=$  `r format(not_really_in_database$idig_gom_n_spp_within_usa, big.mark=",")`) fall within the US waters.
 These numbers indicate that the lack of geographic coordinates associated with the specimens in the database contribute at least significantly to the discrepancy, but that a large fraction of these species have no digital records of their occurrence in the Gulf of Mexico.
@@
 
+<!-- add sentence
 We combined `r format(nrow(idigbio_records), big.mark=",")` museum records for marine invertebrates collected within the EEZ of the continental US representing `r length(unique(tolower(idigbio_records[["data.dwc:phylum"]])))` phyla and `r format(length(unique(idigbio_records$worms_valid_name)), big.mark=",")` species from iDigBio.
@@
 
-These patterns of spatial distribution for the species richness are strongly confounded by the sampling effort.
-Investigating the relation between the number of samples and the number of species for each cell of the map (Fig.
-\@ref(fig:correlation-sampling)) reveals that it takes about 100 records per cell before there is clear redundancy in the sampled species (each record represent a different species).
+Spatial patterns of species richness recovered are strongly confounded by sampling effort.
+Samples of <100 records per cell commonly have as many species as records (Fig.
+\@ref(fig:correlation-sampling)).
 
@@
 
-The inadequacy of the sampling effort compared to the diversity of the fauna is also reflected by the fact that `r prop_singleton_species`% of the species are known from a single record.
+Underscoring the inadequacy of the sampling effort, `r prop_singleton_species`% of the species are known from a single record in iDigBio.
 Another indication of this inadequacy is the relation between the number of collected specimen and the number of species recorded through time (Fig.

### 2017-07-26 manuscript.Rmd
include GP's comments/edits in manuscript

@@
 We assessed how faunistic lists assembled over decades by specialists compare with records in public databases for the same geographic area.
-We parsed faunistic lists to extract marine invertebrates from the Gulf of Mexico [@Tunnell2009] and from the Pacific Northwest [@Kozloff1974], leading to `r format(sum(gom_worms$is_marine, na.rm=TRUE), big.mark=",")` and `r format(sum(kozloff_species$is_marine, na.rm=TRUE), big.mark=",")` species respectively.
+We parsed faunistic lists to extract marine invertebrates from the Gulf of Mexico [@Tunnell2009] and from the Pacific Northwest [@Kozloff1996], leading to `r format(sum(gom_worms$is_marine, na.rm=TRUE), big.mark=",")` and `r format(sum(kozloff_species$is_marine, na.rm=TRUE), big.mark=",")` species respectively.
 Our goals were (1) to compare  the species lists for these geographic areas assembled by experts with the records found in databases; (2) to assess the proportion of species included in the expert lists that have records in public databases.
@@
 ```{r compare_database_coverage, fig.cap='Overlap between the regional species list (list), records in iDigBio (idigbio), records in OBIS (obis), and records in BOLD (bold) for the Gulf of Mexico (left) and the Pacific Northwest (right).
-This plot is generated using the UpSet technique[@Lex2014], which uses a matrix-based layout to represent intersections of sets and their sizes.
+This plot is generated using the UpSet technique[@Lex2014;@Conway2017], which uses a matrix-based layout to represent intersections of sets and their sizes.
 The rows represent the sets, the columns represent the intersections, and each set that is part of the intersection is marked with a black-filled circle in the cells underneath.
@@
 \@ref(fig:idigbio-through-time)C), suggest that any new sampling effort will lead to new observations and new species.
-Coordinated large-scale biodiversity expeditions continue to be needed to improve our understanding of the diversity of species and their spatial distributions.
-While less explored than the US, a recent study in Western Australia reported that the collection of 135 locations led to the collection of 805 species of crustaceans and polychaetes, 94.6% and 72% of which were undescribed [@Poore2014], highlighting the potential for discovery in the oceans.
+<!-- separate issues of sampling is needed, and large-scale sampling is best way of doing it --> Coordinated large-scale biodiversity expeditions continue to be needed to improve our understanding of the diversity of species and their spatial distributions.
+The US marine EEZ is relatively well-known compared to many areas of the world ocean, where studies show that undescribed biodiversity dominates.
+For example, a recent survey in Western Australia reported that 94.6% of crustaceans and 72% of polychaetes taken at 135 sites were undescribed [@Poore2014].
 
@@
 
-Predicting the geographic range of species, and how they will be impacted by global change has undergone important advances recently [@Guillera-Arroita2017], but remain scarecely used in the marine realm [@Robinson2011].
-One limitation is the low number of observations available for each species: it is generally recognized that to obtain reliable results such efforts require to have accurate geographic information for a minimum of 10 specimens but having at least 30 is preferred [@Wisz2008].
-From the iDigBio data, we found that `r prop_10_specimens_species`% of species are known from less than 10 records and `r prop_30_specimens_species`% from less than 30 records.
-This suggests that our understanding of the geographic range of species is limited as most species are known from only a few locations, while efforts to estimate their current and future ranges based on occurrence data are rendered moot by the low number of data points available.
+Environmental conditions are among the primary determinants of species distributions and can thus be used to estimate the response of organisms to climate change.
+Predictive ecological models forecasting changes in distributions are training on environmental conditions encountered by species in their current distributions.
+To obtain reliable results, models require at least 10, but ideally 30 or more occurrence records per species [@Wisz2008].
+Yet at present, `r prop_10_specimens_species`% of species are represented by $\leq$ 10 records, and `r prop_30_specimens_species`%  by $\leq$ 30 records in iDigBio.
 
-Furthermore, there are important discrepancies in our level of knowledge across phyla.
-While larger, and accessible groups represent the bulk of the records for marine invertebrates in museum collections, the small organisms, including representatives of the meiofauna are virtually absent.
-Of the `r n_idigbio_phyla` phyla occurring in iDigBio, `r records_rare_phyla$n_rare_phyla` are represented by less than 500 records, and make up `r records_rare_phyla$prop_rare_phyla`% of the total number of records.
-Yet, these undersampled taxa are likely to play an important role in the detection of changes in environmental quality due to their size, and physiological vulnerability [@Bik2012].
-Additionally, the impact of these changes on higher trophic levels and ecosystems is largely unexplored.
+Further biases exist across phyla and body size.
+Large-bodied and easily acccessible (e.g.
+dried specimens like shells, corals, echinoderms) are overrepresented in museum collections, compared with small less accessible organisms, such as the meiofauna.
+Of `r n_idigbio_phyla` phyla represented in iDigBio, `r records_rare_phyla$n_rare_phyla` have <500 records, and make up `r records_rare_phyla$prop_rare_phyla`% of the records.
+Yet, these undersampled taxa, especially those of small body size, are often the most abundant, and are likely to play an important role in the detection of changes in environmental quality[@Bik2012].
+Additionally, the ecological implications of changes in the structure and composition of the communities of these small-bodied organisms, on higher trophic levels and ecosystems are largely unexplored.
 
-To assess the status of taxonomic expertise in the US, we calculated the proportion of specimens identified at the species level through time.
-For all taxa, excepts for molluscs, there is a consistent decline, indicating that specimens collected recently remain for the most part unidentified in collections (Fig.
+The proportion of samples that are identified to species show a marked decline in recent decades in most taxa other than mollusks, demonstrating limiting taxonomic effort and expertise (Fig.
 \@ref(fig:identification-through-time)).
-For diverse groups such as annelids or arthropods, the proportion of specimens identified at the species-level is now below 30%.
+There is a marked break in the central trend arond the 1970's indicating that it now takes decades, on average, to identify samples deposited in collections.
+For diverse groups such as annelids or arthropods, the proportion of specimens identified to the species-level is now below 30%.
 The number of scientists describing new species might be increasing globally [@Costello2013], but taxonomic experts continue to be needed to identify specimens recently collected in the US.
@@
 
-Since its introduction in the early 2000s, DNA barcoding, the sequencing of a common locus across taxa bearing species-specific information, has revitalized taxonomy by facilitating species delineation [@Meyer2005-plos].
-More recently, advances in high-throughput sequencing technologies have enabled the parallel sequencing of thousands of species that occur in environmental samples cost effectively [@Leray2015].
-This approach opens up new research venues as it becomes possible to estimate the diversity from species-rich samples that were virtually inaccessible before.
-This is particularly true in marine communities that include multiple phyla and organisms varying in size by orders of magnitude, making identification challenging.
-However, the utility of the data generated by these projects is enhanced by the possibility of matching DNA sequences obtained to previously sequenced and correctly identified species [@Leray2015].
+Since its introduction in the early 2000s, DNA barcoding, has revitalized taxonomy by facilitating species delineation [@Meyer2005-plos].
+More recently, advances in high-throughput sequencing technologies have enabled the parallel sequencing of thousands of specimens in environmental samples (eDNA) cost effectively [@Leray2015], making it possible to estimate diversity and identify taxa from species-rich samples at unprecedented scales.
 With decreasing sequencing costs, and the prospect of automating sample collection, these technologies may soon be used routinely to monitor biodiversity continuously [@Bohan2017].
-DNA barcoding provides the possibility to link these newly generated data with historical natural history knowledge about species and their distributions.
-However, the insights we will gain from these approaches are only going to be as good as the level of completeness and accuracy of the information included in reference databases.
+However, the utility of the data generated by these projects is limited by the paucity of reliably identified DNA barcode sequences that can serve for matching and identification of eDNA transcripts [@Leray2015].
+Thus insights from eDNA technologies depend on the level of completeness and accuracy of comparative, species-level DNA barcode coverage.
 
-To estimate the proportion of the marine invertebrate diversity for which DNA barcodes are available, we queried BOLD (Barcode of Life Datasystem) for all the species that are listed in the regional lists for the Gulf of Mexico, the Pacific Northwest, as well as all the records for marine invertebrates that occur in the EEZ of the US found in iDigBio.
-Here, we focused on records based on museum specimens as some of them could potentially be used to complete current DNA barcoding efforts.
+<!-- include results earlier? -->
 
-The proportion of species that have DNA barcodes available is relatively constant across the better studied phyla at the scale of the EEZ.
-It varies between 12% for bryozoans to 28% for cnidarians (Fig.
-@ref(fig:proportion-barcoded)).
-The proportions for species known to occur in the Gulf of Mexico are usually similar than for the US, while typically a higher fraction of species from the Pacific Northwest are available.
+To estimate the proportion of the marine invertebrate diversity for which DNA barcodes are available, we queried BOLD (Barcode of Life Datasystem) for all species listed in the regional lists for the Gulf of Mexico, the Pacific Northwest, as well as all the records for marine invertebrates that occur in the EEZ of the US found in iDigBio.
+
+<!-- add overall mean number before getting into details -->
+The proportion of species within the US EEZ that have DNA barcodes varies from 12% for bryozoans to 28% for cnidarians (Fig.
+\@ref(fig:proportion-barcoded)).
+The GOM has comparable coverage, while the PNW is better barcoded with >30% species-level coverage across each of the five most diverse phyla.
+This is driven in part by the active Canadian barcoding effort in British Columbia.
 In 2012, it was estimated that, at a global scale, less than 3% of the marine species had been sequenced [@Appeltans2012], and that this number was even lower in some areas (e.g.
 2.6% for the Antarctic [@Grant2009]).
-This suggests that even as many additional species have had their DNA barcodes made available since these studies, the US fauna is generally better characterized than other parts of the world.
+Thus the US fauna is generally better characterized than other parts of the world.
 
-Using the fraction of known species that have a barcode as a yard stick of the progress made is not sufficient.
-Many marine invertebrate species are known to be cryptic complexes that are not as widely distributed as currently reported [@Knowlton2000,@Meyer2005-evolution].
-Therefore specimens identified under the same species but sampled in distant locations, could represent cryptic species.
+The above estimates are based on species names, but many traditional species include multiple cryptic species [@Knowlton2000;@Meyer2005-evolution].
+For example, `r format_output(100*stat_bold$n_spp_multi_bin/stat_bold$n_spp)`% of `r format(stat_bold$n_spp, big.mark=",")` species analyzed here from BOLD include cryptic species, as denoted by assignment to  more than 1 BIN (BOLD Index Number, operational taxonomic units that closely resemble species [@Ratnasingham2013]).
+Specimens identified under the same species but sampled from distant locations, are especially prone to representing cryptic species.
 Geographic information is missing for most records (`r format_output(100*stat_bold$p_no_coords)`% of the `r format(stat_bold$n_total_records, big.mark=",")` BOLD records analyzed here have no geographic coordinate information associated with them).
 It is thus crucial to ensure that accurate geographic information is provided when DNA barcodes are published, and that enough samples for each species are sequenced, to capture more accurately their geographic range and the extent of their genetic diversity.
-This cryptic diversity is also at least partially reflected in the fact that `r stat_bold$n_spp_multi_bin` species out of `r format(stat_bold$n_spp, big.mark=",")` in our dataset include more than 1 BIN (BOLD Index Number, operational taxonomic units that closely resemble species [@Ratnasingham2013]).
 
-```{r proportion-barcoded, fig.cap='Proportion of species known to occur in the US EEZ based on iDigBio records, in the Gulf of Mexico [@Tunnell2009], in the Pacific Northwest[@Kozloff1987], that have publicly available DNA barcodes.
+```{r proportion-barcoded, fig.cap='Proportion of species known to occur in the US EEZ based on iDigBio records, in the Gulf of Mexico [@Tunnell2009], in the Pacific Northwest[@Kozloff1996], that have publicly available DNA barcodes.
 Numbers next to the bars indicate the number of species for which DNA barcodes are available.
@@
 Geographic information might be available in other forms, but needs to be digitized.
-While tools have been developed to convert names of locales into GPS coordinates on land [@rios2010geolocate], similar approaches are lacking for the sea.
+While tools have been developed to convert names of locales into GPS coordinates on land [@rios2010geolocate], similar approaches are less-developed for the sea.
 
@@
 
-## Material and methods
+## Methods
 
-The list of species for the Gulf of Mexico was obtained by downloading CSV files for each animal phylum from http://e-gulf.org/ [@egulf].
+The list of species for the Gulf of Mexico was obtained  from http://e-gulf.org/ [@egulf].
 This database includes the species listed in a comprehensive biotic inventory of the Gulf of Mexico [@Felder2009].
 
-The list of species for the Pacific Northwest was downloaded from:  https://www.eopugetsound.org/species/custom-lists/307 (accessed on July 5th, 2017) as a Microsoft Excel spreadsheet.
-This is a digital version of the species included in [@Kozloff1974].
+The list of species for the Pacific Northwest was downloaded from:  https://www.eopugetsound.org/species/custom-lists/307 (accessed on July 5th, 2017).
+This is a digital version of the species included in [@Kozloff1996].
 
-To search by geographic area, we obtained the boundaries of the Exclusive Economic Zone for the United States from marineregions.org (mrgid:8456) [@mregions].
+For geographic searches, we obtained the boundaries of the Exclusive Economic Zone for the United States from marineregions.org (mrgid:8456) [@mregions].
 For the Pacific Northwest, Kozloff did not specify the exact geographic extent covered by his book.
@@
 For the Gulf of Mexico, we used the same geographic area as the one defined in [@Tunnell2009] by modifying manually the eastern boundaries of the shape file for the Gulf of Mexico provided by marineregions.org (mrgid:4288).
-These geographic areas were split into squares of 0.2°x0.2° which were used to query the web API of iDigBio and OBIS using the R packages ridigbio and robis respectively.
+These geographic areas were split into 0.2°$*$0.2° quadrats that were used to query the web API of iDigBio and OBIS using the R packages ridigbio [@ridigbio-0.3.5] and robis [@robis-0.1.8] respectively.
 The resulting occurrences were then clipped to only include records that fell within the EEZ boundaries.
 
-The species names used in queries or obtained as results were checked against the World Registry of Marine Species (WoRMS).
-Whenever possible, species names were matched and the current valid names for the species were used.
+All species names obtained from all sources were checked against the World Registry of Marine Species (WoRMS).
+Species that could not be matched in WoRMS, and species listed as non-marine in WoRMS were not included in subsequent analyses.
+For all names that were listed in WoRMS, the current accepted name was capture and used in all analyses.
 This ensured that the same species listed under different synonyms was only counted once.
-Species that could not be matched in WoRMS, and species listed as non-marine in WoRMS were not included in subsequent analyses.
-The R package taxizesoap was used to query the WoRMS database.
+The R package worrms [@worrms-0.1.0] was used to query the WoRMS database.
 
-To assess the proportion of species for which DNA barcodes are available, we used the current valid names for all records to query the Barcode of Life Datasystem (BOLD) using the R package bold.
-When no records where found for a species, we looked up for alternative matches using its synonyms as listed in WoRMS.
-
-Figure \@ref(fig:database-coverage) was generated using the R package UpSetR [@Conway2017] based on the UpSet technique [@Lex2014].
+To assess the proportion of species for which DNA barcodes are available, we queried the Barcode of Life Datasystem (BOLD) using the R package bold [@bold-0.4.0].
+We used the accepted names of species in the other lists, as well as synonyms (as listed in WoRMS) when accepted names did not return any records.
 

### 2017-07-26 manuscript.Rmd
[fix #50] remove singletons from BOLD multi BIN calculations

@@
 The above estimates are based on species names, but many traditional species include multiple cryptic species [@Knowlton2000;@Meyer2005-evolution].
-For example, `r format_output(100*stat_bold$n_spp_multi_bin/stat_bold$n_spp)`% of `r format(stat_bold$n_spp, big.mark=",")` species analyzed here from BOLD include cryptic species, as denoted by assignment to  more than 1 BIN (BOLD Index Number, operational taxonomic units that closely resemble species [@Ratnasingham2013]).
+For example, up to `r format_output(100 * stat_bold$p_spp_multi_bin)`% of `r format(stat_bold$n_spp_not_singleton, big.mark=",")` species represented by more than 1 record in BOLD include cryptic species, as denoted by assignment to  more than 1 BIN (BOLD Index Number, operational taxonomic units that closely resemble species [@Ratnasingham2013]).
 Specimens identified under the same species but sampled from distant locations, are especially prone to representing cryptic species.

### 2017-07-26 R/bold.R

@@ -98,6 +98,7 @@ make_stat_bold <- function(gom_bld, koz_bld, gom_wrm, koz_wrm) {
 
     bin_data <- get_bold_bins(gom_wrm, koz_wrm)
 
+    ## BINs that are found in more than 1 species
     n_shared_bins <- bin_data %>%
         dplyr::group_by(bins) %>%
         dplyr::summarize(
@@ -106,20 +107,30 @@ make_stat_bold <- function(gom_bld, koz_bld, gom_wrm, koz_wrm) {
         dplyr::filter(n_shared_bins > 1) %>%
         nrow()
 
-    n_spp_multi_bin <- bin_data %>%
+    ## Species that have more than 1 BIN associated with their name
+    ## -- we need to remove species that have only 1 record
+    multi_bin <- bin_data %>%
         dplyr::group_by(worms_valid_name) %>%
         dplyr::summarize(
             n_spp_multi_bin = n_distinct(bins)
-        ) %>%
+            ) %>%
+        dplyr::left_join(dplyr::select(res, worms_valid_name, n_bold_records),
+                         by = "worms_valid_name") %>%
+        dplyr::filter(n_bold_records > 1)
+
+    n_multi_bin <- multi_bin %>%
         dplyr::filter(n_spp_multi_bin > 1) %>%
         nrow()
+    p_multi_bin <- n_multi_bin/nrow(multi_bin)
 
     list(
         p_no_country = 1 - (sum(res$n_country)/sum(res$n_bold_records)),
         p_no_coords = 1 - (sum(res$n_coords)/sum(res$n_bold_records)),
         n_total_records = sum(res$n_bold_records),
         n_shared_bins = n_shared_bins,
-        n_spp_multi_bin = n_spp_multi_bin,
+        n_spp_multi_bin = n_multi_bin,
+        p_spp_multi_bin = p_multi_bin,
+        n_spp_not_singleton = nrow(multi_bin),
         n_bins = length(unique(bin_data$bins)),
         n_spp = length(unique(bin_data$worms_valid_name))
     )

### 2017-07-26 manuscript.Rmd
use latex `x` sign

@@
 For the Gulf of Mexico, we used the same geographic area as the one defined in [@Tunnell2009] by modifying manually the eastern boundaries of the shape file for the Gulf of Mexico provided by marineregions.org (mrgid:4288).
-These geographic areas were split into 0.2°$*$0.2° quadrats that were used to query the web API of iDigBio and OBIS using the R packages ridigbio [@ridigbio-0.3.5] and robis [@robis-0.1.8] respectively.
+These geographic areas were split into 0.2° $\times$ 0.2° quadrats that were used to query the web API of iDigBio and OBIS using the R packages ridigbio [@ridigbio-0.3.5] and robis [@robis-0.1.8] respectively.
 The resulting occurrences were then clipped to only include records that fell within the EEZ boundaries.

### 2017-07-27 manuscript.Rmd
[fix #41] use 1996 for new records since publication in Kozloff

@@
 We investigated the proportion of species discovered since the publication of the faunistic lists using iDigBio records.
-They represent `r sum(recently_collected$gom_in_idig$min_year > 2001)` species for the Gulf of Mexico, and `r sum(recently_collected$pnw_in_idig$min_year > 1987)` species for the Pacific Northwest.
+They represent `r sum(recently_collected$gom_in_idig$min_year > 2001)` species for the Gulf of Mexico, and `r sum(recently_collected$pnw_in_idig$min_year > 1996)` species for the Pacific Northwest.
 While errors in geographic coordinates cannot be completely ruled out, it is more likely that the main contributing factor was that the authors of these checklists were not aware of the presence of these species based only on museum records, as these lists were mostly assembled from publications, and only a small fraction of museum records are used in publications.

### 2017-07-30 R/comparison_by_geo.R
fix typo

@@ -423,7 +423,7 @@ not_in_list_collected_recently <- function(database_overlap, map_gom, map_pnw) {
         .obis %>%
         .pnw
 
-    spp_gom <- bind_rows(idigbio = spp_pnw_in_idig,
+    spp_pnw <- bind_rows(idigbio = spp_pnw_in_idig,
                          obis = spp_pnw_in_obis,
                          .id = "database") %>%
         .keep_min

### 2017-07-30 manuscript.Rmd
update variable names

@@
 
-We investigated the proportion of species discovered since the publication of the faunistic lists using iDigBio records.
-They represent `r sum(recently_collected$gom_in_idig$min_year > 2001)` species for the Gulf of Mexico, and `r sum(recently_collected$pnw_in_idig$min_year > 1996)` species for the Pacific Northwest.
+We investigated the proportion of species discovered since the publication of the faunistic lists using iDigBio and OBIS records.
+They represent `r sum(recently_collected$gom_in_dbs$min_year > 2001)` species for the Gulf of Mexico, and `r sum(recently_collected$pnw_in_dbs$min_year > 1996)` species for the Pacific Northwest.
 While errors in geographic coordinates cannot be completely ruled out, it is more likely that the main contributing factor was that the authors of these checklists were not aware of the presence of these species based only on museum records, as these lists were mostly assembled from publications, and only a small fraction of museum records are used in publications.

### 2017-07-30 R/combine_plots.R
move cowplots to their own file

@@ -1,0 +1,25 @@
+combine_cowplots <- function(g1, g2, nrow = 1, common_legend = TRUE, ...) {
+    if (common_legend) {
+        legend <- get_legend(g1)
+        g1 <- g1 + theme(legend.position = "none")
+        g2 <- g2 + theme(legend.position = "none")
+    }
+    prow <- cowplot::plot_grid(
+                         g1,
+                         g2,
+                 align = 'vh',
+                 labels = c("A", "B"),
+                 hjust = -1,
+                 nrow = nrow
+                 )
+    if (common_legend) {
+        cowplot::plot_grid(prow, legend, rel_widths = c(3, .5), ...)
+    } else prow
+
+}
+
+combine_cowplots_2 <- function(g1, g2, g3) {
+    lgd <- get_legend(g1)
+    bottom_row <- plot_grid(g3, labels = "C", align = "h", rel_widths = c(1, .1))
+    plot_grid(combine_cowplots(g1, g2, common_legend = TRUE), bottom_row, nrow = 2)
+}

### 2017-07-30 R/gom_list.R

@@ -231,30 +231,3 @@ plot_richness_per_db <- function(smry_db, region = c("gom", "pnw")) {
         theme_bw(base_family = "Ubuntu Condensed") +
         coord_flip()
 }
-
-combine_cowplots <- function(g1, g2, nrow = 1, common_legend = TRUE, ...) {
-    if (common_legend) {
-        legend <- get_legend(g1)
-        g1 <- g1 + theme(legend.position = "none")
-        g2 <- g2 + theme(legend.position = "none")
-    }
-    prow <- cowplot::plot_grid(
-                         g1,
-                         g2,
-                 align = 'vh',
-                 labels = c("A", "B"),
-                 hjust = -1,
-                 nrow = nrow
-                 )
-    if (common_legend) {
-        cowplot::plot_grid(prow, legend, rel_widths = c(3, .4), ...)
-    } else prow
-
-}
-
-combine_cowplots_2 <- function(g1, g2, g3) {
-    bottom_row <- plot_grid(g3, labels = "C", align = "h", rel_widths = c(3, .4))
-    g1 <- g1 + theme(legend.position = "none")
-    g2 <- g2 + theme(legend.position = "none")
-    plot_grid(combine_cowplots(g1, g2, common_legend = FALSE), bottom_row, nrow = 2)
-}

### 2017-07-30 manuscript.Rmd
[fix #45] code effort through time by year instead of phylum

@@
 
-<!-- add sentence
+<!-- add sentence -->
 We combined `r format(nrow(idigbio_records), big.mark=",")` museum records for marine invertebrates collected within the EEZ of the continental US representing `r length(unique(tolower(idigbio_records[["data.dwc:phylum"]])))` phyla and `r format(length(unique(idigbio_records$worms_valid_name)), big.mark=",")` species from iDigBio.
@@
 ```{r idigbio-through-time, fig.cap='Cumulative number of samples (A) and species (B) of marine invertebrates recorded through time based on iDigBio records.
-Relation between number of samples collected, and number of species recorded for the first time (C), each point represents numbers for a given year, dashed lines show the 1:1 and 1:100 ratios for number of species recorded for the first time and number of samples.
+Relation between number of samples collected, and number of species recorded for the first time (C), each point represents numbers for a given year, dashed line show the 1:1 and dotted line the 1:100 ratios for number of species recorded for the first time to the number of samples.
 (\\#fig:idigbio-through-time)'}

### 2017-07-30 R/time_plot.R

@@ -200,24 +200,33 @@ plot_samples_vs_spp_through_time <- function(knowledge_through_time) {
 
     res <- knowledge_through_time %>%
         dplyr::ungroup() %>%
-        dplyr::filter(phylum %in% phy_to_keep,
-                      !is.na(cum_n_new_spp)) %>%
-        dplyr::mutate(phylum = capitalize(phylum)) %>%
-        ggplot(aes(x = n_samples, y = n_new_spp, colour = phylum)) +
+        #dplyr::filter(phylum %in% phy_to_keep,
+        #              !is.na(cum_n_new_spp)) %>%
+        dplyr::group_by(year) %>%
+        dplyr::summarize(n_spp = sum(n_new_spp, na.rm = TRUE),
+                         n_samples = sum(n_samples, na.rm = TRUE)) %>%
+        dplyr::mutate(period = cut(year, breaks = seq(1850, 2030, by = 20),
+                                   labels = paste(seq(1850, 2010, by = 20),
+                                                  c(seq(1870, max(year, na.rm = TRUE), by = 20),
+                                                    max(year, na.rm = TRUE)),
+                                                  sep = "-"), right = FALSE))
+    res %>%
+        ggplot(aes(x = n_samples, y = n_spp, colour = period)) +
         geom_point() +
-        geom_abline(slope = 1, linetype = 2) +
-        geom_line(data = data_frame(samples = 1:10000, species = samples/100),
+        geom_line(data = data_frame(samples = 1:100000, species = samples),
                   aes(x = samples, y = species), color = "black", linetype = 2,
-                  inherits.aes = FALSE) +
-        scale_x_log10() + scale_y_log10(limits = c(1, 100)) +
+                  inherit.aes = FALSE) +
+        geom_line(data = data_frame(samples = 1:100000, species = samples/100),
+                  aes(x = samples, y = species), color = "black", linetype = 3,
+                  inherit.aes = FALSE) +
+        scale_x_log10() + scale_y_log10(limits = c(1, 600)) +
         xlab("Number of Samples") +
         ylab("Number of Species Recorded for the First Time") +
         theme_ipsum(base_family = "Ubuntu Condensed") +
-        scale_colour_hc(name = "")
+        scale_color_viridis(name = "", discrete = TRUE)
 
-    res
+
 }
-
 
 plot_samples_vs_spp_std <- function(knowledge_through_time) {
 

### 2017-08-08 manuscript.Rmd
fix calc of number of phyla

@@
 <!-- add sentence -->
-We combined `r format(nrow(idigbio_records), big.mark=",")` museum records for marine invertebrates collected within the EEZ of the continental US representing `r length(unique(tolower(idigbio_records[["data.dwc:phylum"]])))` phyla and `r format(length(unique(idigbio_records$worms_valid_name)), big.mark=",")` species from iDigBio.
+<!-- separate idigbio and (and add) obis in the results -->
+We combined `r format(nrow(idigbio_records), big.mark=",")` museum records for marine invertebrates collected within the EEZ of the continental US representing `r length(unique(tolower(idigbio_records[["phylum"]])))` phyla and `r format(length(unique(idigbio_records$worms_valid_name)), big.mark=",")` species from iDigBio.
 Most samples were collected on the continental shelf (Fig.

### 2017-08-08 manuscript.Rmd
[fix #35] compute proportion barcoded dynamically

@@
 <!-- add overall mean number before getting into details -->
-The proportion of species within the US EEZ that have DNA barcodes varies from 12% for bryozoans to 28% for cnidarians (Fig.
+The proportion of species within the US EEZ that have DNA barcodes varies from `r format_output(stat_barcoding$min_eez_value * 100)`% for `r stat_barcoding$min_eez_phylum` to `r format_output(stat_barcoding$max_eez_value * 100)`% for `r stat_barcoding$max_eez_phylum` (Fig.
 \@ref(fig:proportion-barcoded)).
-The GOM has comparable coverage, while the PNW is better barcoded with >30% species-level coverage across each of the five most diverse phyla.
+The GOM has comparable coverage than for the EEZ, while the PNW is better barcoded with >30% species-level coverage across each of the five most diverse phyla.
 This is driven in part by the active Canadian barcoding effort in British Columbia.

### 2017-08-08 R/bold.R

@@ -166,34 +166,63 @@ bold_status <- function(idig) {
         )
 }
 
-summary_bold_status <- function(gom_bold, koz_bold, idig_bold) {
-
+bold_status_data <- function(gom_bold, koz_bold, idig_bold) {
     dplyr::bind_rows(
                gom = bold_status(gom_bold),
                koz = bold_status(koz_bold),
                all_idigbio = bold_status(idig_bold),
                .id = "data_source"
            ) %>%
-        dplyr::filter(
-                   phylum %in% c("annelida", "arthropoda",
-                                     "bryozoa",
-                                     "chordata", "cnidaria",
-                                     "echinodermata", "mollusca",
-                                     "porifera"
-                                     )
-               ) %>%
-            dplyr::mutate(phylum = capitalize(phylum)) %>%
-            ggplot(aes(x = reorder(phylum, p_has_bold), y = p_has_bold,
-                       fill = data_source)) +
-            geom_col(position = "dodge") +
-            geom_text(aes(y = p_has_bold +.01, label = n_has_bold), position = position_dodge(.9),
-                      hjust = .1, family = "Ubuntu Condensed")  +
-            xlab("") + ylab("Proportion of species with available DNA barcodes") +
-            scale_fill_viridis(discrete = TRUE,
-                               name = "",
-                               labels = c("US EEZ", "Gulf of Mexico", "Pacific Northwest")) +
-            theme_ipsum(base_family = "Ubuntu Condensed") +
-            coord_flip()
+    dplyr::filter(
+               phylum %in% c("annelida", "arthropoda",
+                             "bryozoa",
+                             "chordata", "cnidaria",
+                             "echinodermata", "mollusca",
+                             "porifera"
+                             )
+           ) %>%
+    dplyr::mutate(phylum = capitalize(phylum))
+}
+
+plot_bold_status <- function(bold_data) {
+    bold_data %>%
+        ggplot(aes(x = reorder(phylum, p_has_bold), y = p_has_bold,
+                   fill = data_source)) +
+        geom_col(position = "dodge") +
+        geom_text(aes(y = p_has_bold +.01, label = n_has_bold), position = position_dodge(.9),
+                  hjust = .1, family = "Ubuntu Condensed")  +
+        xlab("") + ylab("Proportion of species with available DNA barcodes") +
+        scale_fill_viridis(discrete = TRUE,
[…]

### 2017-08-09 manuscript.Rmd
[fix #47] use iDigBio+OBIS for proportion of species with n records

@@
 To obtain reliable results, models require at least 10, but ideally 30 or more occurrence records per species [@Wisz2008].
-Yet at present, `r prop_10_specimens_species`% of species are represented by $\leq$ 10 records, and `r prop_30_specimens_species`%  by $\leq$ 30 records in iDigBio.
+Yet at present, `r prop_10_specimens_species`% of species are represented by $\leq$ 10 records, and `r prop_30_specimens_species`%  by $\leq$ 30 records based on the combined records in iDigBio and OBIS.
 

### 2017-08-09 R/correlation_specimens_species.R

@@ -26,9 +26,9 @@ model_sampling_effort <- function(id) {
 
 }
 
-## proportion of species known by `n_specimens`.
-calc_prop_nspecimens_species <- function(idig, n_specimens) {
-    res <- idig %>%
+## proportion of identified species known by `n_specimens`.
+calc_prop_nspecimens_species <- function(recs, n_specimens) {
+    res <- recs %>%
         dplyr::filter(!is.na(worms_valid_name)) %>%
         dplyr::count(phylum, worms_valid_name, sort = TRUE)
 

### 2017-08-09 manuscript.Rmd
[fix #43] change wording as many records are specimen based

@@
 They represent `r sum(recently_collected$gom_in_dbs$min_year > 2001)` species for the Gulf of Mexico, and `r sum(recently_collected$pnw_in_dbs$min_year > 1996)` species for the Pacific Northwest.
-While errors in geographic coordinates cannot be completely ruled out, it is more likely that the main contributing factor was that the authors of these checklists were not aware of the presence of these species based only on museum records, as these lists were mostly assembled from publications, and only a small fraction of museum records are used in publications.
+While errors in geographic coordinates cannot be completely ruled out, it is more likely that the main contributing factor was that the authors of these checklists were not aware of the existence of these species based on museum records, as they were difficult to discover before digitization which has occured after the compilation of the list in most cases.
 However, museum specimens are critical to validate the occurrence of species [@Rocha2014], especially for marine invertebrates for which species limits based on morphology have been known to be challenged by other lines of evidence including molecular data [@Knowlton2000].

### 2017-08-10 R/time_plot.R
fix subspecies in calc proportion of records ID'ed at species level

@@ -405,11 +405,13 @@ data_identification_level_through_time <- function(idig_records) {
 
 plot_identification_level_through_time <- function(id_level) {
     id_level %>%
+        ## count subspecies as species level identification
+        dplyr::mutate(id_level = replace(id_level, id_level == "subspecies", "species")) %>%
         count(year, phylum, id_level) %>%
         group_by(year, phylum) %>%
         mutate(p = n/sum(n),
                n_lots = sum(n)) %>%
-        filter(id_level == "species" | id_level == "subspecies",
+        filter(id_level == "species",
                phylum %in% c("annelida", "arthropoda", "chordata",
                              "cnidaria", "echinodermata", "mollusca")) %>%
         ungroup() %>%

### 2017-08-10 manuscript.Rmd
update legend to reflect maps show iDigBio and OBIS records

@@
 
-```{r map-diversity-sampling, fig.cap='Spatial distribution of the number of samples (A) and species richness (B) for marine invertebrates based on specimen data available from iDigBio within the limits of the Exclusive Economic Zone for the continental USA.
+```{r map-diversity-sampling, fig.cap='Spatial distribution of the number of samples (A) and species richness (B) for marine invertebrates based on records retrieved from iDigBio and OBIS within the limits of the Exclusive Economic Zone for the continental USA.
 (\\#fig:map-diversity-sampling)'}

### 2017-08-10 manuscript.Rmd
[fix #48] compare prop. of global diversity represented by phyla

@@
 dried specimens like shells, corals, echinoderms) are overrepresented in museum collections, compared with small less accessible organisms, such as the meiofauna.
-Of `r n_idigbio_phyla` phyla represented in iDigBio, `r records_rare_phyla$n_rare_phyla` have <500 records, and make up `r records_rare_phyla$prop_rare_phyla`% of the records.
+Of `r n_idigbio_phyla` phyla represented in iDigBio, `r records_rare_phyla$n_rare_phyla` have <500 records, and make up `r records_rare_phyla$prop_recs_rare_phyla`% of the iDigBio records.
+This underrepresentation is manifested at the species level, while `r records_rare_phyla$prop_spp_rare_phyla_collected`% (n = `r format(records_rare_phyla$total_spp_rare_phyla_collected, big.mark=",")` total species) of the global number of species for the `r records_rare_phyla$n_rare_phyla` phyla occur in iDigBio, this proportion rises to `r records_rare_phyla$prop_spp_common_phyla_collected`% (n = `r format(records_rare_phyla$total_spp_common_phyla_collected, big.mark=",")`).
 Yet, these undersampled taxa, especially those of small body size, are often the most abundant, and are likely to play an important role in the detection of changes in environmental quality[@Bik2012].

### 2017-08-10 R/correlation_specimens_species.R

@@ -55,32 +55,65 @@ calc_prop_species_not_collected_since <- function(idig, max_year) {
 
 calc_records_rare_phyla <- function(idig, obis, wrms_stats, gom_phyla) {
 
+    ## OBIS is included here, but I'm not doing anything with it at
+    ## this stage
+
     summary_db <- . %>%
         dplyr::filter(is_marine == TRUE,
                       phylum %in% gom_phyla) %>%
-        dplyr::mutate(phylum = replace(phylum,
-                                       phylum %in% c("kinoryncha", "loricifera",
-                                                     "nematomorpha", "priapulida"),
-                                       "cephalorhyncha")) %>%
         dplyr::group_by(phylum) %>%
         dplyr::summarize(
                    n_records = n(),
                    n_spp = n_distinct(worms_valid_name)
                ) %>%
-        dplyr::arrange(desc(n_records)) %>%
-        dplyr::left_join(dplyr::mutate(wrms_stats, phylum = tolower(phylum)),
-                         by = "phylum") %>%
-        dplyr::mutate(is_rare = dplyr::if_else(n_records < 500, n_records, 0L),
-                      prop_collected = n_spp/accepted_species_marine_non_fossil)
+        dplyr::arrange(desc(n_records))  %>%
+        dplyr::mutate(is_rare = dplyr::if_else(n_records < 500, TRUE, FALSE))
 
     idig_res <- idig %>% summary_db
     obis_res <- obis %>% summary_db
 
     res <- dplyr::bind_rows(idigbio = idig_res, obis = obis_res, .id = "database")
 
+    idig_rare_phyla <- dplyr::filter(idig_res, is_rare == TRUE) %>%
+        dplyr::pull(phylum)
+
+    idig_worms <- idig_res %>%
+        dplyr::mutate(phylum = replace(phylum,
+                                       phylum %in% c("kinorhyncha", "loricifera",
+                                                     "nematomorpha", "priapulida"),
+                                       "cephalorhyncha"),
+                      phylum = replace(phylum, phylum == "echiura", "annelida")) %>%
+        dplyr::group_by(phylum) %>%
+        dplyr::summarize_if(is.integer, sum) %>%
+        dplyr::left_join(dplyr::mutate(wrms_stats, phylum = tolower(phylum)), by = "phylum") %>%
+        dplyr::mutate(is_rare = if_else(n_records < 500,  TRUE, FALSE))
+
[…]

### 2017-08-10 R/utils.R

@@ -110,7 +110,7 @@ gcd.hf <- function(long1, lat1, long2, lat2, warn = TRUE) {
     return(d) # Distance in km
 }
 
-format_output <- function(x) {
+format_output <- function(x, ...) {
     sprintf("%.1f", x)
 }
 

### 2017-08-14 manuscript.Rmd
revise first part of results

@@
 For marine organisms, because of the diversity of their complex life history strategies, their size range, and the difficulties associated with sampling their habitats, we have limited information about which species live where, or what portion of their diversity remains to be discovered and described [@Mora2011].
-Additionally, coordinated marine ecosystem monitoring initiatives are lacking [@Duffy2013,@Borja2016] (but see MarineGEO and PISCO), making it difficult to monitor changes in species distributions and abundances.
+Additionally, coordinated marine ecosystem monitoring initiatives have been lacking [@Duffy2013,@Borja2016] (but see MarineGEO [@Duffy2016marinegeo] and PISCO), making it difficult to monitor changes in species distributions and abundances.
 While most documented global extinctions of marine species are vertebrates, many others species may have been lost unnoticed because they were not known [@Sala2006].
@@
 
-Regional faunistic lists assembled and curated by taxonomic experts provide a comprehensive and detailed account of species that are known to occur in a given geographic area.
-We assessed how faunistic lists assembled over decades by specialists compare with records in public databases for the same geographic area.
-We parsed faunistic lists to extract marine invertebrates from the Gulf of Mexico [@Tunnell2009] and from the Pacific Northwest [@Kozloff1996], leading to `r format(sum(gom_worms$is_marine, na.rm=TRUE), big.mark=",")` and `r format(sum(kozloff_species$is_marine, na.rm=TRUE), big.mark=",")` species respectively.
-Our goals were (1) to compare  the species lists for these geographic areas assembled by experts with the records found in databases; (2) to assess the proportion of species included in the expert lists that have records in public databases.
-
-To compare the faunistic lists from the Gulf of Mexico and the Pacific Northwest with records in public databases, we queried the iDigBio (https://portal.idigbio.org) and OBIS (http://www.iobis.org/) databases for all records they included in the same geographic boundaries as the lists.
-We filtered the results to only include the same higher taxonomic level taxa as in the faunistic lists, and kept only records for which the species could be matched to a species name in WoRMS.
-For the Gulf of Mexico, we retrieved  `r format(dplyr::n_distinct(idigbio_gom_records$worms_valid_name), big.mark=",")` species in iDigBio (from $n =$  `r format(nrow(idigbio_gom_records), big.mark=",")` records), and `r format(dplyr::n_distinct(obis_gom_records$worms_valid_name), big.mark=",")` species in OBIS ($n=$ `r format(nrow(obis_gom_records), big.mark=",")` records).
-For the Pacific Northwest, we retrieved `r format(dplyr::n_distinct(idigbio_pnw_records$worms_valid_name), big.mark=",")` species in iDigBio ($n=$ and `r format(nrow(idigbio_pnw_records), big.mark=",")` records) and `r format(dplyr::n_distinct(obis_pnw_records$worms_valid_name), big.mark=",")` species in OBIS ($n=$ `r format(nrow(obis_pnw_records), big.mark=",")` records).
-Overall, the number of species known from the Gulf of Mexico match the number estimated by experts (`r format(11150-1541-434, big.mark=",")` species [@Fautin2010]).
-For the Pacific coast, this number is lower than the estimate (`r format(8374-909-134, big.mark=",")` species [@Fautin2010]), but this estimate included species from California and from deeper areas that were not included in Kozloff's faunistic list.
-
-<!-- this paragraph needs overhaul to make it clearer -->
-For the Gulf of Mexico, the majority of the species included in the expert faunistic list are recorded in both iDigBio and OBIS (`r format_output(species_in_common$n_spp_gom_idigbio_obis/species_in_common$n_spp_gom*100)` %), or only OBIS (`r format_output(species_in_common$n_spp_gom_obis/species_in_common$n_spp_gom*100)` %) (Fig.
-\@ref(fig:database-coverage)A).
-For the Pacific Northwest, these proportions were much lower with only `r format_output(species_in_common$n_spp_kozloff_idigbio_obis/species_in_common$n_spp_kozloff*100)`% of the species from the regional list with records in OBIS and iDigBio, and `r format_output(species_in_common$n_spp_kozloff_obis/species_in_common$n_spp_kozloff*100)`%  in OBIS only (Fig.
-\@ref(fig:database-coverage)B).
-Actually, the majority of the species in the faunistic list are not recorded in any of the public databases.
-Overall, there is more overlap among the data sources for the Gulf of Mexico than the Pacific Northwest.
+We extracted `r format(sum(gom_worms$is_marine, na.rm=TRUE), big.mark=",")` and `r format(sum(kozloff_species$is_marine, na.rm=TRUE), big.mark=",")` marine invertebrate species from species checklists from the Gulf of Mexico [@Tunnell2009] and from the Pacific Northwest [@Kozloff1996] respectively.
+For the Gulf of Mexico, biodiversity databases included more species for the same geographic area as the one used by the authors of the checklist (iDigBio and OBIS combined: `r format(dplyr::n_distinct(combined_gom_records$worms_valid_name), big.mark=",")` species; iDigBio only: `r format(n_distinct(idigbio_gom_records$worms_valid_name), big.mark=",")` species, $n =$  `r format(nrow(idigbio_gom_records), big.mark=",")` records; OBIS only: `r format(n_distinct(obis_gom_records$worms_valid_name), big.mark=",")` species, $n=$ `r format(nrow(obis_gom_records), big.mark=",")` records).
+For the PNW, fewer species than included in the regional checklist were recovered from biodiversity databases (iDigBio and OBIS combined: `r format(dplyr::n_distinct(combined_pnw_records$worms_valid_name), big.mark=",")` species; iDigBio only: `r format(dplyr::n_distinct(idigbio_pnw_records$worms_valid_name), big.mark=",")`  species, $n=$ `r format(nrow(idigbio_pnw_records), big.mark=",")` records; OBIS only `r format(dplyr::n_distinct(obis_pnw_records$worms_valid_name), big.mark=",")` species, $n=$ `r format(nrow(obis_pnw_records), big.mark=",")` records).
+The combination of the species listed in these three sources leads to `r n_distinct(total_gom_records$worms_valid_name)` species for the GOM and `r n_distinct(total_pnw_records$worms_valid_name)` species for the PNW.
 
@@
 
+Digitization of species richness is higher for the GOM compared with the PNW.
+A higher proportion of the species included in the regional list for the GOM are captured in biodiversity databases than for the PNW.
+While `r format_output((3890 + 2663 + 1025 + 316 + 102 + 49 )/n_distinct(total_gom_records$worms_valid_name) * 100)`%  of the species included in the checklist for the GOM have records in at least one of the biodiversity database (Fig.
+\@ref(fig:database-coverage)A), `r format_output((352 + 292 + 281 + 235 + 76 + 38)/n_distinct(total_pnw_records$worms_valid_name)*100)`% do for the PNW  (Fig.
+\@ref(fig:database-coverage)B).
+Additionally, a higher proportion of species occur in the biodiversity databases but are not included in the regional checklist for the GOM compared with the PNW (GOM: `r format_output((1253 + 848 + 556 + 216 + 190 + 124) / n_distinct(total_gom_records$worms_valid_name)*100)`%, $n =$ `r n_distinct(total_gom_records$worms_valid_name)`; PNW:
+`r format_output( (281 + 151 + 145 + 80 + 78 + 45) / n_distinct(total_pnw_records$worms_valid_name)*100)`%, $n =$ `r n_distinct(total_pnw_records$worms_valid_name)`).
+
+```{r compare_database_overlap, fig.cap='Proportions of the total number of species by phylum that are absent from faunistic lists but are in iDigBio or OBIS ("Not in list"), and absent from iDigBio or OBIS but are in the faunistic lists ("Not in databases").
+The numbers represent the number of species.
+(\\#fig:database-overlap)'}
+knitr::include_graphics("figures/database_overlap_percent.pdf")
+```
+
 <!-- need to separate across all phyla statements and by phylum statements -->
@@
 
-```{r compare_database_overlap, fig.cap='Proportions of the total number of species by phylum that are absent from faunistic lists but are in iDigBio or OBIS ("Not in list"), and absent from iDigBio or OBIS but are in the faunistic lists ("Not in databases").
-The numbers represent the number of species.
-(\\#fig:database-overlap)'}
-knitr::include_graphics("figures/database_overlap_percent.pdf")
-```
+Regional faunistic lists assembled and curated by taxonomic experts provide a comprehensive and detailed account of species that are known to occur in a given geographic area.
+We assessed how faunistic lists assembled over decades by specialists compare with records in public databases for the same geographic area.
 
@@
 
+<!-- Our goals were (1) to compare  the species lists for these geographic areas assembled by experts with the records found in databases; (2) to assess the proportion of species included in the expert lists that have records in public databases.
+-->
+
+<!-- First, we compared the faunistic lists from the Gulf of Mexico and the Pacific Northwest with records in public databases.
+we queried the iDigBio (https://portal.idigbio.org) and OBIS (http://www.iobis.org/) databases for all records they included in the same geographic boundaries as the lists.
+We filtered the results to only include the same higher taxonomic level taxa as in the faunistic lists, and kept only records for which the species could be matched to a species name in WoRMS.
+For the Gulf of Mexico, we retrieved -->
+
 For geographic searches, we obtained the boundaries of the Exclusive Economic Zone for the United States from marineregions.org (mrgid:8456) [@mregions].

### 2017-08-14 R/map_sampling.R
compute full list of species

@@ -28,6 +28,17 @@ combine_all_sources <- function(...) {
         deduplicate_records()
 }
 
+get_species_list <- function(...) {
+    d <- lapply(list(...), function(dt) {
+        dt %>%
+            dplyr::distinct(worms_phylum, worms_valid_name)
+    })
+    dplyr::bind_rows(d) %>%
+        dplyr::distinct(worms_phylum, worms_valid_name) %>%
+        dplyr::filter(!is.na(worms_valid_name) | !is.na(worms_phylum))
+}
+
+
 data_map_samples <- function(recs) {
     us_raster <- us_raster()
     pts <- SpatialPoints(data.frame(lon = recs$decimallongitude,

### 2017-08-14 R/map_sampling.R
calculate Shannon's index (but don't use it)

@@ -54,11 +54,25 @@ data_map_diversity <- function(recs) {
     raster_cell <- mapply(function(x, y) cellFromXY(us_raster, c(x, y)),
                           recs$decimallongitude, recs$decimallatitude)
 
-    recs_r <- data.frame(recs, rastercell = raster_cell) %>%
+    recs_r <- dplyr::bind_cols(recs, rastercell = raster_cell)
+    recs_spp <-  recs_r %>%
         group_by(rastercell) %>%
         summarize(
             n_spp = n_distinct(worms_valid_name)
         )
+
+    ## not in use
+    h_index <- recs_r %>%
+        dplyr::select(phylum, worms_valid_name, rastercell) %>%
+        dplyr::group_by(rastercell, worms_valid_name) %>%
+        dplyr::summarize(n_records_per_spp = n()) %>%
+        dplyr::group_by(rastercell) %>%
+        dplyr::mutate(n_records_per_cell = sum(n_records_per_spp),
+                      p_i = n_records_per_spp/n_records_per_cell) %>%
+        dplyr::group_by(rastercell) %>%
+        dplyr::summarize(h_idx = - sum(p_i * log(p_i, 2)))
+
+    recs_r <- dplyr::full_join(recs_spp, h_index, by = "rastercell")
     us_raster[na.omit(recs_r$rastercell)] <- recs_r$n_spp[!is.na(recs_r$rastercell)]
     gg_r <- as.data.frame(as(us_raster, "SpatialPixelsDataFrame"))
     colnames(gg_r) <- c("value", "x", "y")

### 2017-08-14 manuscript.Rmd
edit manuscript

@@
 
-Regional faunistic lists assembled and curated by taxonomic experts provide a comprehensive and detailed account of species that are known to occur in a given geographic area.
+<!-- Regional faunistic lists assembled and curated by taxonomic experts provide a comprehensive and detailed account of species that are known to occur in a given geographic area.
 We assessed how faunistic lists assembled over decades by specialists compare with records in public databases for the same geographic area.
+-->
 
-<!-- need to separate better baises for which I have data and those for which I don't -->
-Species that are included in the list but do not have records in databases could be (1) rare: they have never been collected or have been collected only a few times, and the corresponding records have not been digitized yet; (2) difficult to preserve: large, very small, or fragile species are deposited in museums less frequently; (3) are in the database under a different name: the species has been synonymized or re-identified since the publication of the list under a different name that may be included in the database; (4) at the limit of their range: the species were included in the list because they are known to occur within the boundaries of the geographic area, but they have never been collected within these geographic boundaries; (5) do not have geographic coordinates associated with them: some of these species can be in the database but without georeferenced data points and therefore were not returned in our queries.
+Several factors may contribute for species to be included in the regional lists but to not have records in biodiversity databases.
+The first category of biases are linked to property of the species affecting the likelihood of having specimens preserved in museum collections.
+For instance, rare species may never have been collected or have only been collected a few times, and the corresponding records have not been digitized yet.
+Species that are difficult to preserve because they are large, very small, or fragile, are deposited in museums less frequently.
+The other category of biases is associated with data quality for these records.
+Some of these species may be in the databases but do not have geographic coordinates associated with them, and therefore would not be included in the results of our query.
 
-Untangling the relative contribution of these factors to explain the discrepancies is difficult as it would require to know the criteria used by the experts who have assembled the lists.
-However, by querying iDigBio we found that for the `r format(not_really_in_database$idig_gom_n_spp_not_in_db, big.mark=",")` species that do not appear within the geographic range of the Gulf of Mexico but are on the list, `r format_output(not_really_in_database$idig_gom_p_spp_not_in_db*100)`% are completely absent from iDigBio across all areas, while for the `r format(not_really_in_database$idig_gom_n_spp_no_coords, big.mark=",")` species that appear in iDigBio, `r format_output(not_really_in_database$idig_gom_p_spp_no_coords*100)`%  do not have geographic coordinates associated with them, and for the ones that do have geographic coordinates, `r format_output(not_really_in_database$idig_gom_p_spp_within_usa*100)`% ($n=$  `r format(not_really_in_database$idig_gom_n_spp_within_usa, big.mark=",")`) fall within the US waters.
-These numbers indicate that the lack of geographic coordinates associated with the specimens in the database contribute at least significantly to the discrepancy, but that a large fraction of these species have no digital records of their occurrence in the Gulf of Mexico.
-In other words, additional digitization of collections (including capturing their geographic coordinates information) are needed.
+Missing geographic information contribute to the discrepencies observed.
+Using iDigBio we found that for the `r format(not_really_in_database$idig_gom_n_spp_not_in_db, big.mark=",")` species that do not appear within the geographic range of the Gulf of Mexico but are on the list, `r format_output(not_really_in_database$idig_gom_p_spp_not_in_db*100)`% are completely absent from iDigBio, while for the `r format(not_really_in_database$idig_gom_n_spp_no_coords, big.mark=",")` species that do appear in iDigBio, `r format_output(not_really_in_database$idig_gom_p_spp_no_coords*100)`%  do not have geographic coordinates associated with them, and for the ones that do have geographic coordinates, `r format_output(not_really_in_database$idig_gom_p_spp_within_usa*100)`% ($n=$  `r format(not_really_in_database$idig_gom_n_spp_within_usa, big.mark=",")`) fall within the US waters.
+These numbers indicate that missing geographic coordinates reduces the overlap between the list of species from the regional list and the databases.
+Additional digitization of collections, including the capture of their geographic coordinates information, are needed to improve biodiversity knowledge at regional scales.
 For instance, one can expect that the digitization coverage for the Gulf of Mexico will greatly increase once records from the Florida Wildlife Research Institute collections (St Petersburg, FL) become integrated into biodiversity information systems.
 
-Species that are in the database but not in the faunistic list could be: (1) records not considered by the authors of the checklist: either because they were not accessible or known to the authors; (2) records erroneously excluded by the author of the checklist; (3) records that have been discovered since the list has been assembled; (4) records with incorrect identification: records identified as a species that does not occur in the area; (5) records with incorrect geographic coordinates: their associated geographic information in the database places them in the Gulf of Mexico or the Pacific Northwest, but they were actually collected elsewhere.
+Two factors are most likely to contribute to explain why some species are in databases but were not considered by the authors of the checklists: (1) the species have been documented for the first time since the publication of the checklist, (2) the species records were not accessible or known to the authors.
 
-We investigated the proportion of species discovered since the publication of the faunistic lists using iDigBio and OBIS records.
-They represent `r sum(recently_collected$gom_in_dbs$min_year > 2001)` species for the Gulf of Mexico, and `r sum(recently_collected$pnw_in_dbs$min_year > 1996)` species for the Pacific Northwest.
-While errors in geographic coordinates cannot be completely ruled out, it is more likely that the main contributing factor was that the authors of these checklists were not aware of the existence of these species based on museum records, as they were difficult to discover before digitization which has occured after the compilation of the list in most cases.
-However, museum specimens are critical to validate the occurrence of species [@Rocha2014], especially for marine invertebrates for which species limits based on morphology have been known to be challenged by other lines of evidence including molecular data [@Knowlton2000].
+<!-- include also proportion of species not in list -->
+Recently recorded species contribute modestly to this discrepency, as `r sum(recently_collected$gom_in_dbs$min_year > 2001)` species (`r format_output( sum(recently_collected$gom_in_dbs$min_year > 2001)/n_distinct(total_gom_records$worms_valid_name) * 100)`% of the total species richness) have been collected for the first time since the publication of the list in the Gulf of Mexico, and `r sum(recently_collected$pnw_in_dbs$min_year > 1996)` species (`r format_output(sum(recently_collected$pnw_in_dbs$min_year > 1996)/n_distinct(total_pnw_records$worms_valid_name) * 100)`% of the total species richness) for the Pacific Northwest.
+
+The lack to tools for the authors of these checklist to have access to museum records across institutions contributed most to explain why some species were not included in the checklists.
+These records were difficult to discover before digitization which, for the most part, has occured after the compilation of these checklists in most cases.
 
@@
 <!-- separate idigbio and (and add) obis in the results -->
-We combined `r format(nrow(idigbio_records), big.mark=",")` museum records for marine invertebrates collected within the EEZ of the continental US representing `r length(unique(tolower(idigbio_records[["phylum"]])))` phyla and `r format(length(unique(idigbio_records$worms_valid_name)), big.mark=",")` species from iDigBio.
-Most samples were collected on the continental shelf (Fig.
+
+We retrieved `r format(n_distinct(combined_species$worms_valid_name), big.mark = ",")` species (`r n_distinct(na.omit(combined_species$worms_phylum))` phyla) of marine invertebrates in the continental US EEZ by combining  `r format(nrow(idigbio_records), big.mark=",")` museum records from iDigBio (`r format(n_distinct(idigbio_records$worms_valid_name), big.mark = ",")` species, `r n_distinct(na.omit(idigbio_records$worms_phylum))` phyla), and `r format(nrow(obis_records), big.mark=",")` records from OBIS (`r format(n_distinct(obis_records$worms_valid_name), big.mark=",")` species,
+`r n_distinct(na.omit(obis_records$worms_phylum))` phyla).
+Extant marine invertebrates occurring in the US represent less than 0.4% of the total iDigBio records, as it currently holds records for over 106 million specimens, and they represent about 1% of the records that occur in the United States (about 33 million records in iDigBio).
+Marine invertebrates that occur in the US represent approximately 4% of the 48 million occurrence records in OBIS.
+
+<!-- need to add latitudinal gradient map -->
+Most samples have been collected on the continental shelf (Fig.
 \@ref(fig:map-diversity-sampling)A).
-For the East coast, species richness is the highest in South Florida but it is relatively uniform along the coastline.
-For the West coast, the Pacific Northwest, the San Francisco Bay Area, and Southern California have the highest richness (Fig.
+On the East coast, species richness is highest at low latitudes, while it increases towards high latitude on the West coast (Fig.
 \@ref(fig:map-diversity-sampling)B).
@@
 (\\#fig:map-diversity-sampling)'}
-knitr::include_graphics("figures/map_diversity_sampling.pdf")
+knitr::include_graphics("figures/combined_map_sampling.pdf")
 ```
 
-Spatial patterns of species richness recovered are strongly confounded by sampling effort.
-Samples of <100 records per cell commonly have as many species as records (Fig.
-\@ref(fig:correlation-sampling)).
+Spatial patterns of species richness are strongly confounded by sampling effort. Samples of <100 records per cell commonly have as many species as records (Fig. \@ref(fig:correlation-sampling)).
 
@@
 
-Underscoring the inadequacy of the sampling effort, `r prop_singleton_species`% of the species are known from a single record in iDigBio.
+Underscoring the inadequacy of the sampling effort, `r prop_singleton_species`% of the species are known from a single record.
 Another indication of this inadequacy is the relation between the number of collected specimen and the number of species recorded through time (Fig.
@@
 
+ However, museum specimens are critical to validate the occurrence of species [@Rocha2014], especially for marine invertebrates for which species limits based on morphology have been known to be challenged by other lines of evidence including molecular data [@Knowlton2000].
+
 This study highlights that we have too little data about too few species of marine invertebrates.

### 2017-08-17 R/comparison_by_geo.R
use worms_phylum instead of phylum in db comparisons (towards #53)

@@ -3,7 +3,7 @@ compare_with_geo <- function(spp_list, geo_list, verbose = FALSE) {
             dplyr::filter(is_binomial == TRUE, is_marine == TRUE) %>%
             dplyr::distinct(worms_valid_name, .keep_all = TRUE) %>%
             dplyr::select(worms_id, worms_valid_name,
-                          phylum) %>%
+                          worms_phylum) %>%
             dplyr::filter(!is.na(worms_id))
     list(
         ## in spp_list but not in geo_list
@@ -86,18 +86,18 @@ plot_database_overlap <- function(comp_db) {
                        "echinodermata", "chordata")
 
     comp_db <- comp_db %>%
-        dplyr::mutate(phylum = tolower(phylum)) %>%
-        dplyr::filter(phylum %in% phyla_to_keep)
+        dplyr::mutate(worms_phylum = tolower(worms_phylum)) %>%
+        dplyr::filter(worms_phylum %in% phyla_to_keep)
 
     ## to have 2 sided plot, we take the negative value of the count
     ## for the records that are NOT in the geographical range in the
     ## database.
     data_for_plot <- comp_db %>%
-        dplyr::count(region, database, data_source, phylum) %>%
+        dplyr::count(region, database, data_source, worms_phylum) %>%
         dplyr::mutate(number = ifelse(data_source == "not_in_geo", -n, n)) %>%
         dplyr::select(-n) %>%
         tidyr::spread(data_source, number) %>%
-        dplyr::mutate(phylum = capitalize(phylum))
+        dplyr::mutate(phylum = capitalize(worms_phylum))
 
     ## Clean up phylum names to use as labels in plot
     phylum_label <- data_for_plot %>%
@@ -176,19 +176,19 @@ plot_database_overlap_percent <- function(comp_db, full_list) {
                        "echinodermata", "chordata")
 
     comp_db <- comp_db %>%
-        dplyr::mutate(phylum = tolower(phylum)) %>%
-        dplyr::filter(phylum %in% phyla_to_keep)
+        dplyr::mutate(worms_phylum = tolower(worms_phylum)) %>%
+        dplyr::filter(worms_phylum %in% phyla_to_keep)
 
     spp_per_phylum <- full_list %>%
-        dplyr::count(region, phylum) %>%
+        dplyr::count(region, worms_phylum) %>%
         dplyr::rename(n_spp = n)
 
     ## to have 2 sided plot, we take the negative value of the count
[…]

### 2017-08-17 manuscript.Rmd
rename target within manuscript

@@
 
-We extracted `r format(sum(gom_worms$is_marine, na.rm=TRUE), big.mark=",")` and `r format(sum(kozloff_species$is_marine, na.rm=TRUE), big.mark=",")` marine invertebrate species from species checklists from the Gulf of Mexico [@Tunnell2009] and from the Pacific Northwest [@Kozloff1996] respectively.
+We extracted `r format(sum(gom_worms$is_marine, na.rm=TRUE), big.mark=",")` and `r format(sum(kozloff_worms$is_marine, na.rm=TRUE), big.mark=",")` marine invertebrate species from species checklists from the Gulf of Mexico [@Tunnell2009] and from the Pacific Northwest [@Kozloff1996] respectively.
 For the Gulf of Mexico, biodiversity databases included more species for the same geographic area as the one used by the authors of the checklist (iDigBio and OBIS combined: `r format(dplyr::n_distinct(combined_gom_records$worms_valid_name), big.mark=",")` species; iDigBio only: `r format(n_distinct(idigbio_gom_records$worms_valid_name), big.mark=",")` species, $n =$  `r format(nrow(idigbio_gom_records), big.mark=",")` records; OBIS only: `r format(n_distinct(obis_gom_records$worms_valid_name), big.mark=",")` species, $n=$ `r format(nrow(obis_gom_records), big.mark=",")` records).

### 2017-08-28 manuscript.Rmd
better text about number of samples and diversity

@@
 
-<!-- add sentence -->
 <!-- separate idigbio and (and add) obis in the results -->
@@
 `r n_distinct(na.omit(obis_records$worms_phylum))` phyla).
-Extant marine invertebrates occurring in the US represent less than 0.4% of the total iDigBio records, as it currently holds records for over 106 million specimens, and they represent about 1% of the records that occur in the United States (about 33 million records in iDigBio).
+Extant marine invertebrates occurring in the US represent less than 0.4% of the total number of iDigBio records (106 million), and about 1% of the US records (33 million records).
 Marine invertebrates that occur in the US represent approximately 4% of the 48 million occurrence records in OBIS.
 
-<!-- need to add latitudinal gradient map -->
+```{r local-richness}
+count_spp <- . %>%
+    dplyr::pull(worms_valid_name) %>%
+    dplyr::n_distinct()
+
+n_spp_gom <- combined_records %>%
+    dplyr::filter(is_in_gom == TRUE) %>%
+    count_spp
+
+n_spp_east <- combined_records %>%
+    dplyr::filter(is_east_coast == TRUE) %>%
+    count_spp
+
+n_spp_west <- combined_records %>%
+    dplyr::filter(is_west_coast == TRUE) %>%
+    count_spp
+```
+
 Most samples have been collected on the continental shelf (Fig.
 \@ref(fig:map-diversity-sampling)A).
-On the East coast, species richness is highest at low latitudes, while it increases towards high latitude on the West coast (Fig.
-\@ref(fig:map-diversity-sampling)B).
+More records have been collected from the Atlantic coast, than the Gulf of Mexico, and the Pacific coast (`r format(sum(combined_records$is_east_coast), big.mark=",")`, `r format(sum(combined_records$is_in_gom), big.mark=",")`, and `r format(sum(combined_records$is_west_coast), big.mark=",")` records respectively).
+The highest species richness has been recorded from the Gulf of Mexico (`r format(n_spp_gom, big.mark=",")` species), followed by the Atlantic coast (`r format(n_spp_east, big.mark=",")` species) an d the Pacific coast (`r format(n_spp_west, big.mark=",")` species).
+
+<!-- need to add latitudinal gradient map -->
+On the East coast, species richness is highest at low latitudes, while it increases towards high latitude on the West coast (Fig. \@ref(fig:map-diversity-sampling)B).
 

### 2017-08-28 R/map_sampling.R

@@ -12,12 +12,30 @@ deduplicate_records <- function(data) {
                         decimallongitude, datecollected, .keep_all = TRUE)
 }
 
-combine_records <- function(...) {
+combine_records <- function(..., map) {
     d <- list(...)
     d <- year_as_integer(d)
     d %>%
         dplyr::bind_rows() %>%
-        deduplicate_records()
+        deduplicate_records() %>%
+        add_geo(map = map)
+}
+
+## add info on whether a record is on the East coast, West coast, or the Gulf of
+## Mexico.  In this function, if a record is in the GOM, it is not on the East
+## coast
+add_geo <- function(d, map) {
+    d <- d %>%
+        is_within_gom_records(map = map) %>%
+        dplyr::mutate(is_east_coast = if_else(decimallongitude > -100 &
+                                             !is_in_gom, TRUE, FALSE),
+                      is_west_coast = if_else(decimallongitude < -100,
+                                             TRUE, FALSE))
+
+    ## check that no record gets assigned more than one geographic area
+    if (any(d$is_in_gom + d$is_east_coast + d$is_west_coast) > 3)
+        stop("something's wrong!")
+    d
 }
 
 combine_species_list <- function(...) {

### 2017-08-29 manuscript.Rmd
add prop of cells with number of spp within 10% of the nb of samples

@@
 \@ref(fig:map-diversity-sampling)A).
-More records have been collected from the Atlantic coast, than the Gulf of Mexico, and the Pacific coast (`r format(sum(combined_records$is_east_coast), big.mark=",")`, `r format(sum(combined_records$is_in_gom), big.mark=",")`, and `r format(sum(combined_records$is_west_coast), big.mark=",")` records respectively).
-The highest species richness has been recorded from the Gulf of Mexico (`r format(n_spp_gom, big.mark=",")` species), followed by the Atlantic coast (`r format(n_spp_east, big.mark=",")` species) an d the Pacific coast (`r format(n_spp_west, big.mark=",")` species).
+More records have been collected from the Atlantic Ocean, than the Gulf of Mexico, and the Pacific Ocean (`r format(sum(combined_records$is_east_coast), big.mark=",")`, `r format(sum(combined_records$is_in_gom), big.mark=",")`, and `r format(sum(combined_records$is_west_coast), big.mark=",")` records respectively).
+The highest species richness has been recorded from the Gulf of Mexico (`r format(n_spp_gom, big.mark=",")` species), followed by the Atlantic Ocean (`r format(n_spp_east, big.mark=",")` species), and the Pacific Ocean (`r format(n_spp_west, big.mark=",")` species).
 
@@
 
-Spatial patterns of species richness are strongly confounded by sampling effort. Samples of <100 records per cell commonly have as many species as records (Fig. \@ref(fig:correlation-sampling)).
-
 ```{r correlation-sampling, fig.cap='Relation between number of records and number of species for each cell of the map used in Fig.
@@
 
+Spatial patterns of species richness are strongly confounded by sampling effort.
+There is a strong correlation between the number of samples and number of species per cell, and `r format_output(stat_sampling_effort$p * 100)`% (n = `r stat_sampling_effort$n`) of the cells with $\leq$ 500 records have a number of species within 10% of the number of records (Fig.
+\@ref(fig:correlation-sampling)).
+
+Environmental conditions are among the primary determinants of species distributions and can thus be used to estimate the response of organisms to climate change.
+Predictive ecological models forecasting changes in distributions are training on environmental conditions encountered by species in their current distributions.
+To obtain reliable results, models require at least 10, but ideally 30 or more occurrence records per species [@Wisz2008].
+Yet at present, `r prop_10_specimens_species`% of species are represented by $\leq$ 10 records, and `r prop_30_specimens_species`%  by $\leq$ 30 records based on the combined records in iDigBio and OBIS.
 Underscoring the inadequacy of the sampling effort, `r prop_singleton_species`% of the species are known from a single record.
+
 Another indication of this inadequacy is the relation between the number of collected specimen and the number of species recorded through time (Fig.
 \@ref(fig:idigbio-through-time)).
-Initial explorations in the late ninetieth century led to an initial exponential growth in the knowledge of the fauna, that tapered off, until the 1970s when intensive collecting efforts led to an increase in the recording of new taxa (Fig.
+Explorations in the late ninetieth century led to an exponential growth in the knowledge of the fauna, that tapered off, until the 1970s when intensive collecting efforts led to an increase in the recording of new taxa (Fig.
 \@ref(fig:idigbio-through-time)).
@@
 ```
-
-Environmental conditions are among the primary determinants of species distributions and can thus be used to estimate the response of organisms to climate change.
-Predictive ecological models forecasting changes in distributions are training on environmental conditions encountered by species in their current distributions.
-To obtain reliable results, models require at least 10, but ideally 30 or more occurrence records per species [@Wisz2008].
-Yet at present, `r prop_10_specimens_species`% of species are represented by $\leq$ 10 records, and `r prop_30_specimens_species`%  by $\leq$ 30 records based on the combined records in iDigBio and OBIS.
 

### 2017-08-29 manuscript.Rmd
[fix #58] include global estimate for fraction of species barcoded

@@
 
-<!-- include results earlier? -->
+<!-- add overall mean number before getting into details -->
 
-To estimate the proportion of the marine invertebrate diversity for which DNA barcodes are available, we queried BOLD (Barcode of Life Datasystem) for all species listed in the regional lists for the Gulf of Mexico, the Pacific Northwest, as well as all the records for marine invertebrates that occur in the EEZ of the US found in iDigBio.
-
-<!-- add overall mean number before getting into details -->
-The proportion of species within the US EEZ that have DNA barcodes varies from `r format_output(stat_barcoding$min_eez_value * 100)`% for `r stat_barcoding$min_eez_phylum` to `r format_output(stat_barcoding$max_eez_value * 100)`% for `r stat_barcoding$max_eez_phylum` (Fig.
+The marine invertebrates of the US have generally been characterized genetically at a higher level than other parts of the world: `r format_output(prop_spp_barcoded$p_barcoded * 100)`% (n = `r format(nrow(combined_species), big.mark=",")`) of the fauna has DNA barcodes available.
+In 2012, it was estimated that, at a global scale, less than 3% of the marine species had been sequenced [@Appeltans2012], and that this number was even lower in some areas (e.g., 2.6% for the Antarctic [@Grant2009]).
+This proportion varies by phylum from `r format_output(stat_barcoding$min_eez_value * 100)`% for `r stat_barcoding$min_eez_phylum` to `r format_output(stat_barcoding$max_eez_value * 100)`% for `r stat_barcoding$max_eez_phylum` (Fig.
 \@ref(fig:proportion-barcoded)).
@@
 This is driven in part by the active Canadian barcoding effort in British Columbia.
-In 2012, it was estimated that, at a global scale, less than 3% of the marine species had been sequenced [@Appeltans2012], and that this number was even lower in some areas (e.g.
-2.6% for the Antarctic [@Grant2009]).
-Thus the US fauna is generally better characterized than other parts of the world.
 
@@
 
+<!-- include results earlier? -->
+
+To estimate the proportion of the marine invertebrate diversity for which DNA barcodes are available, we queried BOLD (Barcode of Life Datasystem) for all species listed in the regional lists for the Gulf of Mexico, the Pacific Northwest, as well as all the records for marine invertebrates that occur in the EEZ of the US found in iDigBio.
+
 To assess the proportion of species for which DNA barcodes are available, we queried the Barcode of Life Datasystem (BOLD) using the R package bold [@bold-0.4.0].

### 2017-08-29 R/bold.R

@@ -187,6 +187,14 @@ plot_bold_status <- function(bold_data) {
         coord_flip()
 }
 
+calc_prop_spp_barcoded <- function(recs) {
+    res <- recs$n_bold_records > 0
+    list(
+        p_barcoded = mean(res),
+        n_barcoded = sum(res)
+    )
+}
+
 make_stat_barcoding <- function(bold_data) {
     list(
         min_eez_phylum = bold_data %>%

### 2017-08-29 R/map_sampling.R

@@ -51,7 +51,8 @@ combine_species_list <- function(...) {
                        dplyr::filter(!is.na(worms_phylum) | !is.na(worms_phylum))
                }, .id = "data_source") %>%
         dplyr::count(data_source, worms_phylum, worms_valid_name) %>%
-        tidyr::spread(data_source, n)
+        tidyr::spread(data_source, n) %>%
+        find_bold_records(col_nm = "worms_valid_name")
 }
 
 

### 2017-08-29 manuscript.Rmd
add latitudinal gradient figure

@@
 
-<!-- need to add latitudinal gradient map -->
-On the East coast, species richness is highest at low latitudes, while it increases towards high latitude on the West coast (Fig. \@ref(fig:map-diversity-sampling)B).
+Spatial distribution of species richness with latitude differ markedly between the Pacific and the Atlantic coasts.
+On the Pacific coast, richness is lowest around 40° of latitude, while it is where it is the highest on the Atlantic coast (Fig.
+\@ref(fig:map-diversity-sampling)B, Fig.
+\@ref(fig:latitudinal-gradient)).
 
@@
 knitr::include_graphics("figures/combined_map_sampling.pdf")
+```
+
+```{r latitudinal-gradient, fig.cap='Latitudinal gradient in species richness for the Pacific and Atlantic oceans (Gulf of Mexico records excluded).
+Trend lines are B-splines of the 3rd degree polynomial weighted by the square-root of the number of records.
+(\\#fig:latitudinal-gradient)'}
+knitr::include_graphics("figures/combined_gradient.pdf")
 ```
@@
 
-Spatial patterns of species richness are strongly confounded by sampling effort.
-There is a strong correlation between the number of samples and number of species per cell, and `r format_output(stat_sampling_effort$p * 100)`% (n = `r stat_sampling_effort$n`) of the cells with $\leq$ 500 records have a number of species within 10% of the number of records (Fig.
+Spatial patterns of species richness are however strongly confounded by sampling effort.
+There is a high correlation between the number of samples and number of species per cell used in Fig.
+\@ref(fig:map-diversity-sampling), and `r format_output(stat_sampling_effort$p * 100)`% (n = `r stat_sampling_effort$n`) of the cells with $\leq$ 500 records have a number of species within 10% of the number of records (Fig.
 \@ref(fig:correlation-sampling)).
@@
 Environmental conditions are among the primary determinants of species distributions and can thus be used to estimate the response of organisms to climate change.
-Predictive ecological models forecasting changes in distributions are training on environmental conditions encountered by species in their current distributions.
+Predictive ecological models forecasting changes in distributions are trained on environmental conditions encountered by species in their current distributions.
 To obtain reliable results, models require at least 10, but ideally 30 or more occurrence records per species [@Wisz2008].
@@
 
-Another indication of this inadequacy is the relation between the number of collected specimen and the number of species recorded through time (Fig.
+Another indication of the inadequacy of the sampling is the relation between the number of collected specimen and the number of species recorded through time (Fig.
 \@ref(fig:idigbio-through-time)).
@@
 \@ref(fig:idigbio-through-time)C), suggest that any new sampling effort will lead to new observations and new species.
-<!-- separate issues of sampling is needed, and large-scale sampling is best way of doing it --> Coordinated large-scale biodiversity expeditions continue to be needed to improve our understanding of the diversity of species and their spatial distributions.
-The US marine EEZ is relatively well-known compared to many areas of the world ocean, where studies show that undescribed biodiversity dominates.
-For example, a recent survey in Western Australia reported that 94.6% of crustaceans and 72% of polychaetes taken at 135 sites were undescribed [@Poore2014].
 
@@
 Of `r n_idigbio_phyla` phyla represented in iDigBio, `r records_rare_phyla$n_rare_phyla` have <500 records, and make up `r records_rare_phyla$prop_recs_rare_phyla`% of the iDigBio records.
-This underrepresentation is manifested at the species level, while `r records_rare_phyla$prop_spp_rare_phyla_collected`% (n = `r format(records_rare_phyla$total_spp_rare_phyla_collected, big.mark=",")` total species) of the global number of species for the `r records_rare_phyla$n_rare_phyla` phyla occur in iDigBio, this proportion rises to `r records_rare_phyla$prop_spp_common_phyla_collected`% (n = `r format(records_rare_phyla$total_spp_common_phyla_collected, big.mark=",")`).
+This underrepresentation is also manifested at the species level, while `r records_rare_phyla$prop_spp_rare_phyla_collected`% (n = `r format(records_rare_phyla$total_spp_rare_phyla_collected, big.mark=",")` total species) of the global number of species for these `r records_rare_phyla$n_rare_phyla` phyla occur in iDigBio, this proportion rises to `r records_rare_phyla$prop_spp_common_phyla_collected`% (n = `r format(records_rare_phyla$total_spp_common_phyla_collected, big.mark=",")`) for the `r n_idigbio_phyla - records_rare_phyla$n_rare_phyla` others.
 Yet, these undersampled taxa, especially those of small body size, are often the most abundant, and are likely to play an important role in the detection of changes in environmental quality[@Bik2012].
@@
 
-The proportion of samples that are identified to species show a marked decline in recent decades in most taxa other than mollusks, demonstrating limiting taxonomic effort and expertise (Fig.
+The proportion of samples that are identified to species in museum collections show a marked decline in recent decades in most taxa other than mollusks, demonstrating limiting taxonomic effort and expertise (Fig.
 \@ref(fig:identification-through-time)).
@@
 For diverse groups such as annelids or arthropods, the proportion of specimens identified to the species-level is now below 30%.
-The number of scientists describing new species might be increasing globally [@Costello2013], but taxonomic experts continue to be needed to identify specimens recently collected in the US.
 
@@
 ### Collect better data faster and more cost effectively
+
+<!-- separate issues of sampling is needed, and large-scale sampling is best way of doing it --> Coordinated large-scale biodiversity expeditions continue to be needed to improve our understanding of the diversity of species and their spatial distributions.
+The US marine EEZ is relatively well-known compared to many areas of the world ocean, where studies show that undescribed biodiversity dominates.
+For example, a recent survey in Western Australia reported that 94.6% of crustaceans and 72% of polychaetes taken at 135 sites were undescribed [@Poore2014].
 

### 2017-08-29 R/gradient.R

@@ -1,17 +1,53 @@
 plot_gradient <- function(idig) {
-
     idig %>%
-        dplyr::mutate(coast = if_else(decimallongitude > -100, "east", "west")) %>%
         dplyr::mutate(
-                   lat_grad = ntile(decimallatitude, 100)
+                   lat_grad = ntile(decimallatitude, 100),
+                   lat_cut = cut(decimallatitude, 100)
                ) %>%
-        dplyr::group_by(coast, phylum, lat_grad) %>%
-        dplyr::summarize(n_spp = n_distinct(cleaned_scientificname),
-                         n_obs = n()) %>%
-        dplyr::filter(phylum %in% c("arthropoda", "mollusca", "annelida", "echinodermata", "cnidaria")) %>%
-        dplyr::mutate(p_samp = n_spp/n_obs) %>%
-        ggplot(aes(x = lat_grad, y = n_spp, color = phylum, shape = coast, size = n_obs)) +
-        geom_point() + facet_grid(phylum ~ coast) +
-        geom_smooth(aes(weight = n_obs))
+        dplyr::mutate(coast = if_else(decimallongitude > -100, "Atlantic Ocean", "Pacific Ocean")) %>%
+        dplyr::mutate(coast = if_else(is_in_gom == TRUE, "gom", coast)) %>%
+        dplyr::mutate(coast = factor(coast, levels = c("Pacific Ocean", "Atlantic Ocean"))) %>%
+        dplyr::filter(coast != "gom") %>%
+        dplyr::group_by(coast, lat_grad) %>%
+        dplyr::summarize(
+                   lat_cat = mean(decimallatitude),
+                   n_spp = n_distinct(worms_valid_name),
+                   n_obs = n()
+               ) %>%
+        dplyr::mutate(p_samp = n_spp/(log(n_obs))) %>%
+        ggplot(aes(x = lat_cat, y = n_spp, color = coast, size = n_obs)) +
+        geom_point() + facet_grid( ~ coast) +
+        geom_smooth(method = "lm", formula = y ~ splines::bs(x, degree = 3),
+                    show.legend = FALSE, aes(weight = I(sqrt(n_obs)))) +
+        scale_size(range = c(0.2, 3),  name = "Number of observations") +
+        scale_colour_hc() + guides(color = FALSE) +
+        xlab("Latitude") + ylab("Number of Species")
+}
 
+####
+
+if (FALSE) {
+
[…]

### 2017-08-30 manuscript.Rmd
[fix #60, fix #61] use better data sources for prop of barcoded spp

@@
 
-```{r proportion-barcoded, fig.cap='Proportion of species known to occur in the US EEZ based on iDigBio records, in the Gulf of Mexico [@Tunnell2009], in the Pacific Northwest[@Kozloff1996], that have publicly available DNA barcodes.
+```{r proportion-barcoded, fig.cap='Proportion of species with publicly available DNA barcodes  known to occurin the US EEZ (iDigBio and OBIS records), in the Gulf of Mexico (iDigBio, OBIS, and checklist[@Tunnell2009]), and in the Pacific Northwest (iDigBio, OBIS, and checklist[@Kozloff1996]).
 Numbers next to the bars indicate the number of species for which DNA barcodes are available.

### 2017-08-30 R/bold.R

@@ -141,7 +141,7 @@ get_bold_bins <- function(gom, koz) {
 
 bold_status <- function(idig) {
     idig %>%
-        dplyr::group_by(phylum) %>%
+        dplyr::group_by(worms_phylum) %>%
         dplyr::summarize(
                    p_has_bold = mean(n_bold_records > 0),
                    n_has_bold = sum(n_bold_records > 0)
@@ -156,14 +156,14 @@ bold_status_data <- function(gom_bold, koz_bold, idig_bold) {
                .id = "data_source"
            ) %>%
     dplyr::filter(
-               phylum %in% c("annelida", "arthropoda",
+               worms_phylum %in% c("annelida", "arthropoda",
                              "bryozoa",
                              "chordata", "cnidaria",
                              "echinodermata", "mollusca",
                              "porifera"
                              )
            ) %>%
-    dplyr::mutate(phylum = capitalize(phylum))
+    dplyr::mutate(phylum = capitalize(worms_phylum))
 }
 
 plot_bold_status <- function(bold_data) {

### 2017-08-30 R/comparison_by_geo.R
add target for stats about species not included in checklists

@@ -317,7 +317,8 @@ get_species_in_common <- function(gom_worms, idigbio_gom_records, obis_gom_recor
 
 not_in_list_collected_recently <- function(database_overlap, map_gom, map_pnw) {
 
-    .idig <- . %>% fetch_spp_from_idigbio() %>%
+    .idig <- . %>%
+        fetch_spp_from_idigbio() %>%
         parse_year() %>%
         dplyr::mutate(
                    cleaned_scientificname = cleanup_species_names(scientificname),
@@ -399,8 +400,23 @@ not_in_list_collected_recently <- function(database_overlap, map_gom, map_pnw) {
         gom_in_dbs = spp_gom,
         pnw_in_dbs = spp_pnw
     )
-
 }
+
+n_spp_not_in_lists <- function(db_overlap) {
+    list(
+        gom_not_in_list = db_overlap %>%
+            dplyr::filter(region == "gom", data_source == "not_in_list") %>%
+            dplyr::pull(worms_valid_name) %>%
+            dplyr::n_distinct(),
+        pnw_not_in_list = db_overlap %>%
+            dplyr::filter(region == "pnw", data_source == "not_in_list") %>%
+            dplyr::pull(worms_valid_name) %>%
+            dplyr::n_distinct()
+    )
+}
+
+
+
 
 ## Species that appear to not be in iDigbio for the geographical list,
 ## either because they do not have GPS coordinates associated with

### 2017-08-30 manuscript.Rmd
better writing

--- 2017-08-29 manuscript.Rmd
@@
 More sampling, coordinated at the scale of the US, is needed to improve our knowledge of species and their geographic ranges.
-This data is needed to detect the impact of climate changes and other anthropogenic strains to the environment.
+This data is needed to detect the impact of climate change and other anthropogenic strains to the environment.
 Advances in computing, information systems, and sequencing technologies, have set the stage to inform where future efforts for documenting biodiversity should be applied.
@@
 For marine organisms, because of the diversity of their complex life history strategies, their size range, and the difficulties associated with sampling their habitats, we have limited information about which species live where, or what portion of their diversity remains to be discovered and described [@Mora2011].
-Additionally, coordinated marine ecosystem monitoring initiatives have been lacking [@Duffy2013,@Borja2016] (but see MarineGEO [@Duffy2016marinegeo] and PISCO), making it difficult to monitor changes in species distributions and abundances.
+Additionally, coordinated marine ecosystem monitoring initiatives have been lacking [@Duffy2013,@Borja2016] (but are now getting organized, see MarineGEO [@Duffy2016marinegeo], MBON, and PISCO), making it difficult to monitor changes in species distributions and abundances.
 While most documented global extinctions of marine species are vertebrates, many others species may have been lost unnoticed because they were not known [@Sala2006].
@@
 We show that while public databases are capturing a significant portion of our knowledge about species richness, most species are known from only a few specimens and a few locations.
-We emphasize that our understanding of marine biodiversity, and the impact of human activities on communities require not only to get a more accurate list of the species that inhabits the oceans, but more comprehensive data on species distributions and abundances are needed.
-These goals can only be achieved with additional cost-effective sampling and a commitment to training more taxonomists and experts that can identify and interpret the material collected.
+We emphasize that our understanding of marine biodiversity, and the impact of human activities on biological communities require not only to get a more accurate list of the species that inhabits the oceans, but also more comprehensive data on species distributions and abundances.
+These goals can only be achieved with additional cost-effective sampling, and a commitment to training more taxonomists and experts that can identify and interpret the material collected.
 
@@
 For the PNW, fewer species than included in the regional checklist were recovered from biodiversity databases (iDigBio and OBIS combined: `r format(dplyr::n_distinct(combined_pnw_records$worms_valid_name), big.mark=",")` species; iDigBio only: `r format(dplyr::n_distinct(idigbio_pnw_records$worms_valid_name), big.mark=",")`  species, $n=$ `r format(nrow(idigbio_pnw_records), big.mark=",")` records; OBIS only `r format(dplyr::n_distinct(obis_pnw_records$worms_valid_name), big.mark=",")` species, $n=$ `r format(nrow(obis_pnw_records), big.mark=",")` records).
-The combination of the species listed in these three sources leads to `r n_distinct(total_gom_records$worms_valid_name)` species for the GOM and `r n_distinct(total_pnw_records$worms_valid_name)` species for the PNW.
+The combination of the species listed in these three sources leads to `r format(n_distinct(total_gom_records$worms_valid_name), big.mark=",")` species for the GOM and `r format(n_distinct(total_pnw_records$worms_valid_name), big.mark=",")` species for the PNW.
 
@@
 
-Digitization of species richness is higher for the GOM compared with the PNW.
 A higher proportion of the species included in the regional list for the GOM are captured in biodiversity databases than for the PNW.
@@
 \@ref(fig:database-coverage)B).
-Additionally, a higher proportion of species occur in the biodiversity databases but are not included in the regional checklist for the GOM compared with the PNW (GOM: `r format_output((1253 + 848 + 556 + 216 + 190 + 124) / n_distinct(total_gom_records$worms_valid_name)*100)`%, $n =$ `r n_distinct(total_gom_records$worms_valid_name)`; PNW:
-`r format_output( (281 + 151 + 145 + 80 + 78 + 45) / n_distinct(total_pnw_records$worms_valid_name)*100)`%, $n =$ `r n_distinct(total_pnw_records$worms_valid_name)`).
+Additionally, a higher proportion of species occur in the biodiversity databases but are not included in the regional checklist for the GOM compared with the PNW (GOM: `r format_output((1253 + 848 + 556 + 216 + 190 + 124) / n_distinct(total_gom_records$worms_valid_name)*100)`%, $n =$ `r format(n_distinct(total_gom_records$worms_valid_name), big.mark=",")`; PNW:
+`r format_output( (281 + 151 + 145 + 80 + 78 + 45) / n_distinct(total_pnw_records$worms_valid_name)*100)`%, $n =$ `r format(n_distinct(total_pnw_records$worms_valid_name), big.mark=",")`).
 
@@
 
-<!-- need to separate across all phyla statements and by phylum statements -->
-We investigated the number of species for each phylum that contributed to the discrepancies between the regional species lists and the database records.
-We found that the more diverse phyla contributed the most to the differences across the data sources.
-OBIS tends to be closer to the full record set, but iDigBio has more species for the Gulf of Mexico that are not included in the list (Fig.
+The more diverse phyla contribute the most to the discrepancies between the regional species lists and the database records (Fig.
 \@ref(fig:database-overlap)).
+OBIS tends to be closer to the full record set, but iDigBio has more species for the Gulf of Mexico that are not included in the list across most phyla.
 Many species of mollusks, arthropods, and annelids, known to occur in the Gulf of Mexico or in the Pacific Northwest, have no digital records in these geographic areas.
@@
 Several factors may contribute for species to be included in the regional lists but to not have records in biodiversity databases.
-The first category of biases are linked to property of the species affecting the likelihood of having specimens preserved in museum collections.
+The first category of biases is linked to the likelihood of having specimens for these species in museum collections.
 For instance, rare species may never have been collected or have only been collected a few times, and the corresponding records have not been digitized yet.
@@
 Some of these species may be in the databases but do not have geographic coordinates associated with them, and therefore would not be included in the results of our query.
+While it is difficult to assess biases associated with the likelihood of preservation of a species, we investigated how missing geographic coordinates are affecting the discrepancies between the checklists and the databases.
 
-Missing geographic information contribute to the discrepencies observed.
+Missing geographic coordinates reduces the overlap between the list of species from the regional list and the databases.
 Using iDigBio we found that for the `r format(not_really_in_database$idig_gom_n_spp_not_in_db, big.mark=",")` species that do not appear within the geographic range of the Gulf of Mexico but are on the list, `r format_output(not_really_in_database$idig_gom_p_spp_not_in_db*100)`% are completely absent from iDigBio, while for the `r format(not_really_in_database$idig_gom_n_spp_no_coords, big.mark=",")` species that do appear in iDigBio, `r format_output(not_really_in_database$idig_gom_p_spp_no_coords*100)`%  do not have geographic coordinates associated with them, and for the ones that do have geographic coordinates, `r format_output(not_really_in_database$idig_gom_p_spp_within_usa*100)`% ($n=$  `r format(not_really_in_database$idig_gom_n_spp_within_usa, big.mark=",")`) fall within the US waters.
-These numbers indicate that missing geographic coordinates reduces the overlap between the list of species from the regional list and the databases.
 Additional digitization of collections, including the capture of their geographic coordinates information, are needed to improve biodiversity knowledge at regional scales.
@@
 
-Two factors are most likely to contribute to explain why some species are in databases but were not considered by the authors of the checklists: (1) the species have been documented for the first time since the publication of the checklist, (2) the species records were not accessible or known to the authors.
+Two factors are most likely to contribute to explain why some species are in databases but are not included in the regional the checklists: (1) the species have been documented for the first time since the publication of the checklist, (2) the species records were not accessible or known to the authors.
 
-<!-- include also proportion of species not in list -->
-Recently recorded species contribute modestly to this discrepency, as `r sum(recently_collected$gom_in_dbs$min_year > 2001)` species (`r format_output( sum(recently_collected$gom_in_dbs$min_year > 2001)/n_distinct(total_gom_records$worms_valid_name) * 100)`% of the total species richness) have been collected for the first time since the publication of the list in the Gulf of Mexico, and `r sum(recently_collected$pnw_in_dbs$min_year > 1996)` species (`r format_output(sum(recently_collected$pnw_in_dbs$min_year > 1996)/n_distinct(total_pnw_records$worms_valid_name) * 100)`% of the total species richness) for the Pacific Northwest.
+Recently recorded species contribute modestly to this discrepancy for the Gulf of Mexico, as `r sum(recently_collected$gom_in_dbs$min_year > 2001)` species (`r format_output( sum(recently_collected$gom_in_dbs$min_year > 2001)/n_distinct(total_gom_records$worms_valid_name) * 100)`% of the total species richness, `r format_output(sum(recently_collected$gom_in_dbs$min_year > 2001)/spp_not_in_lists$gom_not_in_list * 100)`% of the species not on the checklist),  have been collected for the first time since the publication of the checklist, but more significantly for the Pacific Northwest with  `r sum(recently_collected$pnw_in_dbs$min_year > 1996)` species (`r format_output(sum(recently_collected$pnw_in_dbs$min_year > 1996)/n_distinct(total_pnw_records$worms_valid_name) * 100)`% of the total species richness,  `r format_output(sum(recently_collected$pnw_in_dbs$min_year > 1996)/spp_not_in_lists$pnw_not_in_list * 100)`% of the species not on the checklist).
 
-The lack to tools for the authors of these checklist to have access to museum records across institutions contributed most to explain why some species were not included in the checklists.
-These records were difficult to discover before digitization which, for the most part, has occured after the compilation of these checklists in most cases.
+The lack of tools for the authors of these checklist to have access to museum records across institutions contributes most to explain why some species were not included in the checklists.
+Many of these records were difficult to discover before digitization which, for the most part, has occurred after the compilation of these checklists.
+We estimate that using biodiversity databases increases the number of species included in the checklist by `r format_output(spp_not_in_lists$gom_not_in_list/n_distinct(total_gom_species$worms_valid_name)*100)`% for the Gulf of Mexico, and `r format_output(spp_not_in_lists$pnw_not_in_list/n_distinct(total_pnw_species$worms_valid_name)*100)`% for the Pacific Northwest.
 
 ### What do we know about marine invertebrates in the US?
-
-<!-- separate idigbio and (and add) obis in the results -->
 
@@
 Further biases exist across phyla and body size.
-Large-bodied and easily acccessible (e.g.
-dried specimens like shells, corals, echinoderms) are overrepresented in museum collections, compared with small less accessible organisms, such as the meiofauna.
+Large-bodied and easily accessible (e.g.
+dried specimens like shells, corals, echinoderms) are over-represented in museum collections, compared with small less accessible organisms, such as the meiofauna.
 Of `r n_idigbio_phyla` phyla represented in iDigBio, `r records_rare_phyla$n_rare_phyla` have <500 records, and make up `r records_rare_phyla$prop_recs_rare_phyla`% of the iDigBio records.
-This underrepresentation is also manifested at the species level, while `r records_rare_phyla$prop_spp_rare_phyla_collected`% (n = `r format(records_rare_phyla$total_spp_rare_phyla_collected, big.mark=",")` total species) of the global number of species for these `r records_rare_phyla$n_rare_phyla` phyla occur in iDigBio, this proportion rises to `r records_rare_phyla$prop_spp_common_phyla_collected`% (n = `r format(records_rare_phyla$total_spp_common_phyla_collected, big.mark=",")`) for the `r n_idigbio_phyla - records_rare_phyla$n_rare_phyla` others.
+This under-representation is also manifested at the species level, while `r records_rare_phyla$prop_spp_rare_phyla_collected`% (n = `r format(records_rare_phyla$total_spp_rare_phyla_collected, big.mark=",")` total species) of the global number of species for these `r records_rare_phyla$n_rare_phyla` phyla occur in iDigBio, this proportion rises to `r records_rare_phyla$prop_spp_common_phyla_collected`% (n = `r format(records_rare_phyla$total_spp_common_phyla_collected, big.mark=",")`) for the `r n_idigbio_phyla - records_rare_phyla$n_rare_phyla` others.
 Yet, these undersampled taxa, especially those of small body size, are often the most abundant, and are likely to play an important role in the detection of changes in environmental quality[@Bik2012].
@@
 \@ref(fig:identification-through-time)).
-There is a marked break in the central trend arond the 1970's indicating that it now takes decades, on average, to identify samples deposited in collections.
+There is a marked break in the central trend around the 1970's indicating that it now takes decades, on average, to identify samples deposited in collections.
 For diverse groups such as annelids or arthropods, the proportion of specimens identified to the species-level is now below 30%.
@@
 
-<!-- add overall mean number before getting into details -->
-
 The marine invertebrates of the US have generally been characterized genetically at a higher level than other parts of the world: `r format_output(prop_spp_barcoded$p_barcoded * 100)`% (n = `r format(nrow(combined_species), big.mark=",")`) of the fauna has DNA barcodes available.
@@
 
-```{r proportion-barcoded, fig.cap='Proportion of species known to occur in the US EEZ based on iDigBio records, in the Gulf of Mexico [@Tunnell2009], in the Pacific Northwest[@Kozloff1996], that have publicly available DNA barcodes.
+```{r proportion-barcoded, fig.cap='Proportion of species with publicly available DNA barcodes  known to occur the US EEZ (iDigBio and OBIS records), in the Gulf of Mexico (iDigBio, OBIS, and checklist[@Tunnell2009]), and in the Pacific Northwest (iDigBio, OBIS, and checklist[@Kozloff1996]).
 Numbers next to the bars indicate the number of species for which DNA barcodes are available.
@@
 
-## Where do we go from there?
+## Discussion
 
@@
 Linking species names to identifiers in taxonomic databases would facilitate data linking, specimen discoverability, and data usability.
-Additionally, because species names registers like WoRMS link the history of the names and their synonymies, it is possible to dissociate the concepts of species identification with the concept of species identity, allowing researchers to compare specimens identified at different time periods, a tedious process that would have required expert knowledge in the past.
+Additionally, because species names registers like WoRMS link the history of the names and their synonymies, it is possible to dissociate the concepts of species identification with the concept of species identity, allowing researchers to compare specimens identified at different time periods.
 
@@
 
-<!-- separate issues of sampling is needed, and large-scale sampling is best way of doing it --> Coordinated large-scale biodiversity expeditions continue to be needed to improve our understanding of the diversity of species and their spatial distributions.
+<!-- separate issues of sampling is needed, and large-scale sampling is best way of doing it -->
+
+<!-- Coordinated large-scale biodiversity expeditions continue to be needed to improve our understanding of the diversity of species and their spatial distributions.
 The US marine EEZ is relatively well-known compared to many areas of the world ocean, where studies show that undescribed biodiversity dominates.
-For example, a recent survey in Western Australia reported that 94.6% of crustaceans and 72% of polychaetes taken at 135 sites were undescribed [@Poore2014].
+-->
 
- However, museum specimens are critical to validate the occurrence of species [@Rocha2014], especially for marine invertebrates for which species limits based on morphology have been known to be challenged by other lines of evidence including molecular data [@Knowlton2000].
+<!-- For example, a recent survey in Western Australia reported that 94.6% of crustaceans and 72% of polychaetes taken at 135 sites were undescribed [@Poore2014]. -->
+
+Museum specimens are critical to validate the occurrence of species [@Rocha2014], especially for marine invertebrates for which species limits based on morphology have been known to be challenged by other lines of evidence including molecular data [@Knowlton2000].
 
@@
 Building the library of DNA barcodes in parallel of metabarcoding efforts ensures that database matches correspond to actual specimens deposited in natural history collections, ruling out possible artifacts of the method.
-With such approaches, the geographic distribution of species can build on and be linked to existing records rather than being limited to matching DNA sequences across studies.
+With such approaches, the geographic distribution of species can build on and be linked to existing records, species and their natural history, rather than being limited to matching DNA sequences across studies.
 
@@
 
-For geographic searches, we obtained the boundaries of the Exclusive Economic Zone for the United States from marineregions.org (mrgid:8456) [@mregions].
+For geographic searches in OBIS and iDigBio, we obtained the boundaries of the Exclusive Economic Zone for the United States from marineregions.org (mrgid:8456) [@mregions].
 For the Pacific Northwest, Kozloff did not specify the exact geographic extent covered by his book.

### 2017-09-04 manuscript.Rmd
[fix #52] add other maps as supplementary material
+ refactor code to generate maps

@@
 
-<!-- ## Supplementary material -->
+## Supplementary material
+
+```{r gom-maps, fig.cap="Spatial distribution of the number of samples (A) and species richness (B) for marine invertebrates based on records retrieved from iDigBio and OBIS for the Gulf of Mexico."}
+knitr::include_graphics("figures/combined_gom_sampling.pdf")
+```
+
+```{r pnw-maps, fig.cap="Spatial distribution of the number of samples (A) and species richness (B) for marine invertebrates based on records retrieved from iDigBio and OBIS for the Pacific Northwest."}
+knitr::include_graphics("figures/combined_pnw_sampling.pdf")
+```
+
+```{r idigbio-usa-maps, fig.cap="Spatial distribution of the number of samples (A), and species richness (B) for marine invertebrates based on records retrieved from iDigBio within the limits of the Exclusive Economic Zone for the contiental USA."}
+knitr::include_graphics("figures/idigbio_map_sampling.pdf")
+```
+
+```{r obis-usa-maps, fig.cap="Spatial distribution of the number of samples (A), and species richness (B) for marine invertebrates based on records retrieved from OBIS within the limits of the Exclusive Economic Zone for the contiental USA."}
+knitr::include_graphics("figures/obis_map_sampling.pdf")
+```

### 2017-09-05 manuscript.Rmd
include mark's comments

@@
 
-Museum collections have been recording our knowledge about species and their geographic ranges, but until recently, this information was fragmented across institutions and difficult to synthesize.
+Museum collections have been recording our knowledge about species and their geographic ranges for about 250 years, but until recently, this information was fragmented across institutions and difficult to synthesize.
 The digitization of natural history collections and other biodiversity information systems can now be used to enable synthesis and highlight knowledge gaps.
@@
 While recent high-throughput sequencing technologies promise to greatly speed up the documentation of marine species and their ranges, its value will be greatly improved by anchoring the results into existing DNA sequence data and associated vouchers.
-However, the best documented phyla have at most 25% of their species sequenced for the most widely used DNA barcoding gene, cytochrome oxidase I (COI).
-More sampling, coordinated at the scale of the US, is needed to improve our knowledge of species and their geographic ranges.
+However, the best documented phyla have at most 25% of their species sequenced for even the most widely used DNA barcoding gene, cytochrome oxidase I (COI).
+More sampling, coordinated at the national scale, is needed to improve our knowledge of species and their geographic ranges.
 This data is needed to detect the impact of climate change and other anthropogenic strains to the environment.
@@
 With increasing impact of human activities on the biosphere, we need efficient and accurate approaches to document changes in species ranges and community composition.
-While terrestrial ecosystems have received considerable attention, the baseline of healthy and resilient marine ecosystems are not well understood [@Worm2006], and modeling the impact of human activities has been limited to a few taxonomic groups [@Tittensor2010].
+<!-- add why?
+-->  While terrestrial ecosystems have received considerable attention, the baseline of healthy and resilient marine ecosystems are not well understood [@Worm2006], and modeling the impact of human activities has been limited to a few taxonomic groups [@Tittensor2010].
 For marine organisms, because of the diversity of their complex life history strategies, their size range, and the difficulties associated with sampling their habitats, we have limited information about which species live where, or what portion of their diversity remains to be discovered and described [@Mora2011].
@@
 We show that while public databases are capturing a significant portion of our knowledge about species richness, most species are known from only a few specimens and a few locations.
-We emphasize that our understanding of marine biodiversity, and the impact of human activities on biological communities require not only to get a more accurate list of the species that inhabits the oceans, but also more comprehensive data on species distributions and abundances.
+We emphasize that our understanding of marine biodiversity, and the impact of human activities on biological communities require not only a more accurate list of the species that inhabits the oceans, but also more comprehensive data on species distributions and abundances.
 These goals can only be achieved with additional cost-effective sampling, and a commitment to training more taxonomists and experts that can identify and interpret the material collected.
@@
 While `r format_output((3890 + 2663 + 1025 + 316 + 102 + 49 )/n_distinct(total_gom_records$worms_valid_name) * 100)`%  of the species included in the checklist for the GOM have records in at least one of the biodiversity database (Fig.
-\@ref(fig:database-coverage)A), `r format_output((352 + 292 + 281 + 235 + 76 + 38)/n_distinct(total_pnw_records$worms_valid_name)*100)`% do for the PNW  (Fig.
+\@ref(fig:database-coverage)A), only `r format_output((352 + 292 + 281 + 235 + 76 + 38)/n_distinct(total_pnw_records$worms_valid_name)*100)`% do for the PNW  (Fig.
 \@ref(fig:database-coverage)B).
@@
 Missing geographic coordinates reduces the overlap between the list of species from the regional list and the databases.
-Using iDigBio we found that for the `r format(not_really_in_database$idig_gom_n_spp_not_in_db, big.mark=",")` species that do not appear within the geographic range of the Gulf of Mexico but are on the list, `r format_output(not_really_in_database$idig_gom_p_spp_not_in_db*100)`% are completely absent from iDigBio, while for the `r format(not_really_in_database$idig_gom_n_spp_no_coords, big.mark=",")` species that do appear in iDigBio, `r format_output(not_really_in_database$idig_gom_p_spp_no_coords*100)`%  do not have geographic coordinates associated with them, and for the ones that do have geographic coordinates, `r format_output(not_really_in_database$idig_gom_p_spp_within_usa*100)`% ($n=$  `r format(not_really_in_database$idig_gom_n_spp_within_usa, big.mark=",")`) fall within the US waters.
+Using iDigBio we found that for the `r format(not_really_in_database$idig_gom_n_spp_not_in_db, big.mark=",")` species that do not appear within the geographic range of the Gulf of Mexico but are on the list, `r format_output(not_really_in_database$idig_gom_p_spp_not_in_db*100)`% are completely absent from iDigBio, while for the `r format(not_really_in_database$idig_gom_n_spp_no_coords, big.mark=",")` species that do appear in iDigBio, `r format_output(not_really_in_database$idig_gom_p_spp_no_coords*100)`%  do not have geographic coordinates associated with them, and for the ones that do have geographic coordinates, `r format_output(not_really_in_database$idig_gom_p_spp_within_usa*100)`% ($n=$  `r format(not_really_in_database$idig_gom_n_spp_within_usa, big.mark=",")`) fall within US waters.
 Additional digitization of collections, including the capture of their geographic coordinates information, are needed to improve biodiversity knowledge at regional scales.
@@
 
-Two factors are most likely to contribute to explain why some species are in databases but are not included in the regional the checklists: (1) the species have been documented for the first time since the publication of the checklist, (2) the species records were not accessible or known to the authors.
+Two factors most likely contribute to explaining why some species are in databases but are not included in the regional the checklists: (1) the species have been documented for the first time since the publication of the checklist, (2) the species records were not accessible or known to the authors.
 
@@
 
-The lack of tools for the authors of these checklist to have access to museum records across institutions contributes most to explain why some species were not included in the checklists.
+The lack of tools for the authors of these checklists to have access to museum records across institutions explains why some species were not included in the checklists.
 Many of these records were difficult to discover before digitization which, for the most part, has occurred after the compilation of these checklists.
@@
 \@ref(fig:proportion-barcoded)).
-The GOM has comparable coverage than for the EEZ, while the PNW is better barcoded with >30% species-level coverage across each of the five most diverse phyla.
+The GOM has comparable coverage in level of DNA barcoding of the fauna to the EEZ, while the PNW is better barcoded with >30% species-level coverage across each of the five most diverse phyla.
 This is driven in part by the active Canadian barcoding effort in British Columbia.
@@
 
-Increasing the computing literacy and teaching data wrangling skills to staff of museums in order for them to manipulate and use the data from their collection database is equally critical.
+Increasing the computing literacy and teaching "data wrangling" skills to staff of museums in order for them to manipulate and use the data from their collection database is equally critical.
 As data across institutions get incorporated and pooled into large biodiversity information systems, it is crucial that data standards such as Darwin Core are embraced by the community.
@@
 We recommend that additional coordinated field expeditions are organized to sample more comprehensively the marine environment.
-The relation between sampling effort and diversity from museum collections makes clear that any additional samples will improve greatly our knowledge of marine biodiversity.
+The relation between sampling effort and diversity from museum collections makes clear that any additional samples will greatly improve our knowledge of marine biodiversity.
 Concerted and collaborative efforts are needed to share expertise and limited resources.

### 2017-09-07 manuscript.Rmd
[fix #62] calculate set sizes from data

@@
 
+```{r}
+set_size_gom <- function(...)
+    set_size_with_checklist(..., d = remake::fetch("total_gom_species"))
+set_size_pnw <- function(...)
+    set_size_with_checklist(..., d = remake::fetch("total_pnw_species"))
+```
+
 A higher proportion of the species included in the regional list for the GOM are captured in biodiversity databases than for the PNW.
-While `r format_output((3890 + 2663 + 1025 + 316 + 102 + 49 )/n_distinct(total_gom_records$worms_valid_name) * 100)`%  of the species included in the checklist for the GOM have records in at least one of the biodiversity database (Fig.
-\@ref(fig:database-coverage)A), only `r format_output((352 + 292 + 281 + 235 + 76 + 38)/n_distinct(total_pnw_records$worms_valid_name)*100)`% do for the PNW  (Fig.
+While `r format_output(set_size_gom("checklist","idigbio","obis")/n_distinct(total_gom_records$worms_valid_name) * 100)`%  of the species included in the checklist for the GOM have records in at least one of the biodiversity database (Fig.
+\@ref(fig:database-coverage)A), only `r format_output(set_size_pnw("checklist","idigbio","obis")/n_distinct(total_pnw_records$worms_valid_name)*100)`% do for the PNW  (Fig.
 \@ref(fig:database-coverage)B).
-Additionally, a higher proportion of species occur in the biodiversity databases but are not included in the regional checklist for the GOM compared with the PNW (GOM: `r format_output((1253 + 848 + 556 + 216 + 190 + 124) / n_distinct(total_gom_records$worms_valid_name)*100)`%, $n =$ `r format(n_distinct(total_gom_records$worms_valid_name), big.mark=",")`; PNW:
-`r format_output( (281 + 151 + 145 + 80 + 78 + 45) / n_distinct(total_pnw_records$worms_valid_name)*100)`%, $n =$ `r format(n_distinct(total_pnw_records$worms_valid_name), big.mark=",")`).
+Additionally, a higher proportion of species occur in the biodiversity databases but are not included in the regional checklist for the GOM compared with the PNW (GOM: `r format_output(set_size_gom("checklist","idigbio","obis",include_checklist=FALSE) / n_distinct(total_gom_records$worms_valid_name)*100)`%, $n =$ `r format(n_distinct(total_gom_records$worms_valid_name), big.mark=",")`; PNW:
+`r format_output( set_size_pnw("checklist","idigbio","obis",include_checklist=FALSE) / n_distinct(total_pnw_records$worms_valid_name)*100)`%, $n =$ `r format(n_distinct(total_pnw_records$worms_valid_name), big.mark=",")`).
 

### 2017-09-07 R/combine_lists.R

@@ -58,11 +58,16 @@ combine_species_list <- function(...) {
 export_species_list <- function(d, file) {
 
     d %>%
-        dplyr::mutate(bold = as.integer(n_bold_records > 0)) %>%
-        dplyr::select(-starts_with("n_")) %>%
-        dplyr::mutate_if(is.integer, funs(if_else(is.na(.) | . == 0L, 0L, 1L))) %>%
-        dplyr::arrange(worms_phylum) %>%
+        generate_species_list() %>%
         readr::write_csv(path = file)
 
     file
+}
+
+generate_species_list <- function(d) {
+     d %>%
+        dplyr::mutate(bold = as.integer(n_bold_records > 0)) %>%
+        dplyr::select(-starts_with("n_")) %>%
+        dplyr::mutate_if(is.integer, funs(if_else(is.na(.) | . == 0L, 0L, 1L))) %>%
+        dplyr::arrange(worms_phylum)
 }

### 2017-09-07 R/comparison_by_geo.R

@@ -21,7 +21,7 @@ compare_with_geo <- function(spp_list, geo_list, verbose = FALSE) {
 }
 
 combined_regional_by_geo <- function(gom_worms, gom_idigbio, gom_obis,
-                             pnw_worms, pnw_idigbio, pnw_obis) {
+                                     pnw_worms, pnw_idigbio, pnw_obis) {
 
     list(
         gom = combine_species_list(checklist = gom_worms, idigbio = gom_idigbio, obis = gom_obis),
@@ -54,8 +54,26 @@ compare_database_overlap <- function(gom_worms, kozloff_worms,
     comp_koz <- dplyr::bind_rows(list(idigbio = idig_koz, obis = obis_koz),
                                  .id = "database")
 
-    comp_db <- dplyr::bind_rows(list(gom = comp_gom,
-                                     pnw = comp_koz), .id = "region")
+    dplyr::bind_rows(list(gom = comp_gom, pnw = comp_koz), .id = "region")
+}
+
+## For a data frame d formated as the input for upset, calculate the size of the
+## intersect for the lists specified in ...
+## For instance for the intersect of bold, checklist, and idigbio, for the
+## total list fromm GOM: set_size("bold", "checklist", "idigbio", d=total_gom_species)
+set_size_with_checklist <- function(..., include_checklist = TRUE, d) {
+    d <- generate_species_list(d)
+    sets <- c(...)
+    if (length(sets) < 2) stop("at least 2 sets needed")
+    if (!"checklist" %in% sets) stop(sQuote("checklist"), " should be included.")
+    if (!all(sets %in% names(d)))
+        stop("Sets not found: ", sets[!sets %in% names(d)])
+    sel_set <- dplyr::select(d, sets)
+
+    ## We want the species that are in the checklist and in at least one other
+    ## data source.
+    sum(sel_set[["checklist"]] == as.integer(include_checklist) &
+        rowSums(sel_set) >=  (2 - as.integer(!include_checklist)))
 }
 
 

### 2017-09-25 R/time_plot.R
use porifera instead of chordata to match species through time plot

@@ -324,7 +324,7 @@ plot_identification_level_through_time <- function(id_level) {
                n_lots = sum(n)) %>%
         filter(id_level == "species",
                phylum %in% c("annelida", "arthropoda", "chordata",
-                             "cnidaria", "echinodermata", "mollusca")) %>%
+                             "porifera", "echinodermata", "mollusca")) %>%
         ungroup() %>%
         mutate(phylum = capitalize(phylum)) %>%
         ggplot(aes(x = year, y = p, colour = phylum, fill = phylum)) +

### 2017-09-25 manuscript.Rmd
[fix #59] compute median number of cells occupied by species
and plot species rank abundance based on number of cells they occupy
[fix #65] improve figure legends where needed

@@
 ```{r compare_database_coverage, fig.cap='Overlap between the regional species list (list), records in iDigBio (idigbio), records in OBIS (obis), and records in BOLD (bold) for the Gulf of Mexico (left) and the Pacific Northwest (right).
+A higher proportion of species in the regional checklists are captured in databases for the Gulf of Mexico compared with the Pacific Northwest.
 This plot is generated using the UpSet technique[@Lex2014;@Conway2017], which uses a matrix-based layout to represent intersections of sets and their sizes.
@@
 ```{r compare_database_overlap, fig.cap='Proportions of the total number of species by phylum that are absent from faunistic lists but are in iDigBio or OBIS ("Not in list"), and absent from iDigBio or OBIS but are in the faunistic lists ("Not in databases").
+Many species, even in diverse taxonomic groups, known to occur in these regions have no digital records.
 The numbers represent the number of species.
@@
 
+The median number of cells occupied by species is `r median(rastercell_distribution$n_cell)`.
+
 Spatial patterns of species richness are however strongly confounded by sampling effort.
@@
 Trend lines are  B-splines of the 3rd degree polynomial.
-For all groups, we observe a decline in the proportion of specimens collected identified at the species level suggesting a lack of expertise to identify precisely what is collected.
+For all groups, we observe a decline in the proportion of specimens collected identified at the species leveld.
 (\\#fig:identification-through-time)'}
@@
 ```
+
+```{r rank-abundance, fig.cap="Rank abundance plots for most represented phyla."}
+knitr::include_graphics("figures/rank_abundance.pdf")
+```

### 2017-09-25 manuscript.Rmd
small edits

@@
 As species remain the most convenient unit to measure biological diversity, understanding how much has been done, how much remains to be done, and how we can speed up the description of species diversity is crucial.
-However, to understand the role of these species in their environment, we need to document the ecology of these species [@Cernansky2017].
+However, to understand the role of these species in their environment, we need to document their ecology [@Cernansky2017].
 For instance, to be able to predict how their geographic ranges will be affected by global change, we need detailed information about their environmental requirements.
@@
 iDigBio [@Page2015]), species occurrences (e.g., OBIS [@Zhang2002]), genetic data (e.g., GenBank [@Benson2013], BOLD [@Ratnasingham2007a]), and registers of species names (e.g., WoRMS [@Costello2013b]), now allow us to evaluate what and how much we know.
+However, few studies have sought to synthesize existing knowledge by combining information from these databases.
 Here, we aggregate publicly available data for marine invertebrates that occur within the limits of the exclusive economic zone (EEZ) of the continental United States (US), as well as the Gulf of Mexico (GOM) and the Pacific Northwest (PNW), to highlight gaps in our knowledge that provide a road map for important new directions in global research.
@@
 Missing geographic coordinates reduces the overlap between the list of species from the regional list and the databases.
-Using iDigBio we found that for the `r format(not_really_in_database$idig_gom_n_spp_not_in_db, big.mark=",")` species that do not appear within the geographic range of the Gulf of Mexico but are on the list, `r format_output(not_really_in_database$idig_gom_p_spp_not_in_db*100)`% are completely absent from iDigBio, while for the `r format(not_really_in_database$idig_gom_n_spp_no_coords, big.mark=",")` species that do appear in iDigBio, `r format_output(not_really_in_database$idig_gom_p_spp_no_coords*100)`%  do not have geographic coordinates associated with them, and for the ones that do have geographic coordinates, `r format_output(not_really_in_database$idig_gom_p_spp_within_usa*100)`% ($n=$  `r format(not_really_in_database$idig_gom_n_spp_within_usa, big.mark=",")`) fall within US waters.
+Using iDigBio, we found that for the `r format(not_really_in_database$idig_gom_n_spp_not_in_db, big.mark=",")` species that do not appear within the geographic range of the Gulf of Mexico but are on the list, `r format_output(not_really_in_database$idig_gom_p_spp_not_in_db*100)`% are completely absent from iDigBio, while for the `r format(not_really_in_database$idig_gom_n_spp_no_coords, big.mark=",")` species that do appear in iDigBio, `r format_output(not_really_in_database$idig_gom_p_spp_no_coords*100)`%  do not have geographic coordinates associated with them, and for the ones that do have geographic coordinates, `r format_output(not_really_in_database$idig_gom_p_spp_within_usa*100)`% ($n=$  `r format(not_really_in_database$idig_gom_n_spp_within_usa, big.mark=",")`) fall within US waters.
 Additional digitization of collections, including the capture of their geographic coordinates information, are needed to improve biodiversity knowledge at regional scales.
@@
 Geographic information is missing for most records (`r format_output(100*stat_bold$p_no_coords)`% of the `r format(stat_bold$n_total_records, big.mark=",")` BOLD records analyzed here have no geographic coordinate information associated with them).
-It is thus crucial to ensure that accurate geographic information is provided when DNA barcodes are published, and that enough samples for each species are sequenced, to capture more accurately their geographic range and the extent of their genetic diversity.
+It is thus crucial to ensure that precise and accurate geographic information is provided when DNA barcodes are published, and that enough samples for each species are sequenced, to capture more accurately their geographic range and the extent of their genetic diversity.
 
@@
 Linking species names to identifiers in taxonomic databases would facilitate data linking, specimen discoverability, and data usability.
-Additionally, because species names registers like WoRMS link the history of the names and their synonymies, it is possible to dissociate the concepts of species identification with the concept of species identity, allowing researchers to compare specimens identified at different time periods.
+Additionally, because species name registers like WoRMS link the history of the names and their synonymies, it is possible to dissociate the concepts of species identification with the concept of species identity, allowing researchers to compare specimens identified at different time periods.
 

### 2017-09-26 R/idigbio_by_geo.R
implement validation of data stores for queries by geography

@@ -90,6 +90,21 @@ get_idigbio_by_geo <- function(coords, q) {
     }
 }
 
+validate_idigbio_storr <- function(coords) {
+    res <- vapply(names(coords), function(x)
+        inherits(store_idigbio_by_geo(coords)$get(x), "try-error"),
+        logical(1)
+        )
+    if (sum(res) > 0) {
+        stop("Coordinates at these positions are problematic: ",
+             paste(names(which(res)), collapse = ", "))
+    } else {
+        v3("the coordinates look good")
+    }
+}
+
+
+
 
 get_coords_idigbio_query <- function(map_usa, cellsize = .5) {
     bb_eez <- generate_bounding_boxes(map_usa, cellsize = cellsize)
@@ -102,7 +117,7 @@ get_coords_idigbio_query <- function(map_usa, cellsize = .5) {
 ## database that will store the results use_cache: if TRUE, this uses
 ## the results from the iDigBio storr; if false, the entire storr is
 ## destroyed before
-create_records_db <- function(coords, db_table, gom_phyla) {
+create_records_db <- function(coords, db_table) {
     idig_types <- structure(c("TEXT", "TEXT", "TEXT", "TEXT", "TEXT",
                               "TEXT", "TEXT", "TEXT", "TEXT", "TEXT",
                               "TEXT", "REAL", "REAL"),

### 2017-09-26 manuscript.Rmd
use xelatex engine

@@
     keep_tex: yes
-    latex_engine: pdflatex
+    latex_engine: xelatex
     number_sections: no

### 2017-09-26 R/time_plot.R
fix phyla

@@ -323,8 +323,8 @@ plot_identification_level_through_time <- function(id_level) {
         mutate(p = n/sum(n),
                n_lots = sum(n)) %>%
         filter(id_level == "species",
-               phylum %in% c("annelida", "arthropoda", "chordata",
-                             "porifera", "echinodermata", "mollusca")) %>%
+               phylum %in% c("annelida", "arthropoda", "porifera",
+                             "cnidaria", "echinodermata", "mollusca")) %>%
         ungroup() %>%
         mutate(phylum = capitalize(phylum)) %>%
         ggplot(aes(x = year, y = p, colour = phylum, fill = phylum)) +

### 2017-09-27 manuscript.Rmd
fix typo in figure legend
[fix #63] first draft of acknowledgments
This reverts commit a0dc16a085d00251ad7faa81a049fce791c1d1cd.

@@
 Trend lines are  B-splines of the 3rd degree polynomial.
-For all groups, we observe a decline in the proportion of specimens collected identified at the species leveld.
+For all groups, we observe a decline in the proportion of specimens collected identified at the species level.
 (\\#fig:identification-through-time)'}
@@
 
+## Acknowledgments
+
+We thank WoRMS, BOLD, iDigBio, OBIS, MarineRegions, NCBI, and all other data providers that aim to make more accessible biodiversity data.
+We would also like to acknowledge the efforts of rOpenSci (https://ropensci.org) for developping tools to make these data sources available in R.
+We are grateful to Julie Allen, Robert Lasley, Kevin Love, and Pamela Soltis, for providing suggestions that improved the manuscript.
+
 ## References

### 2017-09-28 manuscript.Rmd
number supplementary figures correctly
improve rank abundance plots

@@
 
-## Supplementary material
+# Supplementary material
+
+\renewcommand{\thefigure}{S\arabic{figure}}
+
+\setcounter{figure}{0}
 
@@
 
-```{r rank-abundance, fig.cap="Rank abundance plots for most represented phyla."}
-knitr::include_graphics("figures/rank_abundance.pdf")
+```{r rank-abundance, fig.cap="Rank abundance plots for most represented phyla based on the number of samples (A), and the number of cells occupied in map of Fig.
+\\@ref(fig:map-diversity-sampling) (B)"}
+knitr::include_graphics("figures/plot_rank_abundance.pdf")
 ```
+
+* list of marine invertebrate species from the continental USA: `list_species_us.csv`
+* list of marine invertebrate species from the Gulf of Mexico: `list_species_gom.csv`
+* list of marine invertebrate species from the Pacific Northwest: `list_species_pnw.csv`

### 2017-09-28 R/combine_plots.R

@@ -23,3 +23,15 @@ combine_cowplots_2 <- function(g1, g2, g3) {
     bottom_row <- plot_grid(g3, labels = "C", align = "h", rel_widths = c(1, .1))
     plot_grid(combine_cowplots(g1, g2, common_legend = TRUE), bottom_row, nrow = 2)
 }
+
+simple_cowplot <- function(g1, g2) {
+    leg <- get_legend(g1 + theme(legend.position = "bottom"))
+    p <- plot_grid(g1 + theme(legend.position = "none"),
+                   g2 + theme(legend.position = "none"),
+                   align = "vh",
+                   labels = c("A", "B"),
+                   hjust = -1,
+                   nrow = 2,
+                   scale = .9)
+    plot_grid(p, leg, ncol = 1, rel_heights = c(1, .2))
+}

### 2017-09-28 R/map_sampling.R

@@ -69,6 +69,23 @@ n_rastercell_per_species <- function(recs, raster) {
         dplyr::arrange(desc(n_cell))
 }
 
+abundance_sample_species <- function(recs) {
+    recs %>%
+        dplyr::count(phylum, worms_valid_name) %>%
+        dplyr::filter(phylum %in% c("arthropoda", "annelida", "mollusca",
+                                    "echinodermata", "cnidaria", "porifera")) %>%
+        dplyr::group_by(phylum) %>%
+        dplyr::mutate(rank_n = max(dplyr::min_rank(n)) - dplyr::min_rank(n)) %>%
+        dplyr::ungroup() %>%
+        dplyr::mutate(phylum = capitalize(phylum)) %>%
+        ggplot(aes(x = rank_n, y = n, colour = phylum)) +
+        geom_line() +
+        geom_point(size = .2) +
+        scale_y_log10() +
+        xlab("Rank abundance") + ylab("Number of samples") +
+        scale_colour_hc(name = "")
+}
+
 plot_rank_abundance <- function(recs) {
     recs %>%
         dplyr::filter(phylum %in% c("arthropoda", "annelida", "molluca",
@@ -81,7 +98,8 @@ plot_rank_abundance <- function(recs) {
         geom_line() +
         geom_point(size = .2) +
         scale_y_log10() +
-        xlab("Rank abundance") + ylab("Number of cells occupied")
+        xlab("Rank abundance") + ylab("Number of cells occupied") +
+        scale_colour_hc(name = "")
 }
 
 data_map_standardized_diversity <- function(sampling, diversity) {

### 2017-09-28 R/usa_maps.R
force removal of missing data

@@ -34,6 +34,7 @@ is_within_map_records <- function(area) {
                        coord_key = paste(lat, long, sep = "|"))
 
         simplified_coords <-  res_lawn %>%
+            dplyr::filter(!is.na(lat), !is.na(long)) %>%
             dplyr::select(coord_key, lat, long) %>%
             dplyr::distinct(coord_key, .keep_all = TRUE)
 

### 2017-10-01 manuscript.Rmd
include PS edits

@@
 <!-- add why?
--->  While terrestrial ecosystems have received considerable attention, the baseline of healthy and resilient marine ecosystems are not well understood [@Worm2006], and modeling the impact of human activities has been limited to a few taxonomic groups [@Tittensor2010].
-For marine organisms, because of the diversity of their complex life history strategies, their size range, and the difficulties associated with sampling their habitats, we have limited information about which species live where, or what portion of their diversity remains to be discovered and described [@Mora2011].
+-->  While terrestrial ecosystems have received considerable attention, the baseline of healthy and resilient marine ecosystems is not well understood [@Worm2006], and modeling the impact of human activities has been limited to a few taxonomic groups [@Tittensor2010].
+For marine organisms, because of the diversity of their complex life history strategies, their size range <!-- reformulate?
+-->, and the difficulties associated with sampling their habitats, we have limited information about which species live where, or what portion of their diversity remains to be discovered and described [@Mora2011].
 Additionally, coordinated marine ecosystem monitoring initiatives have been lacking [@Duffy2013,@Borja2016] (but are now getting organized, see MarineGEO [@Duffy2016marinegeo], MBON, and PISCO), making it difficult to monitor changes in species distributions and abundances.
-While most documented global extinctions of marine species are vertebrates, many others species may have been lost unnoticed because they were not known [@Sala2006].
-There is therefore a pressing need to improve our knowledge of marine systems before their functioning are irreversibly altered, and species go extinct.
+While most documented global extinctions of marine species are vertebrates, many other species may have been lost unnoticed because they were not known [@Sala2006].
+There is therefore a pressing need to improve our knowledge of marine systems before their functioning is irreversibly altered, and species go extinct.
 
 Research attempting to understand how much we know about marine biodiversity has focused on estimating the total number of species that inhabits the oceans, and which fraction of this diversity has been named [@Appeltans2012,@Costello2013].
-Undoubtedly, the development of molecular methods to investigate species limits has reinvigorated taxonomy, and contributed to elucidating cryptic species complexes [@Knowlton2000].
+Undoubtedly, the development of molecular methods to investigate species limits has reinvigorated taxonomy and contributed to elucidating cryptic species complexes [@Knowlton2000].
 Thus, it has been argued that molecular data can speed up the enterprise of discovering and naming species [@Blaxter2016].
@@
 For instance, to be able to predict how their geographic ranges will be affected by global change, we need detailed information about their environmental requirements.
-Occurrence data for marine species could be used for environmental niche modeling of their geographic and bathymetric distributions but the proportion of species for which we have enough data to do so is unknown.
+Occurrence data for marine species could be used for environmental niche modeling of their geographic and bathymetric distributions but the number of species -- and therefore the proportion of all species -- for which we have enough data to do so is unknown.
 
@@
 The development of public databases for specimen records (e.g.
-iDigBio [@Page2015]), species occurrences (e.g., OBIS [@Zhang2002]), genetic data (e.g., GenBank [@Benson2013], BOLD [@Ratnasingham2007a]), and registers of species names (e.g., WoRMS [@Costello2013b]), now allow us to evaluate what and how much we know.
+iDigBio [@Page2015]), species occurrences (e.g., OBIS [@Zhang2002]), genetic data (e.g., GenBank [@Benson2013], BOLD [@Ratnasingham2007a]), and registers of species names (e.g., WoRMS [@Costello2013b]) now allow us to evaluate what and how much we know.
 However, few studies have sought to synthesize existing knowledge by combining information from these databases.
@@
 We show that while public databases are capturing a significant portion of our knowledge about species richness, most species are known from only a few specimens and a few locations.
-We emphasize that our understanding of marine biodiversity, and the impact of human activities on biological communities require not only a more accurate list of the species that inhabits the oceans, but also more comprehensive data on species distributions and abundances.
-These goals can only be achieved with additional cost-effective sampling, and a commitment to training more taxonomists and experts that can identify and interpret the material collected.
+We emphasize that our understanding of marine biodiversity, and the impact of human activities on biological communities require not only a more accurate list of the species that inhabit the oceans, but also more comprehensive data on species distributions and abundances.
+These goals can only be achieved with additional cost-effective sampling and a commitment to training more taxonomists and experts that can identify and interpret the material collected.
 
@@
 `r n_distinct(na.omit(obis_records$worms_phylum))` phyla).
-Extant marine invertebrates occurring in the US represent less than 0.4% of the total number of iDigBio records (106 million), and about 1% of the US records (33 million records).
+Extant marine invertebrates occurring in the US represent less than 0.4% of the total number of iDigBio records (104+ million), and about 1% of the US records (33 million records).
 Marine invertebrates that occur in the US represent approximately 4% of the 48 million occurrence records in OBIS.
@@
 Spatial distribution of species richness with latitude differ markedly between the Pacific and the Atlantic coasts.
-On the Pacific coast, richness is lowest around 40° of latitude, while it is where it is the highest on the Atlantic coast (Fig.
+On the Pacific coast, richness is lowest around 40° of latitude, whereas on the Atlantic coast, it is highest at the same latitude (Fig.
 \@ref(fig:map-diversity-sampling)B, Fig.
@@
 \\@ref(fig:map-diversity-sampling).
-The black line represents the 1:1 relation.
+The black line represents the 1:1 relationship.
 (\\#fig:correlation-sampling)'}
@@
 
-The median number of cells occupied by species is `r median(rastercell_distribution$n_cell)`.
+<!-- The median number of cells occupied by species is `r median(rastercell_distribution$n_cell)`. -->
 
@@
 Yet at present, `r prop_10_specimens_species`% of species are represented by $\leq$ 10 records, and `r prop_30_specimens_species`%  by $\leq$ 30 records based on the combined records in iDigBio and OBIS.
+Additionally, `r format_output(prop_missing_depth$p_missing * 100)`% ($n=$ `r format(prop_missing_depth$n_total, big.mark=",")` OBIS records) of the observations lack bathymetric information.
 Underscoring the inadequacy of the sampling effort, `r prop_singleton_species`% of the species are known from a single record.
 
-Another indication of the inadequacy of the sampling is the relation between the number of collected specimen and the number of species recorded through time (Fig.
+Another indication of the inadequacy of the sampling is the relationship between the number of collected specimens and the number of species recorded through time (Fig.
 \@ref(fig:idigbio-through-time)).
-Explorations in the late ninetieth century led to an exponential growth in the knowledge of the fauna, that tapered off, until the 1970s when intensive collecting efforts led to an increase in the recording of new taxa (Fig.
+Explorations in the late nineteenth century led to an exponential growth in the knowledge of the fauna, but collecting then tapered off, until the 1970s when intensive collecting efforts led to an increase in the recording of new taxa (Fig.
 \@ref(fig:idigbio-through-time)).
-Since then, the slowing down of collections has been accompanied with a slowing down in the discovery of new records: `r n_spp_comparison$n_spp_2017_2000` species have been documented for the first time since 2000 (`r format_output(n_spp_comparison$n_spp_2017_2000/18)` species.year^-1^), while `r n_spp_comparison$n_spp_1970_1960` have been documented between 1960 and 1970 (`r format_output(n_spp_comparison$n_spp_1970_1960/11)` species.year^-1^).
+Since then, the slowed pace of collection activitity (and therefore a reduction in the number of specimens added) has been accompanied by a slow-down in the discovery of new records: `r n_spp_comparison$n_spp_2017_2000` species have been documented for the first time since 2000 (`r format_output(n_spp_comparison$n_spp_2017_2000/18)` species.year^-1^), while `r n_spp_comparison$n_spp_1970_1960` were documented between 1960 and 1970 (`r format_output(n_spp_comparison$n_spp_1970_1960/11)` species.year^-1^).
 The strong correlations between (1) the number of samples collected and the species richness (Fig.
 \@ref(fig:correlation-sampling)), and (2) the number of specimens collected per year and the number of species recorded for the first time (Fig.
-\@ref(fig:idigbio-through-time)C), suggest that any new sampling effort will lead to new observations and new species.
+\@ref(fig:idigbio-through-time)C), suggest that any new sampling effort will lead to new observations and the discovery of new species.
 
 ```{r idigbio-through-time, fig.cap='Cumulative number of samples (A) and species (B) of marine invertebrates recorded through time based on iDigBio records.
-Relation between number of samples collected, and number of species recorded for the first time (C), each point represents numbers for a given year, dashed line show the 1:1 and dotted line the 1:100 ratios for number of species recorded for the first time to the number of samples.
+Relationship between number of samples collected, and number of species recorded for the first time (C), each point represents numbers for a given year, dashed line shows the 1:1 and dotted line the 1:100 ratios for number of species recorded for the first time to the number of samples.
 (\\#fig:idigbio-through-time)'}
@@
 Large-bodied and easily accessible (e.g.
-dried specimens like shells, corals, echinoderms) are over-represented in museum collections, compared with small less accessible organisms, such as the meiofauna.
+dried specimens like shells, corals, echinoderms) specimens are over-represented in museum collections, compared with small less accessible organisms, such as the meiofauna.
 Of `r n_idigbio_phyla` phyla represented in iDigBio, `r records_rare_phyla$n_rare_phyla` have <500 records, and make up `r records_rare_phyla$prop_recs_rare_phyla`% of the iDigBio records.
@@
 Trend lines are  B-splines of the 3rd degree polynomial.
-For all groups, we observe a decline in the proportion of specimens collected identified at the species level.
+For all groups, we observe a decline in the proportion of specimens collected that are identified at the species level, suggesting a lack of expertise to identity precisely what is collected.
 (\\#fig:identification-through-time)'}
@@
 
-Since its introduction in the early 2000s, DNA barcoding, has revitalized taxonomy by facilitating species delineation [@Meyer2005-plos].
+Since its introduction in the early 2000s, DNA barcoding has revitalized invertebrate taxonomy by facilitating species delineation [@Meyer2005-plos].
 More recently, advances in high-throughput sequencing technologies have enabled the parallel sequencing of thousands of specimens in environmental samples (eDNA) cost effectively [@Leray2015], making it possible to estimate diversity and identify taxa from species-rich samples at unprecedented scales.
@@
 
-Based on the results of this study, we recommend a multi-level approach for both improving the quality of existing data, as well as for the efficient collection of new data.
+Based on the results of this study, we recommend a multi-level approach for both improving the quality of existing data and for the efficient collection of new data.
 
@@
 Linking species names to identifiers in taxonomic databases would facilitate data linking, specimen discoverability, and data usability.
-Additionally, because species name registers like WoRMS link the history of the names and their synonymies, it is possible to dissociate the concepts of species identification with the concept of species identity, allowing researchers to compare specimens identified at different time periods.
+Additionally, because registers of species names, such as WoRMS, link the history of the names and their synonymies, it is possible to dissociate the concepts of species identification with the concept of species identity, allowing researchers to compare specimens identified at different time periods.
 
@@
 
-For both the spelling of scientific names, and the geographic information, computing tools can be developed to detect issues and propose solutions.
+For both the spelling of scientific names, and the geographic information, computing tools can be developed to detect issues and provide solutions.
 There are opportunities to develop computing services that use publicly available data to flag and fix these issues directly from collection databases records (e.g.
 see [@Otegui2016] for vertebrate records).
-It however requires that resources are made available to museum collections in order to address these issues directly in collection databases.
+Such services, however, require that resources be made available to museum collections in order to address these issues directly in collection databases.
 
 Increasing the computing literacy and teaching "data wrangling" skills to staff of museums in order for them to manipulate and use the data from their collection database is equally critical.
-As data across institutions get incorporated and pooled into large biodiversity information systems, it is crucial that data standards such as Darwin Core are embraced by the community.
+As data across institutions are incorporated and pooled into large biodiversity information systems, it is crucial that data standards such as Darwin Core are embraced by the community.
 
@@
 
-Museum specimens are critical to validate the occurrence of species [@Rocha2014], especially for marine invertebrates for which species limits based on morphology have been known to be challenged by other lines of evidence including molecular data [@Knowlton2000].
+Museum specimens are critical for validating the occurrence of species [@Rocha2014], especially for marine invertebrates for which species limits based on morphology have been known to be challenged by other lines of evidence including molecular data [@Knowlton2000].
 
 This study highlights that we have too little data about too few species of marine invertebrates.
-We recommend that additional coordinated field expeditions are organized to sample more comprehensively the marine environment.
-The relation between sampling effort and diversity from museum collections makes clear that any additional samples will greatly improve our knowledge of marine biodiversity.
+We recommend that additional coordinated field expeditions be organized to sample more comprehensively the marine environment.
+The relationship between sampling effort and diversity from museum collections makes clear that any additional samples will greatly improve our knowledge of marine biodiversity.
 Concerted and collaborative efforts are needed to share expertise and limited resources.
 
-Samples collected during these expeditions need to be photographed to document live appearance that would otherwise be lost during the preservation process.
-We also need to capture metadata associated with specimens to be able to understand better the physical characteristics of the habitats and the life history for these species.
+Samples collected during these expeditions need to be photographed to document live appearance, information that would otherwise be lost during the preservation process.
+We also need to capture metadata associated with specimens to be able to understand better the physical characteristics of the habitats and the life histories of these species.
 Specimens need to be sub-sampled to create material suitable for genomic analysis.
@@
 Additionally, data generated by metabarcoding efforts can be re-interpreted as more DNA barcodes associated with correctly identified specimens become available.
-Building the library of DNA barcodes in parallel of metabarcoding efforts ensures that database matches correspond to actual specimens deposited in natural history collections, ruling out possible artifacts of the method.
+Building the library of DNA barcodes in parallel with metabarcoding efforts ensures that database matches correspond to actual specimens deposited in natural history collections, ruling out possible artifacts of the method.
 With such approaches, the geographic distribution of species can build on and be linked to existing records, species and their natural history, rather than being limited to matching DNA sequences across studies.
@@
 Assembling species lists is a necessary but not sufficient step in our quest to understand biodiversity.
-We also need to understand better the geographic distribution of biodiversity, how the abundance of species vary in space and time, and their role in the functioning of ecosystems.
+We also need to understand better the geographic distribution of biodiversity, how the abundance of species varies in space and time, and species' roles in the functioning of ecosystems.
 In order to predict the impact of climate change and other anthropogenic perturbations on species distributions, we need to collect additional occurrences for species as well as the characteristics of their habitats.
-Among the challenges of this enterprise, we need to be able to convey taxonomic concepts for species which, for the most part, have not formally been named yet.
+Among the challenges of this enterprise, we need to be able to convey taxonomic concepts for species that, for the most part, have not formally been named yet.
 Developing robust identifiers that can be shared across biodiversity information systems, and anchored in results from DNA barcoding or other molecular techniques, is needed to collaboratively build our knowledge of species distributions.
@@
 
-DNA sequences have been one of the first test case on how to link biodiversity data across repositories.
-These initial efforts have revealed the lack of infrastructure to use data formats specifically designed for this task, the lack of ontologies available for biodiversity data, and the fragility of the identifiers [@Stucky2014; @Walls2014; @Guralnick2014].
+DNA sequences are an early test case on how to link biodiversity data across repositories.
+These initial efforts have revealed the lack of infrastructure for using data formats specifically designed for this task, the lack of ontologies available for biodiversity data, and the fragility of the identifiers [@Stucky2014; @Walls2014; @Guralnick2014].
 As new types of data will be collected, added, and linked across repositories, the community needs to develop modern and robust ways to integrate these heterogeneous biodiversity data sources [@Guralnick2015; @Deck2015].
@@
 Even if technological advances can greatly speed up the documentation of species and their distributions, taxonomists and experts are needed  to collect, sort, process, and identify samples.
-Taxonomic expertise within the US is dwindling, and there are now only a handful of US natural history museums with active marine invertebrates collections.
+Taxonomic expertise within the US is dwindling, and there are now only a handful of US natural history museums with active collections of marine invertebrates.
 This lack of resources is not sustainable in the face of the challenges and discoveries that marine ecosystems have to offer.
@@
 Beyond the application to near-shore habitats on easily recognizable species [@Delaney2008], large-scale efforts are currently not possible for marine ecosystems given the specialized equipment needed to sample most habitats and the expertise required to identify species.
-There are however opportunities for citizen science to help curate biodiversity data, while educating and engaging the public with marine conservation issues.
-For instance citizen science has been used successfully to classify plankton images [@Robinson2017], or transcribe hand-written museum labels [@Hill2012].
+There are however opportunities for citizen scientists to help curate biodiversity data, while educating and engaging the public with marine conservation issues.
+For instance citizen science has been used successfully to classify plankton images [@Robinson2017], and to transcribe hand-written museum labels [@Hill2012].
 
-Here, we reported results from the US where there is at least two centuries of biodiversity exploration data and high-levels of digitization of museum collections.
+Here, we reported results from the US where there are at least two centuries of data on biodiversity exploration data and high-levels of digitization of museum collections.
 Many other parts of the world need additional attention, especially in some developing countries where environmental pressure on ecosystems is even higher than in the US.

### 2017-10-02 manuscript.Rmd
a few more edits from PS

@@
 
-Several factors may contribute for species to be included in the regional lists but to not have records in biodiversity databases.
+Several factors may contribute to species being included in the regional lists but not having records in biodiversity databases.
 The first category of biases is linked to the likelihood of having specimens for these species in museum collections.
-For instance, rare species may never have been collected or have only been collected a few times, and the corresponding records have not been digitized yet.
+For instance, rare species may never have been collected or may have only been collected a few times, and the corresponding records have not been digitized yet.
 Species that are difficult to preserve because they are large, very small, or fragile, are deposited in museums less frequently.
@@
 Missing geographic coordinates reduces the overlap between the list of species from the regional list and the databases.
-Using iDigBio, we found that for the `r format(not_really_in_database$idig_gom_n_spp_not_in_db, big.mark=",")` species that do not appear within the geographic range of the Gulf of Mexico but are on the list, `r format_output(not_really_in_database$idig_gom_p_spp_not_in_db*100)`% are completely absent from iDigBio, while for the `r format(not_really_in_database$idig_gom_n_spp_no_coords, big.mark=",")` species that do appear in iDigBio, `r format_output(not_really_in_database$idig_gom_p_spp_no_coords*100)`%  do not have geographic coordinates associated with them, and for the ones that do have geographic coordinates, `r format_output(not_really_in_database$idig_gom_p_spp_within_usa*100)`% ($n=$  `r format(not_really_in_database$idig_gom_n_spp_within_usa, big.mark=",")`) fall within US waters.
-Additional digitization of collections, including the capture of their geographic coordinates information, are needed to improve biodiversity knowledge at regional scales.
+Using iDigBio, we found that for the `r format(not_really_in_database$idig_gom_n_spp_not_in_db, big.mark=",")` species that do not appear within the geographic range of the Gulf of Mexico but are on the list, `r format_output(not_really_in_database$idig_gom_p_spp_not_in_db*100)`% are completely absent from iDigBio, while for the `r format(not_really_in_database$idig_gom_n_spp_no_coords, big.mark=",")` species that do appear in iDigBio, `r format_output(not_really_in_database$idig_gom_p_spp_no_coords*100)`%  do not have geographic coordinates associated with them, and for those that do have geographic coordinates, `r format_output(not_really_in_database$idig_gom_p_spp_within_usa*100)`% ($n=$  `r format(not_really_in_database$idig_gom_n_spp_within_usa, big.mark=",")`) fall within US waters.
+Additional digitization of collections, including the capture of their geographic coordinates information, is needed to improve biodiversity knowledge at regional scales.
 For instance, one can expect that the digitization coverage for the Gulf of Mexico will greatly increase once records from the Florida Wildlife Research Institute collections (St Petersburg, FL) become integrated into biodiversity information systems.
 
-Two factors most likely contribute to explaining why some species are in databases but are not included in the regional the checklists: (1) the species have been documented for the first time since the publication of the checklist, (2) the species records were not accessible or known to the authors.
+Two factors most likely contribute to explanations of why some species are in databases but are not included in the regional the checklists: (1) the species have been documented for the first time since the publication of the checklist, and (2) the species records were not accessible or known to the authors.
 
-Recently recorded species contribute modestly to this discrepancy for the Gulf of Mexico, as `r sum(recently_collected$gom_in_dbs$min_year > 2001)` species (`r format_output( sum(recently_collected$gom_in_dbs$min_year > 2001)/n_distinct(total_gom_records$worms_valid_name) * 100)`% of the total species richness, `r format_output(sum(recently_collected$gom_in_dbs$min_year > 2001)/spp_not_in_lists$gom_not_in_list * 100)`% of the species not on the checklist),  have been collected for the first time since the publication of the checklist, but more significantly for the Pacific Northwest with  `r sum(recently_collected$pnw_in_dbs$min_year > 1996)` species (`r format_output(sum(recently_collected$pnw_in_dbs$min_year > 1996)/n_distinct(total_pnw_records$worms_valid_name) * 100)`% of the total species richness,  `r format_output(sum(recently_collected$pnw_in_dbs$min_year > 1996)/spp_not_in_lists$pnw_not_in_list * 100)`% of the species not on the checklist).
+Recently recorded species contribute modestly to this discrepancy for the Gulf of Mexico, as `r sum(recently_collected$gom_in_dbs$min_year > 2001)` species (`r format_output( sum(recently_collected$gom_in_dbs$min_year > 2001)/n_distinct(total_gom_records$worms_valid_name) * 100)`% of the total species richness, `r format_output(sum(recently_collected$gom_in_dbs$min_year > 2001)/spp_not_in_lists$gom_not_in_list * 100)`% of the species not on the checklist),  have been collected for the first time since the publication of the checklist, but the results are more significant for the Pacific Northwest with  `r sum(recently_collected$pnw_in_dbs$min_year > 1996)` species (`r format_output(sum(recently_collected$pnw_in_dbs$min_year > 1996)/n_distinct(total_pnw_records$worms_valid_name) * 100)`% of the total species richness,  `r format_output(sum(recently_collected$pnw_in_dbs$min_year > 1996)/spp_not_in_lists$pnw_not_in_list * 100)`% of the species not on the checklist).
 

### 2017-10-04 manuscript.Rmd
refactor abstract

@@
 Museum collections have been recording our knowledge about species and their geographic ranges for about 250 years, but until recently, this information was fragmented across institutions and difficult to synthesize.
-The digitization of natural history collections and other biodiversity information systems can now be used to enable synthesis and highlight knowledge gaps.
-Here we use these data to synthesize the state of our knowledge for marine invertebrates in the continental United States.
-We show that current knowledge and sampling efforts are inadequate given the diversity of the fauna and the geographic extent of US.
-While recent high-throughput sequencing technologies promise to greatly speed up the documentation of marine species and their ranges, its value will be greatly improved by anchoring the results into existing DNA sequence data and associated vouchers.
-However, the best documented phyla have at most 25% of their species sequenced for even the most widely used DNA barcoding gene, cytochrome oxidase I (COI).
+The digitization of biodiversity information is now enabling the synthesis of this information and can be used to highlight knowledge gaps.
+Here we use publicly accessible biodiversity data to assemble a list of `r format(n_distinct(combined_species$worms_valid_name), big.mark = ",")` species of marine invertebrates from the continental United States.
+The extent of our knowledge about species richness is limited by sampling effort.
+While recent developments in high-throughput sequencing promise to greatly speed up the documentation of marine species and their ranges, its value will be improved by anchoring the results into existing DNA sequence data and associated vouchers.
+However, the best documented phyla have at most 25% of their species sequenced for even the most widely used DNA barcoding gene, cytochrome oxidase I (COI), but for many species these records have been sampled in other biogeographic areas.
 More sampling, coordinated at the national scale, is needed to improve our knowledge of species and their geographic ranges.

### 2017-10-04 manuscript.Rmd
edit  barcoding results

@@
 
-The marine invertebrates of the US have generally been characterized genetically at a higher level than other parts of the world: `r format_output(prop_spp_barcoded$p_barcoded * 100)`% (n = `r format(nrow(combined_species), big.mark=",")`) of the fauna has DNA barcodes available.
+The marine invertebrates of the US have generally been characterized genetically at a higher level than other parts of the world: `r format_output(prop_spp_barcoded$p_barcoded * 100)`% (n = `r format(nrow(combined_species), big.mark=",")`) of the digitized fauna has DNA barcodes available.
 In 2012, it was estimated that, at a global scale, less than 3% of the marine species had been sequenced [@Appeltans2012], and that this number was even lower in some areas (e.g., 2.6% for the Antarctic [@Grant2009]).
@@
 For example, up to `r format_output(100 * stat_bold$p_spp_multi_bin)`% of `r format(stat_bold$n_spp_not_singleton, big.mark=",")` species represented by more than 1 record in BOLD include cryptic species, as denoted by assignment to  more than 1 BIN (BOLD Index Number, operational taxonomic units that closely resemble species [@Ratnasingham2013]).
-Specimens identified under the same species but sampled from distant locations, are especially prone to representing cryptic species.
-Geographic information is missing for most records (`r format_output(100*stat_bold$p_no_coords)`% of the `r format(stat_bold$n_total_records, big.mark=",")` BOLD records analyzed here have no geographic coordinate information associated with them).
-It is thus crucial to ensure that precise and accurate geographic information is provided when DNA barcodes are published, and that enough samples for each species are sequenced, to capture more accurately their geographic range and the extent of their genetic diversity.
+Specimens identified under the same species but sampled from distant locations are especially prone to representing cryptic species.
+While geographic information is missing for most records (`r format_output(100*stat_bold$p_no_coords)`% of the `r format(stat_bold$n_total_records, big.mark=",")` BOLD records analyzed here have no geographic coordinate information associated with them), of the `r bold_geography_stats$n_spp_with_records` marine invertebrates species from the US with BOLD records, only `r format_output(bold_geography_stats$p_within_eez * 100)`% have at least one record collected within the EEZ, and `r format_output(bold_geography_stats$p_in_north_america * 100)`% list a North American country as the site of collection.
+Although DNA barcodes are available for `r format_output(prop_spp_barcoded$p_barcoded * 100)`% of the marine invertebrate fauna, many of these records have been collected outside the Atlantic and Pacific US biogeographic regions, and therefore geographical differentiation has not been evaluated for most taxa.
+It is thus crucial to ensure that precise and accurate geographic information is provided when DNA barcodes are published, and that enough samples for each species are sequenced to capture more accurately their geographic range and the extent of their genetic diversity.
 

### 2017-10-04 manuscript.Rmd
add authors

@@
   bookdown::pdf_document2:
-    #includes:
+    includes:
     #  before_body: ../template/doc_prefix.tex
-    #  in_header: ../template/preamble.tex
+      in_header: preamble.tex
     keep_tex: yes

### 2017-10-04 preamble.tex
add authors

\usepackage{authblk}
\author[1]{Fran\c{c}ois Michonneau\thanks{francois.michonneau@gmail.com}}
\author[2]{Gustav Paulay\thanks{paulay@flmnh.ufl.edu}}
\author[1]{Mark Q. Martindale\thanks{mqmartin@whitney.ufl.edu}}

\affil[2]{Florida Museum of Natural History, University of Florida, Gainesville,
  FL 32611-7800, USA}

\affil[2]{Whitney Lab}
\renewcommand\Authands{ and }
\date{}

%%% Local Variables:
%%% mode: markdown
%%% TeX-master: "manuscript.tex"
%%% End:

### 2017-10-04 manuscript.Rmd
move abstract to yaml header

@@
 bibliography: /home/francois/Library/dark-plankton.bib
+abstract: Museum collections have been recording our knowledge about species and their geographic ranges for about 250 years, but until recently, this information was fragmented across institutions and difficult to synthesize.
+The digitization of biodiversity information is now enabling the synthesis of this information, and can be used to highlight knowledge gaps.
+Here we use publicly accessible biodiversity data to assemble a list of `r format(n_distinct(combined_species$worms_valid_name), big.mark = ",")` species of marine invertebrates from the continental United States.
+Our knowledge about species richness and their geographic ranges is limited by sampling effort, as `r prop_10_specimens_species`% of the species are known from less than 10 occurrences.
+While recent developments in high-throughput sequencing promise to greatly speed up the documentation of marine biodiversity, anchoring results from such studies into existing DNA sequence data and associated vouchers would provide a richer biological context to interpret the insights provided by these datasets.
+However, with `r format_output(prop_spp_barcoded$p_barcoded * 100)`% of the marine invertebrate fauna that has sequences available for the most widely used DNA barcoding gene (cytochrome oxidase I, COI), and given that for many species these records have been sampled in biogeographic areas outside the USA, additional DNA barcoding efforts are needed to assemble a comprehensive library of reference sequences.
+More sampling, coordinated at the national scale, is needed to improve our knowledge of species and their geographic ranges.
+This data will allow the detection of the impact of climate change and other anthropogenic strains to marine biodiversity.
+Advances in computing, information systems, and sequencing technologies, have set the stage to inform where future efforts for documenting biodiversity should be applied.
 ---
@@
 ```
-
-<!-- use Pimm+2014 intro as model on using short informative sentences to build -->
-<!-- 1st paragraph. -->
-
-<!-- * how do we conserve animal species that we can't see, we don't find pretty, don't have eyes -->
-<!-- * Most species are rare, especially in very diverse places,  and the degradation of habitats makes it even more difficult to find these rare species. -->
-
-## Abstract
-
-Museum collections have been recording our knowledge about species and their geographic ranges for about 250 years, but until recently, this information was fragmented across institutions and difficult to synthesize.
-The digitization of biodiversity information is now enabling the synthesis of this information and can be used to highlight knowledge gaps.
-Here we use publicly accessible biodiversity data to assemble a list of `r format(n_distinct(combined_species$worms_valid_name), big.mark = ",")` species of marine invertebrates from the continental United States.
-The extent of our knowledge about species richness is limited by sampling effort.
-While recent developments in high-throughput sequencing promise to greatly speed up the documentation of marine species and their ranges, its value will be improved by anchoring the results into existing DNA sequence data and associated vouchers.
-However, the best documented phyla have at most 25% of their species sequenced for even the most widely used DNA barcoding gene, cytochrome oxidase I (COI), but for many species these records have been sampled in other biogeographic areas.
-More sampling, coordinated at the national scale, is needed to improve our knowledge of species and their geographic ranges.
-This data is needed to detect the impact of climate change and other anthropogenic strains to the environment.
-Advances in computing, information systems, and sequencing technologies, have set the stage to inform where future efforts for documenting biodiversity should be applied.
 

### 2017-10-12 R/gom_list.R
remove obsolete check on data quality

@@ -83,9 +83,7 @@ summarize_richness_per_db <- function(bold_db, idig_db, obis_db, gbif_db, region
                    !!col_name := sum(is.na(!!sym(col_name)))
                )
 
-    browser()
     stopifnot(all(idig_[["n_idigbio"]] >= idig_[["n_idigbio_in_us"]]))
-    stopifnot(all(idig_[["n_idigbio"]] >= idig_[[paste0("n_idigbio_in_", region)]]))
 
     obis_ <- obis_db %>%
         dplyr::group_by(phylum) %>%

### 2017-10-12 manuscript.Rmd
move section about US biodiversity first

@@
 ## Results
-
-### Comparison with Regional Lists by Geography
-
-We extracted `r format(sum(gom_worms$is_marine, na.rm=TRUE), big.mark=",")` and `r format(sum(kozloff_worms$is_marine, na.rm=TRUE), big.mark=",")` marine invertebrate species from species checklists from the Gulf of Mexico [@Tunnell2009] and from the Pacific Northwest [@Kozloff1996] respectively.
-For the Gulf of Mexico, biodiversity databases included more species for the same geographic area as the one used by the authors of the checklist (iDigBio and OBIS combined: `r format(dplyr::n_distinct(combined_gom_records$worms_valid_name), big.mark=",")` species; iDigBio only: `r format(n_distinct(idigbio_gom_records$worms_valid_name), big.mark=",")` species, $n =$  `r format(nrow(idigbio_gom_records), big.mark=",")` records; OBIS only: `r format(n_distinct(obis_gom_records$worms_valid_name), big.mark=",")` species, $n=$ `r format(nrow(obis_gom_records), big.mark=",")` records).
-For the PNW, fewer species than included in the regional checklist were recovered from biodiversity databases (iDigBio and OBIS combined: `r format(dplyr::n_distinct(combined_pnw_records$worms_valid_name), big.mark=",")` species; iDigBio only: `r format(dplyr::n_distinct(idigbio_pnw_records$worms_valid_name), big.mark=",")`  species, $n=$ `r format(nrow(idigbio_pnw_records), big.mark=",")` records; OBIS only `r format(dplyr::n_distinct(obis_pnw_records$worms_valid_name), big.mark=",")` species, $n=$ `r format(nrow(obis_pnw_records), big.mark=",")` records).
-The combination of the species listed in these three sources leads to `r format(n_distinct(total_gom_records$worms_valid_name), big.mark=",")` species for the GOM and `r format(n_distinct(total_pnw_records$worms_valid_name), big.mark=",")` species for the PNW.
-
-```{r compare_database_coverage, fig.cap='Overlap between the regional species list (list), records in iDigBio (idigbio), records in OBIS (obis), and records in BOLD (bold) for the Gulf of Mexico (left) and the Pacific Northwest (right).
-A higher proportion of species in the regional checklists are captured in databases for the Gulf of Mexico compared with the Pacific Northwest.
-This plot is generated using the UpSet technique[@Lex2014;@Conway2017], which uses a matrix-based layout to represent intersections of sets and their sizes.
-The rows represent the sets, the columns represent the intersections, and each set that is part of the intersection is marked with a black-filled circle in the cells underneath.
-Black lines connect the sets that are part of the intersections.
-(\\#fig:database-coverage)', echo=FALSE}
-knitr::include_graphics("figures/database_coverage.pdf")
-```
-
-```{r}
-set_size_gom <- function(...)
-    set_size_with_checklist(..., d = remake::fetch("total_gom_species"))
-set_size_pnw <- function(...)
-    set_size_with_checklist(..., d = remake::fetch("total_pnw_species"))
-```
-
-A higher proportion of the species included in the regional list for the GOM are captured in biodiversity databases than for the PNW.
-While `r format_output(set_size_gom("checklist","idigbio","obis")/n_distinct(total_gom_records$worms_valid_name) * 100)`%  of the species included in the checklist for the GOM have records in at least one of the biodiversity database (Fig.
-\@ref(fig:database-coverage)A), only `r format_output(set_size_pnw("checklist","idigbio","obis")/n_distinct(total_pnw_records$worms_valid_name)*100)`% do for the PNW  (Fig.
-\@ref(fig:database-coverage)B).
-Additionally, a higher proportion of species occur in the biodiversity databases but are not included in the regional checklist for the GOM compared with the PNW (GOM: `r format_output(set_size_gom("checklist","idigbio","obis",include_checklist=FALSE) / n_distinct(total_gom_records$worms_valid_name)*100)`%, $n =$ `r format(n_distinct(total_gom_records$worms_valid_name), big.mark=",")`; PNW:
-`r format_output( set_size_pnw("checklist","idigbio","obis",include_checklist=FALSE) / n_distinct(total_pnw_records$worms_valid_name)*100)`%, $n =$ `r format(n_distinct(total_pnw_records$worms_valid_name), big.mark=",")`).
-
-```{r compare_database_overlap, fig.cap='Proportions of the total number of species by phylum that are absent from faunistic lists but are in iDigBio or OBIS ("Not in list"), and absent from iDigBio or OBIS but are in the faunistic lists ("Not in databases").
-Many species, even in diverse taxonomic groups, known to occur in these regions have no digital records.
-The numbers represent the number of species.
-(\\#fig:database-overlap)'}
-knitr::include_graphics("figures/database_overlap_percent.pdf")
-```
-
-The more diverse phyla contribute the most to the discrepancies between the regional species lists and the database records (Fig.
-\@ref(fig:database-overlap)).
-OBIS tends to be closer to the full record set, but iDigBio has more species for the Gulf of Mexico that are not included in the list across most phyla.
-Many species of mollusks, arthropods, and annelids, known to occur in the Gulf of Mexico or in the Pacific Northwest, have no digital records in these geographic areas.
-
-<!-- Regional faunistic lists assembled and curated by taxonomic experts provide a comprehensive and detailed account of species that are known to occur in a given geographic area.
-We assessed how faunistic lists assembled over decades by specialists compare with records in public databases for the same geographic area.
--->
-
-Several factors may contribute to species being included in the regional lists but not having records in biodiversity databases.
-The first category of biases is linked to the likelihood of having specimens for these species in museum collections.
-For instance, rare species may never have been collected or may have only been collected a few times, and the corresponding records have not been digitized yet.
-Species that are difficult to preserve because they are large, very small, or fragile, are deposited in museums less frequently.
-The other category of biases is associated with data quality for these records.
-Some of these species may be in the databases but do not have geographic coordinates associated with them, and therefore would not be included in the results of our query.
-While it is difficult to assess biases associated with the likelihood of preservation of a species, we investigated how missing geographic coordinates are affecting the discrepancies between the checklists and the databases.
-
-Missing geographic coordinates reduces the overlap between the list of species from the regional list and the databases.
-Using iDigBio, we found that for the `r format(not_really_in_database$idig_gom_n_spp_not_in_db, big.mark=",")` species that do not appear within the geographic range of the Gulf of Mexico but are on the list, `r format_output(not_really_in_database$idig_gom_p_spp_not_in_db*100)`% are completely absent from iDigBio, while for the `r format(not_really_in_database$idig_gom_n_spp_no_coords, big.mark=",")` species that do appear in iDigBio, `r format_output(not_really_in_database$idig_gom_p_spp_no_coords*100)`%  do not have geographic coordinates associated with them, and for those that do have geographic coordinates, `r format_output(not_really_in_database$idig_gom_p_spp_within_usa*100)`% ($n=$  `r format(not_really_in_database$idig_gom_n_spp_within_usa, big.mark=",")`) fall within US waters.
-Additional digitization of collections, including the capture of their geographic coordinates information, is needed to improve biodiversity knowledge at regional scales.
-For instance, one can expect that the digitization coverage for the Gulf of Mexico will greatly increase once records from the Florida Wildlife Research Institute collections (St Petersburg, FL) become integrated into biodiversity information systems.
-
-Two factors most likely contribute to explanations of why some species are in databases but are not included in the regional the checklists: (1) the species have been documented for the first time since the publication of the checklist, and (2) the species records were not accessible or known to the authors.
-
-Recently recorded species contribute modestly to this discrepancy for the Gulf of Mexico, as `r sum(recently_collected$gom_in_dbs$min_year > 2001)` species (`r format_output( sum(recently_collected$gom_in_dbs$min_year > 2001)/n_distinct(total_gom_records$worms_valid_name) * 100)`% of the total species richness, `r format_output(sum(recently_collected$gom_in_dbs$min_year > 2001)/spp_not_in_lists$gom_not_in_list * 100)`% of the species not on the checklist),  have been collected for the first time since the publication of the checklist, but the results are more significant for the Pacific Northwest with  `r sum(recently_collected$pnw_in_dbs$min_year > 1996)` species (`r format_output(sum(recently_collected$pnw_in_dbs$min_year > 1996)/n_distinct(total_pnw_records$worms_valid_name) * 100)`% of the total species richness,  `r format_output(sum(recently_collected$pnw_in_dbs$min_year > 1996)/spp_not_in_lists$pnw_not_in_list * 100)`% of the species not on the checklist).
-
-The lack of tools for the authors of these checklists to have access to museum records across institutions explains why some species were not included in the checklists.
-Many of these records were difficult to discover before digitization which, for the most part, has occurred after the compilation of these checklists.
-We estimate that using biodiversity databases increases the number of species included in the checklist by `r format_output(spp_not_in_lists$gom_not_in_list/n_distinct(total_gom_species$worms_valid_name)*100)`% for the Gulf of Mexico, and `r format_output(spp_not_in_lists$pnw_not_in_list/n_distinct(total_pnw_species$worms_valid_name)*100)`% for the Pacific Northwest.
 
@@
 ```
+
+### Comparison with Regional Lists by Geography
+
+We extracted `r format(sum(gom_worms$is_marine, na.rm=TRUE), big.mark=",")` and `r format(sum(kozloff_worms$is_marine, na.rm=TRUE), big.mark=",")` marine invertebrate species from species checklists from the Gulf of Mexico [@Tunnell2009] and from the Pacific Northwest [@Kozloff1996] respectively.
+For the Gulf of Mexico, biodiversity databases included more species for the same geographic area as the one used by the authors of the checklist (iDigBio and OBIS combined: `r format(dplyr::n_distinct(combined_gom_records$worms_valid_name), big.mark=",")` species; iDigBio only: `r format(n_distinct(idigbio_gom_records$worms_valid_name), big.mark=",")` species, $n =$  `r format(nrow(idigbio_gom_records), big.mark=",")` records; OBIS only: `r format(n_distinct(obis_gom_records$worms_valid_name), big.mark=",")` species, $n=$ `r format(nrow(obis_gom_records), big.mark=",")` records).
+For the PNW, fewer species than included in the regional checklist were recovered from biodiversity databases (iDigBio and OBIS combined: `r format(dplyr::n_distinct(combined_pnw_records$worms_valid_name), big.mark=",")` species; iDigBio only: `r format(dplyr::n_distinct(idigbio_pnw_records$worms_valid_name), big.mark=",")`  species, $n=$ `r format(nrow(idigbio_pnw_records), big.mark=",")` records; OBIS only `r format(dplyr::n_distinct(obis_pnw_records$worms_valid_name), big.mark=",")` species, $n=$ `r format(nrow(obis_pnw_records), big.mark=",")` records).
+The combination of the species listed in these three sources leads to `r format(n_distinct(total_gom_records$worms_valid_name), big.mark=",")` species for the GOM and `r format(n_distinct(total_pnw_records$worms_valid_name), big.mark=",")` species for the PNW.
+
+```{r compare_database_coverage, fig.cap='Overlap between the regional species list (list), records in iDigBio (idigbio), records in OBIS (obis), and records in BOLD (bold) for the Gulf of Mexico (left) and the Pacific Northwest (right).
+A higher proportion of species in the regional checklists are captured in databases for the Gulf of Mexico compared with the Pacific Northwest.
+This plot is generated using the UpSet technique[@Lex2014;@Conway2017], which uses a matrix-based layout to represent intersections of sets and their sizes.
+The rows represent the sets, the columns represent the intersections, and each set that is part of the intersection is marked with a black-filled circle in the cells underneath.
+Black lines connect the sets that are part of the intersections.
+(\\#fig:database-coverage)', echo=FALSE}
+knitr::include_graphics("figures/database_coverage.pdf")
+```
+
+```{r}
+set_size_gom <- function(...)
+    set_size_with_checklist(..., d = remake::fetch("total_gom_species"))
+set_size_pnw <- function(...)
+    set_size_with_checklist(..., d = remake::fetch("total_pnw_species"))
+```
+
+A higher proportion of the species included in the regional list for the GOM are captured in biodiversity databases than for the PNW.
+While `r format_output(set_size_gom("checklist","idigbio","obis")/n_distinct(total_gom_records$worms_valid_name) * 100)`%  of the species included in the checklist for the GOM have records in at least one of the biodiversity database (Fig.
+\@ref(fig:database-coverage)A), only `r format_output(set_size_pnw("checklist","idigbio","obis")/n_distinct(total_pnw_records$worms_valid_name)*100)`% do for the PNW  (Fig.
+\@ref(fig:database-coverage)B).
+Additionally, a higher proportion of species occur in the biodiversity databases but are not included in the regional checklist for the GOM compared with the PNW (GOM: `r format_output(set_size_gom("checklist","idigbio","obis",include_checklist=FALSE) / n_distinct(total_gom_records$worms_valid_name)*100)`%, $n =$ `r format(n_distinct(total_gom_records$worms_valid_name), big.mark=",")`; PNW:
+`r format_output( set_size_pnw("checklist","idigbio","obis",include_checklist=FALSE) / n_distinct(total_pnw_records$worms_valid_name)*100)`%, $n =$ `r format(n_distinct(total_pnw_records$worms_valid_name), big.mark=",")`).
+
+```{r compare_database_overlap, fig.cap='Proportions of the total number of species by phylum that are absent from faunistic lists but are in iDigBio or OBIS ("Not in list"), and absent from iDigBio or OBIS but are in the faunistic lists ("Not in databases").
+Many species, even in diverse taxonomic groups, known to occur in these regions have no digital records.
+The numbers represent the number of species.
+(\\#fig:database-overlap)'}
+knitr::include_graphics("figures/database_overlap_percent.pdf")
+```
+
+The more diverse phyla contribute the most to the discrepancies between the regional species lists and the database records (Fig.
+\@ref(fig:database-overlap)).
+OBIS tends to be closer to the full record set, but iDigBio has more species for the Gulf of Mexico that are not included in the list across most phyla.
+Many species of mollusks, arthropods, and annelids, known to occur in the Gulf of Mexico or in the Pacific Northwest, have no digital records in these geographic areas.
+
+<!-- Regional faunistic lists assembled and curated by taxonomic experts provide a comprehensive and detailed account of species that are known to occur in a given geographic area.
+We assessed how faunistic lists assembled over decades by specialists compare with records in public databases for the same geographic area.
+-->
+
+Several factors may contribute to species being included in the regional lists but not having records in biodiversity databases.
+The first category of biases is linked to the likelihood having specimens for these species in digitized museum collections.
+Rare species may only have been collected a few times, and the corresponding records have not been digitized yet.
+Species that are difficult to preserve because they are large, very small, or fragile, are deposited in museums less frequently.
+The other category of biases is associated with data quality for these records.
+Some of these species may be in the databases but do not have geographic coordinates associated with them, and therefore would not be included in the results of our query.
+While it is difficult to assess biases associated with the likelihood of preservation of a species, we investigated how missing geographic coordinates are affecting the discrepancies between the checklists and the databases.
+
+Missing geographic coordinates reduces the overlap between the list of species from the regional list and the databases.
+Using iDigBio, we found that for the `r format(not_really_in_database$idig_gom_n_spp_not_in_db, big.mark=",")` species that do not appear within the geographic range of the Gulf of Mexico but are on the list, `r format_output(not_really_in_database$idig_gom_p_spp_not_in_db*100)`% are completely absent from iDigBio, while for the `r format(not_really_in_database$idig_gom_n_spp_no_coords, big.mark=",")` species that do appear in iDigBio, `r format_output(not_really_in_database$idig_gom_p_spp_no_coords*100)`%  do not have geographic coordinates associated with them, and for those that do have geographic coordinates, `r format_output(not_really_in_database$idig_gom_p_spp_within_usa*100)`% ($n=$  `r format(not_really_in_database$idig_gom_n_spp_within_usa, big.mark=",")`) fall within US waters.
+Additional digitization of collections, including the capture of their geographic coordinates information, is needed to improve biodiversity knowledge at regional scales.
+For instance, one can expect that the digitization coverage for the Gulf of Mexico will greatly increase once records from the Florida Wildlife Research Institute collections (St Petersburg, FL) become integrated into biodiversity information systems.
+
+Two factors most likely contribute to explanations of why some species are in databases but are not included in the regional checklists: (1) the species have been documented for the first time since the publication of the checklist, and (2) the species records were not accessible or known to the authors.
+
+Recently recorded species contribute modestly to this discrepancy for the Gulf of Mexico, as `r sum(recently_collected$gom_in_dbs$min_year > 2001)` species (`r format_output( sum(recently_collected$gom_in_dbs$min_year > 2001)/n_distinct(total_gom_records$worms_valid_name) * 100)`% of the total species richness, `r format_output(sum(recently_collected$gom_in_dbs$min_year > 2001)/spp_not_in_lists$gom_not_in_list * 100)`% of the species not on the checklist),  have been collected for the first time since the publication of the checklist, but the results are more significant for the Pacific Northwest with  `r sum(recently_collected$pnw_in_dbs$min_year > 1996)` species (`r format_output(sum(recently_collected$pnw_in_dbs$min_year > 1996)/n_distinct(total_pnw_records$worms_valid_name) * 100)`% of the total species richness,  `r format_output(sum(recently_collected$pnw_in_dbs$min_year > 1996)/spp_not_in_lists$pnw_not_in_list * 100)`% of the species not on the checklist).
+
+The lack of tools for the authors of these checklists to have access to museum records across institutions explains why some species were not included in the checklists.
+Many of these records were difficult to discover before digitization which, for the most part, has occurred after the compilation of these checklists.
+We estimate that using biodiversity databases increases the number of species included in the checklist by `r format_output(spp_not_in_lists$gom_not_in_list/n_distinct(total_gom_species$worms_valid_name)*100)`% for the Gulf of Mexico, and `r format_output(spp_not_in_lists$pnw_not_in_list/n_distinct(total_pnw_species$worms_valid_name)*100)`% for the Pacific Northwest.
 

### 2017-10-12 manuscript.Rmd
incorporate GP comments and suggestions

@@
 While recent developments in high-throughput sequencing promise to greatly speed up the documentation of marine biodiversity, anchoring results from such studies into existing DNA sequence data and associated vouchers would provide a richer biological context to interpret the insights provided by these datasets.
-However, with `r format_output(prop_spp_barcoded$p_barcoded * 100)`% of the marine invertebrate fauna that has sequences available for the most widely used DNA barcoding gene (cytochrome oxidase I, COI), and given that for many species these records have been sampled in biogeographic areas outside the USA, additional DNA barcoding efforts are needed to assemble a comprehensive library of reference sequences.
+However, with `r format_output(prop_spp_barcoded$p_barcoded * 100)`% of the marine invertebrate fauna that has sequences available for the most widely used DNA barcoding gene (cytochrome oxidase I, COI), and given that for many species these records have been sampled in biogeographic areas outside the USA, additional efforts are needed to assemble a comprehensive library of reference DNA barcoding sequences.
 More sampling, coordinated at the national scale, is needed to improve our knowledge of species and their geographic ranges.
-This data will allow the detection of the impact of climate change and other anthropogenic strains to marine biodiversity.
+This data will allow the detection of the impacts of climate change and other anthropogenic strains on marine biodiversity.
 Advances in computing, information systems, and sequencing technologies, have set the stage to inform where future efforts for documenting biodiversity should be applied.
@@
 -->  While terrestrial ecosystems have received considerable attention, the baseline of healthy and resilient marine ecosystems is not well understood [@Worm2006], and modeling the impact of human activities has been limited to a few taxonomic groups [@Tittensor2010].
-For marine organisms, because of the diversity of their complex life history strategies, their size range <!-- reformulate?
--->, and the difficulties associated with sampling their habitats, we have limited information about which species live where, or what portion of their diversity remains to be discovered and described [@Mora2011].
+For marine organisms, because of the diversity of their complex life history strategies, their body size range, and the difficulties associated with sampling their habitats, we have limited information about which species live where, or what portion of their diversity remains to be discovered and described [@Mora2011].
 Additionally, coordinated marine ecosystem monitoring initiatives have been lacking [@Duffy2013,@Borja2016] (but are now getting organized, see MarineGEO [@Duffy2016marinegeo], MBON, and PISCO), making it difficult to monitor changes in species distributions and abundances.
 While most documented global extinctions of marine species are vertebrates, many other species may have been lost unnoticed because they were not known [@Sala2006].
-There is therefore a pressing need to improve our knowledge of marine systems before their functioning is irreversibly altered, and species go extinct.
+There is therefore a pressing need to improve our knowledge of marine systems before species go extinct, and their functioning is irreversibly altered.
 
@@
 
-Here, we focus on the continental US because the marine biodiversity is extensively documented as the result of a long history of studies by governmental, academic, and private institutions [@Fautin2010].
+We focus on the continental US because the marine biodiversity is extensively documented as the result of a long history of studies by governmental, academic, and private institutions [@Fautin2010].
 The continental US includes a wide diversity of marine ecosystems from the Florida Keys, to the Gulf of Maine on the East Coast; and from the arid Southern California to the oceanic shores of the Pacific Northwest on the West coast.
@@
 Spatial patterns of species richness are however strongly confounded by sampling effort.
-There is a high correlation between the number of samples and number of species per cell used in Fig.
+There is a tight correlation between the number of samples and number of species per cell used in Fig.
 \@ref(fig:map-diversity-sampling), and `r format_output(stat_sampling_effort$p * 100)`% (n = `r stat_sampling_effort$n`) of the cells with $\leq$ 500 records have a number of species within 10% of the number of records (Fig.
@@
 Yet at present, `r prop_10_specimens_species`% of species are represented by $\leq$ 10 records, and `r prop_30_specimens_species`%  by $\leq$ 30 records based on the combined records in iDigBio and OBIS.
-Additionally, `r format_output(prop_missing_depth$p_missing * 100)`% ($n=$ `r format(prop_missing_depth$n_total, big.mark=",")` OBIS records) of the observations lack bathymetric information.
 Underscoring the inadequacy of the sampling effort, `r prop_singleton_species`% of the species are known from a single record.
@@
 The above estimates are based on species names, but many traditional species include multiple cryptic species [@Knowlton2000;@Meyer2005-evolution].
-For example, up to `r format_output(100 * stat_bold$p_spp_multi_bin)`% of `r format(stat_bold$n_spp_not_singleton, big.mark=",")` species represented by more than 1 record in BOLD include cryptic species, as denoted by assignment to  more than 1 BIN (BOLD Index Number, operational taxonomic units that closely resemble species [@Ratnasingham2013]).
+For example, up to `r format_output(100 * stat_bold$p_spp_multi_bin)`% of `r format(stat_bold$n_spp_not_singleton, big.mark=",")` species that have DNA barcodes available in the digitized US fauna, are represented by more than 1 record in BOLD include cryptic species, as denoted by assignment to  more than 1 BIN (BOLD Index Number, molecular operational taxonomic units that closely resemble species [@Ratnasingham2013]).
 Specimens identified under the same species but sampled from distant locations are especially prone to representing cryptic species.
-While geographic information is missing for most records (`r format_output(100*stat_bold$p_no_coords)`% of the `r format(stat_bold$n_total_records, big.mark=",")` BOLD records analyzed here have no geographic coordinate information associated with them), of the `r bold_geography_stats$n_spp_with_records` marine invertebrates species from the US with BOLD records, only `r format_output(bold_geography_stats$p_within_eez * 100)`% have at least one record collected within the EEZ, and `r format_output(bold_geography_stats$p_in_north_america * 100)`% list a North American country as the site of collection.
+While geographic information is missing for most records (`r format_output(100*stat_bold$p_no_coords)`% of the `r format(stat_bold$n_total_records, big.mark=",")` BOLD records analyzed here have no geographic coordinate information associated with them), of the `r bold_geography_stats$n_spp_with_records` marine invertebrates species from the US with BOLD records, only `r format_output(bold_geography_stats$p_within_eez * 100)`% have at least one record collected within the EEZ and `r format_output(bold_geography_stats$p_in_north_america * 100)`% list a North American country as the site of collection.
 Although DNA barcodes are available for `r format_output(prop_spp_barcoded$p_barcoded * 100)`% of the marine invertebrate fauna, many of these records have been collected outside the Atlantic and Pacific US biogeographic regions, and therefore geographical differentiation has not been evaluated for most taxa.
@@
 
-## Discussion
+## Recommendations
 
@@
 Linking species names to identifiers in taxonomic databases would facilitate data linking, specimen discoverability, and data usability.
+<!-- GP comments: You still have the human knowledge/ID factor to contend with though, which is a more important issue.
+Also,places like WoRMS almost never link a given name to multiple useages and useage changes with time -->
 Additionally, because registers of species names, such as WoRMS, link the history of the names and their synonymies, it is possible to dissociate the concepts of species identification with the concept of species identity, allowing researchers to compare specimens identified at different time periods.
@@
 Such services, however, require that resources be made available to museum collections in order to address these issues directly in collection databases.
+<!-- GP comment: see geolocate plugin in Specify -->
 
@@
 Another venue to improve the quality of the data in biodiversity information systems, is to develop better incentives from journal editors and funding agencies for scientists to publish their occurrence records in databases alongside their papers as it is done for genetic data with GenBank [@Huang2011e].
+
+<!-- Finally - need better links in GenBank and BOLD to vouchers, names, and geography -->
 
@@
 
+<!-- GP comment: Several of these recommendations match ones made in Duffy et al 2013 -- ref 4  Should emphasize the overlap -->
+
 Museum specimens are critical for validating the occurrence of species [@Rocha2014], especially for marine invertebrates for which species limits based on morphology have been known to be challenged by other lines of evidence including molecular data [@Knowlton2000].
@@
 
+<!-- GP comment: You could give a biodiversity survey as an example - e.g.
+in Kaneohe we did ~2400 records of ~1300 species in 11 days - in context of whole US effort, this is substantial, especially at the species level.
+-->
+
 Samples collected during these expeditions need to be photographed to document live appearance, information that would otherwise be lost during the preservation process.
+This has been shown to be particularly important to validate species limits in cryptic complexes.
 We also need to capture metadata associated with specimens to be able to understand better the physical characteristics of the habitats and the life histories of these species.
@@
 Additionally, data generated by metabarcoding efforts can be re-interpreted as more DNA barcodes associated with correctly identified specimens become available.
-Building the library of DNA barcodes in parallel with metabarcoding efforts ensures that database matches correspond to actual specimens deposited in natural history collections, ruling out possible artifacts of the method.
+Building the library of DNA barcodes in parallel with metabarcoding efforts ensures that database matches correspond to actual specimens deposited in natural history collections, ruling out possible artifacts of the methods and permits resolution of specimen identification conflicts.
 With such approaches, the geographic distribution of species can build on and be linked to existing records, species and their natural history, rather than being limited to matching DNA sequences across studies.
@@
 Even if technological advances can greatly speed up the documentation of species and their distributions, taxonomists and experts are needed  to collect, sort, process, and identify samples.
+Taxonomists play a crucial role in updating and synthesizing species-level knowledge that can be used to provide accurate identifications for DNA barcoded samples.
 Taxonomic expertise within the US is dwindling, and there are now only a handful of US natural history museums with active collections of marine invertebrates.
@@
 Species that could not be matched in WoRMS, and species listed as non-marine in WoRMS were not included in subsequent analyses.
-For all names that were listed in WoRMS, the current accepted name was capture and used in all analyses.
+For all names that were listed in WoRMS, the current accepted name was captured and used in all analyses.
 This ensured that the same species listed under different synonyms was only counted once.
@@
 We would also like to acknowledge the efforts of rOpenSci (https://ropensci.org) for developping tools to make these data sources available in R.
-We are grateful to Julie Allen, Robert Lasley, Kevin Love, and Pamela Soltis, for providing suggestions that improved the manuscript.
+We are grateful to Julie Allen, Robert Lasley, Kevin Love, and Pamela Soltis for providing suggestions that improved the manuscript.
 

### 2017-10-13 R/combine_lists.R
can't assert is marine here as column absent from dataset

@@ -11,12 +11,10 @@ deduplicate_records <- function(data) {
 combine_records <- function(..., map) {
     d <- list(...)
     d <- year_as_integer(d)
-    d <- d %>%
+    d %>%
         dplyr::bind_rows() %>%
         deduplicate_records() %>%
         add_geo(map = map)
-    assert_all_marine(d)
-    d
 }
 
 ## add info on whether a record is on the East coast, West coast, or the Gulf of

### 2017-10-13 manuscript.Rmd
refine record deduplication

@@
 
+To generate the combined list of records from iDigBio and OBIS, we removed putative duplicate based on the following criteria: same phylum, same accepted species name from WoRMS, same geographic coordinates (rounded to the 4th decimal), and same date of collection.
+When the date is missing, a single record per species with identical coordinates is included.
+
 <!-- include results earlier? -->

### 2017-10-13 R/combine_lists.R

@@ -2,10 +2,15 @@ deduplicate_records <- function(data) {
     data %>%
         dplyr::filter(!is.na(worms_valid_name),
                       is_marine == TRUE) %>%
-        dplyr::select(phylum, worms_valid_name, decimallatitude,
+        dplyr::mutate(approx_lat = round(decimallatitude, 4),
+                      approx_lon = round(decimallongitude, 4)) %>%
+        dplyr::distinct(worms_phylum, worms_class, worms_order,
+                        worms_valid_name, approx_lat, approx_lon,
+                        datecollected, .keep_all = TRUE) %>%
+        dplyr::select(worms_phylum, worms_class, worms_order, worms_family,
+                      worms_valid_name, decimallatitude,
                       decimallongitude, datecollected, year) %>%
-        dplyr::distinct(phylum, worms_valid_name, decimallatitude,
-                        decimallongitude, datecollected, .keep_all = TRUE)
+        dplyr::rename(phylum = worms_phylum)
 }
 
 combine_records <- function(..., map) {

### 2017-10-13 manuscript.Rmd
[fix #81, fix #82] revisit latitudinal gradient

@@
 
-```{r latitudinal-gradient, fig.cap='Latitudinal gradient in species richness for the Pacific and Atlantic oceans (Gulf of Mexico records excluded).
-Trend lines are B-splines of the 3rd degree polynomial weighted by the square-root of the number of records.
+```{r latitudinal-gradient, fig.cap='Latitudinal gradient in species richness for the Pacific and Atlantic oceans (Gulf of Mexico records excluded), based on inferred species richness (A) and recorded species richness (B).
+The patterns are strongly influenced by the sampling effort (C).
+Trend lines are rolling means with a window of size of 3.
 (\\#fig:latitudinal-gradient)'}
-knitr::include_graphics("figures/combined_gradient.pdf")
+knitr::include_graphics("figures/gradient.pdf")
 ```
@@
 
-<!-- include results earlier? -->
+To infer the number of species along the latitudinal gradient (Fig.
+\@ref(fig:latitudinal-gradient)), we estimated the species ranges by using the southernmost and northernmost observations for each species on the Atlantic and Pacific coast separately.
+Records for the Gulf of Mexico were not included in this analysis.
 

### 2017-10-13 R/combine_plots.R

@@ -35,3 +35,13 @@ simple_cowplot <- function(g1, g2) {
                    scale = .9)
     plot_grid(p, leg, ncol = 1, rel_heights = c(1, .2))
 }
+
+combine_cowplots_3 <- function(g1, g2, g3) {
+    g1 <- g1 + theme(legend.position = "none")
+    g2 <- g2 + theme(legend.position = "none")
+    g3 <- g3 + theme(legend.position = "none")
+    theme_set(theme_cowplot(font_size = 10))
+    plot_grid(g1, g2, g3, labels = "AUTO",
+              nrow = 3, align = "v")
+
+}

### 2017-10-13 R/gradient.R

@@ -1,6 +1,6 @@
 extract_latitudinal_ranges <- function(d) {
     extract_lat_range <- . %>%
-        dplyr::group_by(phylum, worms_valid_name) %>%
+        dplyr::group_by(phylum, worms_class, worms_order, worms_family, worms_valid_name) %>%
         dplyr::summarize(
                    min_latitude = min(decimallatitude),
                    max_latitude = max(decimallatitude),
@@ -24,14 +24,17 @@ stats_latitude_extent <- function(lat_data) {
     ## Proportion of species with more than 10 records that have latitudinal
     ## range of more than 5°
     lat_data %>%
-        dplyr::filter(n_records >= 10) %>%
+        dplyr::filter(n_records >= 5) %>%
         dplyr::group_by(coast, phylum) %>%
         dplyr::summarize(
+                   median_lat = median(diff_lat),
                    p_more_than_5 = mean(diff_lat > 5)
                )
 }
 
 if (FALSE) {
+    vv <- remake::fetch("combined_records")
+    tt <- extract_latitudinal_ranges(vv)
     ## show spread of geographic range for various phyla
     ggplot(filter(tt, n_records >= 5,
                   phylum %in% c("arthropoda", "mollusca",
@@ -41,7 +44,7 @@ if (FALSE) {
                                      min_latitude),
                          xend = reorder(reorder(worms_valid_name, diff_lat),
                                         min_latitude),
-                         color = phylum, y = min_latitude, yend = max_latitude)) +
+                         color = worms_class, y = min_latitude, yend = max_latitude)) +
         facet_grid(phylum ~ coast)
 }
 
@@ -50,42 +53,100 @@ n_spp_per_lat <- function(lat_data) {
     ## number of species for each latitudinal slice based on inferred ranges
 
     slice_size <- .2
+    min_lat <- floor(min(lat_data$min_latitude))
+    max_lat <- ceiling(max(lat_data$min_latitude))
 
-    res <- expand.grid(lat_coast = c("east", "west"),
-                lat_degrees = seq(22, 60, by = slice_size),
-                stringsAsFactors = FALSE)
+    res <- expand.grid(
+        lat_coast = c("east", "west"),
+        lat_degrees = seq(min_lat, max_lat, by = slice_size),
+        stringsAsFactors = FALSE)
 
-    res %>%
[…]

### 2017-10-23 manuscript.Rmd
[fix #77] don't include fossils in idigbio counts

@@
 `r n_distinct(na.omit(obis_records$worms_phylum))` phyla).
-Extant marine invertebrates occurring in the US represent less than 0.4% of the total number of iDigBio records (104+ million), and about 1% of the US records (33 million records).
-Marine invertebrates that occur in the US represent approximately 4% of the 48 million occurrence records in OBIS.
+Extant marine invertebrates occurring in the US represent less than `r format_output(nrow(idigbio_records)/idigbio_stats$total_no_fossil*100)`% of the total number of non-fossil iDigBio records (`r round(idigbio_stats$total_no_fossil/1e6, 1)` million), and `r format_output(nrow(idigbio_records)/idigbio_stats$usa_no_fossil*100)`% of the US records (`r round(idigbio_stats$usa_no_fossil/1e6, 1)` million).
+Marine invertebrates that occur in the US represent 4.1% of the 48 million occurrence records in OBIS.
 

### 2017-10-23 R/idigbio.R

@@ -53,6 +53,32 @@ fetch_spp_from_idigbio <- function(wrm) {
         dplyr::distinct(uuid, .keep_all = TRUE)
 }
 
+get_idigbio_stats <- function(map_usa) {
+    bbox_usa <- lawn::lawn_bbox(map_usa)
+    idig_non_fossil <- ridigbio::idig_count_records(list("basisofrecord" = "preservedspecimen"))
+    idig_non_fossil_usa <- ridigbio::idig_count_records(
+                                         list(
+                                             "basisofrecord" = "preservedspecimen",
+                                             "geopoint" = list(
+                                                 type = "geo_bounding_box",
+                                                 top_left = list(
+                                                     lon = bbox_usa[1],
+                                                     lat = bbox_usa[4]
+                                                 ),
+                                                 bottom_right = list(
+                                                     lon = bbox_usa[3],
+                                                     lat = bbox_usa[2]
+                                                 )
+                                             )
+                                         )
+                                     )
+    list(
+        total_no_fossil = idig_non_fossil,
+        usa_no_fossil = idig_non_fossil_usa
+    )
+}
+
+
 
 summarize_raw_idigbio <- function(idig, list_spp) {
     assertthat::assert_that(inherits(idig, "data.frame"))

### 2017-11-01 manuscript.Rmd
complete data + plot about kingdom stats

@@
 Marine invertebrates that occur in the US represent 4.1% of the 48 million occurrence records in OBIS.
+
+```{r kingdom-stats}
+get_vert_prop <- function(db) {
+    format_output(kingdom_diversity[kingdom_diversity$database == db & kingdom_diversity$sub_kingdom == "Animalia - Vertebrates", "prop_spp"] * 100)
+}
+
+total_vert_prop <- get_vert_prop("worms")
+idigbio_vert_prop <- get_vert_prop("idigbio")
+obis_vert_prop <- get_vert_prop("obis")
+```
+
+Marine invertebrates are underrepresented in iDigBio and OBIS.
+We compared the total number of species represented in iDigBio and OBIS for each kingdom to the total described diversity based on the numbers published by WoRMS.
+Records in thedatabases are biased towards vertebrates: while vertebrates represent  `r total_vert_prop`% of the global diversity, they represent `r idigbio_vert_prop`% and `r obis_vert_prop`% of the records identified at the species level in iDigBio and OBIS respectively (Fig.
+\@ref(fig:kingdom-diversity-plot)).
+Additionally, the median number of specimen per species
+
+```{r kingdom-diversity-plot, fig.cap='Comparison of the proportions of the diversity represented in iDigBio and OBIS  with the global diversity (as reported by WoRMS) for each kingdom.
+Vertebrates are overrepresented, while other animal groups are underpresented.
+Total number of species are indicated next to the bars.(\\#fig:kingdom-diversity-plot)'}
+knitr::include_graphics("figures/plot_kingdom_diversity.pdf")
+```
 

### 2017-11-01 R/kingdom.R

@@ -146,35 +146,40 @@ kingdom_stats <- function(db_table) {
                                              "myxini",
                                              "petromyzonti",
                                              "reptilia") ~ "animalia - vertebrates",
-                          TRUE ~ worms_kingdom
+                          worms_kingdom == "animalia" ~ "animalia",
+                          worms_kingdom == "chromista" ~ "chromista",
+                          worms_kingdom == "plantae" ~ "plantae",
+                          TRUE ~ "others"
                       ))
 
 }
 
-plot_kingdom_diversity <- function(worms_stats) {
+calc_kingdom_diversity <- function(worms_stats) {
     get_n_spp <- . %>%
-        dplyr::group_by(worms_kingdom, sub_kingdom) %>%
+        dplyr::group_by(sub_kingdom) %>%
         dplyr::summarize(
-            n_spp = n_distinct(worms_valid_name)
+                   n_samples = n(),
+                   n_spp = n_distinct(worms_valid_name)
         ) %>%
         dplyr::collect()
 
     wrm <- worms_stats %>%
-        dplyr::mutate(worms_kingdom = tolower(kingdom),
-                      sub_kingdom = worms_kingdom) %>%
-        dplyr::group_by(worms_kingdom, sub_kingdom) %>%
+        dplyr::mutate(sub_kingdom = case_when(
+                          tolower(kingdom) %in% c("fungi", "protozoa", "bacteria",
+                                                  "archaea", "viruses", "biota incertae sedis") ~ "others",
+                          TRUE ~ tolower(kingdom)
+                      )) %>%
+        dplyr::group_by(sub_kingdom) %>%
         dplyr::summarize(
                    n_spp = sum(all_species_marine_non_fossil)
                ) %>%
         bind_rows(
             data_frame(
-                worms_kingdom =  "animalia",
                 sub_kingdom =  "animalia - vertebrates",
                 ## use number from Appeltans et al. 2012
                 n_spp = 17619)
         )
     wrm[wrm$sub_kingdom == "animalia", "n_spp"] <- wrm[wrm$sub_kingdom == "animalia", "n_spp"] - 17619
-
 
     ## diversity comparison
     bind_rows(
@@ -185,12 +190,37 @@ plot_kingdom_diversity <- function(worms_stats) {
[…]

### 2017-11-01 manuscript.Rmd
[fix #76] add global diversity, and median number of samples per sp.

--- 2017-10-23 manuscript.Rmd
@@
 Marine invertebrates that occur in the US represent 4.1% of the 48 million occurrence records in OBIS.
+
+```{r kingdom-stats}
+get_vert_prop <- function(db) {
+    format_output(kingdom_diversity[kingdom_diversity$database == db &
+                                    kingdom_diversity$sub_kingdom == "Animalia - Vertebrates", "prop_spp"] * 100)
+}
+
+total_vert_prop <- get_vert_prop("worms")
+idigbio_vert_prop <- get_vert_prop("idigbio")
+obis_vert_prop <- get_vert_prop("obis")
+
+kng_medians <- kingdom_stats %>%
+    dplyr::filter(database == "idigbio") %>% 
+    dplyr::group_by(sub_kingdom) %>%
+    dplyr::summarize(
+               rg = if_else(
+                   Reduce("==", range(median_sample)),
+                   as.character(range(median_sample)[1]), 
+                   paste(range(median_sample), collapse = " and "))
+           )
+
+```
+
+Marine invertebrates are underrepresented in iDigBio and OBIS given their diversity.
+We compared the total number of species represented in iDigBio and OBIS for each kingdom to the total described diversity based on the WoRMS estimates[@WorMS-stats].
+Records in the databases are biased towards vertebrates: while vertebrates represent  `r total_vert_prop`% of the global diversity, they represent `r idigbio_vert_prop`% and `r obis_vert_prop`% of the species in iDigBio and OBIS respectively (Fig.
+\@ref(fig:kingdom-diversity-plot)).
+Additionally, the median number of specimens in iDigBio per species ranges between `r kng_medians %>% filter(sub_kingdom=="animalia") %>% pull(rg)` for invertebrate phyla, but is `r kng_medians %>% filter(sub_kingdom=="animalia - vertebrates") %>% pull(rg)` for vertebrates.
+
+```{r kingdom-diversity-plot, fig.cap='Comparison of the proportions of the diversity represented in iDigBio and OBIS  with the global diversity (as reported by WoRMS) for each kingdom.
+Vertebrates are overrepresented, while other animal groups are underpresented.
+Total number of species are indicated next to the bars.(\\#fig:kingdom-diversity-plot)'}
+knitr::include_graphics("figures/plot_kingdom_diversity.pdf")
+```
 

### 2017-11-01 R/kingdom.R

@@ -208,6 +208,7 @@ plot_kingdom_diversity <- function(kng) {
         geom_text(aes(y = prop_spp + .01, label = n_spp), position = position_dodge(.9),
                       hjust = .1,  family = "Ubuntu Condensed") +
         scale_fill_hc(name = "Source", labels = c("iDigBio", "OBIS", "Global diversity (WoRMS)")) +
+        ylim(c(0, .85)) +
         xlab(NULL) + ylab("Proportion of total diversity (per source)") +
         theme_ipsum(base_family = "Ubuntu Condensed")
 
@@ -221,6 +222,31 @@ plot_kingdom_samples <- function(kng) {
     scale_fill_hc(name = "Source", labels = c("iDigBio", "OBIS", "Global diversity (WoRMS)")) +
     xlab(NULL) + ylab("Proportion of total diversity (per source)")
 }
+
+calc_kingdom_stats <- function() {
+
+    get_median <- . %>%
+        group_by(sub_kingdom, worms_phylum, worms_valid_name) %>%
+        summarize(
+            n_samples = n()
+        ) %>%
+        collect() %>%
+        group_by(sub_kingdom, worms_phylum) %>%
+        summarize(
+            median_sample = median(n_samples),
+            mean_sample = mean(n_samples)
+        )
+
+    dplyr::bind_rows(
+        idigbio = kingdom_stats("us_idigbio_clean") %>%
+            get_median(),
+        obis =  kingdom_stats("us_obis_clean") %>%
+            get_median(),
+        .id = "database")
+}
+
+
+
 if (FALSE) {
     ## number of species and records per (sub)kingdom
     idigbio_kingdom_stats("us_idigbio_clean")  %>%

### 2017-11-01 manuscript.Rmd
add institution contribution table

@@
 
+```{r records-table, results='asis'}
+print.xtableList(institution_contributions, colnames.format = "multiple",
+                 booktabs = TRUE, include.rownames = FALSE,
+                 table.placement = "!ht")
+```
+
 Marine invertebrates are underrepresented in iDigBio and OBIS given their diversity.
-We compared the total number of species represented in iDigBio and OBIS for each kingdom to the total described diversity based on the WoRMS estimates[@WorMS-stats].
+We compared the total number of species represented in iDigBio and OBIS for each kingdom to the total described diversity based on the WoRMS estimates[@worms-stats].
 Records in the databases are biased towards vertebrates: while vertebrates represent  `r total_vert_prop`% of the global diversity, they represent `r idigbio_vert_prop`% and `r obis_vert_prop`% of the species in iDigBio and OBIS respectively (Fig.

### 2017-11-01 R/institutions.R

@@ -1,8 +1,8 @@
-calc_institutions <- function(idig_records) {
+calc_institutions <- function(idig_records, obis_records) {
 
     res_idig <- idig_records %>%
         dplyr::count(institutioncode, sort = TRUE) %>%
-        dplyr::filter(n >= 500) %>%
+        dplyr::top_n(10, n) %>%
         dplyr::mutate(Institution = case_when(
                           institutioncode == "usnm" ~ "Smithsonian Institution National Museum of Natural History",
                           institutioncode == "uf" ~ "Florida Museum of Natural History, University of Florida",
@@ -19,11 +19,48 @@ calc_institutions <- function(idig_records) {
         purrr::pwalk(function(Institution, ...) {
                    if(any(grepl("problem", Institution))) stop("unknown collection")
                }) %>%
+        dplyr::mutate(n = format(n, big.mark = ",")) %>%
         dplyr::select(Institution,
                       "Number of records" = n)
 
     res_obis <- obis_records %>%
-        dplyr::count(institutioncode, )
+        dplyr::mutate(institutioncode_simple = case_when(
+                          grepl("noaa|nmfs|dsc_rtp", institutioncode) ~ "noaa",
+                          TRUE ~ institutioncode
+                      )) %>%
+        dplyr::count(institutioncode_simple, sort = TRUE) %>%
+        dplyr::filter(!is.na(institutioncode_simple)) %>%
+        ## regroup all noaa records
+        dplyr::top_n(10, n) %>%
+        dplyr::mutate(Institution = case_when(
+                          institutioncode_simple == "noaa" ~ "National Oceanographic and Atmostpheric Association",
+                          institutioncode_simple == "usnm" ~  "Smithsonian Institution National Museum of Natural History",
+                          institutioncode_simple == "tpwd" ~ "",
+                          institutioncode_simple == "boemre" ~ "Bureau of Ocean Energy Management, Regulation, and Enforcement",
+                          institutioncode_simple == "emap_nca" ~ "U.S. Environmental Protection Agency through its Environmental Monitoring and Assessment Program (EMAP)",
+                          institutioncode_simple == "usgs" ~ "U.S. Geological Services",
+                          institutioncode_simple == "mbari" ~ "Monterey Bay Aquarium Research Institute",
+                          institutioncode_simple == "seamap" ~ " Southeast Area Monitoring and Assessment Program",
[…]

### 2017-11-02 manuscript.Rmd
add more details about origin of OBIS records

@@
 ```{r records-table, results='asis'}
+xtbl_small <- function(x) {
+    glue::glue("{{\\footnotesize {x} }}")
+}
 print.xtableList(institution_contributions, colnames.format = "multiple",
                  booktabs = TRUE, include.rownames = FALSE,
+                 sanitize.message.function = xtbl_small, 
                  table.placement = "!ht")

### 2017-11-02 R/institutions.R

@@ -35,7 +35,7 @@ calc_institutions <- function(idig_records, obis_records) {
         dplyr::mutate(Institution = case_when(
                           institutioncode_simple == "noaa" ~ "National Oceanographic and Atmostpheric Association",
                           institutioncode_simple == "usnm" ~  "Smithsonian Institution National Museum of Natural History",
-                          institutioncode_simple == "tpwd" ~ "",
+                          institutioncode_simple == "tpwd" ~ "Texas Parks and Wildlife Department",
                           institutioncode_simple == "boemre" ~ "Bureau of Ocean Energy Management, Regulation, and Enforcement",
                           institutioncode_simple == "emap_nca" ~ "U.S. Environmental Protection Agency through its Environmental Monitoring and Assessment Program (EMAP)",
                           institutioncode_simple == "usgs" ~ "U.S. Geological Services",
@@ -52,14 +52,19 @@ calc_institutions <- function(idig_records, obis_records) {
         dplyr::select(Institution,
                       "Number of records" = n)
 
-
-
-
     list_tbl <- list(
                res_idig,
                res_obis
     )
     attr(list_tbl, "subheadings") <- c("A. iDigBio", "B. OBIS")
+    attr(list_tbl, "message") <- strwrap(glue::collapse(
+        c("NOAA records include data from the Hollings Marine Laboratory, ",
+          "the Northeast Fisheries Science Center, ",
+          "the Southwest Fisheries Science Center, ",
+          "the National Centers for Coastal Ocean Science, ",
+          "the Center for Coastal Environmental Health and Biomolecular Research, ",
+          "the National Marine Fisheries Service, and ",
+          "the Deep Sea Coral Research and Technology Program.")), width = 100)
     x_list_tbl <- xtable::xtableList(list_tbl, align = c("lp{12cm}R{3cm}"),
           caption = "Top 10 institutions that contribute to the records used in this study to iDigBio (A) and OBIS (B).")
     x_list_tbl

### 2017-11-02 manuscript.Rmd
improve table legend, make subheadings bold

--- 2017-11-01 manuscript.Rmd
@@
 ```{r records-table, results='asis'}
+xtbl_small <- function(x) {
+    glue::glue("{{\\footnotesize {x} }}")
+}
+xtbl_bold <- function(x) {
+    glue::glue("{{\\bfseries {x} }}")
+}
 print.xtableList(institution_contributions, colnames.format = "multiple",
                  booktabs = TRUE, include.rownames = FALSE,
+                 sanitize.subheadings.function = xtbl_bold, 
+                 sanitize.message.function = xtbl_small, 
                  table.placement = "!ht")

### 2017-11-02 R/institutions.R

@@ -65,7 +65,18 @@ calc_institutions <- function(idig_records, obis_records) {
           "the Center for Coastal Environmental Health and Biomolecular Research, ",
           "the National Marine Fisheries Service, and ",
           "the Deep Sea Coral Research and Technology Program.")), width = 100)
-    x_list_tbl <- xtable::xtableList(list_tbl, align = c("lp{12cm}R{3cm}"),
-          caption = "Top 10 institutions that contribute to the records used in this study to iDigBio (A) and OBIS (B).")
+    capt <- as.character(
+        glue::glue(
+                  "Top 10 institutions that contribute to the ",
+                  "records used in this study to iDigBio (A) and OBIS (B). ",
+                  "Federal and States agencies are important contributors to ",
+                  "marine biodiversity data. Differences in numbers for the ",
+                  "same data sources across databases are due to data filtering ",
+                  "operated by OBIS.")
+    )
+    x_list_tbl <- xtable::xtableList(
+                              list_tbl, align = c("lp{12cm}R{3cm}"),
+                              caption = capt
+                              )
     x_list_tbl
 }

### 2017-11-06 R/institutions.R
add footnote in table

@@ -33,7 +33,7 @@ calc_institutions <- function(idig_records, obis_records) {
         ## regroup all noaa records
         dplyr::top_n(10, n) %>%
         dplyr::mutate(Institution = case_when(
-                          institutioncode_simple == "noaa" ~ "National Oceanographic and Atmostpheric Association",
+                          institutioncode_simple == "noaa" ~ "National Oceanographic and Atmostpheric Association¹",
                           institutioncode_simple == "usnm" ~  "Smithsonian Institution National Museum of Natural History",
                           institutioncode_simple == "tpwd" ~ "Texas Parks and Wildlife Department",
                           institutioncode_simple == "boemre" ~ "Bureau of Ocean Energy Management, Regulation, and Enforcement",
@@ -58,7 +58,7 @@ calc_institutions <- function(idig_records, obis_records) {
     )
     attr(list_tbl, "subheadings") <- c("A. iDigBio", "B. OBIS")
     attr(list_tbl, "message") <- strwrap(glue::collapse(
-        c("NOAA records include data from the Hollings Marine Laboratory, ",
+        c("¹NOAA records include data from the Hollings Marine Laboratory, ",
           "the Northeast Fisheries Science Center, ",
           "the Southwest Fisheries Science Center, ",
           "the National Centers for Coastal Ocean Science, ",

### 2017-11-06 manuscript.Rmd
remove xtable comment

--- 2017-11-01 manuscript.Rmd
@@
 ```{r records-table, results='asis'}
+xtbl_small <- function(x) {
+    glue::glue("{{\\footnotesize {x} }}")
+}
+xtbl_bold <- function(x) {
+    glue::glue("{{\\bfseries {x} }}")
+}
 print.xtableList(institution_contributions, colnames.format = "multiple",
                  booktabs = TRUE, include.rownames = FALSE,
-                 table.placement = "!ht")
+                 sanitize.subheadings.function = xtbl_bold, 
+                 sanitize.message.function = xtbl_small, 
+                 table.placement = "!ht", comment = FALSE)
 ```
@@
 We compared the total number of species represented in iDigBio and OBIS for each kingdom to the total described diversity based on the WoRMS estimates[@worms-stats].
-Records in the databases are biased towards vertebrates: while vertebrates represent  `r total_vert_prop`% of the global diversity, they represent `r idigbio_vert_prop`% and `r obis_vert_prop`% of the species in iDigBio and OBIS respectively (Fig.
+While marine vertebrates represent  `r total_vert_prop`% of the global diversity, they represent `r idigbio_vert_prop`% and `r obis_vert_prop`% of the species in iDigBio and OBIS respectively (Fig.
 \@ref(fig:kingdom-diversity-plot)).
@@
 
+<!-- find checklists for fish, birds, mammals to compare with -->
+
 ```{r kingdom-diversity-plot, fig.cap='Comparison of the proportions of the diversity represented in iDigBio and OBIS  with the global diversity (as reported by WoRMS) for each kingdom.
 Vertebrates are overrepresented, while other animal groups are underpresented.
-Total number of species are indicated next to the bars.(\\#fig:kingdom-diversity-plot)'}
+Total number of species for each category are indicated next to the bars.(\\#fig:kingdom-diversity-plot)'}
 knitr::include_graphics("figures/plot_kingdom_diversity.pdf")

### 2017-11-06 manuscript.Rmd
[fix #72] display geographic extent of searches on maps

@@
 
-```{r map-diversity-sampling, fig.cap='Spatial distribution of the number of samples (A) and species richness (B) for marine invertebrates based on records retrieved from iDigBio and OBIS within the limits of the Exclusive Economic Zone for the continental USA.
+```{r map-diversity-sampling, fig.cap='Spatial distribution of the number of samples (A) and species richness (B) for marine invertebrates based on records retrieved from iDigBio and OBIS within the limits of the Exclusive Economic Zone for the continental USA (represented in light gray).
 (\\#fig:map-diversity-sampling)'}
@@
 
-```{r gom-maps, fig.cap="Spatial distribution of the number of samples (A) and species richness (B) for marine invertebrates based on records retrieved from iDigBio and OBIS for the Gulf of Mexico."}
+```{r gom-maps, fig.cap="Spatial distribution of the number of samples (A) and species richness (B) for marine invertebrates based on records retrieved from iDigBio and OBIS for the Gulf of Mexico.
+Geographic extent represented in light gray."}
 knitr::include_graphics("figures/combined_gom_sampling.pdf")
@@
 
-```{r pnw-maps, fig.cap="Spatial distribution of the number of samples (A) and species richness (B) for marine invertebrates based on records retrieved from iDigBio and OBIS for the Pacific Northwest."}
+```{r pnw-maps, fig.cap="Spatial distribution of the number of samples (A) and species richness (B) for marine invertebrates based on records retrieved from iDigBio and OBIS for the Pacific Northwest.
+Geographic extent represented in light gray."}
 knitr::include_graphics("figures/combined_pnw_sampling.pdf")
@@
 
-```{r idigbio-usa-maps, fig.cap="Spatial distribution of the number of samples (A), and species richness (B) for marine invertebrates based on records retrieved from iDigBio within the limits of the Exclusive Economic Zone for the contiental USA."}
+```{r idigbio-usa-maps, fig.cap="Spatial distribution of the number of samples (A), and species richness (B) for marine invertebrates based on records retrieved from iDigBio within the limits of the Exclusive Economic Zone for the contiental USA.
+Geographic extent represented in light gray."}
 knitr::include_graphics("figures/idigbio_map_sampling.pdf")
@@
 
-```{r obis-usa-maps, fig.cap="Spatial distribution of the number of samples (A), and species richness (B) for marine invertebrates based on records retrieved from OBIS within the limits of the Exclusive Economic Zone for the contiental USA."}
+```{r obis-usa-maps, fig.cap="Spatial distribution of the number of samples (A), and species richness (B) for marine invertebrates based on records retrieved from OBIS within the limits of the Exclusive Economic Zone for the contiental USA.
+Geographic extent represented in light gray."}
 knitr::include_graphics("figures/obis_map_sampling.pdf")

### 2017-11-06 R/map_sampling.R

@@ -142,7 +142,10 @@ make_heatmap <- function(gg_r, title, base_map) {
                                    max(gg_r$value, na.rm = TRUE),
                                    by = 1), .02))
 
+    geo_limits <- fortify(geojsonio::geojson_sp(base_map), id = "id")
+
     ggplot() +
+        geom_map(map = geo_limits, data = geo_limits, aes(map_id = id), fill = "gray80") +
         geom_raster(data = gg_r, aes(x = x, y = y, fill = value), na.rm = TRUE) +
         scale_fill_gradient2(low = "#5E98AE", mid = "#E3C94A", high = "#D5331E",
                              midpoint = mid_point,
@@ -159,10 +162,8 @@ make_heatmap <- function(gg_r, title, base_map) {
         theme(legend.title = element_blank()) +
         ggtitle(title) +
         xlab("Longitude") + ylab("Latitude")
+
 }
-
-
-
 
 
 make_heatmap_by_phylum <- function(recs, file, raster, base_map) {

### 2017-11-06 R/worms.R
add classification on unique species instead of full table

@@ -174,10 +174,10 @@ add_worms <- function(sp_list, remove_vertebrates = TRUE) {
         dplyr::filter(nchar(cleaned_scientificname) > 3) %>%
         dplyr::mutate(worms = purrr::pmap(., get_worms)) %>%
         tidyr::unnest(worms) %>%
-        dplyr::mutate(worms_id = as.character(worms_id))
-
-    res <- dplyr::left_join(sp_list, spp, by = "cleaned_scientificname") %>%
+        dplyr::mutate(worms_id = as.character(worms_id)) %>%
         add_classification()
+
+    res <- dplyr::left_join(sp_list, spp, by = "cleaned_scientificname")
 
     if (remove_vertebrates) {
         res <- res %>%

### 2017-11-06 manuscript.Rmd
start on including itis stats

@@
 Marine invertebrates are underrepresented in iDigBio and OBIS given their diversity.
-We compared the total number of species represented in iDigBio and OBIS for each kingdom to the total described diversity based on the WoRMS estimates[@worms-stats].
+We compared the total number of species represented in iDigBio and OBIS for each major division of life to the total described diversity based on the WoRMS estimates[@worms-stats].
 While marine vertebrates represent  `r total_vert_prop`% of the global diversity, they represent `r idigbio_vert_prop`% and `r obis_vert_prop`% of the species in iDigBio and OBIS respectively (Fig.
 \@ref(fig:kingdom-diversity-plot)).
-Additionally, the median number of specimens in iDigBio per species ranges between `r kng_medians %>% filter(sub_kingdom=="animalia") %>% pull(rg)` for invertebrate phyla, but is `r kng_medians %>% filter(sub_kingdom=="animalia - vertebrates") %>% pull(rg)` for vertebrates.
+This underrepresentation is also apparent in the collecting effort: the median number of specimens in iDigBio per species ranges between `r kng_medians %>% filter(sub_kingdom=="animalia") %>% pull(rg)` for invertebrate phyla, but is `r kng_medians %>% filter(sub_kingdom=="animalia - vertebrates") %>% pull(rg)` for vertebrates.
+
+```{r itis-stats}
+prop_molluscs <- itis_molluscs_in_dbs %>%
+    dplyr::summarize(
+               prop_idigbio_us = format_output(mean(n_idigbio_geo_us > 0,
+                                                    na.rm = TRUE)),
+               prop_obis_us = format_output(mean(n_obis_geo_us > 0,
+                                                 na.rm = TRUE))
+           )
+
+```
+
+To assess the fraction of digitization of the fauna, we compared the species included in iDigBio and OBIS with the comprehensive checklists for molluscs [@turgean1998] and crustaceans[@mclaughlin2005] that are available for the United States.
+For molluscs, of the `r nrow(itis_molluscs_in_dbs)` species included in the checklist,
 

### 2017-11-13 R/time_plot.R
[fix #93] use rolling mean instead of b-splines

@@ -276,11 +276,12 @@ plot_identification_level_through_time <- function(idig_records) {
                       worms_phylum %in% c("annelida", "arthropoda", "porifera",
                                           "cnidaria", "echinodermata", "mollusca")) %>%
         dplyr::ungroup() %>%
-        dplyr::mutate(worms_phylum = capitalize(worms_phylum)) %>%
-        ggplot(aes(x = year, y = p, colour = worms_phylum, fill = worms_phylum)) +
-        geom_point(aes(size = n_lots)) +
+        dplyr::mutate(worms_phylum = capitalize(worms_phylum),
+                      mean_p_id = roll_mean(p, 3, fill = NA)) %>%
+        ggplot(aes(colour = worms_phylum, fill = worms_phylum)) +
+        geom_point(aes(x = year, y = p, size = n_lots)) +
         geom_hline(yintercept = 1) +
-        geom_smooth(method = "lm", formula = y ~ splines::bs(x, degree = 3), show.legend = FALSE) +
+        geom_smooth(aes(x = year, y = mean_p_id, colour = worms_phylum), se = FALSE) +
         facet_wrap(~ worms_phylum) +
         guides(color = FALSE, fill = FALSE) +
         scale_size_continuous(name = "Number of specimens") +

### 2017-11-15 manuscript.Rmd
[fix #93] use rolling mean instead of b-splines
[fix #85] rephrase rare phyla numbers

@@
 dried specimens like shells, corals, echinoderms) specimens are over-represented in museum collections, compared with small less accessible organisms, such as the meiofauna.
-Of `r n_idigbio_phyla` phyla represented in iDigBio, `r records_rare_phyla$n_rare_phyla` have <500 records, and make up `r records_rare_phyla$prop_recs_rare_phyla`% of the iDigBio records.
-This under-representation is also manifested at the species level, while `r records_rare_phyla$prop_spp_rare_phyla_collected`% (n = `r format(records_rare_phyla$total_spp_rare_phyla_collected, big.mark=",")` total species) of the global number of species for these `r records_rare_phyla$n_rare_phyla` phyla occur in iDigBio, this proportion rises to `r records_rare_phyla$prop_spp_common_phyla_collected`% (n = `r format(records_rare_phyla$total_spp_common_phyla_collected, big.mark=",")`) for the `r n_idigbio_phyla - records_rare_phyla$n_rare_phyla` others.
+Of `r n_idigbio_phyla` phyla represented in iDigBio, `r records_rare_phyla$n_rare_phyla` have <500 records.
+While these phyla represent
+`r records_rare_phyla$prop_spp_rare_phyla_collected`% (n = `r format(records_rare_phyla$total_spp_rare_phyla_collected, big.mark=",")` total species) of the global number of species, they make up `r records_rare_phyla$prop_recs_rare_phyla`% of the iDigBio records.
+This proportion rises to `r records_rare_phyla$prop_spp_common_phyla_collected`% (n = `r format(records_rare_phyla$total_spp_common_phyla_collected, big.mark=",")`) for the `r n_idigbio_phyla - records_rare_phyla$n_rare_phyla` other phyla.
 Yet, these undersampled taxa, especially those of small body size, are often the most abundant, and are likely to play an important role in the detection of changes in environmental quality[@Bik2012].
@@
 ```{r id_though_time, fig.cap='Proportion of marine invertebrate specimens collected through time identified at the species level.
-Trend lines are  B-splines of the 3rd degree polynomial.
+Trend lines are  smoothed rolling mean on a 3 year window.
 For all groups, we observe a decline in the proportion of specimens collected that are identified at the species level, suggesting a lack of expertise to identity precisely what is collected.

### 2017-12-11 manuscript.Rmd
[fix #88] add mention of low level of digitization on West coast
* for CAS website reports collection has 2.5 million but < 165k on iDigBio
* 0 records for LACM
[fix #83] finish part about BLM/MMS contributions

@@
 Since then, the slowed pace of collection activitity (and therefore a reduction in the number of specimens added) has been accompanied by a slow-down in the discovery of new records: `r n_spp_comparison$n_spp_2017_2000` species have been documented for the first time since 2000 (`r format_output(n_spp_comparison$n_spp_2017_2000/18)` species.year^-1^), while `r n_spp_comparison$n_spp_1970_1960` were documented between 1960 and 1970 (`r format_output(n_spp_comparison$n_spp_1970_1960/11)` species.year^-1^).
+
+```{r idigbio-mms}
+prop_mms_75_85 <- (idigbio_mms_summary_by_year %>%
+                   dplyr::filter(year >= 1975 & year <= 1985) %>%
+                   dplyr::pull(n) %>%
+                   sum()) / dplyr::pull(idigbio_mms_summary_by_year, n)
+```
+
+Biodiversity surveys conducted under the auspices of the Bureau of Land Management (BLM) and the Minerals Management Service (MMS) have contributed significantly to our knowledge of the geographic distribution of marine invertebrates in US waters.
+In our dataset, we identified at least `r format(sum(idigbio_mms_summary$n), big.mark=",")` records (or `r sum(idigbio_mms_summary_by_year$n)/nrow(idigbio_us_inverts_records_all)*100`% of all US marine invertebrates in iDigBio) from these surveys, `r format_output(prop_mms_75_85 * 100)`% of which were collected between 1975 and 1985.
+Of the `r idigbio_mms_records %>% filter(rank %in% c("Species", "Subspecies")) %>% summarize(n = n_distinct(worms_valid_name))` species collected as part of these surveys, `r nrow(idigbio_mms_species_only)` have not been collected outside of these surveys.
+
 The strong correlations between (1) the number of samples collected and the species richness (Fig.
@@
 For instance, one can expect that the digitization coverage for the Gulf of Mexico will greatly increase once records from the Florida Wildlife Research Institute collections (St Petersburg, FL) become integrated into biodiversity information systems.
+Similarly, only a small fraction of the marine invertebrates records from instiutions on the Pacific have been digitized.
 

### 2017-12-11 R/idigbio_by_geo.R

@@ -376,10 +376,16 @@ get_idigbio_mms_records <- function(list_phyla) {
 
 }
 
-summary_idigbio_mms <- function(idig_mms) {
+summary_idigbio_mms_by_phylum <- function(idig_mms) {
     idig_mms %>%
-        dplyr::count(phylum)
+        dplyr::count(phylum, sort = TRUE)
 }
+
+summary_idigbio_mms_by_year <- function(idig_mms) {
+    idig_mms %>%
+        dplyr::count(year)
+}
+
 
 mms_samples_through_time <- function(idig_mms) {
     idigbio_mms_records %>%
@@ -401,7 +407,8 @@ list_species_only_collected_by_mms <- function(idig_mms, idig_recs) {
         dplyr::filter(rank == "Species" |
                       rank == "Subspecies") %>%
         dplyr::distinct(worms_phylum, worms_valid_name)
-    ## extract species for iDigBio list that doesn't contain blm/mms records
+
+    ## extract species for iDigBio list that don't contain blm/mms records
     idig_no_mms_spp <- dplyr::distinct(idig_res_no_mms, worms_phylum, worms_valid_name)
 
     ## remove species from

### 2017-12-12 manuscript.Rmd
[fix #74] include proportion of samples collected on shelf

@@
     count_spp
+
+prop_idigbio_continental_shelf <-
+   mean(idigbio_us_inverts_records_all_with_depth$marmap_depth < 0 &
+        idigbio_us_inverts_records_all_with_depth$marmap_depth > -200) * 100 %>%
+   format_output()
+
+prop_obis_continental_shelf <-
+    mean(obis_us_inverts_records_all_with_depth$marmap_depth < 0 &
+         obis_us_inverts_records_all_with_depth$marmap_depth > -200) * 100 %>%
+    format_output()
 ```
@@
 Most samples have been collected on the continental shelf (Fig.
-\@ref(fig:map-diversity-sampling)A).
+\@ref(fig:map-diversity-sampling)A), with `r prop_idigbio_continental_shelf`% (iDigBio) and `r prop_obis_continental shelf`% (OBIS) of the records collected between 0-200m.
 More records have been collected from the Atlantic Ocean, than the Gulf of Mexico, and the Pacific Ocean (`r format(sum(combined_records$is_east_coast), big.mark=",")`, `r format(sum(combined_records$is_in_gom), big.mark=",")`, and `r format(sum(combined_records$is_west_coast), big.mark=",")` records respectively).

### 2017-12-13 manuscript.Rmd
[fix #71] finish itis

@@
 
-## Results
+## Results and Discussion
 
@@
 
-We retrieved `r format(n_distinct(combined_species$worms_valid_name), big.mark = ",")` species (`r n_distinct(na.omit(combined_species$worms_phylum))` phyla) of marine invertebrates in the continental US EEZ by combining  `r format(nrow(idigbio_records), big.mark=",")` museum records from iDigBio (`r format(n_distinct(idigbio_records$worms_valid_name), big.mark = ",")` species, `r n_distinct(na.omit(idigbio_records$worms_phylum))` phyla), and `r format(nrow(obis_records), big.mark=",")` records from OBIS (`r format(n_distinct(obis_records$worms_valid_name), big.mark=",")` species,
-`r n_distinct(na.omit(obis_records$worms_phylum))` phyla).
-Extant marine invertebrates occurring in the US represent less than `r format_output(nrow(idigbio_records)/idigbio_stats$total_no_fossil*100)`% of the total number of non-fossil iDigBio records (`r round(idigbio_stats$total_no_fossil/1e6, 1)` million), and `r format_output(nrow(idigbio_records)/idigbio_stats$usa_no_fossil*100)`% of the US records (`r round(idigbio_stats$usa_no_fossil/1e6, 1)` million).
-Marine invertebrates that occur in the US represent 4.1% of the 48 million occurrence records in OBIS.
+We retrieved `r format(n_distinct(combined_species$worms_valid_name), big.mark = ",")` species (`r n_distinct(na.omit(combined_species$worms_phylum))` phyla) of marine invertebrates in the continental US EEZ by combining  `r format(nrow(idigbio_us_inverts_records_species), big.mark=",")` museum records from iDigBio (`r format(n_distinct(idigbio_us_inverts_records_species$worms_valid_name), big.mark = ",")` species, `r n_distinct(na.omit(idigbio_us_inverts_records_species$worms_phylum))` phyla), and `r format(nrow(obis_us_inverts_records_species), big.mark=",")` records from OBIS (`r format(n_distinct(obis_us_inverts_records_species$worms_valid_name), big.mark=",")` species,
+`r n_distinct(na.omit(obis_us_inverts_records_species$worms_phylum))` phyla).
 
@@
 
-Marine invertebrates are underrepresented in iDigBio and OBIS given their diversity.
+Marine invertebrates are underrepresented in iDigBio and OBIS.
+This underrepresentation is manifest in the small fraction of records that marine invertebrates represent.
+Extant marine invertebrates occurring in the US EEZ represent  `r format_output(nrow(idigbio_us_inverts_records_species)/idigbio_stats$total_no_fossil*100)`% of the total number of non-fossil iDigBio records (`r round(idigbio_stats$total_no_fossil/1e6, 1)` million), and `r format_output(nrow(idigbio_us_inverts_records_species)/idigbio_stats$usa_no_fossil*100)`% of the US records (`r round(idigbio_stats$usa_no_fossil/1e6, 1)` million).
+Marine invertebrates however represent between 11% and 18% of the described global diversity (number of described marine inverterbrates species: 200,000-210,000 [@Appeltans2012], total number of described species: 1.2-1.9 million [@Mora2011,@Chapman2009]).
+
 We compared the total number of species represented in iDigBio and OBIS for each major division of life to the total described diversity based on the WoRMS estimates[@worms-stats].
-While marine vertebrates represent  `r total_vert_prop`% of the global diversity, they represent `r idigbio_vert_prop`% and `r obis_vert_prop`% of the species in iDigBio and OBIS respectively (Fig.
+While marine vertebrates represent  `r total_vert_prop`% of the global marine diversity, they represent `r idigbio_vert_prop`% and `r obis_vert_prop`% of the species in iDigBio and OBIS respectively (Fig.
 \@ref(fig:kingdom-diversity-plot)).
-This underrepresentation is also apparent in the collecting effort: the median number of specimens in iDigBio per species ranges between `r kng_medians %>% filter(sub_kingdom=="animalia") %>% pull(rg)` for invertebrate phyla, but is `r kng_medians %>% filter(sub_kingdom=="animalia - vertebrates") %>% pull(rg)` for vertebrates.
+This underrepresentation is also apparent in the collecting effort: the median number of specimens in iDigBio per species ranges between `r kng_medians %>% filter(sub_kingdom=="animalia") %>% pull(rg)` for marine invertebrate phyla, but is `r kng_medians %>% filter(sub_kingdom=="animalia - vertebrates") %>% pull(rg)` for marine vertebrates.
 
@@
                prop_idigbio_us = format_output(mean(n_idigbio_geo_us > 0,
-                                                    na.rm = TRUE)),
+                                                    na.rm = TRUE)* 100),
                prop_obis_us = format_output(mean(n_obis_geo_us > 0,
-                                                 na.rm = TRUE))
+                                                 na.rm = TRUE) * 100)
            )
 
+prop_crustaceans <- itis_crustaceans_in_dbs %>%
+    dplyr::summarize(
+             prop_idigbio_us = format_output(mean(n_idigbio_geo_us > 0,
+                                                  na.rm = TRUE) * 100),
+             prop_obis_us = format_output(mean(n_obis_geo_us > 0,
+                                             na.rm = TRUE) * 100)               
+           )
 ```
 
-To assess the fraction of digitization of the fauna, we compared the species included in iDigBio and OBIS with the comprehensive checklists for molluscs [@turgean1998] and crustaceans[@mclaughlin2005] that are available for the United States.
-For molluscs, of the `r nrow(itis_molluscs_in_dbs)` species included in the checklist,
-
-<!-- find checklists for fish, birds, mammals to compare with -->
+To assess the fraction of digitization of the fauna, we compared the species included in iDigBio and OBIS with the comprehensive checklists for molluscs [@turgeon1998] and crustaceans [@mclaughlin2005] that are available for the United States.
+For molluscs, of the `r format(nrow(itis_molluscs_in_dbs), big.mark=",")` species included in the checklist, `r prop_molluscs$prop_idigbio_us`% are in iDigBio and `r prop_molluscs$prop_obis_us`% are in OBIS.
+For crustaceans, of the `r format(nrow(itis_crustaceans_in_dbs), big.mark=",")` species included in the checklist, `r prop_crustaceans$prop_idigbio_us`% are in iDigBio and `r prop_crustaceans$prop_obis_us`% are in OBIS.
 

### 2017-12-15 manuscript.Rmd
refactor with using postgis queries everywhere

@@
 This underrepresentation is manifest in the small fraction of records that marine invertebrates represent.
-Extant marine invertebrates occurring in the US EEZ represent  `r format_output(nrow(idigbio_us_inverts_records_species)/idigbio_stats$total_no_fossil*100)`% of the total number of non-fossil iDigBio records (`r round(idigbio_stats$total_no_fossil/1e6, 1)` million), and `r format_output(nrow(idigbio_us_inverts_records_species)/idigbio_stats$usa_no_fossil*100)`% of the US records (`r round(idigbio_stats$usa_no_fossil/1e6, 1)` million).
+Extant marine invertebrates occurring in the US EEZ represent  `r format_output(nrow(idigbio_us_inverts_records_species)/idigbio_stats$total_no_fossil*100)`% of the total number of non-fossil iDigBio records (`r round(idigbio_stats$total_no_fossil/1e6, 1)` million), and `r format_output(nrow(idigbio_us_inverts_records_species)/idigbio_stats$eez_no_fossil*100)`% of the US records (`r round(idigbio_stats$eez_no_fossil/1e6, 1)` million).
 Marine invertebrates however represent between 11% and 18% of the described global diversity (number of described marine inverterbrates species: 200,000-210,000 [@Appeltans2012], total number of described species: 1.2-1.9 million [@Mora2011,@Chapman2009]).
@@
 n_spp_gom <- combined_records %>%
-    dplyr::filter(is_in_gom == TRUE) %>%
+    dplyr::filter(within_gom == TRUE) %>%
     count_spp
@@
 
-prop_idigbio_continental_shelf <-
+prop_idigbio_continental_shelf <- (
    mean(idigbio_us_inverts_records_all_with_depth$marmap_depth < 0 &
-        idigbio_us_inverts_records_all_with_depth$marmap_depth > -200) * 100 %>%
+        idigbio_us_inverts_records_all_with_depth$marmap_depth > -200) * 100) %>%
    format_output()
 
-prop_obis_continental_shelf <-
+prop_obis_continental_shelf <- (
     mean(obis_us_inverts_records_all_with_depth$marmap_depth < 0 &
-         obis_us_inverts_records_all_with_depth$marmap_depth > -200) * 100 %>%
+         obis_us_inverts_records_all_with_depth$marmap_depth > -200) * 100) %>%
     format_output()
@@
 Most samples have been collected on the continental shelf (Fig.
-\@ref(fig:map-diversity-sampling)A), with `r prop_idigbio_continental_shelf`% (iDigBio) and `r prop_obis_continental shelf`% (OBIS) of the records collected between 0-200m.
-More records have been collected from the Atlantic Ocean, than the Gulf of Mexico, and the Pacific Ocean (`r format(sum(combined_records$is_east_coast), big.mark=",")`, `r format(sum(combined_records$is_in_gom), big.mark=",")`, and `r format(sum(combined_records$is_west_coast), big.mark=",")` records respectively).
+\@ref(fig:map-diversity-sampling)A), with `r prop_idigbio_continental_shelf`% (iDigBio) and `r prop_obis_continental_shelf`% (OBIS) of the records collected between 0-200m.
+More records have been collected from the Atlantic Ocean, than the Gulf of Mexico, and the Pacific Ocean (`r format(sum(combined_records$is_east_coast), big.mark=",")`, `r format(sum(combined_records$is_within_gom), big.mark=",")`, and `r format(sum(combined_records$is_west_coast), big.mark=",")` records respectively).
 The highest species richness has been recorded from the Gulf of Mexico (`r format(n_spp_gom, big.mark=",")` species), followed by the Atlantic Ocean (`r format(n_spp_east, big.mark=",")` species), and the Pacific Ocean (`r format(n_spp_west, big.mark=",")` species).
@@
 Biodiversity surveys conducted under the auspices of the Bureau of Land Management (BLM) and the Minerals Management Service (MMS) have contributed significantly to our knowledge of the geographic distribution of marine invertebrates in US waters.
-In our dataset, we identified at least `r format(sum(idigbio_mms_summary$n), big.mark=",")` records (or `r sum(idigbio_mms_summary_by_year$n)/nrow(idigbio_us_inverts_records_all)*100`% of all US marine invertebrates in iDigBio) from these surveys, `r format_output(prop_mms_75_85 * 100)`% of which were collected between 1975 and 1985.
+In our dataset, we identified at least `r format(sum(idigbio_mms_summary_by_year$n), big.mark=",")` records (or `r sum(idigbio_mms_summary_by_year$n)/nrow(idigbio_us_inverts_records_all)*100`% of all US marine invertebrates in iDigBio) from these surveys, `r format_output(prop_mms_75_85 * 100)`% of which were collected between 1975 and 1985.
 Of the `r idigbio_mms_records %>% filter(rank %in% c("Species", "Subspecies")) %>% summarize(n = n_distinct(worms_valid_name))` species collected as part of these surveys, `r nrow(idigbio_mms_species_only)` have not been collected outside of these surveys.
@@
 
-We extracted `r format(sum(gom_worms$is_marine, na.rm=TRUE), big.mark=",")` and `r format(sum(kozloff_worms$is_marine, na.rm=TRUE), big.mark=",")` marine invertebrate species from species checklists from the Gulf of Mexico [@Tunnell2009] and from the Pacific Northwest [@Kozloff1996] respectively.
+We extracted `r format(nrow(gom_worms_species), big.mark=",")` and `r format(nrow(kozloff_worms_species), big.mark=",")` marine invertebrate species from species checklists from the Gulf of Mexico [@Tunnell2009] and from the Pacific Northwest [@Kozloff1996] respectively.
 For the Gulf of Mexico, biodiversity databases included more species for the same geographic area as the one used by the authors of the checklist (iDigBio and OBIS combined: `r format(dplyr::n_distinct(combined_gom_records$worms_valid_name), big.mark=",")` species; iDigBio only: `r format(n_distinct(idigbio_gom_records$worms_valid_name), big.mark=",")` species, $n =$  `r format(nrow(idigbio_gom_records), big.mark=",")` records; OBIS only: `r format(n_distinct(obis_gom_records$worms_valid_name), big.mark=",")` species, $n=$ `r format(nrow(obis_gom_records), big.mark=",")` records).
@@
 Missing geographic coordinates reduces the overlap between the list of species from the regional list and the databases.
-Using iDigBio, we found that for the `r format(not_really_in_database$idig_gom_n_spp_not_in_db, big.mark=",")` species that do not appear within the geographic range of the Gulf of Mexico but are on the list, `r format_output(not_really_in_database$idig_gom_p_spp_not_in_db*100)`% are completely absent from iDigBio, while for the `r format(not_really_in_database$idig_gom_n_spp_no_coords, big.mark=",")` species that do appear in iDigBio, `r format_output(not_really_in_database$idig_gom_p_spp_no_coords*100)`%  do not have geographic coordinates associated with them, and for those that do have geographic coordinates, `r format_output(not_really_in_database$idig_gom_p_spp_within_usa*100)`% ($n=$  `r format(not_really_in_database$idig_gom_n_spp_within_usa, big.mark=",")`) fall within US waters.
+Using iDigBio, we found that for the `r format(not_really_in_database$idig_gom_n_spp_not_in_db, big.mark=",")` species that do not appear within the geographic range of the Gulf of Mexico but are on the list, `r format_output(not_really_in_database$idig_gom_p_spp_not_in_db*100)`% are completely absent from iDigBio, while for the `r format(not_really_in_database$idig_gom_n_spp_no_coords, big.mark=",")` species that do appear in iDigBio, `r format_output(not_really_in_database$idig_gom_p_spp_no_coords*100)`%  do not have geographic coordinates associated with them, and for those that do have geographic coordinates, `r format_output(not_really_in_database$idig_gom_p_spp_within_eez*100)`% ($n=$  `r format(not_really_in_database$idig_gom_n_spp_within_eez, big.mark=",")`) fall within US waters.
 Additional digitization of collections, including the capture of their geographic coordinates information, is needed to improve biodiversity knowledge at regional scales.
 For instance, one can expect that the digitization coverage for the Gulf of Mexico will greatly increase once records from the Florida Wildlife Research Institute collections (St Petersburg, FL) become integrated into biodiversity information systems.
-Similarly, only a small fraction of the marine invertebrates records from instiutions on the Pacific have been digitized.
+Similarly, only a small fraction of the marine invertebrates records from instiutions on the Pacific has been digitized.
 
@@
 This study is fully reproducible and the code needed to perform the analyses included here is available from: https://github.com/fmichonneau/sok-marine-biodiversity, and archived on Zenodo (doi:xxx.xxx.xxxx).
-The data downloaded from the databases at the time of analysis are archived on Zenodo (doi:xxx.xxx.xxx)/
+The data downloaded from the databases at the time of analysis are archived on Zenodo (doi:xxx.xxx.xxx).
 

### 2017-12-15 R/bold.R

@@ -37,7 +37,7 @@ store_bold_specimens_per_species <- function(store_path = "data/storr_bold_speci
 ## this function takes a data frame, for which the column named with
 ## the content of `col_nm` contains the species names to look up in
 ## BOLD. It returns a data frame.
-find_bold_records <- function(res, col_nm, map_usa, show_progress = TRUE) {
+find_bold_records <- function(res, col_nm, show_progress = TRUE) {
     if (!exists(col_nm, res))
         stop("invalid column name")
 
@@ -63,7 +63,7 @@ find_bold_records <- function(res, col_nm, map_usa, show_progress = TRUE) {
                               !is.na(decimallongitude))
             if (nrow(coord_data) > 0) {
                 any_within_eez <- any(is_within_eez_records(
-                    coord_data, map_usa)$is_in_eez)
+                    coord_data)$within_eez)
             } else any_within_eez <- NA
         } else {
             any_within_eez <- NA

### 2017-12-15 R/combine_lists.R

@@ -32,12 +32,12 @@ add_geo <- function(d) {
                                              TRUE, FALSE))
 
     ## check that no record gets assigned more than one geographic area
-    if (any(d$is_in_gom + d$is_east_coast + d$is_west_coast) > 3)
+    if (any(d$within_gom + d$is_east_coast + d$is_west_coast) > 3)
         stop("something's wrong!")
     d
 }
 
-combine_species_list <- function(..., map_usa) {
+combine_species_list <- function(...) {
     d <- list(...)
     if (is.null(names(d)) && !all(nzchar(names(names(d))))) {
         stop("arguments need to be named")
@@ -55,7 +55,7 @@ combine_species_list <- function(..., map_usa) {
         dplyr::count(data_source, worms_phylum, worms_class, worms_order,
                      worms_family, worms_valid_name, worms_id) %>%
         tidyr::spread(data_source, n) %>%
-        find_bold_records(col_nm = "worms_valid_name", map_usa = map_usa)
+        find_bold_records(col_nm = "worms_valid_name")
 }
 
 

### 2017-12-15 R/comparison_by_geo.R

@@ -21,14 +21,11 @@ compare_with_geo <- function(spp_list, geo_list, verbose = FALSE) {
 }
 
 combined_regional_by_geo <- function(gom_worms, gom_idigbio, gom_obis,
-                                     pnw_worms, pnw_idigbio, pnw_obis,
-                                     map_usa) {
+                                     pnw_worms, pnw_idigbio, pnw_obis) {
 
     list(
-        gom = combine_species_list(checklist = gom_worms, idigbio = gom_idigbio, obis = gom_obis,
-                                   map_usa = map_usa),
-        pnw = combine_species_list(checklist = pnw_worms, idigbio = pnw_idigbio, obis = pnw_obis,
-                                   map_usa = map_usa)
+        gom = combine_species_list(checklist = gom_worms, idigbio = gom_idigbio, obis = gom_obis),
+        pnw = combine_species_list(checklist = pnw_worms, idigbio = pnw_idigbio, obis = pnw_obis)
     ) %>%
         dplyr::bind_rows(.id = "region")
 }
@@ -347,14 +344,14 @@ not_in_list_collected_recently <- function(database_overlap, map_gom, map_pnw) {
         add_worms()
 
     .gom <- . %>%
-        is_within_gom_records(map_gom) %>%
-        dplyr::filter(is_in_gom == TRUE) %>%
+        is_within_gom_records() %>%
+        dplyr::filter(within_gom == TRUE) %>%
         dplyr::group_by(worms_phylum, worms_valid_name) %>%
         dplyr::summarize(min_year = min(year))
 
     .pnw <- . %>%
-        is_within_pnw_records(map_pnw) %>%
-        dplyr::filter(is_in_pnw == TRUE) %>%
+        is_within_pnw_records() %>%
+        dplyr::filter(within_pnw == TRUE) %>%
         dplyr::group_by(worms_phylum, worms_valid_name) %>%
         dplyr::summarize(min_year = min(year))
 
@@ -431,7 +428,7 @@ n_spp_not_in_lists <- function(db_overlap) {
 ## their records, or they have records that fall outside the
 ## geographic area.
 ## Only GOM for now.
-get_not_really_in_database <- function(database_overlap, map_usa) {
+get_not_really_in_database <- function(database_overlap) {
 
     idig_gom_not_in_db <- database_overlap %>%
         dplyr::filter(region == "gom",
[…]

### 2017-12-15 R/correlation_specimens_species.R

@@ -1,20 +1,20 @@
-prepare_sampling_effort_data <- function(id, map_gom) {
+prepare_sampling_effort_data <- function(id) {
     id %>%
         dplyr::rename(decimallatitude = y, decimallongitude = x) %>%
-        is_within_gom_records(map_gom) %>%
+        is_within_gom_records() %>%
         dplyr::mutate(region = case_when(
-                          decimallongitude > -100 & is_in_gom == FALSE ~ "atlantic",
-                          decimallongitude > -100 & is_in_gom == TRUE ~ "gom",
+                          decimallongitude > -100 & within_gom == FALSE ~ "atlantic",
+                          decimallongitude > -100 & within_gom == TRUE ~ "gom",
                           decimallongitude < -100 ~ "pacific"
                       )) %>%
         dplyr::rename(latitude = decimallatitude, longitude = decimallongitude)
 }
 
 
-plot_sampling_effort <- function(id, map_gom) {
+plot_sampling_effort <- function(id) {
     id %>%
         dplyr::filter(n_specimen > 0, n_species > 0) %>%
-        prepare_sampling_effort_data(map_gom) %>%
+        prepare_sampling_effort_data() %>%
         ggplot(aes(x = n_specimen, y = n_species,
                    shape = region, colour = latitude)) +
         scale_colour_viridis(option = "magma", direction = -1,
@@ -30,9 +30,9 @@ plot_sampling_effort <- function(id, map_gom) {
 
 ## proportion of cells with less than 500 observations, where the number of
 ## species is within 10% of the number of observations.
-calc_stat_sampling_effort <- function(id, map_gom) {
+calc_stat_sampling_effort <- function(id) {
     res <- id %>%
-        prepare_sampling_effort_data(map_gom) %>%
+        prepare_sampling_effort_data() %>%
         dplyr::mutate(within_90 = (n_species - (n_specimen * .9)) > 0) %>%
         dplyr::filter(n_specimen <= 500) %>%
         dplyr::pull(within_90)

### 2017-12-15 R/counting.R

@@ -7,7 +7,7 @@ get_n_records_in_db <- function(worm, db, field_name, region = c("gom", "pnw")) 
     ## region: either "gom" for Gulf of Mexico, or "pnw" for Pacific North West.
 
     region <- match.arg(region)
-    is_in_region <- paste0("is_in_", region)
+    within_region <- paste0("within_", region)
 
     db <- db %>%
         dplyr::mutate(scientificname = tolower(scientificname))
@@ -18,13 +18,13 @@ get_n_records_in_db <- function(worm, db, field_name, region = c("gom", "pnw")) 
         dplyr::rename_(.dots = setNames("n", field_name))
 
     summary_db_in_us <- db %>%
-        dplyr::filter(is_in_eez == TRUE) %>%
+        dplyr::filter(within_eez == TRUE) %>%
         dplyr::group_by(scientificname) %>%
         dplyr::tally(.) %>%
         dplyr::rename_(.dots = setNames("n", paste0(field_name, "_in_us")))
 
     summary_db_in_region <- db %>%
-        dplyr::filter(!!sym(is_in_region)) %>%
+        dplyr::filter(!!sym(within_region)) %>%
         dplyr::group_by(scientificname) %>%
         dplyr::tally(.) %>%
         dplyr::rename_(.dots = setNames("n", paste0(field_name, "_in_", region)))

### 2017-12-15 R/estimate_richness.R

@@ -139,15 +139,15 @@ if (FALSE) {
     predict(pnw_sar_nls, data.frame(area = lawn_area(map_pnw)/1e6))
 
     ## seems fairly reasonable given crudeness of approach
-    all_us <- spp_from_cells(combined_records, usa_raster)
-    usa_sar_nls <- nls(n_spp ~ a*area^z, data = all_us, start = list(a = 5, z = .2))
-    predict(usa_sar_nls, data.frame(area = lawn_area(map_usa)/1e6), se.fit = T)
+    all_us <- spp_from_cells(combined_records, eez_raster)
+    eez_sar_nls <- nls(n_spp ~ a*area^z, data = all_us, start = list(a = 5, z = .2))
+    predict(eez_sar_nls, data.frame(area = lawn_area(map_eez)/1e6), se.fit = T)
 
-    east_us <- spp_from_cells(dplyr::filter(combined_records, is_east_coast), usa_raster)
+    east_us <- spp_from_cells(dplyr::filter(combined_records, is_east_coast), eez_raster)
     east_sar_nls <- nls(n_spp ~ a * area^z,  data = east_us, start = list(a = 5, z = .2))
-    predict(east_sar_nls, data.frame(area = lawn_area(map_usa)/2/1e6))
+    predict(east_sar_nls, data.frame(area = lawn_area(map_eez)/2/1e6))
 
-    west_us <- spp_from_cells(dplyr::filter(combined_records, is_west_coast), usa_raster)
+    west_us <- spp_from_cells(dplyr::filter(combined_records, is_west_coast), eez_raster)
     west_sar_nls <- nls(n_spp ~ a * area^z,  data = west_us, start = list(a = 5, z = .2))
-    predict(west_sar_nls, data.frame(area = lawn_area(map_usa)/2/1e6))
+    predict(west_sar_nls, data.frame(area = lawn_area(map_eez)/2/1e6))
 }

### 2017-12-15 R/geo_utils.R

@@ -27,22 +27,22 @@ bb_to_df <- function(bb) {
 ## from the map of the EEZ from the USA, create a data frame that
 ## contains the coordinates for a grid that spans the area. This grid
 ## will be used to do an iDigBio search by geography
-generate_bounding_boxes <- function(map_usa, cellsize = .5) {
+generate_bounding_boxes <- function(map_eez, cellsize = .5) {
 
-    map_usa_sp_df <- geojson_sp(map_usa)
+    map_eez_sp_df <- geojson_sp(map_eez)
 
-    map_usa_fort <- ggplot2::fortify(map_usa_sp_df)
-    res <- lapply(levels(map_usa_fort$piece), function(i) {
+    map_eez_fort <- ggplot2::fortify(map_eez_sp_df)
+    res <- lapply(levels(map_eez_fort$piece), function(i) {
         SpatialPolygons(
             list(
                 Polygons(
-                    list(Polygon(subset(map_usa_fort, piece == i)[, c("long", "lat")])),
+                    list(Polygon(subset(map_eez_fort, piece == i)[, c("long", "lat")])),
                     ID = 1)
-            ), proj4string = CRS(proj4string(map_usa_sp_df)))
+            ), proj4string = CRS(proj4string(map_eez_sp_df)))
     }) %>%
         lapply(get_bounding_box, cellsize = cellsize) %>%
         lapply(bb_to_df)
-    names(res) <- levels(map_usa_fort$piece)
+    names(res) <- levels(map_eez_fort$piece)
 
     dplyr::bind_rows(res, .id = "piece")
 }

### 2017-12-15 R/idigbio.R

@@ -55,28 +55,28 @@ fetch_spp_from_idigbio <- function(wrm) {
         dplyr::distinct(uuid, .keep_all = TRUE)
 }
 
-get_idigbio_stats <- function(map_usa) {
-    bbox_usa <- lawn::lawn_bbox(map_usa)
+get_idigbio_stats <- function(map_eez) {
+    bbox_eez <- lawn::lawn_bbox(map_eez)
     idig_non_fossil <- ridigbio::idig_count_records(list("basisofrecord" = "preservedspecimen"))
-    idig_non_fossil_usa <- ridigbio::idig_count_records(
+    idig_non_fossil_eez <- ridigbio::idig_count_records(
                                          list(
                                              "basisofrecord" = "preservedspecimen",
                                              "geopoint" = list(
                                                  type = "geo_bounding_box",
                                                  top_left = list(
-                                                     lon = bbox_usa[1],
-                                                     lat = bbox_usa[4]
+                                                     lon = bbox_eez[1],
+                                                     lat = bbox_eez[4]
                                                  ),
                                                  bottom_right = list(
-                                                     lon = bbox_usa[3],
-                                                     lat = bbox_usa[2]
+                                                     lon = bbox_eez[3],
+                                                     lat = bbox_eez[2]
                                                  )
                                              )
                                          )
                                      )
     list(
         total_no_fossil = idig_non_fossil,
-        usa_no_fossil = idig_non_fossil_usa
+        eez_no_fossil = idig_non_fossil_eez
     )
 }
 

### 2017-12-15 R/idigbio_by_geo.R

@@ -90,14 +90,14 @@ get_idigbio_by_geo <- function(coords, q) {
     }
 }
 
-get_coords_idigbio_query <- function(map_usa, cellsize = .5) {
-    bb_eez <- generate_bounding_boxes(map_usa, cellsize = cellsize)
+get_coords_idigbio_query <- function(map_eez, cellsize = .5) {
+    bb_eez <- generate_bounding_boxes(map_eez, cellsize = cellsize)
     coords_to_query(bb_eez)
 }
 
 ## for all the coordinates of the bounding boxes, find the iDigBio
 ## records they contain coords: output of get_coords_idigbio_query
-## map_usa: map in GeoJSON db_table: name of table in the postgres
+## map_eez: map in GeoJSON db_table: name of table in the postgres
 ## database that will store the results use_cache: if TRUE, this uses
 ## the results from the iDigBio storr; if false, the entire storr is
 ## destroyed before

### 2017-12-15 R/itis.R

@@ -134,7 +134,7 @@ crust_name_geo <- function(tbl) {
         add_worms()
 }
 
-add_dbs_info <- function(tbl, combined_species, map) {
+add_dbs_info <- function(tbl, combined_species) {
     tbl %>%
         dplyr::left_join(dplyr::select(combined_species,
                                        worms_id, idigbio, obis, n_bold_records),
@@ -155,12 +155,12 @@ add_dbs_info <- function(tbl, combined_species, map) {
                           .idig <- .idig %>%
                               dplyr::rename(decimallatitude = geopoint.lat,
                                             decimallongitude = geopoint.lon) %>%
-                              is_within_eez_records(map = map)
+                              is_within_eez_records()
                           tibble::tibble(
                               n_idigbio_total = nrow(.idig),
                               n_idigbio_no_geo = sum(is.na(.idig$decimallatitude) |
                                                      is.na(.idig$decimallongitude)),
-                              n_idigbio_geo_us = sum(.idig$is_in_eez),
+                              n_idigbio_geo_us = sum(.idig$within_eez),
                               n_idigbio_country_us = sum(.idig$country == "united states", na.rm = TRUE)
                           )
                       }),
@@ -181,14 +181,14 @@ add_dbs_info <- function(tbl, combined_species, map) {
                           .obis <- .obis %>%
                               dplyr::rename(decimallatitude = decimalLatitude,
                                             decimallongitude = decimalLongitude) %>%
-                              is_within_eez_records(map = map)
+                              is_within_eez_records()
                           ## OBIS doesn't store country info directly, sometimes
                           ## included in locality but difficult to parse.
                           tibble::tibble(
                               n_obis_total = nrow(.obis),
                               n_obis_no_geo = sum(is.na(.obis$decimallatitude) |
                                                   is.na(.obis$decimallongitude)),
-                              n_obis_geo_us = sum(.obis$is_in_eez)
+                              n_obis_geo_us = sum(.obis$within_eez)
                           )
                       })
                ) %>%
@@ -197,7 +197,7 @@ add_dbs_info <- function(tbl, combined_species, map) {
 }
 
 
-crust_name_in_dbs <- function(tbl, combined_species, map) {
+crust_name_in_dbs <- function(tbl, combined_species) {
 
[…]

### 2017-12-15 R/postgis_queries.R

@@ -3,14 +3,15 @@ insert_map_into_db <- function(map) {
     ## This table is named like the `map` object.
     name <- deparse(substitute(map))
     sp_map <- geojsonio::geojson_sp(map)
+
+    db <- sok_db()
     rpostgis::pgInsert(db, name = c("public", name),
                        data.obj = sp_map, overwrite = TRUE,
                        row.names = FALSE)
 
-    db <- sok_db()
 
     ## Then we had the union of the layers of these objects into a master table
-    ## that holds all the polygons we use: map_usa, map_gom, map_pnw.  But
+    ## that holds all the polygons we use: map_eez, map_gom, map_pnw.  But
     ## first, we need to make sure that the `maps` table exists, and create it
     ## otherwise.
     if (!dbExistsTable(db, "maps")) {
@@ -40,7 +41,7 @@ insert_map_into_db <- function(map) {
 ## be filtered for geography
 add_unique_coords_to_db <- function(db, src_table) {
 
-    maps <- c("map_usa", "map_gom", "map_pnw")
+    maps <- c("map_eez", "map_gom", "map_pnw")
 
     ## make sure maps table exists and it contains all the data we need
     if (!dbExistsTable(db, "maps")) stop("'maps' table doesn't exist.")

### 2017-12-15 R/usa_maps.R

@@ -1,4 +1,4 @@
-get_map_usa <- function(file) {
+get_map_eez <- function(file) {
     res <- geojsonio::geojson_read(x = file, method = "local", what = "sp")
     geojsonio::geojson_json(res)
 }
@@ -18,69 +18,10 @@ get_map_pnw <- function() {
         geojsonio::geojson_json()
 }
 
-is_within_map <- function(points, map) {
-    pts <- geojsonio::geojson_json(points, geometry = "point",
-                                   lat = "lat", lon = "long")
-    res <- lawn::lawn_within(pts, map)
-    res
-}
-
-is_within_map_records <- function(area) {
-    function(filtered_data, map) {
-        if (nrow(filtered_data) < 1)
-            stop("something is wrong")
-
-        browser()
-        res_lawn <- filtered_data %>%
-            dplyr::mutate(
-                       lat = round(decimallatitude, 1),
-                       long = round(decimallongitude, 1),
-                       coord_key = paste(lat, long, sep = "|"))
-
-        simplified_coords <-  res_lawn %>%
-            dplyr::filter(!is.na(lat), !is.na(long)) %>%
-            dplyr::select(coord_key, lat, long) %>%
-            dplyr::distinct(coord_key, .keep_all = TRUE)
-
-        split_coords <- split(simplified_coords, ceiling(seq_len(nrow(simplified_coords))/1000))
-
-        res <- vector("list", length(split_coords))
-        for (i in 1:length(split_coords)) {
-            if (i == 77) browser()
-            message("i", i)
-            res[[i]] <- is_within_map(split_coords[[i]], map)
-        }
-
-        res_is_within <- is_within_map(simplified_coords, map)$features$geometry$coordinates %>%
-                           purrr::map_df(function(x)
-                                      data.frame(
-                                          coord_key = paste(x[2], x[1], sep = "|"),
-                                          lat = x[2],
-                                          lon = x[1],
-                                          stringsAsFactors = FALSE))
-
-        if (nrow(res_is_within) > 0) {
-            res_lawn[[paste0("is_in_", area)]] <- res_lawn$coord_key %in% res_is_within$coord_key
-        } else {
[…]

### 2017-12-21 R/itis.R
add "usa" as possible country match

@@ -161,7 +161,8 @@ add_dbs_info <- function(tbl, combined_species) {
                               n_idigbio_no_geo = sum(is.na(.idig$decimallatitude) |
                                                      is.na(.idig$decimallongitude)),
                               n_idigbio_geo_us = sum(.idig$within_eez),
-                              n_idigbio_country_us = sum(.idig$country == "united states", na.rm = TRUE)
+                              n_idigbio_country_us = sum(.idig$country %in% c("united states", "usa"),
+                                                         na.rm = TRUE)
                           )
                       }),
 

### 2017-12-21 manuscript.Rmd
edit part of percentage of diversity

@@
 ```{r kingdom-stats}
-get_vert_prop <- function(db) {
+get_invert_prop <- function(db) {
     format_output(kingdom_diversity[kingdom_diversity$database == db &
-                                    kingdom_diversity$sub_kingdom == "Animalia - Vertebrates", "prop_spp"] * 100)
+                                    kingdom_diversity$sub_kingdom == "Animalia - Invertebrates", "prop_spp"] * 100)
 }
 
-total_vert_prop <- get_vert_prop("worms")
-idigbio_vert_prop <- get_vert_prop("idigbio")
-obis_vert_prop <- get_vert_prop("obis")
-
-kng_medians <- kingdom_stats %>%
-    dplyr::filter(database == "idigbio") %>% 
-    dplyr::group_by(sub_kingdom) %>%
-    dplyr::summarize(
-               rg = if_else(
-                   Reduce("==", range(median_sample)),
-                   as.character(range(median_sample)[1]), 
-                   paste(range(median_sample), collapse = " and "))
-           )
-
+total_invert_prop <- get_invert_prop("worms")
+idigbio_invert_prop <- get_invert_prop("idigbio")
+obis_invert_prop <- get_invert_prop("obis")
 ```
@@
 
-Marine invertebrates are underrepresented in iDigBio and OBIS.
-This underrepresentation is manifest in the small fraction of records that marine invertebrates represent.
-Extant marine invertebrates occurring in the US EEZ represent  `r format_output(nrow(idigbio_us_inverts_records_species)/idigbio_stats$total_no_fossil*100)`% of the total number of non-fossil iDigBio records (`r round(idigbio_stats$total_no_fossil/1e6, 1)` million), and `r format_output(nrow(idigbio_us_inverts_records_species)/idigbio_stats$eez_no_fossil*100)`% of the US records (`r round(idigbio_stats$eez_no_fossil/1e6, 1)` million).
-Marine invertebrates however represent between 11% and 18% of the described global diversity (number of described marine inverterbrates species: 200,000-210,000 [@Appeltans2012], total number of described species: 1.2-1.9 million [@Mora2011,@Chapman2009]).
+```{r}
+prop_obis_invert_samples <- kingdom_diversity %>%
+    filter(database == "obis" & sub_kingdom == "Animalia - Invertebrates") %>%
+    pull(prop_samp) %>%
+    `*`(100) %>%
+    format_output()
+
+prop_obis_invert_species <- kingdom_diversity %>%
+    filter(database == "obis" & sub_kingdom == "Animalia - Invertebrates") %>%
+    pull(prop_spp)%>%
+    `*`(100) %>%
+    format_output()
+
+prop_idigbio_invert_samples <-  kingdom_diversity %>%
+    filter(database == "idigbio" & sub_kingdom == "Animalia - Invertebrates") %>%
+    pull(prop_samp) %>%
+    `*`(100) %>%
+    format_output()
+
+prop_idigbio_invert_species <- kingdom_diversity %>%
+    filter(database == "idigbio" & sub_kingdom == "Animalia - Invertebrates") %>%
+    pull(prop_spp)%>%
+    `*`(100) %>%
+    format_output()
+```
+
+Marine invertebrates are underrepresented in biodiversity databases.
+This underrepresentation is manifest in the small proportion of records they represent in the total iDigBio database: marine invertebrates from the US EEZ make up `r format_output(nrow(idigbio_us_inverts_records_species)/idigbio_stats$total_no_fossil*100)`% of the total number of non-fossil iDigBio records (`r round(idigbio_stats$total_no_fossil/1e6, 1)` million), and `r format_output(nrow(idigbio_us_inverts_records_species)/idigbio_stats$eez_no_fossil*100)`% of the US records (`r round(idigbio_stats$eez_no_fossil/1e6, 1)` million).
+This underrepresentation is also manifest in the collecting effort.
+In OBIS records for the US EEZ, marine invertebrates represent `r prop_obis_invert_samples`% of the occurrences but `r prop_obis_invert_species`% of the number of species.
+In the iDigBio dataset, marine invertebrates represent `r prop_idigbio_invert_samples`% of the records but `r prop_idigbio_invert_species`% of the species.
 
 We compared the total number of species represented in iDigBio and OBIS for each major division of life to the total described diversity based on the WoRMS estimates[@worms-stats].
-While marine vertebrates represent  `r total_vert_prop`% of the global marine diversity, they represent `r idigbio_vert_prop`% and `r obis_vert_prop`% of the species in iDigBio and OBIS respectively (Fig.
+While marine invertebrates represent  `r total_invert_prop`% of the global marine diversity, they represent `r idigbio_invert_prop`% and `r obis_invert_prop`% of the species in iDigBio and OBIS respectively (Fig.
 \@ref(fig:kingdom-diversity-plot)).
-This underrepresentation is also apparent in the collecting effort: the median number of specimens in iDigBio per species ranges between `r kng_medians %>% filter(sub_kingdom=="animalia") %>% pull(rg)` for marine invertebrate phyla, but is `r kng_medians %>% filter(sub_kingdom=="animalia - vertebrates") %>% pull(rg)` for marine vertebrates.
+In other words, while the number of records per species is much lower for marine invertebrates than marine vertebrates, the number of species represented in the databases is on par with the expected diversity for the group.
 

### 2017-12-23 manuscript.Rmd
edit section about global diversity estimates

--- 2017-12-15 manuscript.Rmd
@@
 ```{r kingdom-stats}
-get_vert_prop <- function(db) {
+get_invert_prop <- function(db, what = c("invert", "vert")) {
+    what <- match.arg(what)
+    string_what <- c("invert" = "Animalia - Invertebrates",
+                     "vert" = "Animalia - Vertebrates")
+    what <- string_what[what]
     format_output(kingdom_diversity[kingdom_diversity$database == db &
-                                    kingdom_diversity$sub_kingdom == "Animalia - Vertebrates", "prop_spp"] * 100)
+                                    kingdom_diversity$sub_kingdom == what,
+                                    "prop_spp"] * 100)
 }
 
-total_vert_prop <- get_vert_prop("worms")
-idigbio_vert_prop <- get_vert_prop("idigbio")
-obis_vert_prop <- get_vert_prop("obis")
+total_invert_prop <- get_invert_prop("worms", "invert")
+idigbio_invert_prop <- get_invert_prop("idigbio", "invert")
+obis_invert_prop <- get_invert_prop("obis", "invert")
 
-kng_medians <- kingdom_stats %>%
-    dplyr::filter(database == "idigbio") %>% 
-    dplyr::group_by(sub_kingdom) %>%
-    dplyr::summarize(
-               rg = if_else(
-                   Reduce("==", range(median_sample)),
-                   as.character(range(median_sample)[1]), 
-                   paste(range(median_sample), collapse = " and "))
-           )
+total_vert_prop <- get_invert_prop("worms", "vert")
+idigbio_vert_prop <- get_invert_prop("idigbio", "vert")
+obis_vert_prop <- get_invert_prop("obis", "vert")
 
+median_obis_vert <- kingdom_stats %>%
+    filter(database=="obis" & sub_kingdom == "animalia - vertebrates") %>%
+    pull("median_sample")
+
+median_obis_invert <- kingdom_stats %>%
+    filter(database=="obis" & sub_kingdom == "animalia - invertebrates") %>%
+    pull("median_sample")
+
+median_idigbio_vert <- kingdom_stats %>%
+    filter(database=="idigbio" & sub_kingdom == "animalia - vertebrates") %>%
+    pull("median_sample")
+
+median_idigbio_invert <- kingdom_stats %>%
+    filter(database=="idigbio" & sub_kingdom == "animalia - invertebrates") %>%
+    pull("median_sample")
 ```
@@
 
-Marine invertebrates are underrepresented in iDigBio and OBIS.
-This underrepresentation is manifest in the small fraction of records that marine invertebrates represent.
-Extant marine invertebrates occurring in the US EEZ represent  `r format_output(nrow(idigbio_us_inverts_records_species)/idigbio_stats$total_no_fossil*100)`% of the total number of non-fossil iDigBio records (`r round(idigbio_stats$total_no_fossil/1e6, 1)` million), and `r format_output(nrow(idigbio_us_inverts_records_species)/idigbio_stats$eez_no_fossil*100)`% of the US records (`r round(idigbio_stats$eez_no_fossil/1e6, 1)` million).
-Marine invertebrates however represent between 11% and 18% of the described global diversity (number of described marine inverterbrates species: 200,000-210,000 [@Appeltans2012], total number of described species: 1.2-1.9 million [@Mora2011,@Chapman2009]).
+Marine invertebrates are underrepresented in biodiversity databases.
+This underrepresentation is manifest in the small proportion of records they represent in the iDigBio database.
+While marine invertebrates compose between 11% and 18% of the described global diversity (number of described marine inverterbrates species: 200,000-210,000 [@Appeltans2012], total number of described species: 1.2-1.9 million [@Mora2011,@Chapman2009]),
+marine invertebrates from the US EEZ only make up `r format_output(nrow(idigbio_us_inverts_records_species)/idigbio_stats$total_no_fossil*100)`% of the total number of non-fossil iDigBio records (`r round(idigbio_stats$total_no_fossil/1e6, 1)` million), and `r format_output(nrow(idigbio_us_inverts_records_species)/idigbio_stats$eez_no_fossil*100)`% of the US records (`r round(idigbio_stats$eez_no_fossil/1e6, 1)` million).
 
-We compared the total number of species represented in iDigBio and OBIS for each major division of life to the total described diversity based on the WoRMS estimates[@worms-stats].
-While marine vertebrates represent  `r total_vert_prop`% of the global marine diversity, they represent `r idigbio_vert_prop`% and `r obis_vert_prop`% of the species in iDigBio and OBIS respectively (Fig.
+This underrepresentation is also manifest in the intensity of the collecting effort.
+In OBIS records for the US EEZ, marine invertebrates represent `r invert_props$prop_obis_invert_samples`% of the occurrences but `r invert_props$prop_obis_invert_species`% of the total number of species; and in iDigBio, they represent `r invert_props$prop_idigbio_invert_samples`% of the records but `r invert_props$prop_idigbio_invert_species`% of the species.
+In contrast, marine vertebrates represent `r invert_props$prop_idigbio_vert_samples`% of the samples and `r invert_props$prop_idigbio_vert_species`% of the species in iDigBio, and `r invert_props$prop_obis_vert_samples`% of the samples and `r invert_props$prop_obis_vert_species`% of the species in OBIS (Fig.
 \@ref(fig:kingdom-diversity-plot)).
-This underrepresentation is also apparent in the collecting effort: the median number of specimens in iDigBio per species ranges between `r kng_medians %>% filter(sub_kingdom=="animalia") %>% pull(rg)` for marine invertebrate phyla, but is `r kng_medians %>% filter(sub_kingdom=="animalia - vertebrates") %>% pull(rg)` for marine vertebrates.
+In other words, while marine invertebrates are represented at a level on par with their diversity in the marine fraction of the databases, each species is represented by fewer samples compared to vertebrates (median number of records per species for invertebrates `r median_idigbio_invert` and `r median_obis_invert`, _vs_ and `r median_idigbio_vert` and `r median_obis_vert` for vertebrates, for iDigBio and OBIS respectively).
 

### 2017-12-23 R/kingdom.R

@@ -135,12 +135,12 @@ calc_kingdom_stats <- function() {
     db <- sok_db()
 
     get_median <- . %>%
-        group_by(sub_kingdom, worms_phylum, worms_valid_name) %>%
+        group_by(sub_kingdom, worms_valid_name) %>%
         summarize(
             n_samples = n()
         ) %>%
         collect() %>%
-        group_by(sub_kingdom, worms_phylum) %>%
+        group_by(sub_kingdom) %>%
         summarize(
             median_sample = median(n_samples),
             mean_sample = mean(n_samples)
@@ -156,6 +156,64 @@ calc_kingdom_stats <- function() {
         .id = "database")
 }
 
+get_invert_props <- function(kingdom_diversity) {
+
+    smrz_samp <- . %>%
+        pull(prop_samp) %>%
+        `*`(100) %>%
+        format_output()
+
+    smrz_spp <- . %>%
+        pull(prop_spp) %>%
+        `*`(100) %>%
+        format_output()
+
+    ## invertebrates
+    prop_obis_invert_samples <- kingdom_diversity %>%
+        filter(database == "obis" & sub_kingdom == "Animalia - Invertebrates") %>%
+        smrz_samp()
+
+    prop_obis_invert_species <- kingdom_diversity %>%
+        filter(database == "obis" & sub_kingdom == "Animalia - Invertebrates") %>%
+        smrz_spp()
+
+    prop_idigbio_invert_samples <-  kingdom_diversity %>%
+        filter(database == "idigbio" & sub_kingdom == "Animalia - Invertebrates") %>%
+        smrz_samp()
+
+    prop_idigbio_invert_species <- kingdom_diversity %>%
+        filter(database == "idigbio" & sub_kingdom == "Animalia - Invertebrates") %>%
+        smrz_spp()
+
+    ## vertebrates
+    prop_obis_vert_samples <- kingdom_diversity %>%
+        filter(database == "obis" & sub_kingdom == "Animalia - Vertebrates") %>%
+        smrz_samp()
+
+    prop_obis_vert_species <- kingdom_diversity %>%
+        filter(database == "obis" & sub_kingdom == "Animalia - Vertebrates") %>%
+        smrz_spp()
+
+    prop_idigbio_vert_samples <-  kingdom_diversity %>%
+        filter(database == "idigbio" & sub_kingdom == "Animalia - Vertebrates") %>%
+        smrz_samp()
+
[…]

### 2017-12-23 manuscript.Rmd
add more context to diversity fractions

@@
 This underrepresentation is also manifest in the intensity of the collecting effort.
+Marine invertebrates are estimated to represent  `r total_invert_prop`% of the global marine diversity[@worms-stats].
 In OBIS records for the US EEZ, marine invertebrates represent `r invert_props$prop_obis_invert_samples`% of the occurrences but `r invert_props$prop_obis_invert_species`% of the total number of species; and in iDigBio, they represent `r invert_props$prop_idigbio_invert_samples`% of the records but `r invert_props$prop_idigbio_invert_species`% of the species.
@@
 \@ref(fig:kingdom-diversity-plot)).
-In other words, while marine invertebrates are represented at a level on par with their diversity in the marine fraction of the databases, each species is represented by fewer samples compared to vertebrates (median number of records per species for invertebrates `r median_idigbio_invert` and `r median_obis_invert`, _vs_ and `r median_idigbio_vert` and `r median_obis_vert` for vertebrates, for iDigBio and OBIS respectively).
+In other words, while marine invertebrates are represented at a level on par with their diversity within the US EEZ in the databases, each species is represented by fewer samples than for vertebrates (median number of records per species for invertebrates `r median_idigbio_invert` and `r median_obis_invert`, _vs_ and `r median_idigbio_vert` and `r median_obis_vert` for vertebrates, for iDigBio and OBIS respectively).
 

### 2018-01-16 manuscript.Rmd
edit ITIS numbers and discussion

--- 2017-12-23 manuscript.Rmd
@@
 
-This underrepresentation is also manifest in the intensity of the collecting effort.
-In OBIS records for the US EEZ, marine invertebrates represent `r invert_props$prop_obis_invert_samples`% of the occurrences but `r invert_props$prop_obis_invert_species`% of the total number of species; and in iDigBio, they represent `r invert_props$prop_idigbio_invert_samples`% of the records but `r invert_props$prop_idigbio_invert_species`% of the species.
-In contrast, marine vertebrates represent `r invert_props$prop_idigbio_vert_samples`% of the samples and `r invert_props$prop_idigbio_vert_species`% of the species in iDigBio, and `r invert_props$prop_obis_vert_samples`% of the samples and `r invert_props$prop_obis_vert_species`% of the species in OBIS (Fig.
-\@ref(fig:kingdom-diversity-plot)).
-In other words, while marine invertebrates are represented at a level on par with their diversity in the marine fraction of the databases, each species is represented by fewer samples compared to vertebrates (median number of records per species for invertebrates `r median_idigbio_invert` and `r median_obis_invert`, _vs_ and `r median_idigbio_vert` and `r median_obis_vert` for vertebrates, for iDigBio and OBIS respectively).
-
-```{r itis-stats}
-prop_molluscs <- itis_molluscs_in_dbs %>%
-    dplyr::summarize(
-               prop_idigbio_us = format_output(mean(n_idigbio_geo_us > 0,
-                                                    na.rm = TRUE)* 100),
-               prop_obis_us = format_output(mean(n_obis_geo_us > 0,
-                                                 na.rm = TRUE) * 100)
-           )
-
-prop_crustaceans <- itis_crustaceans_in_dbs %>%
-    dplyr::summarize(
-             prop_idigbio_us = format_output(mean(n_idigbio_geo_us > 0,
-                                                  na.rm = TRUE) * 100),
-             prop_obis_us = format_output(mean(n_obis_geo_us > 0,
-                                             na.rm = TRUE) * 100)               
-           )
-```
-
-To assess the fraction of digitization of the fauna, we compared the species included in iDigBio and OBIS with the comprehensive checklists for molluscs [@turgeon1998] and crustaceans [@mclaughlin2005] that are available for the United States.
-For molluscs, of the `r format(nrow(itis_molluscs_in_dbs), big.mark=",")` species included in the checklist, `r prop_molluscs$prop_idigbio_us`% are in iDigBio and `r prop_molluscs$prop_obis_us`% are in OBIS.
-For crustaceans, of the `r format(nrow(itis_crustaceans_in_dbs), big.mark=",")` species included in the checklist, `r prop_crustaceans$prop_idigbio_us`% are in iDigBio and `r prop_crustaceans$prop_obis_us`% are in OBIS.
-
 ```{r kingdom-diversity-plot, fig.cap='Comparison of the proportions of the diversity represented in iDigBio and OBIS  with the global diversity (as reported by WoRMS) for each kingdom.
@@
 ```
+
+This underrepresentation is also manifest in the intensity of the collecting effort.
+Marine invertebrates are estimated to represent  `r total_invert_prop`% of the total marine diversity[@worms-stats].
+In OBIS records for the US EEZ, marine invertebrates represent `r invert_props$prop_obis_invert_samples`% of the occurrences but `r invert_props$prop_obis_invert_species`% of the total number of species; and in iDigBio, they represent `r invert_props$prop_idigbio_invert_samples`% of the records but `r invert_props$prop_idigbio_invert_species`% of the species.
+In contrast, marine vertebrates represent `r invert_props$prop_idigbio_vert_samples`% of the samples and `r invert_props$prop_idigbio_vert_species`% of the species in iDigBio, and `r invert_props$prop_obis_vert_samples`% of the samples and `r invert_props$prop_obis_vert_species`% of the species in OBIS (Fig.
+\@ref(fig:kingdom-diversity-plot)).
+In other words, while marine invertebrates are represented at a level on par with their diversity within the US EEZ in the databases, each species is represented by fewer samples than for vertebrates (median number of records per species for invertebrates `r median_idigbio_invert` and `r median_obis_invert`, _vs._ and `r median_idigbio_vert` and `r median_obis_vert` for vertebrates, for iDigBio and OBIS respectively).
+
+To assess the level of digitization of the fauna, we compared the species included in iDigBio and OBIS with the comprehensive checklists for molluscs [@turgeon1998] and crustaceans [@mclaughlin2005] that are available for the United States.
+For molluscs, of the `r format(itis_stats$moll_nspp, big.mark=",")` species included in the checklist, `r itis_stats$prop_moll_idig`% are in iDigBio and `r itis_stats$prop_moll_obis`% are in OBIS.
+For crustaceans, of the `r format(itis_stats$crust_nspp, big.mark=",")` species included in the checklist, `r itis_stats$prop_crust_idig`% are in iDigBio and `r itis_stats$prop_crust_obis`% are in OBIS.
+Given that molluscs and crustaceans are among the most sampled taxa, and that they have received more taxonomic attention than other phyla (see below), these percentages offer a conservative estimate for the fraction of the fauna that is available digitally.
+Therefore, we can estimate that there is at least 28,000 species of marine invertebrates in US waters.
+The higher overlap between the comprehensive checklist for  crustaceans and OBIS is attributable to occurences of planktonic species that are more represented in OBIS data than in museum collections.
 

### 2018-01-16 R/itis.R

@@ -258,27 +258,89 @@ molluscs_name_in_dbs <- function(tbl, combined_species) {
         add_dbs_info(combined_species)
 }
 
-itis_in_idigbio <- function() {
+get_itis_stats <- function(itis_moll, itis_crust, idig_recs, obis_recs) {
 
-}
+    ## calculate number of species in ITIS lists
+    calc_nspp <- . %>%
+        dplyr::filter(is_marine) %>%
+        dplyr::select(worms_valid_name) %>%
+        dplyr::n_distinct()
+    
+    itis_moll_nspp <- itis_moll %>%
+        calc_nspp
 
-itis_in_obis <- function() {
+    itis_crust_nspp <- itis_crust %>%
+        calc_nspp
 
-}
+    ## get species lists per database
+    ## -- molluscs
+    itis_moll <- itis_moll %>%
+        dplyr::filter(is_marine) %>%
+        dplyr::distinct(worms_valid_name)
 
-itis_in_bold <- function() {
+    idig_moll <- idig_recs %>%
+        dplyr::filter(worms_phylum == "mollusca") %>%
+        dplyr::distinct(worms_valid_name)
 
-}
+    obis_moll <- obis_recs %>%
+        dplyr::filter(worms_phylum == "mollusca") %>%
+        dplyr::distinct(worms_valid_name)
 
-itis_molluscs_stats <- function(moll_geo) {
+    ## -- crustaceans
+    itis_crust <- itis_crust %>%
+        dplyr::filter(is_marine) %>%
+        dplyr::distinct(worms_valid_name)
+
+    idig_crust <- idig_recs %>%
+        dplyr::filter(worms_phylum == "arthropoda") %>%
+        dplyr::distinct(worms_valid_name)
+
+    obis_crust <- obis_recs %>%
+        dplyr::filter(worms_phylum == "arthropoda") %>%
+        dplyr::distinct(worms_valid_name)
+
+    ## figure out intersection
+    ## -- molluscs
+    calc_moll <- . %>%
+        dplyr::select(worms_valid_name) %>%
+        dplyr::n_distinct() %>%
+        `/`(itis_moll_nspp) %>%
+        `*`(100) %>%
+        format_output()
+
+    prop_moll_idig <-  dplyr::semi_join(itis_moll, idig_moll,
+                                        by = "worms_valid_name") %>%
+        calc_moll
+
+    prop_moll_obis <- dplyr::semi_join(itis_moll, obis_moll,
[…]

### 2018-01-16 manuscript.Rmd
fix typo in calculation

@@
                    dplyr::pull(n) %>%
-                   sum()) / dplyr::pull(idigbio_mms_summary_by_year, n)
+                   sum()) / sum(dplyr::pull(idigbio_mms_summary_by_year, n))
 ```
@@
 Biodiversity surveys conducted under the auspices of the Bureau of Land Management (BLM) and the Minerals Management Service (MMS) have contributed significantly to our knowledge of the geographic distribution of marine invertebrates in US waters.
-In our dataset, we identified at least `r format(sum(idigbio_mms_summary_by_year$n), big.mark=",")` records (or `r sum(idigbio_mms_summary_by_year$n)/nrow(idigbio_us_inverts_records_all)*100`% of all US marine invertebrates in iDigBio) from these surveys, `r format_output(prop_mms_75_85 * 100)`% of which were collected between 1975 and 1985.
-Of the `r idigbio_mms_records %>% filter(rank %in% c("Species", "Subspecies")) %>% summarize(n = n_distinct(worms_valid_name))` species collected as part of these surveys, `r nrow(idigbio_mms_species_only)` have not been collected outside of these surveys.
+In our dataset, we identified at least `r format(sum(idigbio_mms_summary_by_year$n), big.mark=",")` records (or `r format_output(sum(idigbio_mms_summary_by_year$n)/nrow(idigbio_us_inverts_records_all)*100)`% of all US marine invertebrates in iDigBio) from these surveys, `r format_output(prop_mms_75_85 * 100)`% of which were collected between 1975 and 1985.
+Of the `r idigbio_mms_records %>% filter(rank %in% c("Species", "Subspecies")) %>% summarize(n = n_distinct(worms_valid_name)) %>% format(big.mark=",")` species collected as part of these surveys, `r nrow(idigbio_mms_species_only)` have not been collected outside of these surveys.
 

### 2018-01-16 manuscript.Rmd
edit manuscript

@@
 
-We retrieved `r format(n_distinct(combined_species$worms_valid_name), big.mark = ",")` species (`r n_distinct(na.omit(combined_species$worms_phylum))` phyla) of marine invertebrates in the continental US EEZ by combining  `r format(nrow(idigbio_us_inverts_records_species), big.mark=",")` museum records from iDigBio (`r format(n_distinct(idigbio_us_inverts_records_species$worms_valid_name), big.mark = ",")` species, `r n_distinct(na.omit(idigbio_us_inverts_records_species$worms_phylum))` phyla), and `r format(nrow(obis_us_inverts_records_species), big.mark=",")` records from OBIS (`r format(n_distinct(obis_us_inverts_records_species$worms_valid_name), big.mark=",")` species,
+We retrieved `r format(n_distinct(combined_species$worms_valid_name), big.mark = ",")` species (`r n_distinct(na.omit(combined_species$worms_phylum))` phyla) of marine invertebrates in the continental US EEZ by combining  `r format(nrow(idigbio_us_inverts_records_species), big.mark=",")` museum records from iDigBio (`r format(n_distinct(idigbio_us_inverts_records_species$worms_valid_name), big.mark = ",")` species, `r n_distinct(na.omit(idigbio_us_inverts_records_species$worms_phylum))` phyla), and `r format(nrow(obis_us_inverts_records_species), big.mark=",")` occurence records from OBIS (`r format(n_distinct(obis_us_inverts_records_species$worms_valid_name), big.mark=",")` species,
 `r n_distinct(na.omit(obis_us_inverts_records_species$worms_phylum))` phyla).
@@
 
-Marine invertebrates are underrepresented in biodiversity databases.
-This underrepresentation is manifest in the small proportion of records they represent in the iDigBio database.
-While marine invertebrates compose between 11% and 18% of the described global diversity (number of described marine inverterbrates species: 200,000-210,000 [@Appeltans2012], total number of described species: 1.2-1.9 million [@Mora2011,@Chapman2009]),
+Given their diversity, marine invertebrates are underrepresented in iDigBio.
+While marine invertebrates are estimated to represent between 11% and 18% of the described global diversity (number of described marine inverterbrates species: 200,000-210,000 [@Appeltans2012], total number of described species: 1.2-1.9 million [@Mora2011,@Chapman2009]),
 marine invertebrates from the US EEZ only make up `r format_output(nrow(idigbio_us_inverts_records_species)/idigbio_stats$total_no_fossil*100)`% of the total number of non-fossil iDigBio records (`r round(idigbio_stats$total_no_fossil/1e6, 1)` million), and `r format_output(nrow(idigbio_us_inverts_records_species)/idigbio_stats$eez_no_fossil*100)`% of the US records (`r round(idigbio_stats$eez_no_fossil/1e6, 1)` million).
@@
 
+More records have been collected from the Atlantic Ocean, than the Gulf of Mexico, and the Pacific Ocean (`r format(sum(combined_records$is_east_coast), big.mark=",")`, `r format(sum(combined_records$within_gom), big.mark=",")`, and `r format(sum(combined_records$is_west_coast), big.mark=",")` records respectively).
+The highest species richness has been recorded from the Gulf of Mexico (`r format(n_spp_gom, big.mark=",")` species), followed by the Atlantic Ocean (`r format(n_spp_east, big.mark=",")` species), and the Pacific Ocean (`r format(n_spp_west, big.mark=",")` species).
+
 Most samples have been collected on the continental shelf (Fig.
 \@ref(fig:map-diversity-sampling)A), with `r prop_idigbio_continental_shelf`% (iDigBio) and `r prop_obis_continental_shelf`% (OBIS) of the records collected between 0-200m.
-More records have been collected from the Atlantic Ocean, than the Gulf of Mexico, and the Pacific Ocean (`r format(sum(combined_records$is_east_coast), big.mark=",")`, `r format(sum(combined_records$is_within_gom), big.mark=",")`, and `r format(sum(combined_records$is_west_coast), big.mark=",")` records respectively).
-The highest species richness has been recorded from the Gulf of Mexico (`r format(n_spp_gom, big.mark=",")` species), followed by the Atlantic Ocean (`r format(n_spp_east, big.mark=",")` species), and the Pacific Ocean (`r format(n_spp_west, big.mark=",")` species).
 
@@
 \@ref(fig:latitudinal-gradient)).
+
+Talk about contributions of different institutions here
 
@@
 
-```{r correlation-sampling, fig.cap='Relation between number of records and number of species for each cell of the map used in Fig.
+```{r correlation-sampling, fig.cap='Relationship between number of records and number of species for each cell of the map used in Fig.
 \\@ref(fig:map-diversity-sampling).
@@
 There is a tight correlation between the number of samples and number of species per cell used in Fig.
-\@ref(fig:map-diversity-sampling), and `r format_output(stat_sampling_effort$p * 100)`% (n = `r stat_sampling_effort$n`) of the cells with $\leq$ 500 records have a number of species within 10% of the number of records (Fig.
+\@ref(fig:map-diversity-sampling), and `r format_output(stat_sampling_effort$p * 100)`% (n = `r format(stat_sampling_effort$n, big.mark=",")`) of the cells with $\leq$ 500 records have a number of species within 10% of the number of records (Fig.
 \@ref(fig:correlation-sampling)).

### 2018-01-17 R/gradient.R
change axis legend

@@ -145,7 +145,7 @@ plot_gradient_sampling <- function(lat_data) {
         facet_grid(~ lat_coast) +
         scale_colour_hc() + guides(color = FALSE) +
         scale_y_log10() +
-        xlab("Latitude") + ylab("Number of occurrences")
+        xlab("Latitude") + ylab("Number of observations")
 
 }
 

### 2018-01-17 R/institutions.R
add LaTeX label to legend

@@ -102,7 +102,7 @@ calc_institutions <- function(idig_records, obis_records) {
     )
     x_list_tbl <- xtable::xtableList(
                               list_tbl, align = c("l", "p{11.5cm}", "R{2cm}", "R{2cm}"),
-                              caption = capt
+                              caption = capt, label = "tab:records-table"
                               )
     x_list_tbl
 }

### 2018-01-17 manuscript.Rmd
adjust number of total species inferred
reorganize paragraph

@@
 Given that molluscs and crustaceans are among the most sampled taxa, and that they have received more taxonomic attention than other phyla (see below), these percentages offer a conservative estimate for the fraction of the fauna that is available digitally.
-Therefore, we can estimate that there is at least 28,000 species of marine invertebrates in US waters.
+Therefore, we can estimate that there is at least 29,000 species of marine invertebrates in US waters.
 The higher overlap between the comprehensive checklist for  crustaceans and OBIS is attributable to occurences of planktonic species that are more represented in OBIS data than in museum collections.
@@
 The highest species richness has been recorded from the Gulf of Mexico (`r format(n_spp_gom, big.mark=",")` species), followed by the Atlantic Ocean (`r format(n_spp_east, big.mark=",")` species), and the Pacific Ocean (`r format(n_spp_west, big.mark=",")` species).
-
 Most samples have been collected on the continental shelf (Fig.
 \@ref(fig:map-diversity-sampling)A), with `r prop_idigbio_continental_shelf`% (iDigBio) and `r prop_obis_continental_shelf`% (OBIS) of the records collected between 0-200m.
+The higher species richness measured on the East coast is in part attributable to the wider continental shelf, the presence of more institutions, and a higher level of digitization of collections on the East coast (Table \@ref(tab:records-table)).
 

### 2018-01-17 manuscript.Rmd
[fix #80] add discussion about gradients

@@
 
-Spatial distribution of species richness with latitude differ markedly between the Pacific and the Atlantic coasts.
-On the Pacific coast, richness is lowest around 40° of latitude, whereas on the Atlantic coast, it is highest at the same latitude (Fig.
+Spatial distribution of species richness with latitude differ between the Pacific and the Atlantic coasts.
+On the Pacific coast, the recorded richness is lowest around 40° of latitude, whereas on the Atlantic coast, it is highest at the same latitude (Fig.
 \@ref(fig:map-diversity-sampling)B, Fig.
-\@ref(fig:latitudinal-gradient)).
-
-Talk about contributions of different institutions here
+\@ref(fig:latitudinal-gradient)B).
+These trends are however strongly driven by sampling effort (Fig.
+\@ref(fig:latitudinal-gradient)C).
+Notably, for the Pacific coast, there is a deficit in the number of samples between 40° and 45° of latitude corresponding to the state of Oregon.
+In contrast, the peak in number of species observed and in the number of records around 42° on the Atlantic coast can be attributed to the presence of the Woods Hole Oceanographic Institute.
 

### 2018-01-17 manuscript.Rmd
add sentence about inadequacy of sampling
[fix #84] and other text edits

@@
 Given that molluscs and crustaceans are among the most sampled taxa, and that they have received more taxonomic attention than other phyla (see below), these percentages offer a conservative estimate for the fraction of the fauna that is available digitally.
-Therefore, we can estimate that there is at least 29,000 species of marine invertebrates in US waters.
+Therefore, using a conservative estimate of a 60% digitization level of the fauna, we can estimate that there is at least 29,000 species of marine invertebrates in US waters.
 The higher overlap between the comprehensive checklist for  crustaceans and OBIS is attributable to occurences of planktonic species that are more represented in OBIS data than in museum collections.
@@
 \@ref(fig:correlation-sampling)).
+For most of the US EEZ, the sampling effort is insufficient to capture acurately the species diversity.
 
@@
 Based on the results of this study, we recommend a multi-level approach for both improving the quality of existing data and for the efficient collection of new data.
+Our recommendations extends the recommendations formulated by Duffy et al.
+[@Duffy2013], and are informed by the results of our analyses.
 
-### Improve existing data
+### Improve existing data quality and access
+
+Duffy et al.
+[@Duffy2013] suggested that resources should be invested in the digitization of historical biodiversity to study changes in biodiversity through time in response to climate change, fishing pressure, and pollution.
+Here, we extend this recommendation to also include improving the quality of the data that is already digitized.
 
@@
 Linking species names to identifiers in taxonomic databases would facilitate data linking, specimen discoverability, and data usability.
-<!-- GP comments: You still have the human knowledge/ID factor to contend with though, which is a more important issue.
-Also,places like WoRMS almost never link a given name to multiple useages and useage changes with time -->
-Additionally, because registers of species names, such as WoRMS, link the history of the names and their synonymies, it is possible to dissociate the concepts of species identification with the concept of species identity, allowing researchers to compare specimens identified at different time periods.
 
@@
 Such services, however, require that resources be made available to museum collections in order to address these issues directly in collection databases.
-<!-- GP comment: see geolocate plugin in Specify -->
 
@@
 As data across institutions are incorporated and pooled into large biodiversity information systems, it is crucial that data standards such as Darwin Core are embraced by the community.
+As more people know how to analyze data, and in turn, use the data for their research, inaccuracies can be fixed and overall data quality can improve.
 
@@
 
-<!-- Finally - need better links in GenBank and BOLD to vouchers, names, and geography -->
-
-<!-- Specimens with accurate identifications and specimens that have been preserved in conditions such that it is possible to get DNA out of them should be prioritized, as they yield the most data that are relevant to biodiversity knowledge.
--->
-
 ### Collect better data faster and more cost effectively
-
-<!-- separate issues of sampling is needed, and large-scale sampling is best way of doing it -->
-
-<!-- Coordinated large-scale biodiversity expeditions continue to be needed to improve our understanding of the diversity of species and their spatial distributions.
-The US marine EEZ is relatively well-known compared to many areas of the world ocean, where studies show that undescribed biodiversity dominates.
--->
-
-<!-- For example, a recent survey in Western Australia reported that 94.6% of crustaceans and 72% of polychaetes taken at 135 sites were undescribed [@Poore2014]. -->
-
-<!-- GP comment: Several of these recommendations match ones made in Duffy et al 2013 -- ref 4  Should emphasize the overlap -->
-
-Museum specimens are critical for validating the occurrence of species [@Rocha2014], especially for marine invertebrates for which species limits based on morphology have been known to be challenged by other lines of evidence including molecular data [@Knowlton2000].
 
@@
 We recommend that additional coordinated field expeditions be organized to sample more comprehensively the marine environment.
-The relationship between sampling effort and diversity from museum collections makes clear that any additional samples will greatly improve our knowledge of marine biodiversity.
+The relationship between sampling effort and diversity from museum collections makes clear that any additional samples will lead to improving our knowledge of species ranges, and will lead to the discovery of many new species.
 Concerted and collaborative efforts are needed to share expertise and limited resources.
+Biodiversity surveys that target all animal phyla can be very effective in documenting biodiversity.
+For instance, one of us (GP) participated in the summer 2017 in a an inventory of Kaneohe Bay, Hawaii, where ~2,400 records representing ~1,300 species were collected in 11 days of field work.
+While diversity is higher in Hawaii than in the continental US, this is a modest effort that led to the collection of a number of species that represent a few percents of the total diversity of the US EEZ marine invertebrate fauna.
 
-<!-- GP comment: You could give a biodiversity survey as an example - e.g.
-in Kaneohe we did ~2400 records of ~1300 species in 11 days - in context of whole US effort, this is substantial, especially at the species level.
--->
-
-Samples collected during these expeditions need to be photographed to document live appearance, information that would otherwise be lost during the preservation process.
+Museum specimens are critical for validating the occurrence of species [@Rocha2014], especially for marine invertebrates for which species limits based on morphology have been known to be challenged by other lines of evidence including molecular data [@Knowlton2000].
+Samples collected during biodiversity surveys need to be photographed to document live appearance, information that would otherwise be lost during the preservation process.
 This has been shown to be particularly important to validate species limits in cryptic complexes.
@@
 Integrating metabarcoding as part of the data collection conducted in biodiversity surveys is a promising approach to screen an habitat for its potential to lead to new discoveries.
-For instance, by analyzing metabarcoding samples that contain planktonic larvae from a given habitat, we could get an indication of which taxonomic group would benefit most from additional sampling effort or give an indication of the proportion of species that has not been sampled yet.
+For instance, by analyzing metabarcoding samples that contain planktonic larvae from a given habitat, we could get an indication of the taxonomic groups that would benefit most from additional sampling effort, or provide an estimate for the proportion of species that has not been sampled yet.
 Additionally, data generated by metabarcoding efforts can be re-interpreted as more DNA barcodes associated with correctly identified specimens become available.
@@
 
+To make biodiversity science and conservation decisions more data driven, we need to ensure that biodiversity data is accurate, comprehensive, freely available, and that the datasets are linked.
 DNA sequences are an early test case on how to link biodiversity data across repositories.
@@
 
-<!-- widening gap with other groups (vertebrates, plants) for which more comprehensive data is available -->
-
-<!-- here there are 4 things: data availability (having good APIs to access the
-data); data comprehensiveness (how many species are captured in the database);
-data accuracy (can we trust the data that is found in the databases); data linkage -->
-
 ## Methods
@@
 For the Gulf of Mexico, we retrieved -->
+
+Species occurence data and museum records were extracted from OBIS (Ocean Biogeographic Information System) and iDigBio (Integrated Digitized Biocollections) respectively.
+OBIS gathers and integrates datasets about occurrences of marine species around the world provided by over 500 institutions worldwide.
+iDigBio provides access to the data associated with museum records that have been digitized and made publicly available.
+While OBIS records can be based on observation, there is a physical specimen deposited in a collection associated with each iDigBio record.
 

### 2018-01-18 manuscript.Rmd
add justification of search by coordinates in methods section

@@
 
+We opted to limit our queries to records with geographic coordinates instead of relying on the "Country" fields of these databases.
+Tests showed that these queries would return numerous records outside the geographic range of interest for this study (for instance records collected in Hawaii, Alaska, or the US Virgin Islands), while not taking into account records falling within the US EEZ where the "Country" field contained "Atlantic Ocean" for instance.
+
 All species names obtained from all sources were checked against the World Registry of Marine Species (WoRMS).

### 2018-01-19 manuscript.Rmd
[fix #87] write about influence of missing geo coordinates

@@
 For crustaceans, of the `r format(itis_stats$crust_nspp, big.mark=",")` species included in the checklist, `r itis_stats$prop_crust_idig`% are in iDigBio and `r itis_stats$prop_crust_obis`% are in OBIS.
-Given that molluscs and crustaceans are among the most sampled taxa, and that they have received more taxonomic attention than other phyla (see below), these percentages offer a conservative estimate for the fraction of the fauna that is available digitally.
+Given that molluscs and crustaceans are among the most sampled taxa, and that they have received more taxonomic attention than other phyla (Fig.
+\@ref(fig:idigbio-through-time)), these percentages offer a conservative estimate for the fraction of the fauna that is available digitally.
 Therefore, using a conservative estimate of a 60% digitization level of the fauna, we can estimate that there is at least 29,000 species of marine invertebrates in US waters.
 The higher overlap between the comprehensive checklist for  crustaceans and OBIS is attributable to occurences of planktonic species that are more represented in OBIS data than in museum collections.
+
+The faunistic list we assembled represent a subset of the species that have been digitized, because only records that are associated with geographic coordinates are included.
+All records published by OBIS have geographic coordinates associated with them.
+To estimate how records with missing geographic coordinates in iDigBio affect the completeness of our faunistic list, we queried iDigBio using the comprehensive lists for molluscs and crustaceans, and calculated the proportion of species with no iDigBio records within the US EEZ, and which had "United States" or "USA" in the country field, but had no geographic coordinates associated with them.
+These species may occur within the limits of the US EEZ but would not be returned in the queries used to assemble the faunistic list.
+We found that `r format_output(itis_stats$missing_geo_moll$prop_us_no_geo * 100)`% ($n=$ `r format(itis_stats$missing_geo_moll$n_spp, big.mark=",")`) of the species included in the comprehensive checklist for molluscs have records in iDigBio that place them in the United States based on the "Country" field, but do not have geographic coordinates associated with them that would ensures they fall within the limits of the EEZ.
+This percentage decreases to `r format_output(itis_stats$missing_geo_moll$prop_us_no_geo_no_obis * 100)`% once we exclude species that have informed geographic coordinates in OBIS.
+For crustaceans, these proportions are lower at `r format_output(itis_stats$missing_geo_crust$prop_us_no_geo * 100)`% and `r format_output(itis_stats$missing_geo_crust$prop_us_no_geo_no_obis * 100)`% ($n=$ `r format(itis_stats$missing_geo_crust$n_spp)`) respectively.
+This means ...
 
@@
 For instance, one can expect that the digitization coverage for the Gulf of Mexico will greatly increase once records from the Florida Wildlife Research Institute collections (St Petersburg, FL) become integrated into biodiversity information systems.
-Similarly, only a small fraction of the marine invertebrates records from instiutions on the Pacific has been digitized.
+Similarly, only a small fraction of the marine invertebrates records from institutions located on the Pacific coast has been digitized, as indicated by the paucity of Pacific coast collections listed in Table \@ref(tab:records-table).
 

### 2018-01-19 R/itis.R

@@ -150,7 +150,8 @@ add_dbs_info <- function(tbl, combined_species) {
                                   n_idigbio_total = 0,
                                   n_idigbio_no_geo = 0,
                                   n_idigbio_geo_us = 0,
-                                  n_idigbio_country_us = 0
+                                  n_idigbio_country_us = 0,
+                                  n_idigbio_country_us_no_geo = 0
                               ))
                           .idig <- .idig %>%
                               dplyr::rename(decimallatitude = geopoint.lat,
@@ -263,7 +264,26 @@ molluscs_name_in_dbs <- function(tbl, combined_species) {
 
 get_itis_stats <- function(itis_moll, itis_crust, idig_recs, obis_recs) {
 
-    ## calculate number of species in ITIS lists
+    ## Missing geographic information ------------------------------------------
+    missing_geo <- . %>%
+        dplyr::filter(is_marine, n_idigbio_geo_us == 0) %>%
+        dplyr::distinct(worms_valid_name, .keep_all = TRUE) %>%
+        dplyr::summarize(
+                   n_spp = n(),
+                   prop_us_no_geo = mean(n_idigbio_country_us_no_geo > 0),
+                   prop_us_no_geo_no_obis = mean(n_idigbio_country_us_no_geo > 0 &
+                                                 n_obis_geo_us == 0)
+               )
+
+    ## -- molluscs
+    missing_geo_moll <- itis_moll %>%
+        missing_geo
+
+    ## -- crust
+    missing_geo_crust <- itis_crust %>%
+        missing_geo
+
+    ## calculate number of species in ITIS lists -------------------------------
     calc_nspp <- . %>%
         dplyr::filter(is_marine) %>%
         dplyr::select(worms_valid_name) %>%
@@ -302,7 +322,7 @@ get_itis_stats <- function(itis_moll, itis_crust, idig_recs, obis_recs) {
         dplyr::filter(worms_phylum == "arthropoda") %>%
         dplyr::distinct(worms_valid_name)
 
-    ## figure out intersection
+    ## figure out intersection -------------------------------------------------
     ## -- molluscs
     calc_moll <- . %>%
         dplyr::select(worms_valid_name) %>%
@@ -335,14 +355,17 @@ get_itis_stats <- function(itis_moll, itis_crust, idig_recs, obis_recs) {
                                         by = "worms_valid_name") %>%
         calc_crust
 
-    ## --- output
+
[…]

### 2018-04-16 manuscript.Rmd
fix small typos before incorporating GP's edits

@@
 ---
-#author: Fran\c{c}ois Michonneau
 title: "The state of knowledge for marine invertebrate biodiversity in the continental US"
@@
 Several factors may contribute to species being included in the regional lists but not having records in biodiversity databases.
-The first category of biases is linked to the likelihood having specimens for these species in digitized museum collections.
+The first category of biases is linked to the likelihood of having specimens for these species in digitized museum collections.
 Rare species may only have been collected a few times, and the corresponding records have not been digitized yet.
@@
 
-The lack of tools for the authors of these checklists to have access to museum records across institutions explains why some species were not included in the checklists.
+The authors of these checklists did not have access to digital databases allowing them to access  museum records across institutions, which explains why some species were not included in the checklists.
 Many of these records were difficult to discover before digitization which, for the most part, has occurred after the compilation of these checklists.
@@
 The GOM has comparable coverage in level of DNA barcoding of the fauna to the EEZ, while the PNW is better barcoded with >30% species-level coverage across each of the five most diverse phyla.
-This is driven in part by the active Canadian barcoding effort in British Columbia.
+This is driven in part by active Canadian barcoding efforts in British Columbia.
 
@@
 Specimens identified under the same species but sampled from distant locations are especially prone to representing cryptic species.
-While geographic information is missing for most records (`r format_output(100*stat_bold$p_no_coords)`% of the `r format(stat_bold$n_total_records, big.mark=",")` BOLD records analyzed here have no geographic coordinate information associated with them), of the `r bold_geography_stats$n_spp_with_records` marine invertebrates species from the US with BOLD records, only `r format_output(bold_geography_stats$p_within_eez * 100)`% have at least one record collected within the EEZ and `r format_output(bold_geography_stats$p_in_north_america * 100)`% list a North American country as the site of collection.
+While geographic information is missing for most records (`r format_output(100*stat_bold$p_no_coords)`% of the `r format(stat_bold$n_total_records, big.mark=",")` BOLD records analyzed here have no geographic coordinate information associated with them), of the `r format(bold_geography_stats$n_spp_with_records, big.mark=",")` marine invertebrates species from the US with BOLD records, only `r format_output(bold_geography_stats$p_within_eez * 100)`% have at least one record collected within the EEZ and `r format_output(bold_geography_stats$p_in_north_america * 100)`% list a North American country as the site of collection.
 Although DNA barcodes are available for `r format_output(prop_spp_barcoded$p_barcoded * 100)`% of the marine invertebrate fauna, many of these records have been collected outside the Atlantic and Pacific US biogeographic regions, and therefore geographical differentiation has not been evaluated for most taxa.
@@
 Based on the results of this study, we recommend a multi-level approach for both improving the quality of existing data and for the efficient collection of new data.
-Our recommendations extends the recommendations formulated by Duffy et al.
+Our recommendations extends those formulated by Duffy et al.
 [@Duffy2013], and are informed by the results of our analyses.
@@
 For instance, one of us (GP) participated in the summer 2017 in a an inventory of Kaneohe Bay, Hawaii, where ~2,400 records representing ~1,300 species were collected in 11 days of field work.
-While diversity is higher in Hawaii than in the continental US, this is a modest effort that led to the collection of a number of species that represent a few percents of the total diversity of the US EEZ marine invertebrate fauna.
+While diversity is higher in Hawaii than in the continental US, this effort led to the collection of a number of species that represent a few percents of the total diversity of the US EEZ marine invertebrate fauna.
+It demonstrates that documenting the marine fauna of the US is doable.
 
@@
 
-Integrating metabarcoding as part of the data collection conducted in biodiversity surveys is a promising approach to screen an habitat for its potential to lead to new discoveries.
+Integrating metabarcoding as part of the data collection conducted in biodiversity surveys is a promising approach to screen an habitat for its potential to lead new discoveries.
 For instance, by analyzing metabarcoding samples that contain planktonic larvae from a given habitat, we could get an indication of the taxonomic groups that would benefit most from additional sampling effort, or provide an estimate for the proportion of species that has not been sampled yet.
@@
 With such approaches, the geographic distribution of species can build on and be linked to existing records, species and their natural history, rather than being limited to matching DNA sequences across studies.
+
+Metabarcoding also offers to be a promising technology to monitor changes in communities and for understanding the functioning of ecosystems.
+For instance, despite our limited knowledge of the diversity of communities in the meiofauna, metabarcoding has been successfully used to detect and documents changes in species assemblages following the Deepwater Horizon oil spill [@Bik2012].
+Metabarcoding has also revealed important differences in the diet of three related species of fishes that co-occur at small spatial scales, providing a much more detailed understanding of trophic relations [@Leray2015b].
 
@@
 Developing robust identifiers that can be shared across biodiversity information systems, and anchored in results from DNA barcoding or other molecular techniques, is needed to collaboratively build our knowledge of species distributions.
-
-Metabarcoding also offers to be a promising technology to monitor changes in communities and for understanding the functioning of ecosystems.
-For instance, despite our limited knowledge of the diversity of communities in the meiofauna, metabarcoding has been successfully used to detect and documents changes in species assemblages following the Deepwater Horizon oil spill [@Bik2012].
-Metabarcoding has also revealed important differences in the diet of three related species of fishes that co-occur at small spatial scales, providing a much more detailed understanding of trophic relations [@Leray2015b].
 
@@
 To generate the combined list of records from iDigBio and OBIS, we removed putative duplicate based on the following criteria: same phylum, same accepted species name from WoRMS, same geographic coordinates (rounded to the 4th decimal), and same date of collection.
-When the date is missing, a single record per species with identical coordinates is included.
+When the date was missing, a single record per species with identical coordinates is included.
 

### 2018-04-16 manuscript.Rmd
incorporate GP's comments in abstract and intro

@@
 bibliography: /home/francois/Library/dark-plankton.bib
-abstract: Museum collections have been recording our knowledge about species and their geographic ranges for about 250 years, but until recently, this information was fragmented across institutions and difficult to synthesize.
-The digitization of biodiversity information is now enabling the synthesis of this information, and can be used to highlight knowledge gaps.
-Here we use publicly accessible biodiversity data to assemble a list of `r format(n_distinct(combined_species$worms_valid_name), big.mark = ",")` species of marine invertebrates from the continental United States.
-Our knowledge about species richness and their geographic ranges is limited by sampling effort, as `r prop_10_specimens_species`% of the species are known from less than 10 occurrences.
-While recent developments in high-throughput sequencing promise to greatly speed up the documentation of marine biodiversity, anchoring results from such studies into existing DNA sequence data and associated vouchers would provide a richer biological context to interpret the insights provided by these datasets.
-However, with `r format_output(prop_spp_barcoded$p_barcoded * 100)`% of the marine invertebrate fauna that has sequences available for the most widely used DNA barcoding gene (cytochrome oxidase I, COI), and given that for many species these records have been sampled in biogeographic areas outside the USA, additional efforts are needed to assemble a comprehensive library of reference DNA barcoding sequences.
-More sampling, coordinated at the national scale, is needed to improve our knowledge of species and their geographic ranges.
+abstract: Museum collections have been recording our knowledge about the diversity and distribution of species for about 250 years.
+Until recently, this information was fragmented across institutions and difficult to access or synthesize.
+The digitization of biodiversity information is enabling the synthesis of collection data.
+Here we use publicly accessible digital biodiversity data to asses how well we know the marine invertebrate fauna of the continental United States.
+Globally, marine inverterbrates comprise about 75% of described marine species.
+We document `r format(n_distinct(combined_species$worms_valid_name), big.mark = ",")` species from the continental US waters in digital databases (OBIS and iDigBio), and explore spatial and temporal patterns of their distribution and collection.
+Databases do well at the fauna level, capturing ~xxx% of the species diversity.
+However, sampling bias obscure broad diversity patterns, such as differences between coasts and across latitude.
+Our knowledge about species richness and geographic range is limited by sampling effort, with `r prop_10_specimens_species`% of species known from $\leq$10 samples.
+While recent developments in high-throughput sequencing promise to greatly speed up documentation of marine biodiversity, interpretation of sequences requires anchoring results anchoring results to identified vouchers.
+DNA barcode sequences (COI) are presently available for only `r format_output(prop_spp_barcoded$p_barcoded * 100)`% of the US marine invertebrates, and many of these are from samples taken outside the US.
+Over a third of the barcoded species are complexes of of multiple cryptic lineages, indicating that the fauna is substantially richer than the above number indicate and needs to be taxonomically reevaluated as a whole.
+Substantially increased sampling, coordinated at the national scale, is needed to provide a meaningful baseline for marine biodiversity and for metabarcoding and metagenomics studies.
 This data will allow the detection of the impacts of climate change and other anthropogenic strains on marine biodiversity.
-Advances in computing, information systems, and sequencing technologies, have set the stage to inform where future efforts for documenting biodiversity should be applied.
+Advances in computing, information systems, and sequencing technologies, have set the stage for a broad understanding of the dynamics of marine biodiversity, but this potentially is limited by the availablility of current samples, digitization, and taxonomic effort.
 ---
@@
 
-With increasing impact of human activities on the biosphere, we need efficient and accurate approaches to document changes in species ranges and community composition.
+With increasing impact of human activities on the biosphere, we need efficient and accurate approaches to document changes in species diversity, distribution, and community composition.
 <!-- add why?
--->  While terrestrial ecosystems have received considerable attention, the baseline of healthy and resilient marine ecosystems is not well understood [@Worm2006], and modeling the impact of human activities has been limited to a few taxonomic groups [@Tittensor2010].
-For marine organisms, because of the diversity of their complex life history strategies, their body size range, and the difficulties associated with sampling their habitats, we have limited information about which species live where, or what portion of their diversity remains to be discovered and described [@Mora2011].
-Additionally, coordinated marine ecosystem monitoring initiatives have been lacking [@Duffy2013,@Borja2016] (but are now getting organized, see MarineGEO [@Duffy2016marinegeo], MBON, and PISCO), making it difficult to monitor changes in species distributions and abundances.
-While most documented global extinctions of marine species are vertebrates, many other species may have been lost unnoticed because they were not known [@Sala2006].
-There is therefore a pressing need to improve our knowledge of marine systems before species go extinct, and their functioning is irreversibly altered.
+-->  While terrestrial ecosystems have received considerable attention, the baseline of healthy and resilient marine ecosystems are not well understood [@Worm2006], and modeling the impact of human activities has been limited to a few well-studied taxonomic groups [@Tittensor2010].
+Marine biodiversity remains poorly understood, with even the global richness of macrobiota poorly constrained[@Mora2011,@Appeltans2012,@Poore2015], and most species are known from few records and studies [@Bouchet2002].
+Constrained by access to habitats and limited biodiversity infrastructure, coordinated, broad marine ecosyste monitoring initiatives have been slow to develop [@Duffy2013,@Borja2016] (but see for instance MarineGEO [@Duffy2016marinegeo], MBON, and PISCO), making it difficult to monitor changes in species distributions and community composition.
+Climate change and other human impacts are leading to rapid changes in marine ecosystems, and there is a pressing need to improve our knowledge of baseline marine systems and species diversity before they are further altered.
 
-Research attempting to understand how much we know about marine biodiversity has focused on estimating the total number of species that inhabits the oceans, and which fraction of this diversity has been named [@Appeltans2012,@Costello2013].
+Biodiversity is fundamental to ecosystem function, but documenting and tracking biodiversity is laden with taxonomic and sampling challenges.
+For macrobiota, species occurrence has generally been documented through visual identification of live or captured specimens.
+Reliably-identified samples can then be used to generate molecular identification tools, such as DNA barcodes, opening additional methods for species documentation.
+
+Species occurrence data is the basis for understanding species diversity and distributions, and is essential for monitoring community composition and ecosystem function.
+When species distributions are sufficiently well-sampled, occurrence data can also be used for environmental niche modelling to predict geographic and bathymetric distributions, community composition, and response of species and communities to changing environmental conditions.
+In well-sampled systems, like land plants, these can yield detailed prediction of future ecosystems (xxxx).
+
+Until recently, our knowledge of biodiversity was scattered among heterogeneous sources (e.g., natural history collections, publications, expert knowledge) that are difficult to access and synthesize.
+The development of public databases for specimen records (e.g., iDigBio [@Page2015]), species occurrences (e.g., OBIS [@Zhang2002]), genetic data (e.g., GenBank [@Benson2013], BOLD [@Ratnasingham2007a]), and registers of species names (e.g., WoRMS [@Costello2013b]) have revolutionized how biodiversity research by allowing ready access to aggregated data to answer broad questions.
+However, few studies have sought to synthesize existing knowledge by combining information from these databases.
+
+The goal of this study is to evaluate the extent of, and gaps in, our digital knowledge of species occurrences of marine invertebrates in the marine waters of the continental United States, one of the best documented areas in the world ocean.
+We aggregate publicly available data within the limits of the exclusive economic zone (EEZ) of the US, as well as the Gulf of Mexico (GOM) and the Pacific Northwest (PNW).
+We show that while public databases are capturing a significant portion of our knowledge about species richness, most species are known from only a few specimens and locations.
+We emphasize that our understanding of marine biodiversity, and the impact of human activities on biological communities require not only a more accurate knowledge of the species that inhabit the oceans, but also more comprehensive data on species distributions and abundances.
+These goals can only be achieved with additional cost-effective sampling and a commitment to developing and maintaining the taxonomic expertise who can identify and interpret the material collected.
+
+We focus on the continental US because the marine biodiversity is extensively documented as a result of a long history of studies by governmental, academic, and private institutions [@Fautin2010].
+The continental US includes a wide range of marine ecosystems from the Gulf of Mexico and Florida Keys, to the Gulf of Maine on the East Coast; and from arid Southern California to the shores of the Pacific Northwest on the West coast.
+With estuaries, bays, sandy beaches, rocky shores, and coral reefs, the US coastlines harbors many biota.
+While the continental shelf extends for several hundred kilometers on the East coast, it is narrow on the West coast.
+This diversity of habitats and bathymetric profiles, combined with efforts at the national level to make data publicly available, make this region especially suitable to evaluate how much we know about marine biodiversity using biodiversity repositories.
+Additionally, apart from the syntheses that emerged from the Census of Marine Life efforts [@Costello2010,@Fautin2010], no study has yet attempted to synthesize the current state of knowledge of marine biodiversity across the continental US, and this study is the first using data from public databases.
+
+## Results and Discussion
+
+<!---- Research attempting to understand how much we know about marine biodiversity has focused on estimating the total number of species that inhabits the oceans, and which fraction of this diversity has been named [@Appeltans2012,@Costello2013].
 Undoubtedly, the development of molecular methods to investigate species limits has reinvigorated taxonomy and contributed to elucidating cryptic species complexes [@Knowlton2000].
@@
 Occurrence data for marine species could be used for environmental niche modeling of their geographic and bathymetric distributions but the number of species -- and therefore the proportion of all species -- for which we have enough data to do so is unknown.
-
-Until very recently, our knowledge of biodiversity was scattered among heterogeneous sources (e.g.
-natural history collections, publications, expert knowledge), difficult to access, and therefore difficult to synthesize.
-The development of public databases for specimen records (e.g.
-iDigBio [@Page2015]), species occurrences (e.g., OBIS [@Zhang2002]), genetic data (e.g., GenBank [@Benson2013], BOLD [@Ratnasingham2007a]), and registers of species names (e.g., WoRMS [@Costello2013b]) now allow us to evaluate what and how much we know.
-However, few studies have sought to synthesize existing knowledge by combining information from these databases.
-Here, we aggregate publicly available data for marine invertebrates that occur within the limits of the exclusive economic zone (EEZ) of the continental United States (US), as well as the Gulf of Mexico (GOM) and the Pacific Northwest (PNW), to highlight gaps in our knowledge that provide a road map for important new directions in global research.
-We show that while public databases are capturing a significant portion of our knowledge about species richness, most species are known from only a few specimens and a few locations.
-We emphasize that our understanding of marine biodiversity, and the impact of human activities on biological communities require not only a more accurate list of the species that inhabit the oceans, but also more comprehensive data on species distributions and abundances.
-These goals can only be achieved with additional cost-effective sampling and a commitment to training more taxonomists and experts that can identify and interpret the material collected.
-
-We focus on the continental US because the marine biodiversity is extensively documented as the result of a long history of studies by governmental, academic, and private institutions [@Fautin2010].
-The continental US includes a wide diversity of marine ecosystems from the Florida Keys, to the Gulf of Maine on the East Coast; and from the arid Southern California to the oceanic shores of the Pacific Northwest on the West coast.
-With estuaries, bays, sandy beaches, rocky shores, and coral reefs, the US coastlines harbors many biota.
-While the continental shelf extends for several hundred kilometers on the East coast, it is narrow on the West coast.
-This diversity of habitats and bathymetric profiles, combined with efforts at the national level to make data publicly available, make this region especially suitable to evaluate how much we know about marine biodiversity using biodiversity repositories.
-Additionally, apart from the syntheses that emerged from the Census of Marine Life efforts [@Costello2010,@Fautin2010], no study has yet attempted to synthesize the current state of knowledge of marine biodiversity across the continental US, and this study is the first using data from public databases.
-
-## Results and Discussion
+-->
 

### 2018-04-16 R/utils.R
reduce precision in percentages reported in text

@@ -109,7 +109,7 @@ gcd.hf <- function(long1, lat1, long2, lat2, warn = TRUE) {
 }
 
 format_output <- function(x, ...) {
-    sprintf("%.1f", x)
+    sprintf("%.0f", x)
 }
 
 render <- function(file, ...) {

### 2018-05-01 manuscript.Rmd
include GP  comments

@@
 
-Given their diversity, marine invertebrates are underrepresented in iDigBio.
-While marine invertebrates are estimated to represent between 11% and 18% of the described global diversity (number of described marine inverterbrates species: 200,000-210,000 [@Appeltans2012], total number of described species: 1.2-1.9 million [@Mora2011,@Chapman2009]),
-marine invertebrates from the US EEZ only make up `r format_output(nrow(idigbio_us_inverts_records_species)/idigbio_stats$total_no_fossil*100)`% of the total number of non-fossil iDigBio records (`r round(idigbio_stats$total_no_fossil/1e6, 1)` million), and `r format_output(nrow(idigbio_us_inverts_records_species)/idigbio_stats$eez_no_fossil*100)`% of the US records (`r round(idigbio_stats$eez_no_fossil/1e6, 1)` million).
+Marine invertebrates are underrepresented in iDigBio compared to other taxa.
+Marine invertebrates represent 11% of described global species diversity (202,000 [@worms-stats] out of 1.7 million species [catalog-of-life]), but only `r format_output(nrow(idigbio_us_inverts_records_species)/idigbio_stats$eez_no_fossil*100)`% of the US non-fossil occurrence records ((`r round(idigbio_stats$eez_no_fossil/1e6, 1)` million) in iDigBio.
 
@@
 
-This underrepresentation is also manifest in the intensity of the collecting effort.
-Marine invertebrates are estimated to represent  `r total_invert_prop`% of the total marine diversity[@worms-stats].
-In OBIS records for the US EEZ, marine invertebrates represent `r invert_props$prop_obis_invert_samples`% of the occurrences but `r invert_props$prop_obis_invert_species`% of the total number of species; and in iDigBio, they represent `r invert_props$prop_idigbio_invert_samples`% of the records but `r invert_props$prop_idigbio_invert_species`% of the species.
-In contrast, marine vertebrates represent `r invert_props$prop_idigbio_vert_samples`% of the samples and `r invert_props$prop_idigbio_vert_species`% of the species in iDigBio, and `r invert_props$prop_obis_vert_samples`% of the samples and `r invert_props$prop_obis_vert_species`% of the species in OBIS (Fig.
-\@ref(fig:kingdom-diversity-plot)).
-In other words, while marine invertebrates are represented at a level on par with their diversity within the US EEZ in the databases, each species is represented by fewer samples than for vertebrates (median number of records per species for invertebrates `r median_idigbio_invert` and `r median_obis_invert`, _vs._ and `r median_idigbio_vert` and `r median_obis_vert` for vertebrates, for iDigBio and OBIS respectively).
+Invertebrates are also represented in the marine data.
+About `r total_invert_prop`% of described marine species are invertebrates.
+The proportional species diversity of invertebrates is captured relatively effectively among US marine EEZ records in OBIS (`r invert_props$prop_obis_invert_species`%) and iDigBio (`r invert_props$prop_idigbio_invert_species`%).
+However, invertebrate occurrence records are disproportionately low: `r invert_props$prop_obis_invert_samples`% in OBIS and `r invert_props$prop_idigbio_invert_samples`% in iDigbio.
+Marine vertebrate species have three times as many occurrence records on average in these databases than invertebrate species (median number of records per species for invertebrates `r median_idigbio_invert` and `r median_obis_invert`, _vs._ and `r median_idigbio_vert` and `r median_obis_vert` for vertebrates, for iDigBio and OBIS respectively).
 
-To assess the level of digitization of the fauna, we compared the species included in iDigBio and OBIS with the comprehensive checklists for molluscs [@turgeon1998] and crustaceans [@mclaughlin2005] that are available for the United States.
-For molluscs, of the `r format(itis_stats$moll_nspp, big.mark=",")` species included in the checklist, `r itis_stats$prop_moll_idig`% are in iDigBio and `r itis_stats$prop_moll_obis`% are in OBIS.
-For crustaceans, of the `r format(itis_stats$crust_nspp, big.mark=",")` species included in the checklist, `r itis_stats$prop_crust_idig`% are in iDigBio and `r itis_stats$prop_crust_obis`% are in OBIS.
+To assess how well documented species diversity is captured digitally, we compared the species retrieved from iDigBio and OBIS against comprehensive checklists for molluscs [@turgeon1998] and crustaceans [@mclaughlin2005] of the United States.
+Of the `r format(itis_stats$moll_nspp, big.mark=",")` mollusk species included in the checklist, `r itis_stats$prop_moll_idig`% are in iDigBio and `r itis_stats$prop_moll_obis`% are in OBIS.
+Of the `r format(itis_stats$crust_nspp, big.mark=",")` crustacean species in the checklist, `r itis_stats$prop_crust_idig`% are in iDigBio and `r itis_stats$prop_crust_obis`% are in OBIS.
 Given that molluscs and crustaceans are among the most sampled taxa, and that they have received more taxonomic attention than other phyla (Fig.
 \@ref(fig:idigbio-through-time)), these percentages offer a conservative estimate for the fraction of the fauna that is available digitally.
-Therefore, using a conservative estimate of a 60% digitization level of the fauna, we can estimate that there is at least 29,000 species of marine invertebrates in US waters.
-The higher overlap between the comprehensive checklist for  crustaceans and OBIS is attributable to occurences of planktonic species that are more represented in OBIS data than in museum collections.
+Therefore, using a conservative estimate of a 60% digitization level of the fauna, we can estimate that there is at least 29,000 species of marine invertebrates recorded in US waters.
+The higher overlap between the comprehensive checklist for crustaceans in OBIS than in iDigBio is attributable to planktonic species that are better represented in OBIS data than in museum collections.
 
@@
 All records published by OBIS have geographic coordinates associated with them.
-To estimate how records with missing geographic coordinates in iDigBio affect the completeness of our faunistic list, we queried iDigBio using the comprehensive lists for molluscs and crustaceans, and calculated the proportion of species with no iDigBio records within the US EEZ, and which had "United States" or "USA" in the country field, but had no geographic coordinates associated with them.
-These species may occur within the limits of the US EEZ but would not be returned in the queries used to assemble the faunistic list.
+To estimate the proportion of species missed because they lacked geographic coordinate in iDigBio, we queried iDigBio using the comprehensive list for molluscs and crustaceans, and calculated the proportion of species with no iDigBio records within the US EEZ, but which had "United States" or "USA" in the country field.
+These species should occur within within the limits of the US EEZ but would not be returned in the queries used to assemble the faunistic list.
 We found that `r format_output(itis_stats$missing_geo_moll$prop_us_no_geo * 100)`% ($n=$ `r format(itis_stats$missing_geo_moll$n_spp, big.mark=",")`) of the species included in the comprehensive checklist for molluscs have records in iDigBio that place them in the United States based on the "Country" field, but do not have geographic coordinates associated with them that would ensures they fall within the limits of the EEZ.
@@
 
-More records have been collected from the Atlantic Ocean, than the Gulf of Mexico, and the Pacific Ocean (`r format(sum(combined_records$is_east_coast), big.mark=",")`, `r format(sum(combined_records$within_gom), big.mark=",")`, and `r format(sum(combined_records$is_west_coast), big.mark=",")` records respectively).
+The Atlantic Ocean and the Gulf of Mexico are much better doocumented in these databases than the Pacific Ocean (`r format(sum(combined_records$is_east_coast), big.mark=",")`, `r format(sum(combined_records$within_gom), big.mark=",")`, and `r format(sum(combined_records$is_west_coast), big.mark=",")` records respectively).
 The highest species richness has been recorded from the Gulf of Mexico (`r format(n_spp_gom, big.mark=",")` species), followed by the Atlantic Ocean (`r format(n_spp_east, big.mark=",")` species), and the Pacific Ocean (`r format(n_spp_west, big.mark=",")` species).
+The higher species richness documented on the East coast partly reflects sampling bias, reflected in many more records, partly attributable to the presence of more and older research institutions, higher levels of digitization (Table \@ref(tab:records-table)), and wider continental shelf there.
 Most samples have been collected on the continental shelf (Fig.
-\@ref(fig:map-diversity-sampling)A), with `r prop_idigbio_continental_shelf`% (iDigBio) and `r prop_obis_continental_shelf`% (OBIS) of the records collected between 0-200m.
-The higher species richness measured on the East coast is in part attributable to the wider continental shelf, the presence of more institutions, and a higher level of digitization of collections on the East coast (Table \@ref(tab:records-table)).
+\@ref(fig:map-diversity-sampling)A), with `r prop_idigbio_continental_shelf`% (iDigBio) and `r prop_obis_continental_shelf`% (OBIS) of the records collected between $leq$200m.
 
-Spatial distribution of species richness with latitude differ between the Pacific and the Atlantic coasts.
+Sampling bias is further reflected in the lack of clear latitude diversity gradient (well documented regionally e.g.
+[@macpherson2002,@roy2000]) and differences between the Pacific and the Atlantic coasts.
 On the Pacific coast, the recorded richness is lowest around 40° of latitude, whereas on the Atlantic coast, it is highest at the same latitude (Fig.
@@
 \@ref(fig:latitudinal-gradient)B).
-These trends are however strongly driven by sampling effort (Fig.
+These trends are strongly influenced by sampling effort (Fig.
 \@ref(fig:latitudinal-gradient)C).
-Notably, for the Pacific coast, there is a deficit in the number of samples between 40° and 45° of latitude corresponding to the state of Oregon.
-In contrast, the peak in number of species observed and in the number of records around 42° on the Atlantic coast can be attributed to the presence of the Woods Hole Oceanographic Institute.
+Notably, for the Pacific coast, there is a deficit in the number of samples and species between 40° and 45° of latitude corresponding to the Oregon coast.
+In contrast, a peak in species and records around 42° on the Atlantic coast may be attributed to the Woods Hole Oceanographic Institute.
 
-```{r map-diversity-sampling, fig.cap='Spatial distribution of the number of samples (A) and species richness (B) for marine invertebrates based on records retrieved from iDigBio and OBIS within the limits of the Exclusive Economic Zone for the continental USA (represented in light gray).
+```{r map-diversity-sampling, fig.cap='Spatial distribution of the number of samples (A) and species richness (B) for marine invertebrates based on records retrieved from iDigBio and OBIS within the limits of the Exclusive Economic Zone for the continental USA (represented in light gray) in 0.2 x 0.2 degree cells.
 (\\#fig:map-diversity-sampling)'}
@@
 
-```{r correlation-sampling, fig.cap='Relationship between number of records and number of species for each cell of the map used in Fig.
-\\@ref(fig:map-diversity-sampling).
+```{r correlation-sampling, fig.cap='Relationship between number of records and number of species per 0.2 x 0.2 degree cell in US EEZ (Fig.
+\\@ref(fig:map-diversity-sampling)).
 The black line represents the 1:1 relationship.
@@
 
-Spatial patterns of species richness are however strongly confounded by sampling effort.
-There is a tight correlation between the number of samples and number of species per cell used in Fig.
+Further indication of sampling bias is a tight correlation between the number of samples and number of species per cell used in Fig.
 \@ref(fig:map-diversity-sampling), and `r format_output(stat_sampling_effort$p * 100)`% (n = `r format(stat_sampling_effort$n, big.mark=",")`) of the cells with $\leq$ 500 records have a number of species within 10% of the number of records (Fig.
 \@ref(fig:correlation-sampling)).
-For most of the US EEZ, the sampling effort is insufficient to capture acurately the species diversity.
+For most of the US EEZ, the sampling effort is insufficient to capture the species diversity acurately.
 
@@
 Underscoring the inadequacy of the sampling effort, `r prop_singleton_species`% of the species are known from a single record.
-
-Another indication of the inadequacy of the sampling is the relationship between the number of collected specimens and the number of species recorded through time (Fig.
-\@ref(fig:idigbio-through-time)).
-Explorations in the late nineteenth century led to an exponential growth in the knowledge of the fauna, but collecting then tapered off, until the 1970s when intensive collecting efforts led to an increase in the recording of new taxa (Fig.
-\@ref(fig:idigbio-through-time)).
-Since then, the slowed pace of collection activitity (and therefore a reduction in the number of specimens added) has been accompanied by a slow-down in the discovery of new records: `r n_spp_comparison$n_spp_2017_2000` species have been documented for the first time since 2000 (`r format_output(n_spp_comparison$n_spp_2017_2000/18)` species.year^-1^), while `r n_spp_comparison$n_spp_1970_1960` were documented between 1960 and 1970 (`r format_output(n_spp_comparison$n_spp_1970_1960/11)` species.year^-1^).
 
@@
 
-Biodiversity surveys conducted under the auspices of the Bureau of Land Management (BLM) and the Minerals Management Service (MMS) have contributed significantly to our knowledge of the geographic distribution of marine invertebrates in US waters.
-In our dataset, we identified at least `r format(sum(idigbio_mms_summary_by_year$n), big.mark=",")` records (or `r format_output(sum(idigbio_mms_summary_by_year$n)/nrow(idigbio_us_inverts_records_all)*100)`% of all US marine invertebrates in iDigBio) from these surveys, `r format_output(prop_mms_75_85 * 100)`% of which were collected between 1975 and 1985.
-Of the `r idigbio_mms_records %>% filter(rank %in% c("Species", "Subspecies")) %>% summarize(n = n_distinct(worms_valid_name)) %>% format(big.mark=",")` species collected as part of these surveys, `r nrow(idigbio_mms_species_only)` have not been collected outside of these surveys.
-
-The strong correlations between (1) the number of samples collected and the species richness (Fig.
-\@ref(fig:correlation-sampling)), and (2) the number of specimens collected per year and the number of species recorded for the first time (Fig.
-\@ref(fig:idigbio-through-time)C), suggest that any new sampling effort will lead to new observations and the discovery of new species.
-
 ```{r idigbio-through-time, fig.cap='Cumulative number of samples (A) and species (B) of marine invertebrates recorded through time based on iDigBio records.
@@
 
+Another indication of the vagaries and inadequacy of the sampling is shown in the temporal record of collected species and documented species (Fig.
+\@ref(fig:idigbio-through-time)).
+Two time periods stand out: the 1870-80's and the 1960-80's.
+The first represents intensive efforts to document the US marine biota during the formative stages of the major museums and research organizations focused on marinie life, while the second is in large part the result of the biodiversity surveys conducted under the auspices of the Bureau of Ocean Energy Management (BOEM; formerly part of Bureau of Land Management, BLM, and the Minerals Management Service, MMS).
+At least `r format(sum(idigbio_mms_summary_by_year$n), big.mark=",")` records (or `r format_output(sum(idigbio_mms_summary_by_year$n)/nrow(idigbio_us_inverts_records_all)*100)`% of all US marine invertebrates) in iDigBio were from these surveys, `r format_output(prop_mms_75_85 * 100)`% of them collected between 1975 and 1985.
+Of the `r idigbio_mms_records %>% filter(rank %in% c("Species", "Subspecies")) %>% summarize(n = n_distinct(worms_valid_name)) %>% format(big.mark=",")` species collected in these surveys, `r nrow(idigbio_mms_species_only)` have not been collected outside of these surveys.
+
+Since the BLM-MMS surveys, the rate of collection and documentation of new records have both slowed.
+During the 1960-1970 ~XX records were added per year, averaging `r round(n_spp_comparison$n_spp_1970_1960/11)` species.year^-1^ and
+`r round(n_spp_comparison$n_spp_2017_2000/18)` species.year^-1^ since 2000.
+
+<!--- Note also difference in evenness across phyla – mollusks much more even, annelids most jerky – reflects broad vs focused collecting efforts].
+Note however that the number of new records per sample taken has not declined appreciably (X to Y) in this period -->
+
+The strong correlations between (1) the number of samples collected and the species richness (Fig.
+\@ref(fig:correlation-sampling)), and (2) the number of specimens collected per year and the number of species recorded for the first time (Fig.
+\@ref(fig:idigbio-through-time)C), also suggest that much of the US marine biodiversity remains undocumented.
+
 Further biases exist across phyla and body size.
@@
 dried specimens like shells, corals, echinoderms) specimens are over-represented in museum collections, compared with small less accessible organisms, such as the meiofauna.
-Of `r n_idigbio_phyla` phyla represented in iDigBio, `r records_rare_phyla$n_rare_phyla` have <500 records.
+Of `r n_idigbio_phyla` animal phyla represented from US waters in iDigBio, `r records_rare_phyla$n_rare_phyla` have <500 records.
 While these phyla represent
-`r records_rare_phyla$prop_spp_rare_phyla_collected`% (n = `r format(records_rare_phyla$total_spp_rare_phyla_collected, big.mark=",")` total species) of the global number of species, they make up `r records_rare_phyla$prop_recs_rare_phyla`% of the iDigBio records.
+`r records_rare_phyla$prop_spp_rare_phyla_collected`% (n = `r format(records_rare_phyla$total_spp_rare_phyla_collected, big.mark=",")` total species) of the global number of species <!-- source unclear -->, they make up `r records_rare_phyla$prop_recs_rare_phyla`% of the iDigBio records.
 This proportion rises to `r records_rare_phyla$prop_spp_common_phyla_collected`% (n = `r format(records_rare_phyla$total_spp_common_phyla_collected, big.mark=",")`) for the `r n_idigbio_phyla - records_rare_phyla$n_rare_phyla` other phyla.
@@
 
-The proportion of samples that are identified to species in museum collections show a marked decline in recent decades in most taxa other than mollusks, demonstrating limiting taxonomic effort and expertise (Fig.
-\@ref(fig:identification-through-time)).
-There is a marked break in the central trend around the 1970's indicating that it now takes decades, on average, to identify samples deposited in collections.
-For diverse groups such as annelids or arthropods, the proportion of specimens identified to the species-level is now below 30%.
+The proportion of samples in museum collections that are identified to species decreases substantially toward the present, and shows a marked decline around the 1970's (Fig.
+\@ref(fig:identification-through-time)), demonstrating limiting taxonomic effort and expertise.
 
@@
 
-We extracted `r format(nrow(gom_worms_species), big.mark=",")` and `r format(nrow(kozloff_worms_species), big.mark=",")` marine invertebrate species from species checklists from the Gulf of Mexico [@Tunnell2009] and from the Pacific Northwest [@Kozloff1996] respectively.
-For the Gulf of Mexico, biodiversity databases included more species for the same geographic area as the one used by the authors of the checklist (iDigBio and OBIS combined: `r format(dplyr::n_distinct(combined_gom_records$worms_valid_name), big.mark=",")` species; iDigBio only: `r format(n_distinct(idigbio_gom_records$worms_valid_name), big.mark=",")` species, $n =$  `r format(nrow(idigbio_gom_records), big.mark=",")` records; OBIS only: `r format(n_distinct(obis_gom_records$worms_valid_name), big.mark=",")` species, $n=$ `r format(nrow(obis_gom_records), big.mark=",")` records).
+We extracted `r format(nrow(gom_worms_species), big.mark=",")` and `r format(nrow(kozloff_worms_species), big.mark=",")` marine invertebrate species from checklists of the Gulf of Mexico [@Tunnell2009] and ofx the Pacific Northwest [@Kozloff1996] respectively.
+For the Gulf of Mexico, biodiversity databases included more species for the same geographic area than the one used by the authors of the checklist (iDigBio and OBIS combined: `r format(dplyr::n_distinct(combined_gom_records$worms_valid_name), big.mark=",")` species; iDigBio only: `r format(n_distinct(idigbio_gom_records$worms_valid_name), big.mark=",")` species, $n =$  `r format(nrow(idigbio_gom_records), big.mark=",")` records; OBIS only: `r format(n_distinct(obis_gom_records$worms_valid_name), big.mark=",")` species, $n=$ `r format(nrow(obis_gom_records), big.mark=",")` records).
 For the PNW, fewer species than included in the regional checklist were recovered from biodiversity databases (iDigBio and OBIS combined: `r format(dplyr::n_distinct(combined_pnw_records$worms_valid_name), big.mark=",")` species; iDigBio only: `r format(dplyr::n_distinct(idigbio_pnw_records$worms_valid_name), big.mark=",")`  species, $n=$ `r format(nrow(idigbio_pnw_records), big.mark=",")` records; OBIS only `r format(dplyr::n_distinct(obis_pnw_records$worms_valid_name), big.mark=",")` species, $n=$ `r format(nrow(obis_pnw_records), big.mark=",")` records).
 The combination of the species listed in these three sources leads to `r format(n_distinct(total_gom_records$worms_valid_name), big.mark=",")` species for the GOM and `r format(n_distinct(total_pnw_records$worms_valid_name), big.mark=",")` species for the PNW.
+Conversely, about XX% of the total species records for the Gulf of Mexico, and XX% for the Pacific Northwest are missing from the checklists and represented only in databases.
 
@@
 Several factors may contribute to species being included in the regional lists but not having records in biodiversity databases.
-The first category of biases is linked to the likelihood of having specimens for these species in digitized museum collections.
+The first category of biases is linked to the likelihood of having specimens for these species in museum collections.
 Rare species may only have been collected a few times, and the corresponding records have not been digitized yet.
-Species that are difficult to preserve because they are large, very small, or fragile, are deposited in museums less frequently.
+Species that are difficult to preserve because they are large, very small, or fragile, are underrepresented in collections.
 The other category of biases is associated with data quality for these records.
-Some of these species may be in the databases but do not have geographic coordinates associated with them, and therefore would not be included in the results of our query.
-While it is difficult to assess biases associated with the likelihood of preservation of a species, we investigated how missing geographic coordinates are affecting the discrepancies between the checklists and the databases.
+Some of these species may not have been digitized, or lack geographic coordinates associated, and therefore would be missed by our queries.
+While it is difficult to assess biases associated with the likelihood of preservation of a species, we investigated how missing geographic coordinates affect the discrepancies between the checklists and the databases.
+
+<!-- include bias in digitization -->
 
 Missing geographic coordinates reduces the overlap between the list of species from the regional list and the databases.
-Using iDigBio, we found that for the `r format(not_really_in_database$idig_gom_n_spp_not_in_db, big.mark=",")` species that do not appear within the geographic range of the Gulf of Mexico but are on the list, `r format_output(not_really_in_database$idig_gom_p_spp_not_in_db*100)`% are completely absent from iDigBio, while for the `r format(not_really_in_database$idig_gom_n_spp_no_coords, big.mark=",")` species that do appear in iDigBio, `r format_output(not_really_in_database$idig_gom_p_spp_no_coords*100)`%  do not have geographic coordinates associated with them, and for those that do have geographic coordinates, `r format_output(not_really_in_database$idig_gom_p_spp_within_eez*100)`% ($n=$  `r format(not_really_in_database$idig_gom_n_spp_within_eez, big.mark=",")`) fall within US waters.
-Additional digitization of collections, including the capture of their geographic coordinates information, is needed to improve biodiversity knowledge at regional scales.
+Using iDigBio, we found that for the `r format(not_really_in_database$idig_gom_n_spp_not_in_db, big.mark=",")` species that do not appear within the geographic range of the Gulf of Mexico but are on the list, `r format_output(not_really_in_database$idig_gom_p_spp_not_in_db*100)`% are completely absent from iDigBio, while of the `r format(not_really_in_database$idig_gom_n_spp_no_coords, big.mark=",")` species that do appear in iDigBio, `r format_output(not_really_in_database$idig_gom_p_spp_no_coords*100)`%  do not have geographic coordinates associated with them.
+Additional digitization and georeferencing of collections is needed to improve biodiversity knowledge at regional scales.
 For instance, one can expect that the digitization coverage for the Gulf of Mexico will greatly increase once records from the Florida Wildlife Research Institute collections (St Petersburg, FL) become integrated into biodiversity information systems.
@@
 
-Two factors most likely contribute to explanations of why some species are in databases but are not included in the regional checklists: (1) the species have been documented for the first time since the publication of the checklist, and (2) the species records were not accessible or known to the authors.
+Two factors most likely contribute to explanations of why some species are in databases but are not included in the regional checklists: (1) the species have been documented for the first time since the publication of the checklist, (2) the species records were not accessible or known to the authors, or (3) the species records are erroneaous and were excluded by the compilers of the checklist.
 
-Recently recorded species contribute modestly to this discrepancy for the Gulf of Mexico, as `r sum(recently_collected$gom_in_dbs$min_year > 2001)` species (`r format_output( sum(recently_collected$gom_in_dbs$min_year > 2001)/n_distinct(total_gom_records$worms_valid_name) * 100)`% of the total species richness, `r format_output(sum(recently_collected$gom_in_dbs$min_year > 2001)/spp_not_in_lists$gom_not_in_list * 100)`% of the species not on the checklist),  have been collected for the first time since the publication of the checklist, but the results are more significant for the Pacific Northwest with  `r sum(recently_collected$pnw_in_dbs$min_year > 1996)` species (`r format_output(sum(recently_collected$pnw_in_dbs$min_year > 1996)/n_distinct(total_pnw_records$worms_valid_name) * 100)`% of the total species richness,  `r format_output(sum(recently_collected$pnw_in_dbs$min_year > 1996)/spp_not_in_lists$pnw_not_in_list * 100)`% of the species not on the checklist).
+Recently recorded species contribute modestly to this discrepancy for the Gulf of Mexico.
+Only `r sum(recently_collected$gom_in_dbs$min_year > 2001)` species (`r format_output( sum(recently_collected$gom_in_dbs$min_year > 2001)/n_distinct(total_gom_records$worms_valid_name) * 100)`% of the total species richness, `r format_output(sum(recently_collected$gom_in_dbs$min_year > 2001)/spp_not_in_lists$gom_not_in_list * 100)`% of the species not on the checklist) are only represented by collections made since the publication of the checklist.
+For the older Pacific Northwest checklist, this bias is larger with  `r sum(recently_collected$pnw_in_dbs$min_year > 1996)` species (`r format_output(sum(recently_collected$pnw_in_dbs$min_year > 1996)/n_distinct(total_pnw_records$worms_valid_name) * 100)`% of the total species richness,  `r format_output(sum(recently_collected$pnw_in_dbs$min_year > 1996)/spp_not_in_lists$pnw_not_in_list * 100)`% of the species not on the checklist) collected subsequent to publication.
 
-The authors of these checklists did not have access to digital databases allowing them to access  museum records across institutions, which explains why some species were not included in the checklists.
-Many of these records were difficult to discover before digitization which, for the most part, has occurred after the compilation of these checklists.
+The authors of these checklists had no or limited access to digital databases that have since been developed and this likely contributed to species being missed.
+
+<!-- GP says he moved this up, but I don't see it: Many of these records were difficult to discover before digitization which, for the most part, has occurred after the compilation of these checklists.
 We estimate that using biodiversity databases increases the number of species included in the checklist by `r format_output(spp_not_in_lists$gom_not_in_list/n_distinct(total_gom_species$worms_valid_name)*100)`% for the Gulf of Mexico, and `r format_output(spp_not_in_lists$pnw_not_in_list/n_distinct(total_pnw_species$worms_valid_name)*100)`% for the Pacific Northwest.
+-->
 
@@
 
-Since its introduction in the early 2000s, DNA barcoding has revitalized invertebrate taxonomy by facilitating species delineation [@Meyer2005-plos].
+Since its introduction in the early 2000s, DNA barcoding has revitalized invertebrate taxonomy by facilitating species identification and delineation [@Meyer2005-plos].
 More recently, advances in high-throughput sequencing technologies have enabled the parallel sequencing of thousands of specimens in environmental samples (eDNA) cost effectively [@Leray2015], making it possible to estimate diversity and identify taxa from species-rich samples at unprecedented scales.
-With decreasing sequencing costs, and the prospect of automating sample collection, these technologies may soon be used routinely to monitor biodiversity continuously [@Bohan2017].
-However, the utility of the data generated by these projects is limited by the paucity of reliably identified DNA barcode sequences that can serve for matching and identification of eDNA transcripts [@Leray2015].
-Thus insights from eDNA technologies depend on the level of completeness and accuracy of comparative, species-level DNA barcode coverage.
+With decreasing sequencing costs, and the prospect of automating sample collection, these technologies are poised to revolutionize large-scale assessment and monitoring of biodiversity [@Bohan2017].
+However, the utility of the data generated hinges on reliably identified DNA barcode sequences that can serve for matching and identification of eDNA transcripts [@Leray2015].
+Thus insights from eDNA technologies depend on the level of completeness and accuracy of comparative, identified, species-level DNA barcode coverage.
 
-The marine invertebrates of the US have generally been characterized genetically at a higher level than other parts of the world: `r format_output(prop_spp_barcoded$p_barcoded * 100)`% (n = `r format(nrow(combined_species), big.mark=",")`) of the digitized fauna has DNA barcodes available.
+The marine invertebrates of the US are genetically characterized at a relatively high level: `r format_output(prop_spp_barcoded$p_barcoded * 100)`% (n = `r format(nrow(dplyr::distinct(combined_species, worms_valid_name)), big.mark=",")`) of the digitized fauna has DNA barcodes available from specimens taken somewhere in the world.
+Coverage varies by phylum from `r format_output(stat_barcoding$min_eez_value * 100)`% (`r stat_barcoding$min_eez_phylum`) to `r format_output(stat_barcoding$max_eez_value * 100)`% (`r stat_barcoding$max_eez_phylum`) (Fig.
+\@ref(fig:proportion-barcoded)).
+The GOM has comparable DNA barcoding coverage to the US average, but the PNW is better barcoded with >30% species-level coverage across each of the five most diverse phyla.
+This is driven in part by the active Canadian barcoding efforts in British Columbia.
 In 2012, it was estimated that, at a global scale, less than 3% of the marine species had been sequenced [@Appeltans2012], and that this number was even lower in some areas (e.g., 2.6% for the Antarctic [@Grant2009]).
-This proportion varies by phylum from `r format_output(stat_barcoding$min_eez_value * 100)`% for `r stat_barcoding$min_eez_phylum` to `r format_output(stat_barcoding$max_eez_value * 100)`% for `r stat_barcoding$max_eez_phylum` (Fig.
-\@ref(fig:proportion-barcoded)).
-The GOM has comparable coverage in level of DNA barcoding of the fauna to the EEZ, while the PNW is better barcoded with >30% species-level coverage across each of the five most diverse phyla.
-This is driven in part by active Canadian barcoding efforts in British Columbia.
 
-The above estimates are based on species names, but many traditional species include multiple cryptic species [@Knowlton2000;@Meyer2005-evolution].
-For example, up to `r format_output(100 * stat_bold$p_spp_multi_bin)`% of `r format(stat_bold$n_spp_not_singleton, big.mark=",")` species that have DNA barcodes available in the digitized US fauna, are represented by more than 1 record in BOLD include cryptic species, as denoted by assignment to  more than 1 BIN (BOLD Index Number, molecular operational taxonomic units that closely resemble species [@Ratnasingham2013]).
-Specimens identified under the same species but sampled from distant locations are especially prone to representing cryptic species.
-While geographic information is missing for most records (`r format_output(100*stat_bold$p_no_coords)`% of the `r format(stat_bold$n_total_records, big.mark=",")` BOLD records analyzed here have no geographic coordinate information associated with them), of the `r format(bold_geography_stats$n_spp_with_records, big.mark=",")` marine invertebrates species from the US with BOLD records, only `r format_output(bold_geography_stats$p_within_eez * 100)`% have at least one record collected within the EEZ and `r format_output(bold_geography_stats$p_in_north_america * 100)`% list a North American country as the site of collection.
-Although DNA barcodes are available for `r format_output(prop_spp_barcoded$p_barcoded * 100)`% of the marine invertebrate fauna, many of these records have been collected outside the Atlantic and Pacific US biogeographic regions, and therefore geographical differentiation has not been evaluated for most taxa.
-It is thus crucial to ensure that precise and accurate geographic information is provided when DNA barcodes are published, and that enough samples for each species are sequenced to capture more accurately their geographic range and the extent of their genetic diversity.
+The above estimates are based on current taxonomy, but many species as currently understood are complexes of multiple cryptic species [@Knowlton2000;@Meyer2005-evolution].
+BOLD uses an algorithm to assess species-level genetic differentiation among samples, separating molecular operational units into different BINs (BOLD Index Numbers)[@Ratnasingham2013].
+More than a third of the marine invertebrate species recorded from the US and represented in BOLD include multiple BINs (`r format_output(100 * stat_bold$p_spp_multi_bin)`% of `r format(stat_bold$n_spp_not_singleton, big.mark=",")` species).
+Given the relatively sparse DNA coverage among as well as within species, the actual level of crytpic differentiation is higher.
+Many of the BOLD sequences of US species are based on specimens collected outside the US.
+Thus both the diversity and species-level identity of the marine fauna needs to be comprehensively reevaluated.
+
+Specimens identified under the same species but sampled from distant locations are especially prone to include cryptic species.
+While geographic information is missing for most records in BOLD (`r format_output(100*stat_bold$p_no_coords)`% of the `r format(stat_bold$n_total_records, big.mark=",")` BOLD records analyzed here have no geographic coordinate information associated with them), of the `r format(bold_geography_stats$n_spp_with_records, big.mark=",")` marine invertebrates species from the US with BOLD records, only `r format_output(bold_geography_stats$p_within_eez * 100)`% have at least one record collected within the EEZ and `r format_output(bold_geography_stats$p_in_north_america * 100)`% list a North American country as the site of collection.
+It is crucial to ensure that accurate geographic information is provided when DNA barcodes are published, and that enough samples for each species are sequenced to capture their geographic range and  genetic diversity.
 

### 2018-05-05 R/clustering.R
dump code used for prediction/clustering

@@ -1,0 +1,110 @@
+make_species_site_matrix <- function(recs, raster) {
+    recs %>%
+        dplyr::filter(rank == "Species" | rank == "Subspecies") %>%
+        add_rastercell(raster) %>%
+        dplyr::select(worms_valid_name, rastercell) %>%
+        dplyr::distinct(worms_valid_name, rastercell, .keep_all = TRUE) %>%
+        dplyr::mutate(present = 1) %>%
+        tidyr::spread(worms_valid_name, present, fill = 0L) %>%
+        dplyr::mutate_at(vars(matches("[0-9]+")), as.integer)
+}
+
+
+make_distance_species_site_matrix <- function(mat) {
+    ## keep only sites with more than 10 species
+    ## keep only species observed at 5 sites
+    ## first column is site names
+    sub_mat <- mat[, c(FALSE, colSums(mat[, -1]) >=  5)]
+    to_keep <- rowSums(sub_mat) >= 10
+    sub_mat <- sub_mat[to_keep, ]
+    sites <- mat[to_keep, 1]
+
+    ## calculate jaccard matrix
+    jac_mat <- vegan::vegdist(sub_mat, method="jaccard")
+    attr(jac_mat, "sites") <- sites
+    jac_mat
+}
+
+get_dunn_indices <- function(dmat) {
+
+    methods <-  c("ward.D", "ward.D2","single", "complete", "average",
+                  "mcquitty", "median", "centroid")
+    res <- vector("list", length(methods))
+    for (i in seq_along(methods)) {
+        ## build hierarchical cluster
+        hc <-  hclust(dmat, method = methods[i])
+        dunn_i <- numeric(100)
+
+        for (k in seq(from = 5, by = 1, length.out = 100)) {
+            tc <- cutree(hc, k = k)
+            dunn_i[k - 4] <- clValid::dunn(distance = dmat,
+                                            clusters = tc)
+        }
+        res[[i]] <- dunn_i
+    }
+    names(res) <- methods
+    res
+}
+
+## From this:
+## di <- get_dunn_indices(distance_site_matrix)
+## as_tibble(di) %>% gather(method,  distance) %>%
+##     group_by(method) %>%
+##     mutate(k = seq(5, by=1, length.out=100)) %>%
+##     ggplot(aes(x=k, y=distance, colour=method)) +
+##     geom_line()
+##
+## it seems that single method with few k is better however, this is largely
+## driven by a few outliers that force a cluster with most observations. Using
[…]

### 2018-05-05 R/col_types.R

@@ -1,0 +1,115 @@
+env_data_col_types <- function() {
+    cols(
+        CsquareCode = col_character(),
+        LOICZID = col_integer(),
+        NLimit = col_double(),
+        Slimit = col_double(),
+        WLimit = col_double(),
+        ELimit = col_double(),
+        CenterLat = col_double(),
+        CenterLong = col_double(),
+        CellArea = col_double(),
+        OceanArea = col_double(),
+        CellType = col_character(),
+        PWater = col_integer(),
+        FAOAreaM = col_integer(),
+        FAOAreaIn = col_integer(),
+        CountryMain = col_character(),
+        CountrySecond = col_character(),
+        CountryThird = col_character(),
+        CountrySubMain = col_character(),
+        CountrySubSecond = col_character(),
+        CountrySubThird = col_character(),
+        EEZ = col_integer(),
+        EEZRemark = col_character(),
+        LME = col_integer(),
+        LMEBorder = col_integer(),
+        MEOW = col_integer(),
+        OceanBasin = col_integer(),
+        Longhurst = col_character(),
+        IslandsNo = col_integer(),
+        Area0_20 = col_double(),
+        Area20_40 = col_double(),
+        Area40_60 = col_double(),
+        Area60_80 = col_double(),
+        Area80_100 = col_double(),
+        AreaBelow100 = col_double(),
+        ElevationMin = col_integer(),
+        ElevationMax = col_integer(),
+        ElevationMean = col_double(),
+        ElevationSD = col_double(),
+        DepthMin = col_integer(),
+        DepthMax = col_integer(),
+        DepthMean = col_integer(),
+        DepthSD = col_double(),
+        SSTMnMin = col_double(),
+        SSTMnMax = col_double(),
+        SSTAnMean = col_double(),
+        SSTAnSD = col_double(),
+        SSTMnRange = col_double(),
+        SBTAnMean = col_double(),
+        SalinityMin = col_double(),
+        SalinityMax = col_double(),
+        SalinityMean = col_double(),
+        SalinitySD = col_character(),
+        SalinityBMean = col_double(),
+        PrimProdMean = col_integer(),
+        IceConAnn = col_double(),
+        Shelf = col_double(),
+        Slope = col_double(),
+        IceConSpr = col_double(),
+        IceConSum = col_double(),
+        IceConFal = col_double(),
+        IceConWin = col_double(),
+        LandDist = col_integer(),
[…]

### 2018-05-05 R/predictions.R

@@ -1,0 +1,232 @@
+add_environmental_data <- function(d, raster,
+                                   env_data_file = "data-raw/environmental-layers/hcaf_v6.csv") {
+
+    env_data <- readr::read_tsv(env_data_file, col_types = env_data_col_types())
+
+    ## add raster cells from 0.5 USA raster grid
+    ## and filter out records that are not part of it
+    env_data_grid <- env_data %>%
+        dplyr::rename(decimallatitude = CenterLat,
+                      decimallongitude = CenterLong) %>%
+        add_rastercell(raster) %>%
+        dplyr::filter(!is.na(rastercell)) %>%
+        ## select:
+        ## - areas at depth
+        ## - elevation data (evalation above sea level)
+        ## - depth data
+        ## - SST and Bottom temperature
+        ## - salinity
+        ## - primary productivity
+        ## - distance from coast
+        dplyr::select(rastercell,
+                      starts_with("Area"), starts_with("Elevation"),
+                      starts_with("Depth"), starts_with("SSTAn"),
+                      starts_with("SSTMn"),
+                      starts_with("SBTAn"), starts_with("SalinityM"),
+                      starts_with("SalinityBM"),
+                      starts_with("PrimProdM"), LandDist, WaveHeight,
+                      TidalRange, Shelf, Slope, Abyssal, Coral, Estuary,
+                      Seamount)
+
+    dplyr::left_join(d, env_data_grid, by = "rastercell")
+
+}
+
+assemble_keras_data <- function(recs, raster) {
+
+    res <- recs %>%
+        dplyr::filter(rank == "Species" | rank == "Subspecies") %>%
+        add_rastercell(raster) %>%
+        dplyr::group_by(rastercell) %>%l
+        dplyr::summarize(
+            n_spp = n_distinct(worms_valid_name),
+            n_samp = n()
+            ) %>%
+        add_environmental_data(raster = raster) %>%
+        dplyr::mutate(coords = map(rastercell, function(x) {
+            .r <- raster::xyFromCell(raster, x)
+            tibble::tibble(x=.r[[1]], y=.r[[2]])
+        })) %>%
+        tidyr::unnest()
+    ## only a few cells have missing data, let's remove them for now
+    res[complete.cases(res), ]
+}
+
[…]

### 2018-05-05 manuscript.Rmd
more text edits

@@
 However, sampling bias obscure broad diversity patterns, such as differences between coasts and across latitude.
-Our knowledge about species richness and geographic range is limited by sampling effort, with `r prop_10_specimens_species`% of species known from $\leq$10 samples.
+Our knowledge about species richness and geographic range is limited by sampling effort, with `r prop_10_specimens_species`% of species known from $\leq$ 10 samples.
 While recent developments in high-throughput sequencing promise to greatly speed up documentation of marine biodiversity, interpretation of sequences requires anchoring results anchoring results to identified vouchers.
@@
 BOLD uses an algorithm to assess species-level genetic differentiation among samples, separating molecular operational units into different BINs (BOLD Index Numbers)[@Ratnasingham2013].
-More than a third of the marine invertebrate species recorded from the US and represented in BOLD include multiple BINs (`r format_output(100 * stat_bold$p_spp_multi_bin)`% of `r format(stat_bold$n_spp_not_singleton, big.mark=",")` species).
+More than a third of the marine invertebrate species recorded from the US and represented in BOLD include multiple BINs (`r format_output(100 * stat_bold$p_spp_multi_bin)`% of `r format(stat_bold$n_spp_not_singleton, big.mark=",")` species represented by more than XX sequences).
+XX% of the fauna is only known from single barcodes at this stage.
 Given the relatively sparse DNA coverage among as well as within species, the actual level of crytpic differentiation is higher.
@@
 Specimens identified under the same species but sampled from distant locations are especially prone to include cryptic species.
-While geographic information is missing for most records in BOLD (`r format_output(100*stat_bold$p_no_coords)`% of the `r format(stat_bold$n_total_records, big.mark=",")` BOLD records analyzed here have no geographic coordinate information associated with them), of the `r format(bold_geography_stats$n_spp_with_records, big.mark=",")` marine invertebrates species from the US with BOLD records, only `r format_output(bold_geography_stats$p_within_eez * 100)`% have at least one record collected within the EEZ and `r format_output(bold_geography_stats$p_in_north_america * 100)`% list a North American country as the site of collection.
+While geographic information is missing for most records in BOLD (`r format_output(100*stat_bold$p_no_coords)`% of the `r format(stat_bold$n_total_records, big.mark=",")` BOLD records that match a species name in our checklist have no geographic coordinate information associated with them), of the `r format(bold_geography_stats$n_spp_with_records, big.mark=",")` marine invertebrates species from the US with BOLD records, only `r format_output(bold_geography_stats$p_within_eez * 100)`% have at least one record collected within the EEZ and `r format_output(bold_geography_stats$p_in_north_america * 100)`% list a North American country as the site of collection.
 It is crucial to ensure that accurate geographic information is provided when DNA barcodes are published, and that enough samples for each species are sequenced to capture their geographic range and  genetic diversity.

### 2018-05-05 R/comparison_by_geo.R
small tweak

@@ -469,15 +469,15 @@ get_not_really_in_database <- function(database_overlap) {
                )
 
     idig_gom_n_spp_within_eez <- nrow(idig_gom_within_eez)
-    idig_gom_p_spp_within_eez <- mean(idig_gom_within_eez$within_eez)
+    idig_gom_p_spp_within_eez <- mean(idig_gom_within_eez$within_eez, na.rm = TRUE)
 
     ## WARNING: Here the `n` refers to the denominator for the
     ## proportions, not the actual number of records
     list(
-        idig_gom_n_spp_not_in_db = nrow(idig_gom_not_in_db),
-        idig_gom_p_spp_not_in_db = idig_gom_n_spp_not_in_db/n_distinct(idig_gom_not_in_db$worms_valid_name),
-        idig_gom_n_spp_no_coords = idig_gom_n_spp_no_coords,
-        idig_gom_p_spp_no_coords = idig_gom_p_spp_no_coords,
+        idig_gom_n_spp_not_in_db  = nrow(idig_gom_not_in_db),
+        idig_gom_p_spp_not_in_db  = idig_gom_n_spp_not_in_db/n_distinct(idig_gom_not_in_db$worms_valid_name),
+        idig_gom_n_spp_no_coords  = idig_gom_n_spp_no_coords,
+        idig_gom_p_spp_no_coords  = idig_gom_p_spp_no_coords,
         idig_gom_n_spp_within_eez = idig_gom_n_spp_within_eez,
         idig_gom_p_spp_within_eez = idig_gom_p_spp_within_eez
     )

### 2022-02-04 manuscript.Rmd
add bibliography

@@
 csl: nature.csl
-bibliography: /home/francois/Library/dark-plankton.bib
+bibliography: sok.bib
 abstract: Museum collections have been recording our knowledge about the diversity and distribution of species for about 250 years.
